### 2023-02-20 bg1.Rmd
added a script

---
title: "Importance of background gene list for enrichment analysis"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/background

## Intro

I previously wrote about the importance of a background gene list for
accurate and correct enrichment analysis.

Here I will demonstrate how important this is, as well as providing a
guide for how to make your own background gene list.

For this guide I will be using data from a previous analysis, which is
deposited at NCBI GEO and SRA under accession numbers: GSE109140 and SRP128998.

The data was presented in a [paper by Felisbino et al](https://pubmed.ncbi.nlm.nih.gov/33495488/)

```{r,libs}

library("getDEE2")
library("DESeq2")
library("kableExtra")
library("vioplot")
library("clusterProfiler")
library("eulerr")

```

## Fetch data

Using RNAseq data from [DEE2](https://pubmed.ncbi.nlm.nih.gov/30942868/),
which can be accessed from dee2.io

```{r,fetch}

mdat <- getDEE2Metadata("hsapiens")

# get sample sheet
ss <- subset(mdat,SRP_accession=="SRP128998")

# fetch the whole set of RNA-seq data
x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
mx <- x$GeneCounts
rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)

# vpa subset no filtering
ss1 <- ss[grep("low",ss$Experiment_title),]
ss1$vpa <- grepl("VPA",ss1$Experiment_title)

ss1 %>% 
  kbl(caption="sample sheet for VPA treatment in hepatocytes") %>%
  kable_paper("hover", full_width = F)

mx1 <- mx[,which(colnames(mx) %in% ss1$SRR_accession)] 
dim(mx1)

```

## Analysis of read counts

Although I know this data set is good as I was involved in the project,
it is a good idea to show at least the number of reads for each sample.

```{r,counts1}

par(mar=c(5,7,5,1))
barplot(rev(colSums(mx1)),horiz=TRUE,las=1,main="number of reads per sample in SRP128998")

```

Now we can have a look at the read count distribution for one of the samples.

```{r,dist1}

# look at the distribution of counts
l1 <- lapply(1:ncol(mx1),function(i) log10( mx1[,i] +1 ))
names(l1) <- colnames(mx1)
l1 <- rev(l1)

hist(l1[[6]],main="Distribution of read counts in SRR6467485",
  xlab="log10(readcounts+1)",breaks=30)

abline(v=1,col="red",lty=2)

```

We can see that most genes have a very low read count, consistent
with most genes being undetectable in any sample.

Based on this typical distribution of read counts, I would consider
any gene with fewer than 10 reads as not detected.
This could be considered a strict detection threshold.
Setting a strict threshold like this could be helpful in selecting
genes that are likely to validate with qRT-PCR.

I will run DESeq2 with and without this threshold to see how
it impacts the results.

## Differential expression analysis with and without filtering

To see whether this has any effect on downstream analysis, DESeq2 will
be run on the same dataset without any filtering, with light filtering and with
strict filtering.

Here I define light filter threshold as the lowest baseline expression level that
DE genes start to appear.
I know this contrast will yield thousands of DE genes, so this is a reasonable
way to set the detection threshold.
The DE genes with lowest baseline expression showed mean read counts of about 2.

```{r,de1}

# DESeq2 without any fitering
dds <- DESeqDataSetFromMatrix(countData = mx1 , colData = ss1, design = ~ vpa )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz<-cbind(as.data.frame(z),mx1)
dge1<-as.data.frame(zz[order(zz$pvalue),])

head(dge1,10) %>% 
  kbl(caption="Top DE genes for VPA treatment (unfiltered)") %>%
  kable_paper("hover", full_width = F)

sig1 <- subset(dge1,padj<=0.05)
message("number of significant genes")
nrow(sig1)

plot(rowMeans( log10(dge1[,7:12]+1) ),dge1$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  xlab="log10(mean read count)",ylab="log2(fold change)")

points(rowMeans( log10(sig1[,7:12]+1) ),sig1$log2FoldChange,
  cex=0.6,pch=19,col="red")

abline(v=1,lty=2)

dge1up <- rownames(subset(dge1,padj<=0.05 & log2FoldChange>0))
dge1up <- unique(sapply(strsplit(dge1up," "),"[[",2))

dge1dn <- rownames(subset(dge1,padj<=0.05 & log2FoldChange<0))
dge1dn <- unique(sapply(strsplit(dge1dn," "),"[[",2))

dge1bg <- rownames(dge1)
dge1bg <- unique(sapply(strsplit(dge1bg," "),"[[",2))

```

look at the lowest basemean genes

```{r,low}

sig1 <- sig1[order(sig1$baseMean),]
head(sig1,10) %>%
  kbl(caption="lowest basemean genes") %>%
  kable_paper("hover", full_width = F)

```

The counts look quite low.
Not sure that these DEGs could be validated with qRT-PCR.

## Now run DESeq2 with filtering

```{r,de2}

mx2 <- mx1[which(rowMeans(mx1)>=10),]
dim(mx2)
dds <- DESeqDataSetFromMatrix(countData = mx2 , colData = ss1, design = ~ vpa )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz<-cbind(as.data.frame(z),mx2)
dge2<-as.data.frame(zz[order(zz$pvalue),])

head(dge2,10) %>%
  kbl(caption="Top DE genes for VPA treatment (filtered)") %>%
  kable_paper("hover", full_width = F)

sig2 <- subset(dge2,padj<=0.05)
nrow(sig2)

plot(log10(dge2$baseMean),dge2$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  xlab="log10(base mean)",ylab="log2 fold change",
  main="strict filter")

points(log10(sig2$baseMean),sig2$log2FoldChange,
  cex=0.6,pch=19,col="red")

dge2up <- rownames(subset(dge2,padj<=0.05 & log2FoldChange>0))
dge2up <- unique(sapply(strsplit(dge2up," "),"[[",2))

dge2dn <- rownames(subset(dge2,padj<=0.05 & log2FoldChange<0))
dge2dn <- unique(sapply(strsplit(dge2dn," "),"[[",2))

dge2bg <- rownames(dge2)
dge2bg <- unique(sapply(strsplit(dge2bg," "),"[[",2))

```

## Now run ORA based enrichment analysis

I will run enrichment analysis using the correct correct background list,
and compare the results to the incorrect whole genome background.

```{r,gs}

if (! file.exists("ReactomePathways.gmt")) {
  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
    destfile="ReactomePathways.gmt.zip")
  unzip("ReactomePathways.gmt.zip")
}

genesets2 <- read.gmt("ReactomePathways.gmt")

```

### ORA with correct background list.

Firstly with the upregulated genes

```{r,ora2up}

ora2_up <- as.data.frame(enricher(gene = dge2up , 
  universe = dge2bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora2_up$geneID <- NULL
ora2_ups <- rownames(subset(ora2_up,p.adjust<0.05))

gr <- as.numeric(sapply(strsplit(ora2_up$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora2_up$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora2_up$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora2_up$BgRatio,"/"),"[[",2))

ora2_up$es <- gr/br
ora2_up <- ora2_up[order(-ora2_up$es),]
head(ora2_up)

topup2 <- rev(head(ora2_up$es,20))
names(topup2) <- rev(head(ora2_up$ID,20))

```

Now with the downregulated genes

```{r,ora2dn}

ora2_dn <- as.data.frame(enricher(gene = dge2dn ,
  universe = dge2bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora2_dn$geneID <- NULL
ora2_dns <- rownames(subset(ora2_dn,p.adjust<0.05))

gr <- as.numeric(sapply(strsplit(ora2_dn$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora2_dn$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora2_dn$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora2_dn$BgRatio,"/"),"[[",2))

ora2_dn$es <- gr/br
ora2_dn <- ora2_dn[order(-ora2_dn$es),]
head(ora2_dn)

topdn2 <- head(ora2_dn$es,20)
names(topdn2) <- head(ora2_dn$ID,20)

```

Make a barplot

```{r,topbarplot2}

par(mar=c(5,20,5,1))

cols <- c(rep("blue",20),rep("red",20))

barplot(c(topdn2,topup2),
  horiz=TRUE,las=1,cex.names=0.65,col=cols,
  main="top DE Reactomes",
  xlab="ES")

mtext("correct background")

```

### ORA with whole genome background list.

Firstly with the upregulated genes

```{r,ora1up}

ora1_up <- as.data.frame(enricher(gene = dge1up ,
  universe = dge1bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora1_up$geneID <- NULL
ora1_ups <- rownames(subset(ora1_up,p.adjust<0.05))

gr <- as.numeric(sapply(strsplit(ora1_up$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora1_up$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora1_up$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora1_up$BgRatio,"/"),"[[",2))

ora1_up$es <- gr/br
ora1_up <- ora1_up[order(-ora1_up$es),]
head(ora1_up)

topup1 <- rev(head(ora1_up$es,20))
names(topup1) <- rev(head(ora1_up$ID,20))

```

Now with the downregulated genes

```{r,ora1dn}

ora1_dn <- as.data.frame(enricher(gene = dge1dn ,
  universe = dge1bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora1_dn$geneID <- NULL
ora1_dns <- rownames(subset(ora1_dn,p.adjust<0.05))

gr <- as.numeric(sapply(strsplit(ora1_dn$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora1_dn$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora1_dn$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora1_dn$BgRatio,"/"),"[[",2))

ora1_dn$es <- gr/br
ora1_dn <- ora1_dn[order(-ora1_dn$es),]
head(ora1_dn)

topdn1 <- head(ora1_dn$es,20)
names(topdn1) <- head(ora1_dn$ID,20)

```

Make a barplot

```{r,topbarplot1}

par(mar=c(5,20,5,1))

cols <- c(rep("blue",20),rep("red",20))

barplot(c(topdn1,topup1),
  horiz=TRUE,las=1,cex.names=0.65,col=cols,
  main="top DE Reactomes",
  xlab="ES")

mtext("whole genome background")

```

## Compare pathway results

```{r,venn1}

v1 <- list("correct up"=ora2_ups, "wg up"=ora1_ups)

v2 <- list("correct dn"=ora2_dns, "wg dn"=ora1_dns)

par(mar=c(10,10,10,10))

par(mfrow=c(2,1))
plot(euler(v1),quantities = TRUE)
plot(euler(v2),quantities = TRUE)
dev.off()

```

## Session information

```{r,session}

sessionInfo()

```

### 2023-02-23 bg2.Rmd
fix minor bugs

--- 2023-02-20 bg1.Rmd
@@
 For this guide I will be using data from a previous analysis, which is
-deposited at NCBI GEO and SRA under accession numbers: GSE109140 and SRP128998.
+deposited at NCBI GEO and SRA under accession numbers: GSE55123 and SRP038101.
 
-The data was presented in a [paper by Felisbino et al](https://pubmed.ncbi.nlm.nih.gov/33495488/)
+The data was presented in a [paper by Lund et al, 2014](https://pubmed.ncbi.nlm.nih.gov/25315154/)
+
+The experiment is designed to investigate the effect of Azacitidine treatment on AML3 cells.
 
@@
 
+suppressPackageStartupMessages({
   library("getDEE2")
@@
   library("eulerr")
+})
 
@@
 
-Using RNAseq data from [DEE2](https://pubmed.ncbi.nlm.nih.gov/30942868/),
+Using RNAseq data from [DEE2](https://pubmed.ncbi.nlm.nih.gov/25315154/),
 which can be accessed from dee2.io
@@
 # get sample sheet
-ss <- subset(mdat,SRP_accession=="SRP128998")
+ss <- subset(mdat,SRP_accession=="SRP038101")
 
@@
 rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)
+dim(mx)
 
-# vpa subset no filtering
-ss1 <- ss[grep("low",ss$Experiment_title),]
-ss1$vpa <- grepl("VPA",ss1$Experiment_title)
+# aza no filtering
+ss$trt <- grepl("Treated",ss$Experiment_title)
 
-ss1 %>% 
-  kbl(caption="sample sheet for VPA treatment in hepatocytes") %>%
+ss %>%
+  kbl(caption="sample sheet for Aza treatment in AML3 cells") %>%
   kable_paper("hover", full_width = F)
-
-mx1 <- mx[,which(colnames(mx) %in% ss1$SRR_accession)] 
-dim(mx1)
 
@@
 par(mar=c(5,7,5,1))
-barplot(rev(colSums(mx1)),horiz=TRUE,las=1,main="number of reads per sample in SRP128998")
+barplot(rev(colSums(mx)),horiz=TRUE,las=1,main="number of reads per sample in SRP038101")
 
@@
 # look at the distribution of counts
-l1 <- lapply(1:ncol(mx1),function(i) log10( mx1[,i] +1 ))
-names(l1) <- colnames(mx1)
-l1 <- rev(l1)
+l <- lapply(1:ncol(mx),function(i) log10( mx[,i] +1 ))
+names(l) <- colnames(mx)
+l <- rev(l)
 
-hist(l1[[6]],main="Distribution of read counts in SRR6467485",
+blah()
+hist(l[[6]],main="Distribution of read counts in SRR1171523",
   xlab="log10(readcounts+1)",breaks=30)
@@
 # DESeq2 without any fitering
-dds <- DESeqDataSetFromMatrix(countData = mx1 , colData = ss1, design = ~ vpa )
+dds <- DESeqDataSetFromMatrix(countData = mx , colData = ss, design = ~ trt )
 res <- DESeq(dds)
-z<- results(res)
+z <- results(res)
 vsd <- vst(dds, blind=FALSE)
-zz<-cbind(as.data.frame(z),mx1)
-dge1<-as.data.frame(zz[order(zz$pvalue),])
+zz <- cbind(as.data.frame(z),mx)
+de <- as.data.frame(zz[order(zz$pvalue),])
 
-head(dge1,10) %>% 
-  kbl(caption="Top DE genes for VPA treatment (unfiltered)") %>%
+head(de,10) %>%
+  kbl(caption="Top DE genes for Aza treatment (unfiltered)") %>%
   kable_paper("hover", full_width = F)
 
-sig1 <- subset(dge1,padj<=0.05)
-message("number of significant genes")
-nrow(sig1)
+DET=nrow(mx)
+NSIG=nrow(sig)
+NUP=nrow(subset(sig,log2FoldChange>0))
+NDN=nrow(subset(sig,log2FoldChange<0))
 
-plot(rowMeans( log10(dge1[,7:12]+1) ),dge1$log2FoldChange,
+HEADER=paste(DET,"genes included;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")
+
+plot(log10(de$baseMean) ,de$log2FoldChange,
   cex=0.6,pch=19,col="darkgray",
-  xlab="log10(mean read count)",ylab="log2(fold change)")
+  main="no detection filter",
+  xlab="log10(basemean)",ylab="log2(fold change)")
 
-points(rowMeans( log10(sig1[,7:12]+1) ),sig1$log2FoldChange,
+points(log10(sig$baseMean) ,sig$log2FoldChange,
   cex=0.6,pch=19,col="red")
+
+mtext(HEADER)
 
@@
 
-dge1up <- rownames(subset(dge1,padj<=0.05 & log2FoldChange>0))
-dge1up <- unique(sapply(strsplit(dge1up," "),"[[",2))
+deup <- rownames(subset(de,padj<=0.05 & log2FoldChange>0))
+deup <- unique(sapply(strsplit(deup," "),"[[",2))
 
-dge1dn <- rownames(subset(dge1,padj<=0.05 & log2FoldChange<0))
-dge1dn <- unique(sapply(strsplit(dge1dn," "),"[[",2))
+dedn <- rownames(subset(de,padj<=0.05 & log2FoldChange<0))
+dedn <- unique(sapply(strsplit(dedn," "),"[[",2))
 
-dge1bg <- rownames(dge1)
-dge1bg <- unique(sapply(strsplit(dge1bg," "),"[[",2))
+all <- rownames(de)
+all <- unique(sapply(strsplit(all," "),"[[",2))
 
@@
 
-sig1 <- sig1[order(sig1$baseMean),]
-head(sig1,10) %>%
+sig <- sig[order(sig$baseMean),]
+
+head(sig,10) %>%
   kbl(caption="lowest basemean genes") %>%
@@
 
-mx2 <- mx1[which(rowMeans(mx1)>=10),]
-dim(mx2)
-dds <- DESeqDataSetFromMatrix(countData = mx2 , colData = ss1, design = ~ vpa )
+mxf <- mx[which(rowMeans(mx)>=10),]
+dim(mxf)
+dds <- DESeqDataSetFromMatrix(countData = mxf , colData = ss, design = ~ trt )
 res <- DESeq(dds)
-z<- results(res)
+z <- results(res)
 vsd <- vst(dds, blind=FALSE)
-zz<-cbind(as.data.frame(z),mx2)
-dge2<-as.data.frame(zz[order(zz$pvalue),])
+zz <-cbind(as.data.frame(z),mxf)
+def <-as.data.frame(zz[order(zz$pvalue),])
 
-head(dge2,10) %>%
-  kbl(caption="Top DE genes for VPA treatment (filtered)") %>%
+head(def,10) %>%
+  kbl(caption="Top DE genes for Aza treatment (filtered)") %>%
   kable_paper("hover", full_width = F)
 
-sig2 <- subset(dge2,padj<=0.05)
-nrow(sig2)
+sigf <- subset(def,padj<=0.05)
 
-plot(log10(dge2$baseMean),dge2$log2FoldChange,
+DET=nrow(mxf)
+NSIG=nrow(sigf)
+NUP=nrow(subset(sigf,log2FoldChange>0))
+NDN=nrow(subset(sigf,log2FoldChange<0))
+
+HEADER=paste(DET,"detected genes;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")
+
+plot(log10(def$baseMean) ,def$log2FoldChange,
   cex=0.6,pch=19,col="darkgray",
-  xlab="log10(base mean)",ylab="log2 fold change",
-  main="strict filter")
+  main="strict filter",
+  xlab="log10(basemean)",ylab="log2(fold change)")
 
-points(log10(sig2$baseMean),sig2$log2FoldChange,
+points(log10(sigf$baseMean) ,sigf$log2FoldChange,
   cex=0.6,pch=19,col="red")
 
-dge2up <- rownames(subset(dge2,padj<=0.05 & log2FoldChange>0))
-dge2up <- unique(sapply(strsplit(dge2up," "),"[[",2))
+mtext(HEADER)
 
-dge2dn <- rownames(subset(dge2,padj<=0.05 & log2FoldChange<0))
-dge2dn <- unique(sapply(strsplit(dge2dn," "),"[[",2))
+defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
+defup <- unique(sapply(strsplit(defup," "),"[[",2))
 
-dge2bg <- rownames(dge2)
-dge2bg <- unique(sapply(strsplit(dge2bg," "),"[[",2))
+defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
+defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
+
+bg <- rownames(def)
+bg <- unique(sapply(strsplit(bg," "),"[[",2))
 
@@
 
-ora2_up <- as.data.frame(enricher(gene = dge2up , 
-  universe = dge2bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+ora2_up <- as.data.frame(enricher(gene = defup ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
@@
 ora2_up$geneID <- NULL
-ora2_ups <- rownames(subset(ora2_up,p.adjust<0.05))
+ora2_up <- subset(ora2_up,p.adjust<0.05 & Count >=10)
+ora2_ups <- rownames(ora2_up)
 
@@
 ora2_up <- ora2_up[order(-ora2_up$es),]
-head(ora2_up)
+ora2_up$Description=NULL
+
+head(ora2_up) %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment (filtered)") %>%
+  kable_paper("hover", full_width = F)
 
@@
 
-ora2_dn <- as.data.frame(enricher(gene = dge2dn ,
-  universe = dge2bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+ora2_dn <- as.data.frame(enricher(gene = defdn ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
@@
 ora2_dn$geneID <- NULL
-ora2_dns <- rownames(subset(ora2_dn,p.adjust<0.05))
+ora2_dn <- subset(ora2_dn,p.adjust<0.05 & Count >=10)
+ora2_dns <- rownames(ora2_dn)
 
@@
 ora2_dn <- ora2_dn[order(-ora2_dn$es),]
-head(ora2_dn)
+ora2_dn$Description=NULL
+
+head(ora2_dn) %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment (filtered)") %>%
+  kable_paper("hover", full_width = F)
 
@@
 
-ora1_up <- as.data.frame(enricher(gene = dge1up ,
-  universe = dge1bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+ora1_up <- as.data.frame(enricher(gene = defup ,
+  universe = all,  maxGSSize = 5000, TERM2GENE = genesets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
@@
 ora1_up$geneID <- NULL
-ora1_ups <- rownames(subset(ora1_up,p.adjust<0.05))
+ora1_up <- subset(ora1_up,p.adjust<0.05 & Count >=10)
+ora1_ups <- rownames(ora1_up)
 
@@
 ora1_up <- ora1_up[order(-ora1_up$es),]
-head(ora1_up)
+ora1_up$Description=NULL
+
+head(ora1_up) %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment (unfiltered)") %>%
+  kable_paper("hover", full_width = F)
 
@@
 
-ora1_dn <- as.data.frame(enricher(gene = dge1dn ,
-  universe = dge1bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+ora1_dn <- as.data.frame(enricher(gene = defdn ,
+  universe = all,  maxGSSize = 5000, TERM2GENE = genesets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
@@
 ora1_dn$geneID <- NULL
-ora1_dns <- rownames(subset(ora1_dn,p.adjust<0.05))
+ora1_dn <- subset(ora1_dn,p.adjust<0.05 & Count >=10)
+ora1_dns <- rownames(ora1_dn)
 
@@
 ora1_dn <- ora1_dn[order(-ora1_dn$es),]
-head(ora1_dn)
+ora1_dn$Description=NULL
+
+head(ora1_dn) %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment (unfiltered)") %>%
+  kable_paper("hover", full_width = F)
 
@@
 
-```{r,venn1}
+```{r,venn1,fig.width=6,fig.height=6}
 
@@
 plot(euler(v2),quantities = TRUE)
-dev.off()
+
+```
+
+Jaccard index
+
+```{r,jaccard}
+
+jaccard <- function(a, b) {
+    intersection = length(intersect(a, b))
+    union = length(a) + length(b) - intersection
+    return (intersection/union)
+}
+
+correct <- c(ora2_ups,ora2_dns)
+incorrect <- c(ora1_ups,ora1_dns)
+
+jaccard(correct,incorrect)
 
@@
 ## Session information
+
+For reproducibility
+
+<br><details><summary><b>Click HERE to show session info</b></summary><br><p>
 
@@
 ```
+
+</details>

### 2023-06-20 example_fixbg.Rmd
clusterprofiler bug

---
title: "Examining background gene lists in clusterProfiler"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/enrichment_recipe

## Introduction

This guide is a Rmarkdown script that conducts differential expression and enrichment analysis,
which are very popular workflows for transcriptome data.

The goal of this work is to understand how ClusterProfiler manages the background list,
as we have observed some weird behaviour.
In particular we see that when we provide a custom background, those genes do not appear in the
final analysis unless they are also included in the gene list.

In the code chunk below called `libs`, you can add and remove required R library dependancies.
Check that the libraries listed here match the Dockerfile, otherwise you might get errors.

```{r,libs}

suppressPackageStartupMessages({
  library("getDEE2")
  library("DESeq2")
  library("kableExtra")
  library("clusterProfiler")
  library("fgsea")
  library("eulerr")
  library("gplots")
})

```

For this guide I will be using bulk RNA-seq data from a previous study, which is
deposited at NCBI GEO and SRA under accession numbers: GSE55123 and SRP038101 (Lund et al, 2014).
The experiment is designed to investigate the effect of Azacitidine treatment on AML3 cells.

The raw data have been processed by the DEE2 project, and the summary gene expression counts are
available at the dee2.io website, and programmatically with the getDEE2 bioconductor package
(Ziemann et al, 2019).

Alternatively, you could fetch data from another resource like NCBI GEO, Zenodo or from the host
storage drive.

```{r,fetch}

mdat <- getDEE2Metadata("hsapiens")

# get sample sheet
ss <- subset(mdat,SRP_accession=="SRP038101")

# fetch the whole set of RNA-seq data
x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
mx <- x$GeneCounts
rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)
dim(mx)

# aza no filtering
ss$trt <- grepl("Treated",ss$Experiment_title)

ss %>%
  kbl(caption="sample sheet for Aza treatment in AML3 cells") %>%
  kable_paper("hover", full_width = F)

```

## Data quality control

QC is important, even if you are using public transcriptome data.
For RNA-seq it is a good idea to show the number of reads for each sample.

```{r,counts1}

par(mar=c(5,7,5,1))
barplot(rev(colSums(mx)),horiz=TRUE,las=1,main="number of reads per sample in SRP038101")

```

Now make a MDS plot.

```{r,mds1}

mds <- cmdscale(dist(t(mx)))

# expand plot area
XMIN=min(mds[,1])*1.3
XMAX=max(mds[,1])*1.3
YMIN=min(mds[,2])*1.3
YMAX=max(mds[,2])*1.3

cols <- as.character(grepl("Treated",ss$Experiment_title))
cols <- gsub("FALSE","lightblue",cols)
cols <- gsub("TRUE","pink",cols)
plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(XMIN,XMAX),ylim=c(YMIN,YMAX),
  pch=19,cex=2,col=cols, main="MDS plot")
text(cmdscale(dist(t(mx))), labels=colnames(mx) )

```

## Differential expression analysis

I will be using DESeq2 for DE analysis, the count matrix is prefiltered using
a detection threshold of 10 reads per sample across all samples.
All genes that meet the detection threshold will comprise the background list.
The first 6 rows of the count matrix are shows.

```{r,de1}

mxf <- mx[which(rowMeans(mx)>=10),]
dim(mxf)

head(mxf,6) %>%
  kbl(caption="Count matrix format") %>%
  kable_paper("hover", full_width = F)

dds <- DESeqDataSetFromMatrix(countData = mxf , colData = ss, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <-cbind(as.data.frame(z),mxf)
def <-as.data.frame(zz[order(zz$pvalue),])

head(def,10) %>%
  kbl(caption="Top DE genes for Aza treatment") %>%
  kable_paper("hover", full_width = F)

```

Make a smear plot.

```{r,smearplot}

sigf <- subset(def,padj<=0.05)

DET=nrow(mxf)
NSIG=nrow(sigf)
NUP=nrow(subset(sigf,log2FoldChange>0))
NDN=nrow(subset(sigf,log2FoldChange<0))

HEADER=paste(DET,"detected genes;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")

plot(log10(def$baseMean) ,def$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  main="5-azacitidine treatment in AML3 cells",
  xlab="log10(basemean)",ylab="log2(fold change)")

points(log10(sigf$baseMean) ,sigf$log2FoldChange,
  cex=0.6,pch=19,col="red")

mtext(HEADER)

```

In the next sections I will run enrichment analysis with over-representation analysis (ORA) test
and compare it to functional class scoring.
I will also investigate some strange behviour of the ORA tool clusterprofiler.

## ORA with Clusterprofiler custom background with otherwise default analysis

I've compiled a reporting checklist:

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | via DEE2 at accession SRP038101 (human) |
| Other parameters | Min gene set size of 10 |

Here I provide a custom background gene list.

Get the gene sets loaded in R.

```{r,gs}
# from https://reactome.org/download/current/ReactomePathways.gmt.zip
genesets2 <- read.gmt("ReactomePathways_2023-03-06.gmt")
gsets <- gmtPathways("ReactomePathways_2023-03-06.gmt")

message(paste("number of genes described in the annotation set:",length(unique(genesets2$gene))))

```

Now filter the gene names into three lists, up-regulated, down-regulated and
background.
Background is simply all genes that were detected.

```{r,subset1}

defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
defup <- unique(sapply(strsplit(defup," "),"[[",2))

defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
defdn <- unique(sapply(strsplit(defdn," "),"[[",2))

bg <- rownames(def)
bg <- unique(sapply(strsplit(bg," "),"[[",2))

message(paste("number of genes in background:",length(bg)))

```

Enrichment firstly with upregulated genes

```{r,oraup}

ora_up <- as.data.frame(enricher(gene = defup ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora_up$geneID <- NULL
ora_up <- subset(ora_up,p.adjust<0.05 & Count >=10)
ora_ups <- rownames(ora_up)

gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))

ora_up$es <- gr/br
ora_up <- ora_up[order(-ora_up$es),]
ora_up$Description=NULL

head(ora_up) %>%
  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment") %>%
  kable_paper("hover", full_width = F)

topup2 <- rev(head(ora_up$es,10))
names(topup2) <- rev(head(ora_up$ID,10))

```

The reported background size does not match the dataset.

Now repeat with the downregulated genes

```{r,oradn}

n_pw=10

ora_dn <- as.data.frame(enricher(gene = defdn ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora_dn$geneID <- NULL
ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
ora_dns <- rownames(ora_dn)

gr <- as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",2))

ora_dn$es <- gr/br
ora_dn <- ora_dn[order(-ora_dn$es),]
ora_dn$Description=NULL

head(ora_dn,n_pw) %>%
  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment") %>%
  kable_paper("hover", full_width = F)

topdn2 <- head(ora_dn$es,n_pw)
names(topdn2) <- head(ora_dn$ID,n_pw)

```

Make a barplot

```{r,topbarplot2}

par(mar=c(5,20,5,1))

cols <- c(rep("blue",n_pw),rep("red",n_pw))

barplot(c(topdn2,topup2),
  horiz=TRUE,las=1,cex.names=0.65,col=cols,
  main="top DE Reactomes",
  xlab="ES",
  cex.main=0.9)

mtext("ORA test")

```

## ORA with Clusterprofiler with custom background and special gene set list to fix the potential bug

I've compiled a reporting checklist:

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | via DEE2 at accession SRP038101 (human) |
| Other parameters | Min gene set size of 10 |

Here I provide a background gene list for cluterprofiler to use, like a user should.

I have also done some modification to the Reactome gene sets, to en

Now filter the gene names into three lists, up-regulated, down-regulated and
background.
Background is simply all genes that were detected.

```{r,subsetfix}

defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
defup <- unique(sapply(strsplit(defup," "),"[[",2))

defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
defdn <- unique(sapply(strsplit(defdn," "),"[[",2))

bg <- rownames(def)
bg <- unique(sapply(strsplit(bg," "),"[[",2))
message(paste("number of genes in background:",length(bg)))

```

Adding all detected genes to the background appears to improve results!

```{r,gs2fix}

bgdf <- data.frame("background",bg)
colnames(bgdf) <- c("term","gene")
genesets2 <- rbind(genesets2,bgdf)

```

Enrichment firstly with upregulated genes

```{r,oraupfix}

orafix_up <- as.data.frame(enricher(gene = defup ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

orafix_up$geneID <- NULL
orafix_up <- subset(orafix_up,p.adjust<0.05 & Count >=10)
orafix_ups <- rownames(orafix_up)

gr <- as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",2))

orafix_up$es <- gr/br
orafix_up <- orafix_up[order(-orafix_up$es),]
orafix_up$Description=NULL

head(orafix_up) %>%
  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment") %>%
  kable_paper("hover", full_width = F)

topup2 <- rev(head(orafix_up$es,10))
names(topup2) <- rev(head(orafix_up$ID,10))

```

Now with the downregulated genes

```{r,oradnfix}

n_pw=10

orafix_dn <- as.data.frame(enricher(gene = defdn ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

orafix_dn$geneID <- NULL
orafix_dn <- subset(orafix_dn,p.adjust<0.05 & Count >=10)
orafix_dns <- rownames(orafix_dn)

gr <- as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",2))

orafix_dn$es <- gr/br
orafix_dn <- orafix_dn[order(-orafix_dn$es),]
orafix_dn$Description=NULL

head(orafix_dn,n_pw) %>%
  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment") %>%
  kable_paper("hover", full_width = F)

topdn2 <- head(orafix_dn$es,n_pw)
names(topdn2) <- head(orafix_dn$ID,n_pw)

```

Make a barplot

```{r,topbarplot2fix}

par(mar=c(5,20,5,1))

cols <- c(rep("blue",n_pw),rep("red",n_pw))

barplot(c(topdn2,topup2),
  horiz=TRUE,las=1,cex.names=0.65,col=cols,
  main="top DE Reactomes",
  xlab="ES",
  cex.main=0.9)

mtext("ORA fix test")

```

## Enrichment analysis with functional class scoring

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | FGSEA (check version at foot of report) |
| Statistical test used | Kolmogorov-Smirnov test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | via DEE2 at accession SRP038101 (human) |
| Other parameters | Min gene set size of 10 |

```{r,fgsea}

def2 <- def
def2$genename <- sapply(strsplit(rownames(def2)," "),"[[",2)
def2 <- def2[,c("stat","genename")]
def2 <- aggregate(. ~ genename, def2, sum)
stat <- def2$stat
names(stat) <- def2$genename
fgseaRes <- fgsea(pathways=gsets, stats=stat, minSize=10, nPermSimple = 10000)
fgseaRes <- fgseaRes[order(fgseaRes$pval),]

```

```{r,fgsea_sig}

fgsea_up <- subset(fgseaRes,padj<0.05 & ES>0)
fgsea_ups <- fgsea_up$pathway
fgsea_dn <- subset(fgseaRes,padj<0.05 & ES<0)
fgsea_dns <- fgsea_dn$pathway

fgsea_up <-  data.frame(fgsea_up[order(-fgsea_up$ES),])

fgsea_dn <-  data.frame(fgsea_dn[order(fgsea_dn$ES),])

head(fgsea_up,n_pw) %>%
  kbl(row.names = FALSE, caption="FGSEA:Top upregulated pathways in Aza treatment") %>%
  kable_paper("hover", full_width = F)

head(fgsea_dn,n_pw) %>%
  kbl(row.names = FALSE, caption="FGSEA:Top downregulated pathways in Aza treatment") %>%
  kable_paper("hover", full_width = F)

fgsea_up <- head(fgsea_up,n_pw)
fgsea_up_vec <- fgsea_up$ES
names(fgsea_up_vec) <- fgsea_up$pathway
fgsea_up_vec <- rev(fgsea_up_vec)

fgsea_dn <- head(fgsea_dn,n_pw)
fgsea_dn_vec <- fgsea_dn$ES * -1
names(fgsea_dn_vec) <- fgsea_dn$pathway

```

Make a barplot

```{r,topbarplot3}

par(mar=c(5,20,5,1))

cols <- c(rep("blue",n_pw),rep("red",n_pw))

barplot(c(fgsea_dn_vec,fgsea_up_vec),
  horiz=TRUE,las=1,cex.names=0.65,col=cols,
  main="top DE Reactomes",
  xlab="ES",
  cex.main=0.9)

mtext("FCS test")

```

## Compare pathway results

```{r,venn1,fig.width=8,fig.height=5}

v1 <- list("ORA up"=ora_ups, "ORA fix up"=orafix_ups, "FCS up"=fgsea_ups)
v2 <- list("ORA dn"=ora_dns, "ORA fix dn"=orafix_dns, "FCS dn"=fgsea_dns)
v3 <- list("ORA"=union(ora_ups,ora_dns),
  "ORA fix"=union(orafix_ups,orafix_dns),
  "FCS"=union(fgsea_ups,fgsea_dns))

par(mar=c(10,10,10,10))
par(mfrow=c(2,1))
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="upregulated genes")
plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="downregulated genes")

plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated genes")

```

Jaccard index comparing ORA and FCS.

```{r,jaccard1}

jaccard <- function(a, b) {
    intersection = length(intersect(a, b))
    union = length(a) + length(b) - intersection
    return (intersection/union)
}

ora <- c(ora_ups,ora_dns)
fcs <- c(fgsea_ups,fgsea_dns)

jaccard(ora,fcs)

```

Jaccard index comparing ORA and ORA fix.

```{r,jaccard2}

ora <- c(ora_ups,ora_dns)
orafix <- c(orafix_ups,orafix_dns)

jaccard(ora,orafix)

```

Jaccard index comparing ORA fix and FCS.

```{r,jaccard3}

orafix <- c(orafix_ups,orafix_dns)
fcs <- c(fgsea_ups,fgsea_dns)

jaccard(orafix,fcs)

```

## Drill in to a specific pathway

In this case the top upregulated and downregulated pathways

```{r,eplot}

upname <- head(fgsea_up,1)$pathway
plotEnrichment(gsets[[upname]], stat, gseaParam = 1, ticksSize = 0.2)

dnname <- head(fgsea_dn,1)$pathway
plotEnrichment(gsets[[dnname]], stat, gseaParam = 1, ticksSize = 0.2)

```

Now a heatmap of these gene sets.

```{r,heat}

# reads per million normalisation
rpm <- apply(mxf,2, function(x) { x/sum(x) * 1000000} )
colnames(rpm) <- c("UNTR1","UNTR2","UNTR3","TRT1","TRT2","TRT3")
gnames_up <- gsets[[which(names(gsets) == upname)]]
gnames_dn <- gsets[[which(names(gsets) == dnname)]]
gene_ids <- rownames(mxf)
gene_names <- sapply(strsplit(gene_ids," "),"[[",2)
rpm_up <- rpm[which(gene_names %in% gnames_up),]
rownames(rpm_up) <- sapply(strsplit(rownames(rpm_up)," "),"[[",2)
rpm_dn <- rpm[which(gene_names %in% gnames_dn),]
rownames(rpm_dn) <- sapply(strsplit(rownames(rpm_dn)," "),"[[",2)
colsidecols <- c("blue","blue","blue","red","red","red")
heatmap.2(rpm_up,scale="row",trace="none",margins=c(10,15),main=upname,ColSideColors=colsidecols)
heatmap.2(rpm_dn,scale="row",trace="none",margins=c(10,15),main=dnname,ColSideColors=colsidecols)

```

## Session information

For reproducibility

<br><details><summary><b>Click HERE to show session info</b></summary><br><p>

```{r,session}

sessionInfo()

```

</details>

### 2023-06-20 example_fixbg.Rmd
look at DAVID data

@@
 ```{r,subset1}
+
+defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
+defup <- unique(sapply(strsplit(defup," "),"[[",1))
+
+defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
+defdn <- unique(sapply(strsplit(defdn," "),"[[",1))
+
+bg <- rownames(def)
+bg <- unique(sapply(strsplit(bg," "),"[[",1))
+
+# write ensembl IDs for DAVID 
+writeLines(text=bg,con="ebg.txt")
+writeLines(text=defup,con="eup.txt")
+writeLines(text=defdn,con="edn.txt")
 

### 2023-07-08 manuscript/manuscript.Rmd
start paper

---
title: "Importance of using the correct background list for over-representation analysis"
author: "Mark Ziemann and Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: references.bib
csl: plos-computational-biology.csl
---

Alternate title: *Problems in over-representation analysis software*

[Mark Ziemann<sup>1</sup>*](https://orcid.org/0000-0002-7688-6974),
[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352)

**Affiliations**

1. Deakin University, Geelong, Australia, School of Life and Environmental Sciences.

2. Vellore Institute of Technology, Vellore, India.

(*) Corresponding author: m.ziemann@deakin.edu.au

## Abstract

* Over-representation analysis (ORA) is the most common form of functional enrichment and one of the
most used techniques in computational biology, yet there are common methodological issues that
undermine its validity.

* These problems are typically ascribed to errors in analysis and reporting by the end-users,
however software design is another contributor.

* Here, we identify and characterise a potential bug popular ORA software
package and web tool.

* We show that this bug deleteriously affects the results of ORA tests, and we present a simple
work-around to circumvent it.

## Introduction

```{r,libs,echo=FALSE}

library("DiagrammeR")

```

Functional enrichment analysis is a family of computational approaches used to interpret large omics
data.
The most commonly used method is over-representation analysis (ORA) which involves the selection of
genes of interest, followed by a statistical test to evaluate whether genes of certain annotated
categories are over-represented in the set of interest.
The set of interest may be defined in different ways, such as genes that meet a differential expression
significance threshold, or those that have been clustered into a group based on co-expression patterns.
There are various sources of annotated categories, including gene ontologies, protein-coding features,
gene co-expression and signalling or metabolic pathways.
Providers of such annotations include but are not limited to the gene ontology consortium, Reactome,
KEGG, WikiPathways and MSigDB.

Central to ORA tests is the selection of a list of background genes.
This is also sometimes called the "universe" or the "reference" list.
These are defined as any gene that is detected robustly, and has a chance at making the set of interest.
For RNA-seq studies, this could be a threshold detection limit based either on raw counts or counts per
million.

The first use of ORA-based functional enrichment that we are aware of is a 1999 report that used the
hypergeometric test to determine the chance probability of observing over-representation of certain
annotated functions in co-regulated genes in yeast [@Tavazoie1999-kf].
This general approach is the basis of most ORA-based significance testing and can be represented by the
following equation:

$$
P=1-\sum_{i=0}^{k-1}\frac{ {f \choose i} {g-f \choose n-1} }{g \choose n}
$$

Where (P) is the probability of observing at least (k) genes from a functional category within the
list of selected genes with size (n).
Here, (g) is the total number of detected genes, and (f) is the number of detected genes belonging to a
functional category.

In a hypothetical example, there were 134 genes classified as detected.
Of these, 38 were defined as upregulated in response to a stimulus based on a differential expression
test.
From the 38, 8 genes belonged to the "cytokine signaling pathway" category.
This category contains 26 genes in total, but only 18 were detected in this study.

So the background can be defined as 18/134 and the foreground as 8/38 and the enrichment score
expressed as (8/38)/(18/135)=1.58

In this example:

*k*=8

*n*=26

*g*=134

*f*=18

We can calculate the probability of getting exactly 7 members of the set.

$$
P=\frac{ {18 \choose 7} {134-18 \choose 26-1} }{134 \choose 26}
$$

$$
P=\frac{ 480700 * 1.71e29 }{ 3.84e+33 } = 1/21.4 = 0.047
$$

This needs to be repeated for other values of *i*, namely 6 to 0 and then sum all of these.

```{r,hyperdist}
res<-unlist(lapply(0:38, function(i) {  (choose(18,i) * choose((134-18),(38-1)) / choose(134,38)) }))
plot(res,type="b")
```

## Acknowledgements

This research was supported by use of the Nectar Research Cloud, a collaborative Australian research
platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).

## Availability of materials

* Code repository:

* Example Docker image:

## Bibliography

### 2023-07-09 manuscript/manuscript.Rmd
intro work

@@
 title: "Importance of using the correct background list for over-representation analysis"
-author: "Mark Ziemann and Anusuiya Bora"
+author: "Mark Ziemann, Barry Schroeter, Anusuiya Bora"
 date: "`r Sys.Date()`"
@@
 [Mark Ziemann<sup>1</sup>*](https://orcid.org/0000-0002-7688-6974),
+[Barry Schroeter<sup>1</sup>]()
 [Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352)
@@
 
+The fold enrichment score (ES) is commonly used to express the degree of enrichment and is a ratio-of-ratios.
+The foreground ratio is the number of selected genes belonging a functional class (A) divided by the
+number of detected genes in the functional class (B).
+The background ratio is the number of selected genes not belonging to the functional class (C) divided
+by the number of detected genes not in the functional class (D).
+
+$$
+ES=\frac{ A / B}{ C / D}
+$$
+
+Both equations illustrate that correct definition of the number of background genes is important to
+obtain accurate enrichment scores and significance values.
+A common error in ORA of differential gene expression is to use all genes in the genome as the
+background @Wijesooriya2022-li.
+This leads to a severe over-estimation of the number of background genes as there are ~45k annotated
+genes in the human genome (according to Ensembl v109), but RNA-seq robustly detects only ~15k genes.
+This leads to inflation of p-values and results that share only ~40% similarity with properly conducted
+ORA analysis.
+This issue is even more severe for other omics, where the number of detected genes is lower, for example
+proteomics and single cell RNA-seq where there are only a few thousand detected analytes.
+To address this issue, most major ORA tools allow for a user-provided background list of detected genes.
+
+While testing some of these tools, we observed an unusual behaviour whereby the number of background
+genes reported by the tool varied significantly from the length of the gene list we provided.
+Moreover, this number also varied when we used a different library of gene function annotations.
+Puzzled by this behaviour, we browsed the computational biology forums and were met with other confused
+users wondering why the number of genes in the background list would vary like this.
+The tool developers kindly clarified the situation, by stating that the tool removed genes from the
+entire analysis if they were not members of a functional category.
+We thought this behaviour odd, because it has the potential to significantly alter the test being
+conducted, away from it's original application.
+
+In this paper, we examine the effect of this filtering step on ORA of real RNA-seq data, use simulated
+RNA-seq data to determine the impact upon precision and recall, and provide practical recommendations
+to end users to obtain the most accurate results.
+
 In a hypothetical example, there were 134 genes classified as detected.

### 2023-07-23 analysis/example_proteomics.Rmd
starting

--- 2023-06-20 example_fixbg.Rmd
@@
 ---
-title: "Examining background gene lists in clusterProfiler"
+title: "Examining background gene lists in clusterProfiler - proteomics example"
 author: "Mark Ziemann & Anusuiya Bora"
@@
 
-For this guide I will be using bulk RNA-seq data from a previous study, which is
-deposited at NCBI GEO and SRA under accession numbers: GSE55123 and SRP038101 (Lund et al, 2014).
-The experiment is designed to investigate the effect of Azacitidine treatment on AML3 cells.
-
-The raw data have been processed by the DEE2 project, and the summary gene expression counts are
-available at the dee2.io website, and programmatically with the getDEE2 bioconductor package
-(Ziemann et al, 2019).
-
-Alternatively, you could fetch data from another resource like NCBI GEO, Zenodo or from the host
-storage drive.
+For this report I will be using proteomics data from Mulvey et al 2021 (PMID: 34599159),
+who looked at the effect of LPS on THP-1 cell lines.
+Although in this paper they looked at numerous timepoints, I will be looking only at the 24hr time
+point where the number of significantly differentially expressed proteins was the greatest.
+The data were obtained from Supplementary Table 1 and saved to the /data folder.
 
@@
 
-mdat <- getDEE2Metadata("hsapiens")
-
-# get sample sheet
-ss <- subset(mdat,SRP_accession=="SRP038101")
-
-# fetch the whole set of RNA-seq data
-x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
-mx <- x$GeneCounts
-rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)
-dim(mx)
-
-# aza no filtering
-ss$trt <- grepl("Treated",ss$Experiment_title)
-
-ss %>%
-  kbl(caption="sample sheet for Aza treatment in AML3 cells") %>%
-  kable_paper("hover", full_width = F)
+def <- read.table("../data/lps_24h.tsv",sep="\t",header=TRUE)
 
@@
 
-## Data quality control
-
-QC is important, even if you are using public transcriptome data.
-For RNA-seq it is a good idea to show the number of reads for each sample.
-
-```{r,counts1}
-
-par(mar=c(5,7,5,1))
-barplot(rev(colSums(mx)),horiz=TRUE,las=1,main="number of reads per sample in SRP038101")
-
-```
-
-Now make a MDS plot.
-
-```{r,mds1}
-
-mds <- cmdscale(dist(t(mx)))
-
-# expand plot area
-XMIN=min(mds[,1])*1.3
-XMAX=max(mds[,1])*1.3
-YMIN=min(mds[,2])*1.3
-YMAX=max(mds[,2])*1.3
-
-cols <- as.character(grepl("Treated",ss$Experiment_title))
-cols <- gsub("FALSE","lightblue",cols)
-cols <- gsub("TRUE","pink",cols)
-plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
-  xlim=c(XMIN,XMAX),ylim=c(YMIN,YMAX),
-  pch=19,cex=2,col=cols, main="MDS plot")
-text(cmdscale(dist(t(mx))), labels=colnames(mx) )
-
-```
-
-## Differential expression analysis
-
-I will be using DESeq2 for DE analysis, the count matrix is prefiltered using
-a detection threshold of 10 reads per sample across all samples.
-All genes that meet the detection threshold will comprise the background list.
-The first 6 rows of the count matrix are shows.
-
-```{r,de1}
-
-mxf <- mx[which(rowMeans(mx)>=10),]
-dim(mxf)
-
-head(mxf,6) %>%
-  kbl(caption="Count matrix format") %>%
-  kable_paper("hover", full_width = F)
-
-dds <- DESeqDataSetFromMatrix(countData = mxf , colData = ss, design = ~ trt )
-res <- DESeq(dds)
-z <- results(res)
-vsd <- vst(dds, blind=FALSE)
-zz <-cbind(as.data.frame(z),mxf)
-def <-as.data.frame(zz[order(zz$pvalue),])
-
-head(def,10) %>%
-  kbl(caption="Top DE genes for Aza treatment") %>%
-  kable_paper("hover", full_width = F)
-
-```
-
-Make a smear plot.
-
-```{r,smearplot}
-
-sigf <- subset(def,padj<=0.05)
-
-DET=nrow(mxf)
-NSIG=nrow(sigf)
-NUP=nrow(subset(sigf,log2FoldChange>0))
-NDN=nrow(subset(sigf,log2FoldChange<0))
-
-HEADER=paste(DET,"detected genes;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")
-
-plot(log10(def$baseMean) ,def$log2FoldChange,
-  cex=0.6,pch=19,col="darkgray",
-  main="5-azacitidine treatment in AML3 cells",
-  xlab="log10(basemean)",ylab="log2(fold change)")
-
-points(log10(sigf$baseMean) ,sigf$log2FoldChange,
-  cex=0.6,pch=19,col="red")
-
-mtext(HEADER)
-
-```
+# KEGG pathway analysis
 
@@
 I will also investigate some strange behviour of the ORA tool clusterprofiler.
+
+## ORA with Clusterprofiler custom background with otherwise default analysis
+
+I've compiled a reporting checklist:
+
+| Reporting criteria | Method/resource used |
+| --- | --- |
+| Origin of gene sets | KEGG (2023-06-16) |
+| Tool used | ClusterProfiler (check version at foot of report) |
+| Statistical test used | hypergeometric test |
+| P-value correction for multiple comparisons | FDR method |
+| Background list | Genes with >=10 reads per sample on average across all samples |
+| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| ORA directionality | Separate tests for up- and down-regulation |
+| Data availability | via DEE2 at accession SRP038101 (human) |
+| Other parameters | Min gene set size of 10 |
+
+Here I provide a custom background gene list.
+
+Get the gene sets loaded in R.
+These are KEGG gene sets
+
+```{r,gs1}
+# from https://www.gsea-msigdb.org/gsea/msigdb/human/collections.jsp 16th June 2023
+genesets2 <- read.gmt("../ref/c2.cp.kegg.v2023.1.Hs.symbols.gmt")
+gsets <- gmtPathways("../ref/c2.cp.kegg.v2023.1.Hs.symbols.gmt")
+
+message(paste("number of genes described in the annotation set:",length(unique(genesets2$gene))))
+
+```
+
+Now filter the gene names into three lists, up-regulated, down-regulated and
+background.
+Background is simply all genes that were detected.
+
+```{r,subset1}
+
+defup <- unique(subset(def,adj.P.Val<=0.05 & logFC>0)$GN)
+
+defdn <- unique(subset(def,adj.P.Val<=0.05 & logFC<0)$GN)
+
+bg <- unique(def$GN)
+
+message(paste("number of proteins in background:",length(bg)))
+
+```
+
+Enrichment firstly with upregulated genes
+
+```{r,oraup1}
+
+n_pw=10
+
+ora_up <- as.data.frame(enricher(gene = defup ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora_up$geneID <- NULL
+ora_up <- subset(ora_up,p.adjust<0.05 & Count >=10)
+ora_ups <- rownames(ora_up)
+
+gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))
+
+ora_up$es <- gr/br
+ora_up <- ora_up[order(-ora_up$es),]
+ora_up$Description=NULL
+
+head(ora_up) %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in LPS treatment") %>%
+  kable_paper("hover", full_width = F)
+
+topup2 <- rev(head(ora_up$es,10))
+names(topup2) <- rev(head(ora_up$ID,10))
+
+```
+
+The reported background size does not match the dataset.
+
+Now repeat with the downregulated genes
+
+```{r,oradn1}
+
+ora_dn <- as.data.frame(enricher(gene = defdn ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora_dn$geneID <- NULL
+ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
+ora_dns <- rownames(ora_dn)
+
+gr <- as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",2))
+
+ora_dn$es <- gr/br
+ora_dn <- ora_dn[order(-ora_dn$es),]
+ora_dn$Description=NULL
+
+head(ora_dn,n_pw) %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in LPS treatment") %>%
+  kable_paper("hover", full_width = F)
+
+topdn2 <- head(ora_dn$es,n_pw)
+names(topdn2) <- head(ora_dn$ID,n_pw)
+
+```
+
+Make a barplot
+
+```{r,topbarplot1}
+
+par(mar=c(5,20,5,1))
+
+cols <- c(rep("blue",n_pw),rep("red",n_pw))
+
+barplot(c(topdn2,topup2),
+  horiz=TRUE,las=1,cex.names=0.65,col=cols,
+  main="top DE KEGGs",
+  xlab="ES",
+  cex.main=0.9)
+
+mtext("ORA test")
+
+```
+
+## ORA with Clusterprofiler with custom background and special gene set list to fix the potential bug
+
+I've compiled a reporting checklist:
+
+| Reporting criteria | Method/resource used |
+| --- | --- |
+| Origin of gene sets | KEGG (2023-06-16) |
+| Tool used | ClusterProfiler (check version at foot of report) |
+| Statistical test used | hypergeometric test |
+| P-value correction for multiple comparisons | FDR method |
+| Background list | Genes with >=10 reads per sample on average across all samples |
+| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| ORA directionality | Separate tests for up- and down-regulation |
+| Data availability | via DEE2 at accession SRP038101 (human) |
+| Other parameters | Min gene set size of 10 + BUG FIX |
+
+Here I provide a background gene list for cluterprofiler to use, like a user should.
+
+I have also done some modification to the KEGG gene sets, to en
+
+Now filter the gene names into three lists, up-regulated, down-regulated and
+background.
+Background is simply all genes that were detected.
+
+Adding all detected genes to the background appears to improve results!
+
+```{r,gs2fix1}
+
+bgdf <- data.frame("background",bg)
+colnames(bgdf) <- c("term","gene")
+genesets2 <- rbind(genesets2,bgdf)
+
+```
+
+Enrichment firstly with upregulated genes
+
+```{r,oraupfix1}
+
+n_pw=10
+
+orafix_up <- as.data.frame(enricher(gene = defup ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+orafix_up$geneID <- NULL
+orafix_up <- subset(orafix_up,p.adjust<0.05 & Count >=10)
+orafix_ups <- rownames(orafix_up)
+
+gr <- as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",2))
+
+orafix_up$es <- gr/br
+orafix_up <- orafix_up[order(-orafix_up$es),]
+orafix_up$Description=NULL
+
+head(orafix_up) %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in LPS treatment") %>%
+  kable_paper("hover", full_width = F)
+
+topup2 <- rev(head(orafix_up$es,10))
+names(topup2) <- rev(head(orafix_up$ID,10))
+
+```
+
+Now with the downregulated genes
+
+```{r,oradnfix1}
+
+orafix_dn <- as.data.frame(enricher(gene = defdn ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+orafix_dn$geneID <- NULL
+orafix_dn <- subset(orafix_dn,p.adjust<0.05 & Count >=10)
+orafix_dns <- rownames(orafix_dn)
+
+gr <- as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",2))
+
+orafix_dn$es <- gr/br
+orafix_dn <- orafix_dn[order(-orafix_dn$es),]
+orafix_dn$Description=NULL
+
+head(orafix_dn,n_pw) %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in LPS treatment") %>%
+  kable_paper("hover", full_width = F)
+
+topdn2 <- head(orafix_dn$es,n_pw)
+names(topdn2) <- head(orafix_dn$ID,n_pw)
+
+```
+
+Make a barplot
+
+```{r,topbarplot2fix1}
+
+par(mar=c(5,20,5,1))
+
+cols <- c(rep("blue",n_pw),rep("red",n_pw))
+
+barplot(c(topdn2,topup2),
+  horiz=TRUE,las=1,cex.names=0.65,col=cols,
+  main="top DE KEGGs",
+  xlab="ES",
+  cex.main=0.9)
+
+mtext("ORA fix test")
+
+```
+
+## Compare pathway results
+
+```{r,venn1,fig.width=8,fig.height=5}
+
+v1 <- list("ORA up"=ora_ups, "ORA fix up"=orafix_ups)
+v2 <- list("ORA dn"=ora_dns, "ORA fix dn"=orafix_dns)
+v3 <- list("ORA"=union(ora_ups,ora_dns),
+  "ORA fix"=union(orafix_ups,orafix_dns))
+
+par(mar=c(10,10,10,10))
+par(mfrow=c(2,1))
+plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="upregulated KEGG pathways")
+plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="downregulated KEGG pathways")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
+
+```
+
+Jaccard index comparing ORA and ORA fix.
+
+```{r,jaccard1}
+
+jaccard <- function(a, b) {
+    intersection = length(intersect(a, b))
+    union = length(a) + length(b) - intersection
+    return (intersection/union)
+}
+
+ora <- c(ora_ups,ora_dns)
+orafix <- c(orafix_ups,orafix_dns)
+
+jaccard(ora,orafix)
+
+```
+
+List gene sets identified only in the "fixed" analysis.
+
+```{r,examine1}
+
+ora <- union(ora_ups,ora_dns)
+
+orafix <- union(orafix_ups,orafix_dns)
+
+setdiff(orafix,ora)
+
+```
+
+# Reactome Gene sets
 
@@
 Get the gene sets loaded in R.
+These are Reactome gene sets
 
-```{r,gs}
-# from https://reactome.org/download/current/ReactomePathways.gmt.zip
-genesets2 <- read.gmt("ReactomePathways_2023-03-06.gmt")
-gsets <- gmtPathways("ReactomePathways_2023-03-06.gmt")
+```{r,gs2}
+
+genesets2 <- read.gmt("../ref/ReactomePathways_2023-03-06.gmt")
+gsets <- gmtPathways("../ref/ReactomePathways_2023-03-06.gmt")
 
@@
 
-Now filter the gene names into three lists, up-regulated, down-regulated and
-background.
-Background is simply all genes that were detected.
-
-```{r,subset1}
-
-defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
-defup <- unique(sapply(strsplit(defup," "),"[[",2))
-
-defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
-defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
-
-bg <- rownames(def)
-bg <- unique(sapply(strsplit(bg," "),"[[",2))
-
-message(paste("number of genes in background:",length(bg)))
-
-```
-
 Enrichment firstly with upregulated genes
 
-```{r,oraup}
+```{r,oraup2}
+
+n_pw=10
 
@@
 head(ora_up) %>%
-  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment") %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in LPS treatment") %>%
   kable_paper("hover", full_width = F)
@@
 
-The reported background size does not match the dataset.
-
 Now repeat with the downregulated genes
 
-```{r,oradn}
-
-n_pw=10
+```{r,oradn2}
 
@@
 head(ora_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment") %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in LPS treatment") %>%
   kable_paper("hover", full_width = F)
@@
 
-```{r,topbarplot2}
+```{r,topbarplot22}
 
@@
 | Data availability | via DEE2 at accession SRP038101 (human) |
-| Other parameters | Min gene set size of 10 |
-
-Here I provide a background gene list for cluterprofiler to use, like a user should.
-
-I have also done some modification to the Reactome gene sets, to en
-
-Now filter the gene names into three lists, up-regulated, down-regulated and
-background.
-Background is simply all genes that were detected.
-
-```{r,subsetfix}
-
-defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
-defup <- unique(sapply(strsplit(defup," "),"[[",2))
-
-defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
-defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
-
-bg <- rownames(def)
-bg <- unique(sapply(strsplit(bg," "),"[[",2))
-message(paste("number of genes in background:",length(bg)))
-
-```
+| Other parameters | Min gene set size of 10 + BUG FIX |
 
@@
 
-```{r,gs2fix}
+```{r,gs2fix2}
 
@@
 
-```{r,oraupfix}
+```{r,oraupfix2}
 
@@
 head(orafix_up) %>%
-  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment") %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in LPS treatment") %>%
   kable_paper("hover", full_width = F)
@@
 
-```{r,oradnfix}
+```{r,oradnfix2}
 
@@
 head(orafix_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment") %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in LPS treatment") %>%
   kable_paper("hover", full_width = F)
@@
 
-```{r,topbarplot2fix}
+```{r,topbarplot2fix2}
 
@@
 
-## Enrichment analysis with functional class scoring
+## Compare pathway results
 
-| Reporting criteria | Method/resource used |
-| --- | --- |
-| Origin of gene sets | Reactome (2023-03-06) |
-| Tool used | FGSEA (check version at foot of report) |
-| Statistical test used | Kolmogorov-Smirnov test |
-| P-value correction for multiple comparisons | FDR method |
-| Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | DESeq2 FDR<0.05 |
-| ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | via DEE2 at accession SRP038101 (human) |
-| Other parameters | Min gene set size of 10 |
+```{r,venn2,fig.width=8,fig.height=5}
 
-```{r,fgsea}
+v1 <- list("ORA up"=ora_ups, "ORA fix up"=orafix_ups)
+v2 <- list("ORA dn"=ora_dns, "ORA fix dn"=orafix_dns)
+v3 <- list("ORA"=union(ora_ups,ora_dns),
+  "ORA fix"=union(orafix_ups,orafix_dns))
 
-def2 <- def
-def2$genename <- sapply(strsplit(rownames(def2)," "),"[[",2)
-def2 <- def2[,c("stat","genename")]
-def2 <- aggregate(. ~ genename, def2, sum)
-stat <- def2$stat
-names(stat) <- def2$genename
-fgseaRes <- fgsea(pathways=gsets, stats=stat, minSize=10, nPermSimple = 10000)
-fgseaRes <- fgseaRes[order(fgseaRes$pval),]
+par(mar=c(10,10,10,10))
+par(mfrow=c(2,1))
+plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="upregulated Reactomes")
+plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="downregulated Reactomes")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
 
@@
 
-```{r,fgsea_sig}
+Jaccard index comparing ORA and ORA fix.
 
-fgsea_up <- subset(fgseaRes,padj<0.05 & ES>0)
-fgsea_ups <- fgsea_up$pathway
-fgsea_dn <- subset(fgseaRes,padj<0.05 & ES<0)
-fgsea_dns <- fgsea_dn$pathway
-
-fgsea_up <-  data.frame(fgsea_up[order(-fgsea_up$ES),])
-
-fgsea_dn <-  data.frame(fgsea_dn[order(fgsea_dn$ES),])
-
-head(fgsea_up,n_pw) %>%
-  kbl(row.names = FALSE, caption="FGSEA:Top upregulated pathways in Aza treatment") %>%
-  kable_paper("hover", full_width = F)
-
-head(fgsea_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="FGSEA:Top downregulated pathways in Aza treatment") %>%
-  kable_paper("hover", full_width = F)
-
-fgsea_up <- head(fgsea_up,n_pw)
-fgsea_up_vec <- fgsea_up$ES
-names(fgsea_up_vec) <- fgsea_up$pathway
-fgsea_up_vec <- rev(fgsea_up_vec)
-
-fgsea_dn <- head(fgsea_dn,n_pw)
-fgsea_dn_vec <- fgsea_dn$ES * -1
-names(fgsea_dn_vec) <- fgsea_dn$pathway
-
-```
-
-Make a barplot
-
-```{r,topbarplot3}
-
-par(mar=c(5,20,5,1))
-
-cols <- c(rep("blue",n_pw),rep("red",n_pw))
-
-barplot(c(fgsea_dn_vec,fgsea_up_vec),
-  horiz=TRUE,las=1,cex.names=0.65,col=cols,
-  main="top DE Reactomes",
-  xlab="ES",
-  cex.main=0.9)
-
-mtext("FCS test")
-
-```
-
-## Compare pathway results
-
-```{r,venn1,fig.width=8,fig.height=5}
-
-v1 <- list("ORA up"=ora_ups, "ORA fix up"=orafix_ups, "FCS up"=fgsea_ups)
-v2 <- list("ORA dn"=ora_dns, "ORA fix dn"=orafix_dns, "FCS dn"=fgsea_dns)
-v3 <- list("ORA"=union(ora_ups,ora_dns),
-  "ORA fix"=union(orafix_ups,orafix_dns),
-  "FCS"=union(fgsea_ups,fgsea_dns))
-
-par(mar=c(10,10,10,10))
-par(mfrow=c(2,1))
-plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="upregulated genes")
-plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="downregulated genes")
-
-plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated genes")
-
-```
-
-Jaccard index comparing ORA and FCS.
-
-```{r,jaccard1}
+```{r,jaccard2}
 
@@
 ora <- c(ora_ups,ora_dns)
-fcs <- c(fgsea_ups,fgsea_dns)
-
-jaccard(ora,fcs)
-
-```
-
-Jaccard index comparing ORA and ORA fix.
-
-```{r,jaccard2}
-
-ora <- c(ora_ups,ora_dns)
 orafix <- c(orafix_ups,orafix_dns)
@@
 
-Jaccard index comparing ORA fix and FCS.
+List gene sets identified only in the "fixed" analysis.
 
-```{r,jaccard3}
+```{r,examine2}
 
-orafix <- c(orafix_ups,orafix_dns)
-fcs <- c(fgsea_ups,fgsea_dns)
+ora <- union(ora_ups,ora_dns)
 
-jaccard(orafix,fcs)
+orafix <- union(orafix_ups,orafix_dns)
 
-```
-
-## Drill in to a specific pathway
-
-In this case the top upregulated and downregulated pathways
-
-```{r,eplot}
-
-upname <- head(fgsea_up,1)$pathway
-plotEnrichment(gsets[[upname]], stat, gseaParam = 1, ticksSize = 0.2)
-
-dnname <- head(fgsea_dn,1)$pathway
-plotEnrichment(gsets[[dnname]], stat, gseaParam = 1, ticksSize = 0.2)
-
-```
-
-Now a heatmap of these gene sets.
-
-```{r,heat}
-
-# reads per million normalisation
-rpm <- apply(mxf,2, function(x) { x/sum(x) * 1000000} )
-colnames(rpm) <- c("UNTR1","UNTR2","UNTR3","TRT1","TRT2","TRT3")
-gnames_up <- gsets[[which(names(gsets) == upname)]]
-gnames_dn <- gsets[[which(names(gsets) == dnname)]]
-gene_ids <- rownames(mxf)
-gene_names <- sapply(strsplit(gene_ids," "),"[[",2)
-rpm_up <- rpm[which(gene_names %in% gnames_up),]
-rownames(rpm_up) <- sapply(strsplit(rownames(rpm_up)," "),"[[",2)
-rpm_dn <- rpm[which(gene_names %in% gnames_dn),]
-rownames(rpm_dn) <- sapply(strsplit(rownames(rpm_dn)," "),"[[",2)
-colsidecols <- c("blue","blue","blue","red","red","red")
-heatmap.2(rpm_up,scale="row",trace="none",margins=c(10,15),main=upname,ColSideColors=colsidecols)
-heatmap.2(rpm_dn,scale="row",trace="none",margins=c(10,15),main=dnname,ColSideColors=colsidecols)
+setdiff(orafix,ora)
 

### 2023-07-23 analysis/example_rnaseq.Rmd
starting

--- 2023-06-20 example_fixbg.Rmd
@@
 ---
-title: "Examining background gene lists in clusterProfiler"
+title: "Examining background gene lists in clusterProfiler - RNA-seq example"
 author: "Mark Ziemann & Anusuiya Bora"
@@
 
+```{r,writedata}
+
+write.table(def,file="../data/aza_rnaseq.tsv")
+
+```
+
+# KEGG pathway analysis
+
 In the next sections I will run enrichment analysis with over-representation analysis (ORA) test
@@
 | --- | --- |
-| Origin of gene sets | Reactome (2023-03-06) |
+| Origin of gene sets | KEGG (2023-06-16) |
 | Tool used | ClusterProfiler (check version at foot of report) |
@@
 Get the gene sets loaded in R.
+These are KEGG gene sets
 
-```{r,gs}
-# from https://reactome.org/download/current/ReactomePathways.gmt.zip
-genesets2 <- read.gmt("ReactomePathways_2023-03-06.gmt")
-gsets <- gmtPathways("ReactomePathways_2023-03-06.gmt")
+```{r,gs1}
+# from https://www.gsea-msigdb.org/gsea/msigdb/human/collections.jsp 16th June 2023
+genesets2 <- read.gmt("../ref/c2.cp.kegg.v2023.1.Hs.symbols.gmt")
+gsets <- gmtPathways("../ref/c2.cp.kegg.v2023.1.Hs.symbols.gmt")
 
@@
 
-```{r,oraup}
+```{r,oraup1}
+
+n_pw=10
 
@@
 
-```{r,oradn}
-
-n_pw=10
+```{r,oradn1}
 
@@
 
-```{r,topbarplot2}
+```{r,topbarplot1}
 
@@
   horiz=TRUE,las=1,cex.names=0.65,col=cols,
-  main="top DE Reactomes",
+  main="top DE KEGGs",
   xlab="ES",
@@
 | Data availability | via DEE2 at accession SRP038101 (human) |
-| Other parameters | Min gene set size of 10 |
+| Other parameters | Min gene set size of 10 + BUG FIX |
 
@@
 
-I have also done some modification to the Reactome gene sets, to en
+I have also done some modification to the KEGG gene sets, to en
 
@@
 
-```{r,subsetfix}
+```{r,subsetfix1}
 
@@
 
-```{r,gs2fix}
+```{r,gs2fix1}
 
@@
 
-```{r,oraupfix}
+```{r,oraupfix1}
+
+n_pw=10
 
@@
 
-```{r,oradnfix}
+```{r,oradnfix1}
+
+orafix_dn <- as.data.frame(enricher(gene = defdn ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+orafix_dn$geneID <- NULL
+orafix_dn <- subset(orafix_dn,p.adjust<0.05 & Count >=10)
+orafix_dns <- rownames(orafix_dn)
+
+gr <- as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",2))
+
+orafix_dn$es <- gr/br
+orafix_dn <- orafix_dn[order(-orafix_dn$es),]
+orafix_dn$Description=NULL
+
+head(orafix_dn,n_pw) %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment") %>%
+  kable_paper("hover", full_width = F)
+
+topdn2 <- head(orafix_dn$es,n_pw)
+names(topdn2) <- head(orafix_dn$ID,n_pw)
+
+```
+
+Make a barplot
+
+```{r,topbarplot2fix1}
+
+par(mar=c(5,20,5,1))
+
+cols <- c(rep("blue",n_pw),rep("red",n_pw))
+
+barplot(c(topdn2,topup2),
+  horiz=TRUE,las=1,cex.names=0.65,col=cols,
+  main="top DE KEGGs",
+  xlab="ES",
+  cex.main=0.9)
+
+mtext("ORA fix test")
+
+```
+
+## Compare pathway results
+
+```{r,venn1,fig.width=8,fig.height=5}
+
+v1 <- list("ORA up"=ora_ups, "ORA fix up"=orafix_ups)
+v2 <- list("ORA dn"=ora_dns, "ORA fix dn"=orafix_dns)
+v3 <- list("ORA"=union(ora_ups,ora_dns),
+  "ORA fix"=union(orafix_ups,orafix_dns))
+
+par(mar=c(10,10,10,10))
+par(mfrow=c(2,1))
+plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="upregulated KEGG pathways")
+plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="downregulated KEGG pathways")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
+
+```
+
+Jaccard index comparing ORA and ORA fix.
+
+```{r,jaccard1}
+
+jaccard <- function(a, b) {
+    intersection = length(intersect(a, b))
+    union = length(a) + length(b) - intersection
+    return (intersection/union)
+}
+
+ora <- c(ora_ups,ora_dns)
+orafix <- c(orafix_ups,orafix_dns)
+
+jaccard(ora,orafix)
+
+```
+
+List gene sets identified only in the "fixed" analysis.
+
+```{r,examine1}
+
+ora <- union(ora_ups,ora_dns)
+
+orafix <- union(orafix_ups,orafix_dns)
+
+setdiff(orafix,ora)
+
+```
+
+# Reactome Gene sets
+
+## ORA with Clusterprofiler custom background with otherwise default analysis
+
+I've compiled a reporting checklist:
+
+| Reporting criteria | Method/resource used |
+| --- | --- |
+| Origin of gene sets | Reactome (2023-03-06) |
+| Tool used | ClusterProfiler (check version at foot of report) |
+| Statistical test used | hypergeometric test |
+| P-value correction for multiple comparisons | FDR method |
+| Background list | Genes with >=10 reads per sample on average across all samples |
+| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| ORA directionality | Separate tests for up- and down-regulation |
+| Data availability | via DEE2 at accession SRP038101 (human) |
+| Other parameters | Min gene set size of 10 |
+
+Here I provide a custom background gene list.
+
+Get the gene sets loaded in R.
+These are Reactome gene sets
+
+```{r,gs2}
+
+genesets2 <- read.gmt("../ref/ReactomePathways_2023-03-06.gmt")
+gsets <- gmtPathways("../ref/ReactomePathways_2023-03-06.gmt")
+
+message(paste("number of genes described in the annotation set:",length(unique(genesets2$gene))))
+
+```
+
+Enrichment firstly with upregulated genes
+
+```{r,oraup2}
+
+n_pw=10
+
+ora_up <- as.data.frame(enricher(gene = defup ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora_up$geneID <- NULL
+ora_up <- subset(ora_up,p.adjust<0.05 & Count >=10)
+ora_ups <- rownames(ora_up)
+
+gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))
+
+ora_up$es <- gr/br
+ora_up <- ora_up[order(-ora_up$es),]
+ora_up$Description=NULL
+
+head(ora_up) %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment") %>%
+  kable_paper("hover", full_width = F)
+
+topup2 <- rev(head(ora_up$es,10))
+names(topup2) <- rev(head(ora_up$ID,10))
+
+```
+
+Now repeat with the downregulated genes
+
+```{r,oradn2}
+
+ora_dn <- as.data.frame(enricher(gene = defdn ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora_dn$geneID <- NULL
+ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
+ora_dns <- rownames(ora_dn)
+
+gr <- as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",2))
+
+ora_dn$es <- gr/br
+ora_dn <- ora_dn[order(-ora_dn$es),]
+ora_dn$Description=NULL
+
+head(ora_dn,n_pw) %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment") %>%
+  kable_paper("hover", full_width = F)
+
+topdn2 <- head(ora_dn$es,n_pw)
+names(topdn2) <- head(ora_dn$ID,n_pw)
+
+```
+
+Make a barplot
+
+```{r,topbarplot22}
+
+par(mar=c(5,20,5,1))
+
+cols <- c(rep("blue",n_pw),rep("red",n_pw))
+
+barplot(c(topdn2,topup2),
+  horiz=TRUE,las=1,cex.names=0.65,col=cols,
+  main="top DE Reactomes",
+  xlab="ES",
+  cex.main=0.9)
+
+mtext("ORA test")
+
+```
+
+## ORA with Clusterprofiler with custom background and special gene set list to fix the potential bug
+
+I've compiled a reporting checklist:
+
+| Reporting criteria | Method/resource used |
+| --- | --- |
+| Origin of gene sets | Reactome (2023-03-06) |
+| Tool used | ClusterProfiler (check version at foot of report) |
+| Statistical test used | hypergeometric test |
+| P-value correction for multiple comparisons | FDR method |
+| Background list | Genes with >=10 reads per sample on average across all samples |
+| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| ORA directionality | Separate tests for up- and down-regulation |
+| Data availability | via DEE2 at accession SRP038101 (human) |
+| Other parameters | Min gene set size of 10 + BUG FIX |
+
+Adding all detected genes to the background appears to improve results!
+
+```{r,gs2fix2}
+
+bgdf <- data.frame("background",bg)
+colnames(bgdf) <- c("term","gene")
+genesets2 <- rbind(genesets2,bgdf)
+
+```
+
+Enrichment firstly with upregulated genes
+
+```{r,oraupfix2}
+
+orafix_up <- as.data.frame(enricher(gene = defup ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+orafix_up$geneID <- NULL
+orafix_up <- subset(orafix_up,p.adjust<0.05 & Count >=10)
+orafix_ups <- rownames(orafix_up)
+
+gr <- as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",2))
+
+orafix_up$es <- gr/br
+orafix_up <- orafix_up[order(-orafix_up$es),]
+orafix_up$Description=NULL
+
+head(orafix_up) %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment") %>%
+  kable_paper("hover", full_width = F)
+
+topup2 <- rev(head(orafix_up$es,10))
+names(topup2) <- rev(head(orafix_up$ID,10))
+
+```
+
+Now with the downregulated genes
+
+```{r,oradnfix2}
 

### 2023-07-30 analysis/example_scrnaseq.Rmd
work

--- 2023-07-23 analysis/example_rnaseq.Rmd
@@
 ---
-title: "Examining background gene lists in clusterProfiler - RNA-seq example"
+title: "Examining background gene lists in clusterProfiler - scRNA-seq example"
 author: "Mark Ziemann & Anusuiya Bora"
@@
 
-This guide is a Rmarkdown script that conducts differential expression and enrichment analysis,
-which are very popular workflows for transcriptome data.
-
-The goal of this work is to understand how ClusterProfiler manages the background list,
-as we have observed some weird behaviour.
-In particular we see that when we provide a custom background, those genes do not appear in the
-final analysis unless they are also included in the gene list.
-
-In the code chunk below called `libs`, you can add and remove required R library dependancies.
-Check that the libraries listed here match the Dockerfile, otherwise you might get errors.
+Load libraries and dataset.
 
 ```{r,libs}
+
+load("muscat_vignette.RData")
+
+ls()
 
@@
 
-For this guide I will be using bulk RNA-seq data from a previous study, which is
-deposited at NCBI GEO and SRA under accession numbers: GSE55123 and SRP038101 (Lund et al, 2014).
-The experiment is designed to investigate the effect of Azacitidine treatment on AML3 cells.
+For this report I will use a SingleCellExperiment (SCE) containing 10x droplet-based
+scRNA-seq PBCM data from 8 Lupus patients obtained before and after 6h-treatment with IFN-β
+ (Kang et al. 2018).
 
-The raw data have been processed by the DEE2 project, and the summary gene expression counts are
-available at the dee2.io website, and programmatically with the getDEE2 bioconductor package
-(Ziemann et al, 2019).
+The complete raw data, as well as gene and cell metadata is available through the NCBI GEO,
+accession number GSE96583.
 
-Alternatively, you could fetch data from another resource like NCBI GEO, Zenodo or from the host
-storage drive.
+Downstream analysis was done according to the Muscat vignette.
+
+Focusing on CD4 T-cells.
 
@@
 
-mdat <- getDEE2Metadata("hsapiens")
+names(tbl)
 
-# get sample sheet
-ss <- subset(mdat,SRP_accession=="SRP038101")
+message("Total detected genes in each cluster and differentially expressed")
 
-# fetch the whole set of RNA-seq data
-x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
-mx <- x$GeneCounts
-rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)
-dim(mx)
+lapply(tbl, function(x) {
+  nsig <- nrow(subset(x,p_adj.loc<0.05))
+  ntot <- nrow(x)
+  c(ntot,nsig)
+} )
 
-# aza no filtering
-ss$trt <- grepl("Treated",ss$Experiment_title)
-
-ss %>%
-  kbl(caption="sample sheet for Aza treatment in AML3 cells") %>%
+head(tbl[["CD4 T cells"]])  %>%
+  kbl(caption="Example of muscat dataset. Control vs IFN-beta treatment") %>%
   kable_paper("hover", full_width = F)
 
-```
-
-## Data quality control
-
-QC is important, even if you are using public transcriptome data.
-For RNA-seq it is a good idea to show the number of reads for each sample.
-
-```{r,counts1}
-
-par(mar=c(5,7,5,1))
-barplot(rev(colSums(mx)),horiz=TRUE,las=1,main="number of reads per sample in SRP038101")
-
-```
-
-Now make a MDS plot.
-
-```{r,mds1}
-
-mds <- cmdscale(dist(t(mx)))
-
-# expand plot area
-XMIN=min(mds[,1])*1.3
-XMAX=max(mds[,1])*1.3
-YMIN=min(mds[,2])*1.3
-YMAX=max(mds[,2])*1.3
-
-cols <- as.character(grepl("Treated",ss$Experiment_title))
-cols <- gsub("FALSE","lightblue",cols)
-cols <- gsub("TRUE","pink",cols)
-plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
-  xlim=c(XMIN,XMAX),ylim=c(YMIN,YMAX),
-  pch=19,cex=2,col=cols, main="MDS plot")
-text(cmdscale(dist(t(mx))), labels=colnames(mx) )
-
-```
-
-## Differential expression analysis
-
-I will be using DESeq2 for DE analysis, the count matrix is prefiltered using
-a detection threshold of 10 reads per sample across all samples.
-All genes that meet the detection threshold will comprise the background list.
-The first 6 rows of the count matrix are shows.
-
-```{r,de1}
-
-mxf <- mx[which(rowMeans(mx)>=10),]
-dim(mxf)
-
-head(mxf,6) %>%
-  kbl(caption="Count matrix format") %>%
-  kable_paper("hover", full_width = F)
-
-dds <- DESeqDataSetFromMatrix(countData = mxf , colData = ss, design = ~ trt )
-res <- DESeq(dds)
-z <- results(res)
-vsd <- vst(dds, blind=FALSE)
-zz <-cbind(as.data.frame(z),mxf)
-def <-as.data.frame(zz[order(zz$pvalue),])
-
-head(def,10) %>%
-  kbl(caption="Top DE genes for Aza treatment") %>%
-  kable_paper("hover", full_width = F)
-
-```
-
-Make a smear plot.
-
-```{r,smearplot}
-
-sigf <- subset(def,padj<=0.05)
-
-DET=nrow(mxf)
-NSIG=nrow(sigf)
-NUP=nrow(subset(sigf,log2FoldChange>0))
-NDN=nrow(subset(sigf,log2FoldChange<0))
-
-HEADER=paste(DET,"detected genes;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")
-
-plot(log10(def$baseMean) ,def$log2FoldChange,
-  cex=0.6,pch=19,col="darkgray",
-  main="5-azacitidine treatment in AML3 cells",
-  xlab="log10(basemean)",ylab="log2(fold change)")
-
-points(log10(sigf$baseMean) ,sigf$log2FoldChange,
-  cex=0.6,pch=19,col="red")
-
-mtext(HEADER)
-
-```
-
-```{r,writedata}
-
-write.table(def,file="../data/aza_rnaseq.tsv")
+def <- tbl[["CD4 T cells"]]
 
@@
 | Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| Gene Selection Criteria | Muscat FDR<0.05 CD4 T-cells |
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | via DEE2 at accession SRP038101 (human) |
-| Other parameters | Min gene set size of 10 |
+| Data availability | R/Experiment hub "Kang" or GSE96583. |
+| Other parameters | Min gene set size of 5 |
 
@@
 
-defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
-defup <- unique(sapply(strsplit(defup," "),"[[",2))
+defup <- subset(def,p_adj.loc<=0.05 & logFC>0)$gene
 
-defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
-defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
+defdn <- subset(def,p_adj.loc<=0.05 & logFC<0)$gene
 
-bg <- rownames(def)
-bg <- unique(sapply(strsplit(bg," "),"[[",2))
+bg <- unique(def$gene)
 
@@
 ora_up$geneID <- NULL
-ora_up <- subset(ora_up,p.adjust<0.05 & Count >=10)
+ora_up <- subset(ora_up,p.adjust<0.05 & Count >=5)
 ora_ups <- rownames(ora_up)
@@
 head(ora_up) %>%
-  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment") %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in IFNB treatment") %>%
   kable_paper("hover", full_width = F)
@@
 head(ora_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment") %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in IFNB treatment") %>%
   kable_paper("hover", full_width = F)
@@
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | via DEE2 at accession SRP038101 (human) |
+| Data availability | R/Experiment hub "Kang" or GSE96583. |
 | Other parameters | Min gene set size of 10 + BUG FIX |
@@
 
-defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
-defup <- unique(sapply(strsplit(defup," "),"[[",2))
+defup <- unique(subset(def,p_adj.loc<=0.05 & logFC>0)$gene)
 
-defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
-defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
+defdn <- unique(subset(def,p_adj.loc<=0.05 & logFC<0)$gene)
 
-bg <- rownames(def)
-bg <- unique(sapply(strsplit(bg," "),"[[",2))
+bg <- unique(def$gene)
 message(paste("number of genes in background:",length(bg)))
@@
 head(orafix_up) %>%
-  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment") %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in IFNB treatment") %>%
   kable_paper("hover", full_width = F)
@@
 head(orafix_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment") %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in IFNB treatment") %>%
   kable_paper("hover", full_width = F)
@@
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | via DEE2 at accession SRP038101 (human) |
+| Data availability | R/Experiment hub "Kang" or GSE96583. |
 | Other parameters | Min gene set size of 10 |
@@
 ```{r,oraupfix2}
+
+n_pw=10
 

### 2023-08-02 analysis/example_rnaseq.Rmd
write gene names out

@@
 
+writeLines(text=defup,con="aza_up.txt")
+writeLines(text=defdn,con="aza_dn.txt")
+writeLines(text=bg,con="aza_bg.txt")
+
 ```
@@
 ora_dn$geneID <- NULL
-ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
+ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=5)
 ora_dns <- rownames(ora_dn)
@@
 orafix_up$geneID <- NULL
-orafix_up <- subset(orafix_up,p.adjust<0.05 & Count >=10)
+orafix_up <- subset(orafix_up,p.adjust<0.05 & Count >=5)
 orafix_ups <- rownames(orafix_up)
@@
 orafix_dn$geneID <- NULL
-orafix_dn <- subset(orafix_dn,p.adjust<0.05 & Count >=10)
+orafix_dn <- subset(orafix_dn,p.adjust<0.05 & Count >=5)
 orafix_dns <- rownames(orafix_dn)

### 2023-08-02 analysis/example_scrnaseq.Rmd
write gene names out

@@
 
+```{r,orafunc}
+
+ora <- function(fg,bg,genesets,n_pw=10,mincount=5) {
+
+ora_res <- as.data.frame(enricher(gene = fg ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora_res$geneID <- NULL
+ora_res <- subset(ora_res,p.adjust<0.05 & Count >=mincount)
+ora_res_sets <- rownames(ora_res)
+
+gr <- as.numeric(sapply(strsplit(ora_res$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_res$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora_res$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_res$BgRatio,"/"),"[[",2))
+
+ora_res$es <- gr/br
+ora_res <- ora_res[order(-ora_res$es),]
+ora_res$Description=NULL
+return(ora_res)
+}
+
+orafix <- function(fg,bg,genesets,n_pw=10,mincount=5) {
+
+bgdf <- data.frame("background",bg)
+colnames(bgdf) <- c("term","gene")
+genesets_fix <- rbind(genesets,bgdf)
+
+ora_res <- as.data.frame(enricher(gene = fg ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets_fix,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora_res$geneID <- NULL
+ora_res <- subset(ora_res,p.adjust<0.05 & Count >=mincount)
+ora_res_sets <- rownames(ora_res)
+
+gr <- as.numeric(sapply(strsplit(ora_res$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_res$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora_res$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_res$BgRatio,"/"),"[[",2))
+
+ora_res$es <- gr/br
+ora_res <- ora_res[order(-ora_res$es),]
+ora_res$Description=NULL
+return(ora_res)
+}
+
+```
+
 For this report I will use a SingleCellExperiment (SCE) containing 10x droplet-based
@@
   kable_paper("hover", full_width = F)
-
-def <- tbl[["CD4 T cells"]]
 
@@
 
-defup <- subset(def,p_adj.loc<=0.05 & logFC>0)$gene
+def <- lapply(tbl , function(x) {
+  bg <- unique(x$gene)
+  up <- unique(subset(x,p_adj.loc<0.05&logFC>0)$gene)
+  dn <- unique(subset(x,p_adj.loc<0.05&logFC<0)$gene)
+  return(list("bg"=bg,"up"=up,"dn"=dn))
+} )
 
-defdn <- subset(def,p_adj.loc<=0.05 & logFC<0)$gene
-
-bg <- unique(def$gene)
-
-message(paste("number of genes in background:",length(bg)))
+str(def)
 
@@
 
-n_pw=10
+ups <- lapply(1:length(def),function(i) {
+  ora(def[[i]]$up,def[[i]]$bg,genesets=genesets2)
+})
 
-ora_up <- as.data.frame(enricher(gene = defup ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+up_sets <- unlist(lapply(1:length(ups) , function(i) {
+  nm <- names(def)[[i]]
+  sets <- ups[[i]]$ID
+  if ( length(sets) >0 ) { paste(nm,sets) }
+} ) )
 
-ora_up$geneID <- NULL
-ora_up <- subset(ora_up,p.adjust<0.05 & Count >=5)
-ora_ups <- rownames(ora_up)
-
-gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))
-
-br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))
-
-ora_up$es <- gr/br
-ora_up <- ora_up[order(-ora_up$es),]
-ora_up$Description=NULL
-
-head(ora_up) %>%
-  kbl(row.names = FALSE, caption="Top upregulated pathways in IFNB treatment") %>%
-  kable_paper("hover", full_width = F)
-
-topup2 <- rev(head(ora_up$es,10))
-names(topup2) <- rev(head(ora_up$ID,10))
+length(up_sets)
+up_sets
 
 ```
-
-The reported background size does not match the dataset.
 
@@
 
-ora_dn <- as.data.frame(enricher(gene = defdn ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+dns <- lapply(1:length(def),function(i) {
+  ora(def[[i]]$dn,def[[i]]$bg,genesets=genesets2)
+})
 
-ora_dn$geneID <- NULL
-ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
-ora_dns <- rownames(ora_dn)
+dn_sets <- unlist(lapply(1:length(dns) , function(i) {
+  nm <- names(def)[[i]]
+  sets <- dns[[i]]$ID
+  if ( length(sets) >0 ) { paste(nm,sets) }
+} ) )
 
-gr <- as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",2))
-
-br <- as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",2))
-
-ora_dn$es <- gr/br
-ora_dn <- ora_dn[order(-ora_dn$es),]
-ora_dn$Description=NULL
-
-head(ora_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="Top downregulated pathways in IFNB treatment") %>%
-  kable_paper("hover", full_width = F)
-
-topdn2 <- head(ora_dn$es,n_pw)
-names(topdn2) <- head(ora_dn$ID,n_pw)
-
-```
-
-Make a barplot
-
-```{r,topbarplot1}
-
-par(mar=c(5,20,5,1))
-
-cols <- c(rep("blue",n_pw),rep("red",n_pw))
-
-barplot(c(topdn2,topup2),
-  horiz=TRUE,las=1,cex.names=0.65,col=cols,
-  main="top DE KEGGs",
-  xlab="ES",
-  cex.main=0.9)
-
-mtext("ORA test")
+length(dn_sets)
+dn_sets
 
@@
 
-```{r,subsetfix1}
+Enrichment firstly with upregulated genes
 
-defup <- unique(subset(def,p_adj.loc<=0.05 & logFC>0)$gene)
+```{r,orafixup1}
 
-defdn <- unique(subset(def,p_adj.loc<=0.05 & logFC<0)$gene)
+fixups <- lapply(1:length(def),function(i) {
+  orafix(def[[i]]$up,def[[i]]$bg,genesets=genesets2)
+})
 
-bg <- unique(def$gene)
-message(paste("number of genes in background:",length(bg)))
+fixup_sets <- unlist(lapply(1:length(fixups) , function(i) {
+  nm <- names(def)[[i]]
+  sets <- fixups[[i]]$ID
+  if ( length(sets) >0 ) { paste(nm,sets) }
+} ) )
+
+length(fixup_sets)
+fixup_sets
 
@@
 
-Adding all detected genes to the background appears to improve results!
+Now repeat with the downregulated genes
 
-```{r,gs2fix1}
+```{r,orafixdn1}
 
-bgdf <- data.frame("background",bg)
-colnames(bgdf) <- c("term","gene")
-genesets2 <- rbind(genesets2,bgdf)
+fixdns <- lapply(1:length(def),function(i) {
+  orafix(def[[i]]$dn,def[[i]]$bg,genesets=genesets2)
+})
 
-```
+fixdn_sets <- unlist(lapply(1:length(fixdns) , function(i) {
+  nm <- names(def)[[i]]
+  sets <- fixdns[[i]]$ID
+  if ( length(sets) >0 ) { paste(nm,sets) }
+} ) )
 
-Enrichment firstly with upregulated genes
-
-```{r,oraupfix1}
-
-n_pw=10
-
-orafix_up <- as.data.frame(enricher(gene = defup ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
-
-orafix_up$geneID <- NULL
-orafix_up <- subset(orafix_up,p.adjust<0.05 & Count >=10)
-orafix_ups <- rownames(orafix_up)
-
-gr <- as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",2))
-
-br <- as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",2))
-
-orafix_up$es <- gr/br
-orafix_up <- orafix_up[order(-orafix_up$es),]
-orafix_up$Description=NULL
-
-head(orafix_up) %>%
-  kbl(row.names = FALSE, caption="Top upregulated pathways in IFNB treatment") %>%
-  kable_paper("hover", full_width = F)
-
-topup2 <- rev(head(orafix_up$es,10))
-names(topup2) <- rev(head(orafix_up$ID,10))
-
-```
-
-Now with the downregulated genes
-
-```{r,oradnfix1}
-
-orafix_dn <- as.data.frame(enricher(gene = defdn ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
-
-orafix_dn$geneID <- NULL
-orafix_dn <- subset(orafix_dn,p.adjust<0.05 & Count >=10)
-orafix_dns <- rownames(orafix_dn)
-
-gr <- as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",2))
-
-br <- as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",2))
-
-orafix_dn$es <- gr/br
-orafix_dn <- orafix_dn[order(-orafix_dn$es),]
-orafix_dn$Description=NULL
-
-head(orafix_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="Top downregulated pathways in IFNB treatment") %>%
-  kable_paper("hover", full_width = F)
-
-topdn2 <- head(orafix_dn$es,n_pw)
-names(topdn2) <- head(orafix_dn$ID,n_pw)
-
-```
-
-Make a barplot
-
-```{r,topbarplot2fix1}
-
-par(mar=c(5,20,5,1))
-
-cols <- c(rep("blue",n_pw),rep("red",n_pw))
-
-barplot(c(topdn2,topup2),
-  horiz=TRUE,las=1,cex.names=0.65,col=cols,
-  main="top DE KEGGs",
-  xlab="ES",
-  cex.main=0.9)
-
-mtext("ORA fix test")
+length(fixdn_sets)
+fixdn_sets
 
@@
 
-v1 <- list("ORA up"=ora_ups, "ORA fix up"=orafix_ups)
-v2 <- list("ORA dn"=ora_dns, "ORA fix dn"=orafix_dns)
-v3 <- list("ORA"=union(ora_ups,ora_dns),
-  "ORA fix"=union(orafix_ups,orafix_dns))
+v1 <- list("ORA up"=up_sets, "ORA fix up"=fixup_sets)
+v2 <- list("ORA dn"=dn_sets, "ORA fix dn"=fixdn_sets)
+v3 <- list("ORA"=union(up_sets,dn_sets),
+  "ORA fix"=union(fixup_sets,fixdn_sets))
 
@@
 
-ora <- c(ora_ups,ora_dns)
-orafix <- c(orafix_ups,orafix_dns)
+ora <- c(up_sets,dn_sets)
+orafix <- c(fixup_sets,fixdn_sets)
 
 jaccard(ora,orafix)
-
-```
-
-List gene sets identified only in the "fixed" analysis.
-
-```{r,examine1}
-
-ora <- union(ora_ups,ora_dns)
-
-orafix <- union(orafix_ups,orafix_dns)
-
-setdiff(orafix,ora)
 
@@
 
-n_pw=10
+ups <- lapply(1:length(def),function(i) {
+  ora(def[[i]]$up,def[[i]]$bg,genesets=genesets2)
+})
 
-ora_up <- as.data.frame(enricher(gene = defup ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+up_sets <- unlist(lapply(1:length(ups) , function(i) {
+  nm <- names(def)[[i]]
+  sets <- ups[[i]]$ID
+  if ( length(sets) >0 ) { paste(nm,sets) }
+} ) )
 
-ora_up$geneID <- NULL
-ora_up <- subset(ora_up,p.adjust<0.05 & Count >=5)
-ora_ups <- rownames(ora_up)
-
-gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))
-
-br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))
-
-ora_up$es <- gr/br
-ora_up <- ora_up[order(-ora_up$es),]
-ora_up$Description=NULL
-
-head(ora_up) %>%
-  kbl(row.names = FALSE, caption="Top upregulated pathways in IFNB treatment") %>%
-  kable_paper("hover", full_width = F)
-
-topup2 <- rev(head(ora_up$es,10))
-names(topup2) <- rev(head(ora_up$ID,10))
+length(up_sets)
+up_sets
 
@@
 
-ora_dn <- as.data.frame(enricher(gene = defdn ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+dns <- lapply(1:length(def),function(i) {
+  ora(def[[i]]$dn,def[[i]]$bg,genesets=genesets2)
+})
 
-ora_dn$geneID <- NULL
-ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
-ora_dns <- rownames(ora_dn)
+dn_sets <- unlist(lapply(1:length(dns) , function(i) {
+  nm <- names(def)[[i]]
+  sets <- dns[[i]]$ID
+  if ( length(sets) >0 ) { paste(nm,sets) }
+} ) )
 
-gr <- as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",2))
-
-br <- as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",2))
-
-ora_dn$es <- gr/br
-ora_dn <- ora_dn[order(-ora_dn$es),]
-ora_dn$Description=NULL
-
-head(ora_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="Top downregulated pathways in IFNB treatment") %>%
-  kable_paper("hover", full_width = F)
-
-topdn2 <- head(ora_dn$es,n_pw)
-names(topdn2) <- head(ora_dn$ID,n_pw)
-
-```
-
-Make a barplot
-
-```{r,topbarplot22}
-
-par(mar=c(5,20,5,1))
-
-cols <- c(rep("blue",n_pw),rep("red",n_pw))
-
-barplot(c(topdn2,topup2),
-  horiz=TRUE,las=1,cex.names=0.65,col=cols,
-  main="top DE Reactomes",
-  xlab="ES",
-  cex.main=0.9)
-
-mtext("ORA test")
+length(dn_sets)
+dn_sets
 
@@
 
-Adding all detected genes to the background appears to improve results!
+```{r,orafixup2}
 
-```{r,gs2fix2}
+fixups <- lapply(1:length(def),function(i) {
+  orafix(def[[i]]$up,def[[i]]$bg,genesets=genesets2)
+})
 
-bgdf <- data.frame("background",bg)
-colnames(bgdf) <- c("term","gene")
-genesets2 <- rbind(genesets2,bgdf)
+fixup_sets <- unlist(lapply(1:length(fixups) , function(i) {
+  nm <- names(def)[[i]]
+  sets <- fixups[[i]]$ID
+  if ( length(sets) >0 ) { paste(nm,sets) }
+} ) )
+
+length(fixup_sets)
+fixup_sets
 
@@
 
-Enrichment firstly with upregulated genes
+Now repeat with the downregulated genes
 
-```{r,oraupfix2}
+```{r,orafixdn2}
 
-n_pw=10
+fixdns <- lapply(1:length(def),function(i) {
+  orafix(def[[i]]$dn,def[[i]]$bg,genesets=genesets2)
+})
 
-orafix_up <- as.data.frame(enricher(gene = defup ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+fixdn_sets <- unlist(lapply(1:length(fixdns) , function(i) {
+  nm <- names(def)[[i]]
+  sets <- fixdns[[i]]$ID
+  if ( length(sets) >0 ) { paste(nm,sets) }
+} ) )
 
-orafix_up$geneID <- NULL
-orafix_up <- subset(orafix_up,p.adjust<0.05 & Count >=10)
-orafix_ups <- rownames(orafix_up)
-
-gr <- as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",2))
-
-br <- as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",2))
-
-orafix_up$es <- gr/br
-orafix_up <- orafix_up[order(-orafix_up$es),]
-orafix_up$Description=NULL
-
-head(orafix_up) %>%
-  kbl(row.names = FALSE, caption="Top upregulated pathways in IFNB treatment") %>%
-  kable_paper("hover", full_width = F)
-
-topup2 <- rev(head(orafix_up$es,10))
-names(topup2) <- rev(head(orafix_up$ID,10))
-
-```
-
-Now with the downregulated genes
-
-```{r,oradnfix2}
-
-n_pw=10
-
-orafix_dn <- as.data.frame(enricher(gene = defdn ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
-
-orafix_dn$geneID <- NULL
-orafix_dn <- subset(orafix_dn,p.adjust<0.05 & Count >=10)
-orafix_dns <- rownames(orafix_dn)
-
-gr <- as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",2))
-
-br <- as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",2))
-
-orafix_dn$es <- gr/br
-orafix_dn <- orafix_dn[order(-orafix_dn$es),]
-orafix_dn$Description=NULL
-
-head(orafix_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="Top downregulated pathways in IFNB treatment") %>%
-  kable_paper("hover", full_width = F)
-
-topdn2 <- head(orafix_dn$es,n_pw)
-names(topdn2) <- head(orafix_dn$ID,n_pw)
-
-```
-
-Make a barplot
-
-```{r,topbarplot2fix2}
-
-par(mar=c(5,20,5,1))
-
-cols <- c(rep("blue",n_pw),rep("red",n_pw))
-
-barplot(c(topdn2,topup2),
-  horiz=TRUE,las=1,cex.names=0.65,col=cols,
-  main="top DE Reactomes",
-  xlab="ES",
-  cex.main=0.9)
-
-mtext("ORA fix test")
+length(fixdn_sets)
+fixdn_sets
 
@@
 
-ora <- c(ora_ups,ora_dns)
-orafix <- c(orafix_ups,orafix_dns)
+ora <- c(up_sets,dn_sets)
+orafix <- c(fixup_sets,fixdn_sets)
 

### 2023-08-30 analysis/example_proteomics.Rmd
addition

@@
 suppressPackageStartupMessages({
-  library("getDEE2")
   library("DESeq2")
@@
 
-For this report I will be using proteomics data from Mulvey et al 2021 (PMID: 34599159),
-who looked at the effect of LPS on THP-1 cell lines.
-Although in this paper they looked at numerous timepoints, I will be looking only at the 24hr time
-point where the number of significantly differentially expressed proteins was the greatest.
-The data were obtained from Supplementary Table 1 and saved to the /data folder.
+For this report I will be using proteomics data from Mackmull et al (2015) where they looked at the
+effect of HDACi on protein expression in nuclear and cytosolic compartments.
+Here I am using the data from the TSA treated nuclear fractions after 48 hrs exposure.
+The data were obtained from Supplementary Table 2 and saved to the /data folder.
 
@@
 
-def <- read.table("../data/lps_24h.tsv",sep="\t",header=TRUE)
+def <- read.table("../data/tsa_48hr_nuclear_frac.tsv",sep="\t",header=TRUE)
 
@@
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | via DEE2 at accession SRP038101 (human) |
+| Data availability | Mackmull et al (2015) |
 | Other parameters | Min gene set size of 10 |
@@
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | via DEE2 at accession SRP038101 (human) |
+| Data availability | Mackmull et al (2015) |
 | Other parameters | Min gene set size of 10 + BUG FIX |

### 2023-08-30 analysis/example_proteomics.Rmd
updates

@@
 
-barplot(c(topdn2,topup2),
+top <- c(topdn2,topup2)
+
+if ( length(top) > 1 ) {
+  barplot(c(top),
     horiz=TRUE,las=1,cex.names=0.65,col=cols,
@@
     cex.main=0.9)
-
   mtext("ORA test")
+}
 
@@
     cex.main=0.9)
-
-mtext("ORA fix test")
+  mtext("ORA fix")
+}
 
@@
 plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
+
+png("prot_kegg_euler.png")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
+dev.off()
 
@@
 plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
+
+png("prot_reactome_euler.png")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
+dev.off()
 

### 2023-08-30 analysis/example_rnaseq.Rmd
update
updates

@@
 
-writeLines(text=defup,con="aza_up.txt")
-writeLines(text=defdn,con="aza_dn.txt")
-writeLines(text=bg,con="aza_bg.txt")
+writeLines(text=unique(defup),con="aza_up.txt")
+writeLines(text=unique(defdn),con="aza_dn.txt")
+writeLines(text=unique(bg),con="aza_bg.txt")
 
@@
 plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
+
+png("rna_kegg_euler")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
+dev.off()
 
@@
     cex.main=0.9)
-
-mtext("ORA fix test")
-
+  mtext("ORA fix")
+}
 ```
@@
 plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
+
+png("rna_reactome_euler")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
+dev.off()
 

### 2023-08-30 analysis/example_scrnaseq.Rmd
addition
updates

@@
 
+if ( nrow(ora_res > 1 ) ) {
+
 ora_res$geneID <- NULL
@@
 
+}
+
 orafix <- function(fg,bg,genesets,n_pw=10,mincount=5) {
@@
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+if ( nrow(ora_res > 1 ) ) {
 
@@
 return(ora_res)
+}
+
 }
@@
 
+png("scrna_kegg_euler.png")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
+dev.off()
+
 ```
@@
 
-ora <- c(up_sets,dn_sets)
-orafix <- c(fixup_sets,fixdn_sets)
+orasets <- c(up_sets,dn_sets)
+orafixsets <- c(fixup_sets,fixdn_sets)
 
-jaccard(ora,orafix)
+jaccard(orasets,orafixsets)
 
@@
 
+png("scrna_reactome_euler.png")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
+dev.off()
+
 ```
@@
 
-ora <- union(ora_ups,ora_dns)
-
-orafix <- union(orafix_ups,orafix_dns)
-
 setdiff(orafix,ora)

### 2023-08-30 analysis/example_proteomics.Rmd
prot

@@
 
-For this report I will be using proteomics data from Mackmull et al (2015) where they looked at the
-effect of HDACi on protein expression in nuclear and cytosolic compartments.
-Here I am using the data from the TSA treated nuclear fractions after 48 hrs exposure.
-The data were obtained from Supplementary Table 2 and saved to the /data folder.
+For this report I will be using proteomics data from Stukalov et al (https://pubmed.ncbi.nlm.nih.gov/33845483/),
+who profiled the effect of SARS-COV2 infection on A549 cells.
+
+Here we used the 24hr profile, as there were relatively more changes found.
+
+Probable contaminants were removed, as well as viral proteins.
+This left 5825 proteins, including 165 downregulated and 44 upregulated.
+
+The data were obtained from Supplementary Table 5 and saved to the /data folder.
 
@@
 
-def <- read.table("../data/tsa_48hr_nuclear_frac.tsv",sep="\t",header=TRUE)
+def <- read.table("../data/sarscov2_prot.tsv",sep="\t",header=TRUE)
 
@@
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | Mackmull et al (2015) |
+| Data availability | Stukalov et al 2021: PMID 33845483 |
 | Other parameters | Min gene set size of 10 |
@@
 
-defup <- unique(subset(def,adj.P.Val<=0.05 & logFC>0)$GN)
+defup <- unique(subset(def,change.SARS_CoV2.24h_vs_mock.24h=="+")$gene_name)
 
-defdn <- unique(subset(def,adj.P.Val<=0.05 & logFC<0)$GN)
+defdn <- unique(subset(def,change.SARS_CoV2.24h_vs_mock.24h=="-")$gene_name)
 
-bg <- unique(def$GN)
+bg <- unique(def$gene_name)
 
@@
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | Mackmull et al (2015) |
+| Data availability | Data availability | Stukalov et al 2021: PMID 33845483 |
 | Other parameters | Min gene set size of 10 + BUG FIX |
@@
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | Mackmull et al (2015) |
+| Data availability | Data availability | Stukalov et al 2021: PMID 33845483 |
 | Other parameters | Min gene set size of 10 |

### 2023-08-30 analysis/example_rnaseq.Rmd
prot

@@
 
-png("rna_kegg_euler")
+png("rna_kegg_euler.png")
 plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
@@
 
-png("rna_reactome_euler")
+png("rna_reactome_euler.png")
 plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")

### 2023-09-06 analysis/example_meth.Rmd
analysis for fig1

--- 2023-08-30 analysis/example_proteomics.Rmd
@@
 ---
-title: "Examining background gene lists in clusterProfiler - proteomics example"
+title: "Examining background gene lists in clusterProfiler - methylation example"
 author: "Mark Ziemann & Anusuiya Bora"
@@
 
-Source: https://github.com/markziemann/enrichment_recipe
+Source: https://github.com/markziemann/background
 
@@
   library("gplots")
+  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
 })
@@
 
-For this report I will be using proteomics data from Mackmull et al (2015) where they looked at the
-effect of HDACi on protein expression in nuclear and cytosolic compartments.
-Here I am using the data from the TSA treated nuclear fractions after 48 hrs exposure.
-The data were obtained from Supplementary Table 2 and saved to the /data folder.
+For this report I will be using Epigenetics data from an epigenome wide study for smokers versus non-smokers.
+PubmedID:27651444
+Data was downloaded from EWAS Catalog:
+http://ewascatalog.org/?study=27651444_smoking_current_vs_never_smoking
 
@@
 
-def <- read.table("../data/tsa_48hr_nuclear_frac.tsv",sep="\t",header=TRUE)
+x <- read.table("../data/27651444_smoking_current_vs_never_smoking_16938071371108067.tsv",header=TRUE,sep="\t")
+
+```
+
+Get background gene list.
+
+```{r,bg}
+
+anno = getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
+bg <- anno$UCSC_RefGene_Name
+bg <- unique(unlist(strsplit(bg,";")))
+length(bg)
+
+```
+
+Get up and down-methylated list.
+
+```{r,dm1}
+
+defup <- subset(x,beta>0)$gene
+defup <- defup[defup!="-"]
+defup <- unique(unlist(strsplit(defup,";")))
+length(defup)
+
+defdn <- subset(x,beta<0)$gene
+defdn <- defdn[defdn!="-"]
+defdn <- unique(unlist(strsplit(defdn,";")))
+length(defdn)
+
+```
+
+Gene sets.
+Get the gene sets loaded in R.
+These are KEGG gene sets
+
+```{r,gs1}
+# from https://www.gsea-msigdb.org/gsea/msigdb/human/collections.jsp 16th June 2023
+genesets2 <- read.gmt("../ref/c2.cp.kegg.v2023.1.Hs.symbols.gmt")
+gsets <- gmtPathways("../ref/c2.cp.kegg.v2023.1.Hs.symbols.gmt")
+
+message(paste("number of genes described in the annotation set:",length(unique(genesets2$gene))))
 
@@
 | P-value correction for multiple comparisons | FDR method |
-| Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| Background list | Genes with one or more probes on the 450K array |
+| Gene Selection Criteria | Metaanalysis FDR<0.05 |
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | Mackmull et al (2015) |
+| Data availability | PMID: 27651444 |
 | Other parameters | Min gene set size of 10 |
@@
 
-Get the gene sets loaded in R.
-These are KEGG gene sets
-
-```{r,gs1}
-# from https://www.gsea-msigdb.org/gsea/msigdb/human/collections.jsp 16th June 2023
-genesets2 <- read.gmt("../ref/c2.cp.kegg.v2023.1.Hs.symbols.gmt")
-gsets <- gmtPathways("../ref/c2.cp.kegg.v2023.1.Hs.symbols.gmt")
-
-message(paste("number of genes described in the annotation set:",length(unique(genesets2$gene))))
-
-```
-
-Now filter the gene names into three lists, up-regulated, down-regulated and
-background.
-Background is simply all genes that were detected.
-
-```{r,subset1}
-
-defup <- unique(subset(def,adj.P.Val<=0.05 & logFC>0)$GN)
-
-defdn <- unique(subset(def,adj.P.Val<=0.05 & logFC<0)$GN)
-
-bg <- unique(def$GN)
-
-message(paste("number of proteins in background:",length(bg)))
-
-```
-
-Enrichment firstly with upregulated genes
+Enrichment firstly with upmethylated genes
 
@@
 ora_up <- as.data.frame(enricher(gene = defup ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
@@
 head(ora_up) %>%
-  kbl(row.names = FALSE, caption="Top upregulated pathways in LPS treatment") %>%
+  kbl(row.names = FALSE, caption="Top up-methylated pathways in LPS treatment") %>%
   kable_paper("hover", full_width = F)
@@
 
-Now repeat with the downregulated genes
+Now repeat with the downmethylated genes
 
@@
 ora_dn <- as.data.frame(enricher(gene = defdn ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
@@
 head(ora_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="Top downregulated pathways in LPS treatment") %>%
+  kbl(row.names = FALSE, caption="Top downmethylated pathways in LPS treatment") %>%
   kable_paper("hover", full_width = F)
@@
 | P-value correction for multiple comparisons | FDR method |
-| Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| Background list | Genes with one or more probes on the 450K array |
+| Gene Selection Criteria | Metaanalysis FDR<0.05 |
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | Mackmull et al (2015) |
+| Data availability | Data availability | PMID: 276514443 |
 | Other parameters | Min gene set size of 10 + BUG FIX |
@@
 I have also done some modification to the KEGG gene sets, to en
-
-Now filter the gene names into three lists, up-regulated, down-regulated and
-background.
-Background is simply all genes that were detected.
 
@@
 
-Enrichment firstly with upregulated genes
+Enrichment firstly with upmethylated genes
 
@@
 orafix_up <- as.data.frame(enricher(gene = defup ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
@@
 head(orafix_up) %>%
-  kbl(row.names = FALSE, caption="Top upregulated pathways in LPS treatment") %>%
+  kbl(row.names = FALSE, caption="Top upmethylated pathways in LPS treatment") %>%
   kable_paper("hover", full_width = F)
@@
 
-Now with the downregulated genes
+Now with the downmethylated genes
 
@@
 orafix_dn <- as.data.frame(enricher(gene = defdn ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
@@
 head(orafix_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="Top downregulated pathways in LPS treatment") %>%
+  kbl(row.names = FALSE, caption="Top downmethylated pathways in LPS treatment") %>%
   kable_paper("hover", full_width = F)
@@
 par(mfrow=c(2,1))
-plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="upregulated KEGG pathways")
-plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="downregulated KEGG pathways")
-plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
+plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="up-methylated KEGG pathways")
+plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="down-methylated KEGG pathways")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-methylated KEGG pathways")
 
-png("prot_kegg_euler.png")
-plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
+svg("meth_kegg_euler.svg")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-methylated KEGG pathways")
 dev.off()
@@
 | P-value correction for multiple comparisons | FDR method |
-| Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| Background list | Genes with one or more probes on the 450K array |
+| Gene Selection Criteria | Metaanalysis FDR<0.05 |
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | Mackmull et al (2015) |
+| Data availability | Data availability | PMID: 27651444 |
 | Other parameters | Min gene set size of 10 |
@@
 head(ora_up) %>%
-  kbl(row.names = FALSE, caption="Top upregulated pathways in LPS treatment") %>%
+  kbl(row.names = FALSE, caption="Top upmethylated pathways in LPS treatment") %>%
   kable_paper("hover", full_width = F)
@@
 | P-value correction for multiple comparisons | FDR method |
-| Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| Background list | Genes with one or more probes on the 450K array |
+| Gene Selection Criteria | Metaanalysis FDR<0.05 |
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | Mackmull et al (2015) |
+| Data availability | Data availability | PMID: 27651444 |
 | Other parameters | Min gene set size of 10 + BUG FIX |
@@
 par(mfrow=c(2,1))
-plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="upregulated Reactomes")
-plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="downregulated Reactomes")
-plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
+plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="up-methylated Reactomes")
+plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="down-methylated Reactomes")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-methylated Reactomes")
 
-png("prot_reactome_euler.png")
-plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
+svg("meth_reactome_euler.svg")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-methylated Reactomes")
 dev.off()

### 2023-09-06 analysis/example_scrnaseq2.Rmd
analysis for fig1

--- 2023-08-02 analysis/example_scrnaseq.Rmd
@@
 
-load("muscat_vignette.RData")
-
-ls()
+tbl <- readRDS("scDEres.Rds")
 
@@
 ora_res <- as.data.frame(enricher(gene = fg ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets,
+  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+if ( nrow(ora_res > 1 ) ) {
 
@@
 ora_res <- as.data.frame(enricher(gene = fg ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets_fix,
+  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets_fix,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+if ( nrow(ora_res > 1 ) ) {
 
@@
 
+}
+
 ```
@@
 
-Downstream analysis was done according to the Muscat vignette.
-
-Focusing on CD4 T-cells.
+Downstream analysis was done according to the vignette from Seurat.
+Link: https://satijalab.org/seurat/archive/v3.1/immune_alignment.html
 
@@
 
-lapply(tbl, function(x) {
-  nsig <- nrow(subset(x,p_adj.loc<0.05))
+x <- lapply(tbl, function(x) {
+  nsig <- nrow(subset(x,p_val_adj<0.05))
   ntot <- nrow(x)
@@
 } )
+
+x
+
+x <- do.call(rbind,x)
+colnames(x) <- c("ngenes","nsig")
+x <- as.data.frame(x)
+
+x <- subset(x,ngenes>=800 & nsig>=100)
+
+x
+
+x <- rownames(x)
+
+tbl <- tbl[names(tbl) %in% x]
+
+str(tbl)
 
@@
 | P-value correction for multiple comparisons | FDR method |
-| Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | Muscat FDR<0.05 CD4 T-cells |
+| Background list | As defined by Seurat |
+| Gene Selection Criteria | Seurat FDR<0.05 |
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | R/Experiment hub "Kang" or GSE96583. |
+| Data availability | SeuratData ifnb |
 | Other parameters | Min gene set size of 5 |
@@
 def <- lapply(tbl , function(x) {
-  bg <- unique(x$gene)
-  up <- unique(subset(x,p_adj.loc<0.05&logFC>0)$gene)
-  dn <- unique(subset(x,p_adj.loc<0.05&logFC<0)$gene)
+  bg <- unique(rownames(x))
+  up <- unique(rownames(subset(x,p_val_adj<0.05 & avg_log2FC>0)))
+  dn <- unique(rownames(subset(x,p_val_adj<0.05 & avg_log2FC<0)))
   return(list("bg"=bg,"up"=up,"dn"=dn))
@@
 | P-value correction for multiple comparisons | FDR method |
-| Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| Background list | As defined by Seurat |
+| Gene Selection Criteria | Seurat FDR<0.05 |
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | R/Experiment hub "Kang" or GSE96583. |
+| Data availability | SeuratData ifnb |
 | Other parameters | Min gene set size of 10 + BUG FIX |
@@
 
+svg("scrna2_kegg_euler.svg")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
+dev.off()
+
 ```
@@
 | P-value correction for multiple comparisons | FDR method |
-| Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| Background list |  As defined by Seurat |
+| Gene Selection Criteria | Seurat FDR<0.05 |
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | R/Experiment hub "Kang" or GSE96583. |
+| Data availability | SeuratData ifnb |
 | Other parameters | Min gene set size of 10 |
@@
 
+svg("scrna2_reactome_euler.svg")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
+dev.off()
+
 ```

### 2023-09-06 analysis/sim/run_simpw.R
working on the simulation study

@@ -1,0 +1,41 @@
+library("tidyverse")
+library("parallel")
+library("edgeR")
+library("DESeq2")
+library("limma")
+library("stringi")
+
+source("simpw_func.R")
+
+# obtain count data
+a<-countData()
+
+# generate some gene sets
+gsets<-randomGeneSets(a)
+
+
+# test
+# x<-agg_dge(a,3,40000000,0.3,0.05,1,10,"deseq2",gsets) ; str(x)
+
+###############################################
+# run simulations over a range of parameters
+###############################################
+SIMS=10
+unlink("simpw_res_running.tsv")
+res=NULL
+for ( FRAC_DE in c(0.05)) {
+  for (FC in c(1)) {
+    for (N_REPS in c(5)) {
+      for (DGE_FUNC in c("deseq2")) {
+        for ( SUM_COUNT in c(50000000)) {
+          for ( VARIANCE in c(0.5)) {
+            x<-agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets)
+            x<-as.data.frame(do.call(rbind, x))
+            write.table(x,file="simpw_res_running.tsv",quote=F,sep='\t',append=T)
+            res=c(res,x)
+          }
+        }
+      }
+    }
+  }
+}

### 2023-09-06 analysis/sim/simpw_func.R

@@ -1,0 +1,498 @@
+library("tidyverse")
+library("parallel")
+library("edgeR")
+library("DESeq2")
+library("limma")
+library("stringi")
+library("mitch")
+library("fgsea")
+
+########################################
+# get some counts
+########################################
+countData<-function() {
+# a is orig expression data
+a<-read.table("https://raw.githubusercontent.com/markziemann/simde/master/start_data/ERR2539161/ERR2539161.se.tsv")
+
+# GI is the gene information. It has gene name information for mapping accessions to gene symbols in GMT files
+gi<-read.table("https://raw.githubusercontent.com/markziemann/simde/master/start_data/ERR2539161/GeneInfo.tsv",header=T,row.names=1)
+
+# merge gene names
+aa<-merge(a,gi,by=0)
+aa<-aa[,-c(4:7)]
+aa<-aggregate(. ~ GeneSymbol,aa,function(x) sum(as.numeric(as.character(x))))
+aa$Row.names=NULL
+rownames(aa)<-aa$GeneSymbol
+aa$GeneSymbol=NULL
+a<-aa[which(aa$ERR2539161>=10),,drop=F]
+a
+}
+
+########################################
+# generate some gene sets
+########################################
+randomGeneSets<-function(a){
+gsets<-sapply( rep(50,1000) , function(x) {list(as.character(sample(rownames(a),x))) } )
+names(gsets)<-stri_rand_strings(length(gsets), 15, pattern = "[A-Za-z]")
+gsets
+}
+
+########################################
+# simulate some gene expression data
+########################################
+simrna<-function(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,gsets) {
+
+# N_REPS=5 ; SUM_COUNT=10000000 ; VARIANCE=0.2 ; FRAC_DE=0.05 ; FC=1 ; GMT="ReactomePathways.gmt"
+
+library("edgeR")
+
+df = NULL
+for (k in paste0("data",1:(N_REPS*2)))  {
+        b<-thinCounts(a,target.size=SUM_COUNT)
[…]

### 2023-09-07 analysis/sim/run_simpw.R
working on the simulation

@@ -15,12 +15,13 @@ gsets<-randomGeneSets(a)
 
 
 # test
-# x<-agg_dge(a,3,40000000,0.3,0.05,1,10,"deseq2",gsets) ; str(x)
+# x<-agg_dge(a=a,N_REPS=5,SUM_COUNT=50000000,VARIANCE=0.4,FRAC_DE=0.1,FC=1,SIMS=8,"deseq2",gsets)
+# str(x)
 
 ###############################################
 # run simulations over a range of parameters
 ###############################################
-SIMS=10
+SIMS=8
 unlink("simpw_res_running.tsv")
 res=NULL
 for ( FRAC_DE in c(0.05)) {
@@ -28,7 +29,7 @@ for ( FRAC_DE in c(0.05)) {
     for (N_REPS in c(5)) {
       for (DGE_FUNC in c("deseq2")) {
         for ( SUM_COUNT in c(50000000)) {
-          for ( VARIANCE in c(0.5)) {
+          for ( VARIANCE in c(0.2)) {
             x<-agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets)
             x<-as.data.frame(do.call(rbind, x))
             write.table(x,file="simpw_res_running.tsv",quote=F,sep='\t',append=T)

### 2023-09-07 analysis/sim/simpw_func.R

@@ -6,6 +6,7 @@ library("limma")
 library("stringi")
 library("mitch")
 library("fgsea")
+library("clusterProfiler")
 
 ########################################
 # get some counts
@@ -57,7 +58,7 @@ for (k in paste0("data",1:(N_REPS*2)))  {
 gsets_sub<-which(unlist( lapply(gsets,function(x) { length(which(rownames(a) %in% as.character(unlist(x)))) >10 }  ) ) )
 gsets<-gsets[which(names(gsets) %in% names(gsets_sub))]
 
-#Number of differential genes
+#Number of differential gene sets
 NDIF=round(length(gsets)*FRAC_DE)
 
 if (VARIANCE>0) {
@@ -83,13 +84,13 @@ if (NDIF>0) {
 
   # now find a list of genes inside the pathways
   UP_DE<-unique(unlist(unname(UP_LIST)))
+
   # select the ones that are also in the profile
   UP_DE<-UP_DE[which(UP_DE %in% row.names(df))]
 
   # same for down genes
   DN_DE<-unique(unlist(unname(DN_LIST)))
   DN_DE<-DN_DE[which(DN_DE %in% row.names(df))]
-
 
   ITX<-intersect(UP_DE,DN_DE)
   # need to eliminate the overlapping ones for simplicity
@@ -288,16 +289,18 @@ x
 ##################################
 # clusterprofiler default function
 ##################################
-run_clusterprofiler_default <-function(x,DGE_FUNC,gsets,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS){
+# Note that clusterprofiler requires different gene set format
+run_clusterprofiler_default <-function(x,DGE_FUNC,gsets2,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS){
 
-dge<-sapply(x,"[",6)
+dge <- sapply(x,"[",6)
 
-ups<-lapply(dge, function(x) { rownames(subset(x, padj<0.05 & log2FoldChange > 0)) } )
[…]

### 2023-09-07 analysis/sim/run_simpw.R
updates now working

@@ -4,14 +4,13 @@ library("edgeR")
 library("DESeq2")
 library("limma")
 library("stringi")
-
 source("simpw_func.R")
 
 # obtain count data
 a<-countData()
 
 # generate some gene sets
-gsets<-randomGeneSets(a)
+gsets<-randomGeneSets(a,setsize=30,nsets=500)
 
 
 # test
@@ -29,7 +28,7 @@ for ( FRAC_DE in c(0.05)) {
     for (N_REPS in c(5)) {
       for (DGE_FUNC in c("deseq2")) {
         for ( SUM_COUNT in c(50000000)) {
-          for ( VARIANCE in c(0.2)) {
+          for ( VARIANCE in c(0.2,0.3,0.4)) {
             x<-agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets)
             x<-as.data.frame(do.call(rbind, x))
             write.table(x,file="simpw_res_running.tsv",quote=F,sep='\t',append=T)

### 2023-09-07 analysis/sim/simpw_func.R

@@ -32,8 +32,8 @@ a
 ########################################
 # generate some gene sets
 ########################################
-randomGeneSets<-function(a){
-gsets<-sapply( rep(50,1000) , function(x) {list(as.character(sample(rownames(a),x))) } )
+randomGeneSets<-function(a,setsize,nsets){
+gsets<-sapply( rep(setsize,nsets) , function(x) {list(as.character(sample(rownames(a),x))) } )
 names(gsets)<-stri_rand_strings(length(gsets), 15, pattern = "[A-Za-z]")
 gsets
 }
@@ -359,6 +359,78 @@ x
 }
 
 ##################################
+# clusterprofiler fixed function
+##################################
+# Note that clusterprofiler requires different gene set format
+run_clusterprofiler_fixed <-function(x,DGE_FUNC,gsets2,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS){
+
+dge <- sapply(x,"[",6)
+
+ups <- lapply(dge, function(x) { rownames(subset(x, padj<0.05 & log2FoldChange > 0)) } )
+dns <- lapply(dge, function(x) { rownames(subset(x, padj<0.05 & log2FoldChange < 0)) } )
+bgs <- lapply(dge, function(x) { rownames(x) } )
+
+l_ups <- sapply(ups,length)
+l_dns <- sapply(dns,length)
+geneset_sizes <- sapply( gsets2 , length )
+
+# calculate number of genes in sets that are up and downregulated
+n_dns=n_ups=p_ups=p_dns=obs_up=obs_dn=NULL
+n_dns=n_ups=p_ups=p_dns=obs_up=obs_dn=list()
+
+for (d in 1:length(dge)) {
+
+# clusterprofiler UP
+ora_up <- as.data.frame(enricher(gene = ups[[d]] ,
+  universe = bgs[[d]],  maxGSSize = 500000, TERM2GENE = gsets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora_up$geneID <- NULL
+ora_up <- subset(ora_up,p.adjust<0.05 & Count >=10)
[…]

### 2023-09-07 analysis/sim/run_simpw.R
more work completed on simulation

@@ -20,7 +20,7 @@ gsets<-randomGeneSets(a,setsize=30,nsets=500)
 ###############################################
 # run simulations over a range of parameters
 ###############################################
-SIMS=8
+SIMS=16
 unlink("simpw_res_running.tsv")
 res=NULL
 for ( FRAC_DE in c(0.05)) {
@@ -28,7 +28,7 @@ for ( FRAC_DE in c(0.05)) {
     for (N_REPS in c(5)) {
       for (DGE_FUNC in c("deseq2")) {
         for ( SUM_COUNT in c(50000000)) {
-          for ( VARIANCE in c(0.2,0.3,0.4)) {
+          for ( VARIANCE in c(0.2,0.25,0.3,0.35,0.4)) {
             x<-agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets)
             x<-as.data.frame(do.call(rbind, x))
             write.table(x,file="simpw_res_running.tsv",quote=F,sep='\t',append=T)

### 2023-09-07 analysis/sim/simpw_func.R

@@ -310,23 +310,26 @@ for (d in 1:length(dge)) {
 
 # clusterprofiler UP
 ora_up <- as.data.frame(enricher(gene = ups[[d]] ,
-  universe = bgs[[d]],  maxGSSize = 500000, TERM2GENE = gsets2,
+  universe = bgs[[d]],  minGSSize=2, maxGSSize = 500000, TERM2GENE = gsets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
 
 ora_up$geneID <- NULL
-ora_up <- subset(ora_up,p.adjust<0.05 & Count >=10)
+ora_up <- subset(ora_up,p.adjust<0.05 )
 ora_ups <- rownames(ora_up)
 obs_up[[d]] <- ora_ups
 
 # clusterprofiler DOWN
 ora_dn <- as.data.frame(enricher(gene = dns[[d]] ,
-  universe = bgs[[d]],  maxGSSize = 500000, TERM2GENE = gsets2,
+  universe = bgs[[d]],  minGSSize=2, maxGSSize = 500000, TERM2GENE = gsets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
 
 ora_dn$geneID <- NULL
-ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
+ora_dn <- subset(ora_dn,p.adjust<0.05 )
 ora_dns <- rownames(ora_dn)
 obs_dn[[d]] <- ora_dns
+
+x[[d]][[13]] <- ora_ups
+x[[d]][[14]] <- ora_dns
 
 }
 
@@ -382,7 +385,7 @@ for (d in 1:length(dge)) {
 
 # clusterprofiler UP
 ora_up <- as.data.frame(enricher(gene = ups[[d]] ,
-  universe = bgs[[d]],  maxGSSize = 500000, TERM2GENE = gsets2,
+  universe = bgs[[d]],  minGSSize=2, maxGSSize = 500000, TERM2GENE = gsets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
 
 ora_up$geneID <- NULL
@@ -392,13 +395,16 @@ obs_up[[d]] <- ora_ups
 
 # clusterprofiler DOWN
[…]

### 2023-09-07 analysis/sim/run_simpw.R
bugfix

@@ -20,7 +20,7 @@ gsets<-randomGeneSets(a,setsize=30,nsets=500)
 ###############################################
 # run simulations over a range of parameters
 ###############################################
-SIMS=16
+SIMS=8
 unlink("simpw_res_running.tsv")
 res=NULL
 for ( FRAC_DE in c(0.05)) {
@@ -28,7 +28,7 @@ for ( FRAC_DE in c(0.05)) {
     for (N_REPS in c(5)) {
       for (DGE_FUNC in c("deseq2")) {
         for ( SUM_COUNT in c(50000000)) {
-          for ( VARIANCE in c(0.2,0.25,0.3,0.35,0.4)) {
+          for ( VARIANCE in c(0.2,0.3,0.4)) {
             x<-agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets)
             x<-as.data.frame(do.call(rbind, x))
             write.table(x,file="simpw_res_running.tsv",quote=F,sep='\t',append=T)

### 2023-09-07 analysis/sim/simpw_func.R

@@ -389,7 +389,7 @@ ora_up <- as.data.frame(enricher(gene = ups[[d]] ,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
 
 ora_up$geneID <- NULL
-ora_up <- subset(ora_up,p.adjust<0.05 & Count >=10)
+ora_up <- subset(ora_up,p.adjust<0.05 & Count)
 ora_ups <- rownames(ora_up)
 obs_up[[d]] <- ora_ups
 
@@ -399,7 +399,7 @@ ora_dn <- as.data.frame(enricher(gene = dns[[d]] ,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
 
 ora_dn$geneID <- NULL
-ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
+ora_dn <- subset(ora_dn,p.adjust<0.05 & Count)
 ora_dns <- rownames(ora_dn)
 obs_dn[[d]] <- ora_dns
 

### 2023-09-07 analysis/sim/run_simpw.R
working on reimplementing 1 sided fisher test

@@ -12,7 +12,6 @@ a<-countData()
 # generate some gene sets
 gsets<-randomGeneSets(a,setsize=30,nsets=500)
 
-
 # test
 # x<-agg_dge(a=a,N_REPS=5,SUM_COUNT=50000000,VARIANCE=0.4,FRAC_DE=0.1,FC=1,SIMS=8,"deseq2",gsets)
 # str(x)
@@ -21,21 +20,27 @@ gsets<-randomGeneSets(a,setsize=30,nsets=500)
 # run simulations over a range of parameters
 ###############################################
 SIMS=8
-unlink("simpw_res_running.tsv")
-res=NULL
-for ( FRAC_DE in c(0.05)) {
-  for (FC in c(1)) {
-    for (N_REPS in c(5)) {
-      for (DGE_FUNC in c("deseq2")) {
-        for ( SUM_COUNT in c(50000000)) {
-          for ( VARIANCE in c(0.2,0.3,0.4)) {
-            x<-agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets)
-            x<-as.data.frame(do.call(rbind, x))
-            write.table(x,file="simpw_res_running.tsv",quote=F,sep='\t',append=T)
-            res=c(res,x)
-          }
-        }
-      }
-    }
-  }
-}
+FRAC_DE=0.05
+FC=1
+N_REPS=5
+DGE_FUNC="deseq2"
+SUM_COUNT=3e7
+VARIANCE=c(0.2,0.3,0.4,0.5)
+
+mygrid <- expand.grid(FRAC_DE,FC,N_REPS,DGE_FUNC,SUM_COUNT,VARIANCE)
+colnames(mygrid) <- c("FRAC_DE","FC","N_REPS","DGE_FUNC","SUM_COUNT","VARIANCE")
+
+mygrid
+
+res <- lapply(1:nrow(mygrid), function(i) {
+  FRAC_DE=mygrid[i,"FRAC_DE"]
+  FC=mygrid[i,"FC"]
+  N_REPS=mygrid[i,"N_REPS"]
+  DGE_FUNC=as.character(mygrid[i,"DGE_FUNC"])
+  SUM_COUNT=mygrid[i,"SUM_COUNT"]
+  VARIANCE=mygrid[i,"VARIANCE"]
[…]

### 2023-09-07 analysis/sim/simpw_func.R

@@ -285,6 +285,90 @@ x
 
 }
 
+##################################
+# 1-sided fisher test function
+##################################
+run_fisher <- function(x,DGE_FUNC,gsets,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS){
+
+dge<-sapply(x,"[",6)
+
+ups<-lapply(dge, function(x) { rownames(subset(x, padj<0.05 & log2FoldChange > 0)) } )
+dns<-lapply(dge, function(x) { rownames(subset(x, padj<0.05 & log2FoldChange < 0)) } )
+bgs <- lapply(dge, function(x) { rownames(x) })
+
+l_ups<-sapply(ups,length)
+l_dns<-sapply(dns,length)
+geneset_sizes<-sapply( gsets , length )
+
+# calculate number of genes in sets that are up and downregulated
+upreg_incat=gsdet=unreg_incat=upreg_not_incat=unreg_not_incat=NULL
+upreg_incat=gsdet=unreg_incat=upreg_not_incat=unreg_not_incat=list()
+
+dnreg_incat=gsdet=unreg_incat=dnreg_not_incat=unreg_not_incat=NULL
+dnreg_incat=gsdet=unreg_incat=dnreg_not_incat=unreg_not_incat=list()
+
+
+for (d in 1:length(dge)) {
+  universe =length(rownames(dge[[d]]))
+
+  upreg_incat[[d]]<-sapply( 1:length(gsets),function(x){length(which(gsets[[x]] %in% ups[[d]] ))} )
+  gsdet[[d]] <- sapply( 1:length(gsets),function(x){length(which(gsets[[x]] %in% bgs[[d]] ))} )
+  unreg_incat[[d]] <- gsdet[[d]] - upreg_incat[[d]]
+  upreg_not_incat[[d]]<-sapply( 1:length(gsets),function(x){length(which(!ups[[d]] %in% gsets[[x]] ))} )
+  unreg_not_incat[[d]] <- universe - upreg_incat[[d]] - unreg_incat[[d]] - upreg_not_incat[[d]]
+
+  p_ups[[d]] <- sapply( 1:length(gsets),function(x){
[…]

### 2023-09-08 analysis/sim/run_simpw.Rmd
progress

---
title: "Examining background gene lists using simulation - comparison of different enrichment techniques"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/background

```{r,libs}

library("tidyverse")
library("parallel")
library("edgeR")
library("DESeq2")
library("limma")
library("stringi")
library("kableExtra")
source("simpw_func.R")

```

# Intro

TODO

# Get count data

```{r,getcounts}

a<-countData()

```

# Generate gene sets

```{r,genesetgeneration}

gsets<-randomGeneSets(a,setsize=50,nsets=500)

```

# run simulations over a range of parameters

```{r,setparameters}

SIMS=16
FRAC_DE=0.05
FC=1
N_REPS=4
DGE_FUNC="deseq2"
SUM_COUNT=3e7
VARIANCE=c(0.25,0.3,0.35,0.4,0.45)

mygrid <- expand.grid(FRAC_DE,FC,N_REPS,DGE_FUNC,SUM_COUNT,VARIANCE)
colnames(mygrid) <- c("FRAC_DE","FC","N_REPS","DGE_FUNC","SUM_COUNT","VARIANCE")

mygrid

```

Now run the analysis.

```{r,sims}

res <- lapply(1:nrow(mygrid), function(i) {
  FRAC_DE=mygrid[i,"FRAC_DE"]
  FC=mygrid[i,"FC"]
  N_REPS=mygrid[i,"N_REPS"]
  DGE_FUNC=as.character(mygrid[i,"DGE_FUNC"])
  SUM_COUNT=mygrid[i,"SUM_COUNT"]
  VARIANCE=mygrid[i,"VARIANCE"]
  x <- agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets)
  as.data.frame(do.call(rbind, x))
})

```

Now show the results.

```{r,results}

res %>% kbl(caption="simulation_results") %>% kable_paper("hover", full_width = F)

```

# Session information

```{r,session}

sessionInfo()

```

### 2023-09-08 analysis/sim/run_simpw.R

@@ -1,46 +1,0 @@
-library("tidyverse")
-library("parallel")
-library("edgeR")
-library("DESeq2")
-library("limma")
-library("stringi")
-source("simpw_func.R")
-
-# obtain count data
-a<-countData()
-
-# generate some gene sets
-gsets<-randomGeneSets(a,setsize=30,nsets=500)
-
-# test
-# x<-agg_dge(a=a,N_REPS=5,SUM_COUNT=50000000,VARIANCE=0.4,FRAC_DE=0.1,FC=1,SIMS=8,"deseq2",gsets)
-# str(x)
-
-###############################################
-# run simulations over a range of parameters
-###############################################
-SIMS=8
-FRAC_DE=0.05
-FC=1
-N_REPS=5
-DGE_FUNC="deseq2"
-SUM_COUNT=3e7
-VARIANCE=c(0.2,0.3,0.4,0.5)
-
-mygrid <- expand.grid(FRAC_DE,FC,N_REPS,DGE_FUNC,SUM_COUNT,VARIANCE)
-colnames(mygrid) <- c("FRAC_DE","FC","N_REPS","DGE_FUNC","SUM_COUNT","VARIANCE")
-
-mygrid
-
-res <- lapply(1:nrow(mygrid), function(i) {
-  FRAC_DE=mygrid[i,"FRAC_DE"]
-  FC=mygrid[i,"FC"]
-  N_REPS=mygrid[i,"N_REPS"]
-  DGE_FUNC=as.character(mygrid[i,"DGE_FUNC"])
-  SUM_COUNT=mygrid[i,"SUM_COUNT"]
-  VARIANCE=mygrid[i,"VARIANCE"]
-  x <- agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets)
-  as.data.frame(do.call(rbind, x))
-})
-
-res

### 2023-09-08 analysis/sim/simpw_func.R

@@ -303,34 +303,35 @@ geneset_sizes<-sapply( gsets , length )
 # calculate number of genes in sets that are up and downregulated
 upreg_incat=gsdet=unreg_incat=upreg_not_incat=unreg_not_incat=NULL
 upreg_incat=gsdet=unreg_incat=upreg_not_incat=unreg_not_incat=list()
-
 dnreg_incat=gsdet=unreg_incat=dnreg_not_incat=unreg_not_incat=NULL
 dnreg_incat=gsdet=unreg_incat=dnreg_not_incat=unreg_not_incat=list()
-
+p_ups=p_dns=NULL
+p_ups=p_dns=list()
 
 for (d in 1:length(dge)) {
   universe =length(rownames(dge[[d]]))
 
-  upreg_incat[[d]]<-sapply( 1:length(gsets),function(x){length(which(gsets[[x]] %in% ups[[d]] ))} )
-  gsdet[[d]] <- sapply( 1:length(gsets),function(x){length(which(gsets[[x]] %in% bgs[[d]] ))} )
+  upreg_incat[[d]]<-sapply( 1:length(gsets),function(y){length(which(gsets[[y]] %in% ups[[d]] ))} )
+  gsdet[[d]] <- sapply( 1:length(gsets),function(y){length(which(gsets[[y]] %in% bgs[[d]] ))} )
   unreg_incat[[d]] <- gsdet[[d]] - upreg_incat[[d]]
-  upreg_not_incat[[d]]<-sapply( 1:length(gsets),function(x){length(which(!ups[[d]] %in% gsets[[x]] ))} )
+  upreg_not_incat[[d]]<-sapply( 1:length(gsets),function(y){length(which(!ups[[d]] %in% gsets[[y]] ))} )
   unreg_not_incat[[d]] <- universe - upreg_incat[[d]] - unreg_incat[[d]] - upreg_not_incat[[d]]
 
-  p_ups[[d]] <- sapply( 1:length(gsets),function(x){
-    mx <- matrix(c(unreg_incat[[d]][[x]],unreg_not_incat[[d]][[x]],upreg_incat[[d]][[x]] ,upreg_not_incat[[d]][[x]]),ncol=2)
+  p_ups[[d]] <- sapply( 1:length(gsets),function(y){
[…]

### 2023-09-08 analysis/sim/simpw_func.R
try dose

@@ -336,12 +336,12 @@ for (d in 1:length(dge)) {
     fres$p
   })
 
-  x[[d]][[17]]<-names(gsets[which(p.adjust(p_ups[[d]],method="fdr")<0.05)])
-  x[[d]][[18]]<-names(gsets[which(p.adjust(p_dns[[d]],method="fdr")<0.05)])
+  x[[d]][[19]]<-names(gsets[which(p.adjust(p_ups[[d]],method="fdr")<0.05)])
+  x[[d]][[20]]<-names(gsets[which(p.adjust(p_dns[[d]],method="fdr")<0.05)])
 }
 
-obs_up<-sapply(x,"[",17)
-obs_dn<-sapply(x,"[",18)
+obs_up<-sapply(x,"[",19)
+obs_dn<-sapply(x,"[",20)
 
 gt_up<-sapply(x,"[",4)
 gt_up<-lapply( gt_up , names)
@@ -369,7 +369,6 @@ attr(x,'fisher_res') <-data.frame(N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_
 x
 
 }
-
 
 ##################################
 # clusterprofiler default function
@@ -475,7 +474,7 @@ gsets3 <- rbind(gsets2,bgdf)
 # clusterprofiler UP
 ora_up <- as.data.frame(enricher(gene = ups[[d]] ,
   universe = bgs[[d]],  minGSSize=2, maxGSSize = 500000, TERM2GENE = gsets3,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1 ))
 
 ora_up$geneID <- NULL
 ora_up <- subset(ora_up,p.adjust<0.05 & Count)
@@ -485,7 +484,7 @@ obs_up[[d]] <- ora_ups
 # clusterprofiler DOWN
 ora_dn <- as.data.frame(enricher(gene = dns[[d]] ,
   universe = bgs[[d]],  minGSSize=2, maxGSSize = 500000, TERM2GENE = gsets3,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1 ))
 
 ora_dn$geneID <- NULL
[…]

### 2024-05-29 dataprep/bulkrna1.Rmd
bulkRNAseq data

---
title: "Example bulk RNA-seq analysis 1"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/background

## Intro

Here we are performing an analysis of some gene expression data to demonstrate the difference between ORA and FCS methods and to highlight the differences caused by improper background gene set use.

The dataset being used is SRP128998 and we are comparing the cells grown in normal glucose condition (control) to the high glucose condition (case).

Data are obtained from http://dee2.io/

```{r,begin}

suppressPackageStartupMessages({
  library("getDEE2")
  library("DESeq2")
  library("clusterProfiler")
  library("mitch")
  library("kableExtra")
  library("eulerr")
  library("biomaRt")
})

```

## Get expression data and make an MDS plot

```{r,getdata}

name = "SRP128998"
mdat <- getDEE2Metadata("hsapiens")
samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
samplesheet <- samplesheet[order(samplesheet$SRR_accession),]
samplesheet$trt <- as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
samplesheet$VPA <- as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))
s1 <- subset(samplesheet,VPA==0)

s1 %>%
  kbl(caption = "sample sheet") %>%
  kable_paper("hover", full_width = F)

w <- getDEE2("hsapiens", samplesheet$SRR_accession,
  metadata=mdat,legacy = TRUE)

x <- Tx2Gene(w)
x <- x$Tx2Gene

# save the genetable for later
gt <- w$GeneInfo[,1,drop=FALSE]
gt$accession <- rownames(gt)

# counts
x1 <- x[,which(colnames(x) %in% s1$SRR_accession)]

```

Here show the number of genes in the annotation set, and those detected above the detection threshold.

```{r,filter}

# filter out lowly expressed genes
x1 <- x1[which(rowSums(x1)/ncol(x1)>=(10)),]
nrow(x)
nrow(x1)

```

Now multidimensional scaling (MDS) plot to show the correlation between the datasets.
If the control and case datasets are clustered separately, then it is likely that there will be many differentially expressed genes with FDR<0.05.

```{r,mds}

plot(cmdscale(dist(t(x1))), xlab="Coordinate 1", ylab="Coordinate 2", pch=19, col=s1$trt, main="MDS")

```

## Differential expression

Now run DESeq2 for control vs case.

```{r,deseq2}

y <- DESeqDataSetFromMatrix(countData = round(x1), colData = s1, design = ~ trt)
y <- DESeq(y)
de <- results(y)
de <- as.data.frame(de[order(de$pvalue),])
rownames(de) <- sapply(strsplit(rownames(de),"\\."),"[[",1)
head(de) %>% kbl() %>% kable_paper("hover", full_width = F)

```

Now let's have a look at some of the charts showing differential expression.
In particular, an MA plot and volcano plot.

```{r,deplots,fig.width=7,fig.height=7}

maplot <- function(de,contrast_name) {
  sig <-subset(de, padj < 0.05 )
  up <-rownames(subset(de, padj < 0.05 & log2FoldChange > 0))
  dn <-rownames(subset(de, padj < 0.05 & log2FoldChange < 0))
  GENESUP <- length(up)
  GENESDN <- length(dn)
  DET=nrow(de)
  SUBHEADER = paste(GENESUP, "up, ", GENESDN, "down", DET, "detected")
  ns <-subset(de, padj > 0.05 )
  plot(log2(de$baseMean),de$log2FoldChange, 
       xlab="log2 basemean", ylab="log2 foldchange",
       pch=19, cex=0.5, col="dark gray",
       main=contrast_name, cex.main=0.7)
  points(log2(sig$baseMean),sig$log2FoldChange,
         pch=19, cex=0.5, col="red")
  mtext(SUBHEADER,cex = 0.7)
}

make_volcano <- function(de,name) {
    sig <- subset(de,padj<0.05)
    N_SIG=nrow(sig)
    N_UP=nrow(subset(sig,log2FoldChange>0))
    N_DN=nrow(subset(sig,log2FoldChange<0))
    DET=nrow(de)
    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn", DET, "detected")
    plot(de$log2FoldChange,-log10(de$padj),cex=0.5,pch=19,col="darkgray",
        main=name, xlab="log2 FC", ylab="-log10 pval", xlim=c(-6,6))
    mtext(HEADER)
    grid()
    points(sig$log2FoldChange,-log10(sig$padj),cex=0.5,pch=19,col="red")
}

maplot(de,name)
make_volcano(de,name)

```

## Need to add gene symbol

```{r,symbol}

mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

genes <- getBM(filters= "ensembl_gene_id",
  attributes= c("ensembl_gene_id","hgnc_symbol"),
  values=rownames(de), mart= mart)

m <- merge(de,genes,by.x=0,by.y="ensembl_gene_id")
rownames(m) <- paste(m$Row.names,m$hgnc_symbol)
m$Row.names = m$hgnc_symbol = NULL
dim(de)
dim(m)

```

## Save table

```{r,save}

saveRDS(m,"bulkrna1.Rds")

```

## Session information

```{r,session}

sessionInfo()

```

### 2024-05-29 dataprep/bulkrna2.Rmd
bulkRNAseq data

--- 2024-05-29 dataprep/bulkrna1.Rmd
@@
 ---
-title: "Example bulk RNA-seq analysis 1"
+title: "Example bulk RNA-seq analysis 2"
 author: "Mark Ziemann"
@@
 
-The dataset being used is SRP128998 and we are comparing the cells grown in normal glucose condition (control) to the high glucose condition (case).
+The dataset being used is SRP038101 and we are comparing the cells grown in normal condition (control) to those grown with addition of Azacitidine (case).
 
@@
 
-name = "SRP128998"
+name = "SRP038101"
 mdat <- getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
+samplesheet <- mdat[grep("SRP038101",mdat$SRP_accession),]
 samplesheet <- samplesheet[order(samplesheet$SRR_accession),]
-samplesheet$trt <- as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
-samplesheet$VPA <- as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))
-s1 <- subset(samplesheet,VPA==0)
+samplesheet$trt <- as.factor(c(1,1,1,0,0,0))
+s1 <- samplesheet
 
-s1 %>%
-  kbl(caption = "sample sheet") %>%
-  kable_paper("hover", full_width = F)
+s1 %>% kbl(caption = "sample sheet") %>% kable_paper("hover", full_width = F)
 
-w <- getDEE2("hsapiens", samplesheet$SRR_accession,
-  metadata=mdat,legacy = TRUE)
-
+w <- getDEE2("hsapiens",samplesheet$SRR_accession,metadata=mdat,legacy = TRUE)
 x <- Tx2Gene(w)
@@
 # counts 
-x1 <- x[,which(colnames(x) %in% s1$SRR_accession)]
+x1 <- x[,which(colnames(x) %in% samplesheet$SRR_accession)]
 
@@
 
-saveRDS(m,"bulkrna1.Rds")
+saveRDS(m,"bulkrna2.Rds")
 

### 2024-05-29 dataprep/bulkrna3.Rmd
bulkRNAseq data

--- 2024-05-29 dataprep/bulkrna2.Rmd
@@
 ---
-title: "Example bulk RNA-seq analysis 2"
+title: "Example bulk RNA-seq analysis 3"
 author: "Mark Ziemann"
@@
 
-The dataset being used is SRP038101 and we are comparing the cells grown in normal condition (control) to those grown with addition of Azacitidine (case).
+The dataset being used is SRP096178 and we are comparing the cells grown in normal condition (control) to those grown with addition of SAHA (case).
 
@@
 
-name = "SRP038101"
+name = "SRP037718"
 mdat <- getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP038101",mdat$SRP_accession),]
+samplesheet <- mdat[grep("SRP037718",mdat$SRP_accession),]
 samplesheet <- samplesheet[order(samplesheet$SRR_accession),]
+
 samplesheet$trt <- as.factor(c(1,1,1,0,0,0))
@@
 
-w <- getDEE2("hsapiens",samplesheet$SRR_accession,metadata=mdat,legacy = TRUE)
+w <- getDEE2("hsapiens",s1$SRR_accession,metadata=mdat,legacy = TRUE)
 x <- Tx2Gene(w)
@@
 
-saveRDS(m,"bulkrna2.Rds")
+saveRDS(m,"bulkrna3.Rds")
 

### 2024-05-29 dataprep/bulkrna4.Rmd
bulkRNAseq data

--- 2024-05-29 dataprep/bulkrna2.Rmd
@@
 ---
-title: "Example bulk RNA-seq analysis 2"
+title: "Example bulk RNA-seq analysis 4"
 author: "Mark Ziemann"
@@
 
-The dataset being used is SRP038101 and we are comparing the cells grown in normal condition (control) to those grown with addition of Azacitidine (case).
+The dataset being used is SRP038101 and we are comparing the cells expressing a set7kd shRNA construct (case) compared to the scrambled construct (control).
 
@@
 
-name = "SRP038101"
+name = "SRP096177"
 mdat <- getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP038101",mdat$SRP_accession),]
+samplesheet <- mdat[grep("SRP096177",mdat$SRP_accession),]
 samplesheet <- samplesheet[order(samplesheet$SRR_accession),]
@@
 
-saveRDS(m,"bulkrna2.Rds")
+saveRDS(m,"bulkrna4.Rds")
 

### 2024-05-29 dataprep/bulkrna5.Rmd
bulkRNAseq data

--- 2024-05-29 dataprep/bulkrna1.Rmd
@@
 ---
-title: "Example bulk RNA-seq analysis 1"
+title: "Example bulk RNA-seq analysis 5"
 author: "Mark Ziemann"
@@
 
-The dataset being used is SRP128998 and we are comparing the cells grown in normal glucose condition (control) to the high glucose condition (case).
+The dataset being used is SRP097759 and we are comparing the effect of control construct versus SAHH overexpression construct.
 
@@
 
-name = "SRP128998"
+name = "SRP097759"
 mdat <- getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
+samplesheet <- mdat[grep("SRP097759",mdat$SRP_accession),]
 samplesheet <- samplesheet[order(samplesheet$SRR_accession),]
-samplesheet$trt <- as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
-samplesheet$VPA <- as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))
-s1 <- subset(samplesheet,VPA==0)
+samplesheet$trt <- as.factor(c(0,0,1,1))
+s1 <- samplesheet
 
@@
 
-saveRDS(m,"bulkrna1.Rds")
+saveRDS(m,"bulkrna5.Rds")
 

### 2024-05-29 dataprep/bulkrna6.Rmd
bulkRNAseq data

--- 2024-05-29 dataprep/bulkrna2.Rmd
@@
 ---
-title: "Example bulk RNA-seq analysis 2"
+title: "Example bulk RNA-seq analysis 6"
 author: "Mark Ziemann"
@@
 
-The dataset being used is SRP038101 and we are comparing the cells grown in normal condition (control) to those grown with addition of Azacitidine (case).
+The dataset being used is SRP253951 and we are comparing A549 cells with and without infection with SARS-CoV-2.
 
@@
 
-name = "SRP038101"
+name = "SRP253951"
 mdat <- getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP038101",mdat$SRP_accession),]
-samplesheet <- samplesheet[order(samplesheet$SRR_accession),]
-samplesheet$trt <- as.factor(c(1,1,1,0,0,0))
+samplesheet <- mdat[grep("SRP253951",mdat$SRP_accession),]
+
+samplesheet <- mdat[which(mdat$SRX_accession %in% c("SRX8089264","SRX8089265","SRX8089266","SRX8089267","SRX8089268","SRX8089269")),]
+
+samplesheet$trt <- factor(c(0,0,0,1,1,1))
 s1 <- samplesheet
@@
 
-saveRDS(m,"bulkrna2.Rds")
+saveRDS(m,"bulkrna6.Rds")
 

### 2024-05-29 dataprep/bulkrna7.Rmd
bulkRNAseq data

--- 2024-05-29 dataprep/bulkrna2.Rmd
@@
 ---
-title: "Example bulk RNA-seq analysis 2"
+title: "Example bulk RNA-seq analysis 7"
 author: "Mark Ziemann"
@@
 
-The dataset being used is SRP038101 and we are comparing the cells grown in normal condition (control) to those grown with addition of Azacitidine (case).
+The dataset being used is SRP068733 and we are comparing the healthy endothelial cells with a scrambled siRNA to cells treated with a p300 targeting siRNA.
 
@@
 
-name = "SRP038101"
+name = "SRP068733"
 mdat <- getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP038101",mdat$SRP_accession),]
+samplesheet <- mdat[grep("SRP068733",mdat$SRP_accession),]
 samplesheet <- samplesheet[order(samplesheet$SRR_accession),]
-samplesheet$trt <- as.factor(c(1,1,1,0,0,0))
+SRRvec <- c("SRR3112216","SRR3112217","SRR3112218","SRR3112219","SRR3112220","SRR3112221")
+samplesheet <- samplesheet[which(samplesheet$SRR_accession %in% SRRvec),]
+samplesheet$trt <- as.factor(c(0,0,0,1,1,1)) 
 s1 <- samplesheet
@@
 
-w <- getDEE2("hsapiens",samplesheet$SRR_accession,metadata=mdat,legacy = TRUE)
+w <- getDEE2("hsapiens",SRRvec,metadata=mdat,legacy = TRUE)
 x <- Tx2Gene(w)
@@
 
-saveRDS(m,"bulkrna2.Rds")
+saveRDS(m,"bulkrna7.Rds")
 

### 2024-05-31 analysis/analyse_rnaseq.Rmd
prob1 RNAseq done

---
title: "Examining background gene lists in clusterProfiler - RNA-seq example"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/background

## Introduction

This guide is a Rmarkdown script that conducts differential expression and enrichment analysis,
which are very popular workflows for transcriptome data.

The goal of this work is to understand how ClusterProfiler manages the background list,
as we have observed some weird behaviour.
In particular we see that when we provide a custom background, those genes do not appear in the
final analysis unless they are also included in the gene list.

In the code chunk below called `libs`, you can add and remove required R library dependancies.
Check that the libraries listed here match the Dockerfile, otherwise you might get errors.

```{r,libs}

suppressPackageStartupMessages({
  library("getDEE2")
  library("DESeq2")
  library("kableExtra")
  library("clusterProfiler")
  library("fgsea")
  library("eulerr")
  library("gplots")
  library("parallel")
})

```

For this guide I will be using bulk RNA-seq data from several previous studies.

## Load data

Gene sets were obtained from MSigDB.

```{r,load_data}

myfiles <- list.files("../dataprep",pattern="*Rds",full.names=TRUE)
d <- lapply(myfiles,readRDS)
names(d) <- basename(myfiles)

mygmts <- list.files("../gmt",pattern="*gmt",full.names=TRUE)
gs <- lapply(mygmts,gmtPathways)
names(gs) <- basename(mygmts)

names(gs) <- gsub(".v2023.2.Hs.symbols.gmt","",names(gs))

```

## Basic geneset analysis

```{r,gs1}

lapply(gs,length)

lapply(gs,function(x) {
  y <- unlist(lapply(x, length ))
  summary(y)
} )

lapply(gs,function(x) {
  y <- unique(unlist(x))
  length(y)
} )

```

## Enrichment with Clusterprofiler

Here is a function that conducts over-representation analysis with ClusterProfiler.
Genes that are not annotated as having any gene set category are discarded.

```{r,orabad}

# df = deseq output
# df must have "ensID geneSymbol" as rownames

# gs = gene sets
# gs format is list of symbols
orabad <- function( df, gs ) {

  def <- df

  gs1 <- stack(gs)
  colnames(gs1) <- c("gene","term")
  gs1 <- gs1[,c(2,1)]

  defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
  defup <- defup[lapply(strsplit(defup," "),length)==2]
  defup <- unique(sapply(strsplit(defup," "),"[[",2))

  if ( length(defup)<250 ) {
    defup <- head(rownames(subset(def,log2FoldChange>0)),250)
    defup <- defup[lapply(strsplit(defup," "),length)==2]
    defup <- unique(sapply(strsplit(defup," "),"[[",2))
  }

  defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
  defdn <- defdn[lapply(strsplit(defdn," "),length)==2]
  defdn <- unique(sapply(strsplit(defdn," "),"[[",2))

  if ( length(defdn)<250 ) {
    defdn <- head(rownames(subset(def,log2FoldChange<0)),250)
    defdn <- defdn[lapply(strsplit(defdn," "),length)==2]
    defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
  }

  bg <- rownames(def)
  bg <- bg[lapply(strsplit(bg," "),length)==2]
  bg <- unique(sapply(strsplit(bg," "),"[[",2))

  ora_up <- as.data.frame(enricher(gene = defup ,
    universe = bg,  maxGSSize = 500000, TERM2GENE = gs1,
    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

  ora_up$geneID <- NULL
  ora_ups <- rownames(ora_up)

  gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
    as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))

  br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
    as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))

  ora_up$es <- gr/br
  ora_up <- ora_up[order(-ora_up$es),]
  ora_up$Description=NULL

  ora_dn <- as.data.frame(enricher(gene = defdn ,
    universe = bg,  maxGSSize = 500000, TERM2GENE = gs1,
    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

  ora_dn$geneID <- NULL
  ora_dns <- rownames(ora_dn)

  gr <- as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",1)) /
    as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",2))

  br <- as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",1)) /
    as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",2))

  ora_dn$es <- gr/br
  ora_dn <- ora_dn[order(-ora_dn$es),]
  ora_dn$Description=NULL

  result <- list("orabad_up"=ora_up,"orabad_dn"=ora_dn)
  return(result)
}

```

This function fixes the problem with excluding un-annotated genes.

```{r,oragood}

oragood <- function( df, gs ) {

  def <- df

  gs1 <- stack(gs)
  colnames(gs1) <- c("gene","term")
  gs1 <- gs1[,c(2,1)]

  defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
  defup <- defup[lapply(strsplit(defup," "),length)==2]
  defup <- unique(sapply(strsplit(defup," "),"[[",2))

  if ( length(defup)<250 ) {
    defup <- head(rownames(subset(def,log2FoldChange>0)),250)
    defup <- defup[lapply(strsplit(defup," "),length)==2]
    defup <- unique(sapply(strsplit(defup," "),"[[",2))
  }

  defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
  defdn <- defdn[lapply(strsplit(defdn," "),length)==2]
  defdn <- unique(sapply(strsplit(defdn," "),"[[",2))

  if ( length(defdn)<250 ) {
    defdn <- head(rownames(subset(def,log2FoldChange<0)),250)
    defdn <- defdn[lapply(strsplit(defdn," "),length)==2]
    defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
  }

  bg <- rownames(def)
  bg <- bg[lapply(strsplit(bg," "),length)==2]
  bg <- unique(sapply(strsplit(bg," "),"[[",2))

  bgdf <- data.frame("background",bg)
  names(bgdf) <- c("term","gene")
  gs1 <- rbind(gs1,bgdf)

  ora_up <- as.data.frame(enricher(gene = defup ,
    universe = bg,  maxGSSize = 500000, TERM2GENE = gs1,
    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

  ora_up$geneID <- NULL
  ora_ups <- rownames(ora_up)

  gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
    as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))

  br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
    as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))

  ora_up$es <- gr/br
  ora_up <- ora_up[order(-ora_up$es),]
  ora_up$Description=NULL

  ora_dn <- as.data.frame(enricher(gene = defdn ,
    universe = bg,  maxGSSize = 500000, TERM2GENE = gs1,
    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

  ora_dn$geneID <- NULL
  ora_dns <- rownames(ora_dn)

  gr <- as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",1)) /
    as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",2))

  br <- as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",1)) /
    as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",2))

  ora_dn$es <- gr/br
  ora_dn <- ora_dn[order(-ora_dn$es),]
  ora_dn$Description=NULL

  result <- list("oragood_up"=ora_up,"oragood_dn"=ora_dn)
  return(result)
}

```

## Compare the results of bad and good analysis

```{r,compare}

# bad=ORA analysis, a list with results of up and down regulated pathways
# good=same as bad, except fixed the background problem
compare_ora_plots <- function(bad,good){
  badup <- bad$orabad_up
  baddn <- bad$orabad_dn
  goodup <- good$oragood_up
  gooddn <- good$oragood_dn
  mup <- merge(badup,goodup,by=0)
  MAX=max(c(mup$es.x,mup$es.y))

  plot(mup$es.x, mup$es.y,
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original", ylab="corrected", main="Enrichment score")
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)

  MAX=max(-log10(c(mup$pvalue.x,mup$pvalue.y)))

  plot(-log10(mup$pvalue.x), -log10(mup$pvalue.y),
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original", ylab="corrected", main="-log10 p-value")
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)

  mdn <- merge(baddn,gooddn,by=0)
  MAX=max(c(mdn$es.x,mdn$es.y))

  plot(mdn$es.x, mdn$es.y,
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original", ylab="corrected", main="Enrichment score")
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)

  MAX=max(-log10(c(mdn$pvalue.x,mdn$pvalue.y)))

  plot(-log10(mdn$pvalue.x), -log10(mdn$pvalue.y),
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original", ylab="corrected", main="-log10 p-value")
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)
}

compare_ora <- function(bad,good){
  badup <- bad$orabad_up
  baddn <- bad$orabad_dn
  goodup <- good$oragood_up
  gooddn <- good$oragood_dn
  mup <- merge(badup,goodup,by=0)
  mdn <- merge(baddn,gooddn,by=0)
  bup <- subset(mup,p.adjust.x<0.05)$Row.names
  gup <- subset(mup,p.adjust.y<0.05)$Row.names
  bdn <- subset(mdn,p.adjust.x<0.05)$Row.names
  gdn <- subset(mdn,p.adjust.y<0.05)$Row.names
  sig_bad <- unique(c(bup,bdn))
  sig_good <- unique(c(gup,gdn))
  nsig_bad <- length(sig_bad)
  nsig_good <- length(sig_good)
  jac <- length(intersect(sig_bad,sig_good)) / length(union(sig_bad,sig_good))
  result=c("nsig_bad"=nsig_bad,"nsig_good"=nsig_good,"jaccard"=jac)
  return(result)
}

```

## Run some analyses

```{r,run1}

params <- expand.grid(1:length(d),1:length(gs))
params2 <- expand.grid(names(d),names(gs))

colnames(params) <- c("d","gs")

ora_dat <- mclapply(1:nrow(params), function(i) {
  j=params[i,1]
  k=params[i,2]
  bad <- orabad(d[[j]],gs[[k]])
  good <- oragood(d[[j]],gs[[k]])
  res <- compare_ora(bad,good)
  return(res)
} , mc.cores=8)

ora_dat <- do.call(rbind,ora_dat)

res1 <- cbind(params2,ora_dat)

summary(res1$nsig_bad)
summary(res1$nsig_good)
summary(res1$jaccard)

# NUMBER OF SIGNIFICANT SETS
ns1 <- matrix( res1$nsig_bad , ncol=9 )
rownames(ns1) <- names(d)
colnames(ns1) <- names(gs)

ns1 %>%
  kbl(row.names = TRUE, caption="significant genesets in original analysis") %>%
  kable_paper("hover", full_width = F)

ns1 <- apply(ns1,2,function(x) { x <- x[!is.na(x)] ; x <- x[is.finite(x)] ;  mean(x) } )

ns2 <- matrix( res1$nsig_good , ncol=9 )
rownames(ns2) <- names(d)
colnames(ns2) <- names(gs)

ns2 %>%
  kbl(row.names = TRUE, caption="significant genesets in corrected analysis") %>%
  kable_paper("hover", full_width = F)

ns2 <- apply(ns2,2,function(x) { x <- x[!is.na(x)] ; x <- x[is.finite(x)] ;  mean(x) } )
ns2

df <- data.frame(ns1,ns2)
df <- df[order(df$ns1),]

par(mar=c(c(5.1, 10.1, 4.1, 2.1) ))
barplot(t(df),beside=TRUE,horiz=TRUE,las=1,
  xlab="no. significant sets", main="Comparison of original and corrected data",
  col=c("gray30","gray80"))

legend("bottomright", inset=.02, legend=c("original","corrected"),
  fill=c("gray30","gray80"), horiz=TRUE, cex=1)

# RATIO
rat <- matrix( (res1$nsig_good+0) / (res1$nsig_bad+0) , ncol=9 )
rownames(rat) <- names(d)
colnames(rat) <- names(gs)

signif(rat,3) %>%
  kbl(row.names = TRUE, caption="Ratio of significant genesets Corrected:Original") %>%
  kable_paper("hover", full_width = F)

rat2 <- apply(rat,2,function(x) { x <- x[!is.na(x)] ; x <- x[is.finite(x)] ;  mean(x) } )
rat2

barplot(sort(rat2),horiz=TRUE,las=1,main="increase in significant sets after correction",xlab="fold increase")

# JACCARD
jac <- matrix( res1$jaccard , ncol=9 )
rownames(jac) <- names(d)
colnames(jac) <- names(gs)

signif(jac,3) %>%
  kbl(row.names = FALSE, caption="Jaccard similarity Corrected:Original") %>%
  kable_paper("hover", full_width = F)

jac2 <- apply(jac,2,function(x) { x <- x[!is.na(x)] ; x <- x[is.finite(x)] ;  mean(x) } )
jac2

barplot(sort(jac2),horiz=TRUE,las=1,xlim=c(0,1),
  main="Jaccard similarity Corrected:Original",xlab="fold increase")

```

## Session information

For reproducibility

```{r,session}

sessionInfo()

```

### 2024-05-31 analysis/analyse_rnaseq.Rmd
fix

@@
 
-par(mar=c(c(5.1, 10.1, 4.1, 2.1) ))
+par(mar=c(c(5.1, 7.1, 4.1, 2.1) ))
 barplot(t(df),beside=TRUE,horiz=TRUE,las=1,
@@
 signif(jac,3) %>%
-  kbl(row.names = FALSE, caption="Jaccard similarity Corrected:Original") %>%
+  kbl(row.names = TRUE, caption="Jaccard similarity Corrected:Original") %>%
   kable_paper("hover", full_width = F)

### 2024-05-31 analysis/analyse_rnaseq.Rmd
show enrichment and logpval discrepancy

@@
 
+res1 %>%
+  kbl(row.names = TRUE, caption="all results") %>%
+  kable_paper("hover", full_width = F)
+
 summary(res1$nsig_bad)
@@
 
+## Make charts showing changes in pvalues
+
+For bulk RNA-seq data set #1, run enrichment with all the gene set libraries.
+
+```{r,charts1}
+
+names(gs)
+
+lapply(1:length(gs), function(i) {
+  bad <- orabad(d[[1]],gs[[i]])
+  good <- oragood(d[[1]],gs[[i]])
+  compare_ora_plots(bad,good)
+} )
+
+```
+
 ## Session information

### 2024-05-31 analysis/analyse_rnaseq1.Rmd
minsetsize

--- 2024-05-31 analysis/analyse_rnaseq.Rmd
@@
   ora_up <- as.data.frame(enricher(gene = defup ,
-    universe = bg,  maxGSSize = 500000, TERM2GENE = gs1,
+    universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = gs1,
     pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
@@
   ora_dn <- as.data.frame(enricher(gene = defdn ,
-    universe = bg,  maxGSSize = 500000, TERM2GENE = gs1,
+    universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = gs1,
     pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

### 2024-05-31 analysis/analyse_rnaseq2.Rmd
added fdr analysis

--- 2024-05-31 analysis/analyse_rnaseq1.Rmd
@@
 
+  gs1 <- gs1[which(gs1$gene %in% bg),]
+
   ora_up <- as.data.frame(enricher(gene = defup ,
@@
 
-This function fixes the problem with excluding un-annotated genes.
+This function fixes the problem with not inlcuding sets for FDR correction.
 
@@
 
-  bgdf <- data.frame("background",bg)
-  names(bgdf) <- c("term","gene")
-  gs1 <- rbind(gs1,bgdf)
+  gs1 <- gs1[which(gs1$gene %in% bg),]
+  terms <- names(which(table(gs1$term)>5))
+  gs1 <- gs1[gs1$term %in% terms,]
 
@@
 
+  nsets <- length(which(table(gs1$term)>5))
+  nres <- nrow(ora_up)
+  diff <- nsets - nres
+  pvals <- c(ora_up$pvalue,rep(1,diff))
+  ora_up$p.adjust <- p.adjust(pvals,method="fdr")[1:nrow(ora_up)]
+
   ora_dn <- as.data.frame(enricher(gene = defdn ,
@@
   ora_dn$Description=NULL
+
+  nsets <- length(which(table(gs1$term)>5))
+  nres <- nrow(ora_dn)
+  diff <- nsets - nres
+  pvals <- c(ora_dn$pvalue,rep(1,diff))
+  ora_dn$p.adjust <- p.adjust(pvals,method="fdr")[1:nrow(ora_dn)]
 
@@
 } , mc.cores=8)
+
+ora_dat <- lapply(1:3, function(i) {
+  j=params[i,1]
+  k=params[i,2]
+  bad <- orabad(d[[j]],gs[[k]])
+  good <- oragood(d[[j]],gs[[k]])
+  res <- compare_ora(bad,good)
+  return(res)
+} )
 

### 2024-05-31 analysis/analyse_rnaseq1.Rmd
fix
better gene set stats

@@
 
+# number of sets
 lapply(gs,length)
 
+# set size summary
 lapply(gs,function(x) {
@@
 
+# gene coverage
 lapply(gs,function(x) {
@@
   length(y)
+} )
+
+# total number of annotations
+lapply(gs, function(x) {
+  sum(unlist(lapply(x,length)))
 } )
@@
 barplot(sort(jac2),horiz=TRUE,las=1,xlim=c(0,1),
-  main="Jaccard similarity Corrected:Original",xlab="fold increase")
+  main="Jaccard similarity Corrected:Original",xlab="Jaccard index")
 

### 2024-05-31 analysis/analyse_rnaseq2.Rmd
fix
better gene set stats

@@
   bg <- unique(sapply(strsplit(bg," "),"[[",2))
-
-  gs1 <- gs1[which(gs1$gene %in% bg),]
 
@@
   mup <- merge(badup,goodup,by=0)
-  MAX=max(c(mup$es.x,mup$es.y))
 
-  plot(mup$es.x, mup$es.y,
+  MAX=max(-log10(c(mup$p.adjust.x,mup$p.adjust.y)))
+
+  plot(-log10(mup$p.adjust.x), -log10(mup$p.adjust.y),
     xlim=c(0,MAX), ylim=c(0,MAX),
-    xlab="original", ylab="corrected", main="Enrichment score")
-  grid()
-  abline(a = 0, b = 1,lwd=2,lty=2)
-
-  MAX=max(-log10(c(mup$pvalue.x,mup$pvalue.y)))
-
-  plot(-log10(mup$pvalue.x), -log10(mup$pvalue.y),
-    xlim=c(0,MAX), ylim=c(0,MAX),
-    xlab="original", ylab="corrected", main="-log10 p-value")
+    xlab="original", ylab="corrected", main="-log10 FDR")
   grid()
@@
   mdn <- merge(baddn,gooddn,by=0)
-  MAX=max(c(mdn$es.x,mdn$es.y))
 
-  plot(mdn$es.x, mdn$es.y,
+  MAX=max(-log10(c(mdn$p.adjust.x,mdn$p.adjust.y)))
+
+  plot(-log10(mdn$p.adjust.x), -log10(mdn$p.adjust.y),
     xlim=c(0,MAX), ylim=c(0,MAX),
-    xlab="original", ylab="corrected", main="Enrichment score")
-  grid()
-  abline(a = 0, b = 1,lwd=2,lty=2)
-
-  MAX=max(-log10(c(mdn$pvalue.x,mdn$pvalue.y)))
-
-  plot(-log10(mdn$pvalue.x), -log10(mdn$pvalue.y),
-    xlim=c(0,MAX), ylim=c(0,MAX),
-    xlab="original", ylab="corrected", main="-log10 p-value")
+    xlab="original", ylab="corrected", main="-log10 FDR")
   grid()
@@
   gooddn <- good$oragood_dn
-  mup <- merge(badup,goodup,by=0)
-  mdn <- merge(baddn,gooddn,by=0)
-  bup <- subset(mup,p.adjust.x<0.05)$Row.names
-  gup <- subset(mup,p.adjust.y<0.05)$Row.names
-  bdn <- subset(mdn,p.adjust.x<0.05)$Row.names
-  gdn <- subset(mdn,p.adjust.y<0.05)$Row.names
+  bup <- rownames(subset(badup,p.adjust<0.05))
+  gup <- rownames(subset(goodup,p.adjust<0.05))
+  bdn <- rownames(subset(baddn,p.adjust<0.05))
+  gdn <- rownames(subset(gooddn,p.adjust<0.05))
   sig_bad <- unique(c(bup,bdn))
@@
 } , mc.cores=8)
-
-ora_dat <- lapply(1:3, function(i) {
-  j=params[i,1]
-  k=params[i,2]
-  bad <- orabad(d[[j]],gs[[k]])
-  good <- oragood(d[[j]],gs[[k]])
-  res <- compare_ora(bad,good)
-  return(res)
-} )
 
@@
 barplot(sort(jac2),horiz=TRUE,las=1,xlim=c(0,1),
-  main="Jaccard similarity Corrected:Original",xlab="fold increase")
+  main="Jaccard similarity Corrected:Original",xlab="Jaccard index")
+
+# overall
+oa <- matrix(paste(res1$nsig_bad,res1$nsig_good,signif(res1$jaccard,2)),ncol=9)
+rownames(oa) <- names(d)
+colnames(oa) <- names(gs)
+
+oa %>%
+  kbl(row.names = TRUE, caption="Original:Corrected:Jaccard") %>%
+  kable_paper("hover", full_width = F)
 

### 2024-05-31 old/analysis/sim/simpw_func.R
sim

@@ -1,0 +1,1087 @@
+library("tidyverse")
+library("parallel")
+library("edgeR")
+library("DESeq2")
+library("limma")
+library("stringi")
+library("mitch")
+library("fgsea")
+library("clusterProfiler")
+
+########################################
+# get some counts
+########################################
+countData<-function() {
+# a is orig expression data
+a<-read.table("https://raw.githubusercontent.com/markziemann/simde/master/start_data/ERR2539161/ERR2539161.se.tsv")
+
+# GI is the gene information. It has gene name information for mapping accessions to gene symbols in GMT files
+gi<-read.table("https://raw.githubusercontent.com/markziemann/simde/master/start_data/ERR2539161/GeneInfo.tsv",header=T,row.names=1)
+
+# merge gene names
+aa<-merge(a,gi,by=0)
+aa<-aa[,-c(4:7)]
+aa<-aggregate(. ~ GeneSymbol,aa,function(x) sum(as.numeric(as.character(x))))
+aa$Row.names=NULL
+rownames(aa)<-aa$GeneSymbol
+aa$GeneSymbol=NULL
+a<-aa[which(aa$ERR2539161>=10),,drop=F]
+a
+}
+
+########################################
+# generate some gene sets
+########################################
+randomGeneSets<-function(a,setsize,nsets){
+gsets<-sapply( rep(setsize,nsets) , function(x) {list(as.character(sample(rownames(a),x))) } )
+names(gsets)<-stri_rand_strings(length(gsets), 15, pattern = "[A-Za-z]")
+gsets
+}
+
+########################################
+# simulate some gene expression data
+########################################
+simrna<-function(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,gsets) {
+
+# N_REPS=5 ; SUM_COUNT=10000000 ; VARIANCE=0.2 ; FRAC_DE=0.05 ; FC=1 ; GMT="ReactomePathways.gmt"
+
+library("edgeR")
+
+df = NULL
+for (k in paste0("data",1:(N_REPS*2)))  {
[…]

### 2024-05-31 analysis/analyse_rnaseq2.Rmd
effect of gene list length

@@
 # gs format is list of symbols
-orabad <- function( df, gs ) {
+orabad <- function( df, gs, ngenes=2000 ) {
 
@@
 
-  defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
+  defup <- head(rownames(subset(def,log2FoldChange>0)),ngenes)
   defup <- defup[lapply(strsplit(defup," "),length)==2]
@@
 
-  if ( length(defup)<250 ) {
-    defup <- head(rownames(subset(def,log2FoldChange>0)),250)
-    defup <- defup[lapply(strsplit(defup," "),length)==2]
-    defup <- unique(sapply(strsplit(defup," "),"[[",2))
-  }
-
-  defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
+  defdn <- head(rownames(subset(def,log2FoldChange<0)),ngenes)
   defdn <- defdn[lapply(strsplit(defdn," "),length)==2]
   defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
-
-  if ( length(defdn)<250 ) {
-    defdn <- head(rownames(subset(def,log2FoldChange<0)),250)
-    defdn <- defdn[lapply(strsplit(defdn," "),length)==2]
-    defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
-  }
 
@@
 
-oragood <- function( df, gs ) {
+oragood <- function( df, gs, ngenes=2000 ) {
 
@@
 
+Now run an exhaustive analysis, changing the gene list length from 125 to 2000.
+
+```{r,run2}
+
+ngenes <- c(125,250,500,1000,2000)
+params <- expand.grid(1:length(d),1:length(gs),ngenes)
+params2 <- expand.grid(names(d),names(gs),ngenes)
+
+colnames(params) <- c("d","gs","ngenes")
+
+ora_dat <- mclapply(1:nrow(params), function(i) {
+  j=params[i,1]
+  k=params[i,2]
+  l=params[i,3]
+  bad <- orabad(d[[j]],gs[[k]],l)
+  good <- oragood(d[[j]],gs[[k]],l)
+  res <- compare_ora(bad,good)
+  return(res)
+} , mc.cores=8)
+
+ora_dat <- do.call(rbind,ora_dat)
+
+res1 <- cbind(params2,ora_dat)
+
+res1 %>%
+  kbl(row.names = TRUE, caption="all results") %>%
+  kable_paper("hover", full_width = F)
+
+summary(res1$nsig_bad)
+summary(res1$nsig_good)
+summary(res1$jaccard)
+
+lapply(ngenes, function(n) {
+  res2 <- subset(res1,Var3==n) ; summary(res2$nsig_bad)
+} )
+
+lapply(ngenes, function(n) {
+  res2 <- subset(res1,Var3==n) ; summary(res2$nsig_good)
+} )
+
+lapply(ngenes, function(n) {
+  res2 <- subset(res1,Var3==n) ; summary(res2$jaccard)
+} )
+
+# NUMBER OF SIGNIFICANT SETS
+
+ns1 <- lapply(ngenes ,function(n) {
+  res2 <- subset(res1,Var3==n)
+  ns <- matrix( res2$nsig_bad , ncol=9 )
+  rownames(ns) <- names(d)
+  colnames(ns) <- names(gs)
+  return(ns)
+} )
+
+names(ns1) <- ngenes
+
+ns1
+
+ns1 <- lapply(ns1,function(y) {
+  res <- apply(y,2,function(x) {
+    x <- x[!is.na(x)]
+    x <- x[is.finite(x)]
+    mean(x)
+  } )
+  return(res)
+})
+
+ns1
+
+ns2 <- lapply(ngenes ,function(n) {
+  res2 <- subset(res1,Var3==n)
+  ns <- matrix( res2$nsig_good , ncol=9 )
+  rownames(ns) <- names(d)
+  colnames(ns) <- names(gs)
+  return(ns)
+} )
+
+names(ns2) <- ngenes
+
+ns2
+
+ns2 <- lapply(ns2,function(y) {
+  res <- apply(y,2,function(x) {
+    x <- x[!is.na(x)]
+    x <- x[is.finite(x)]
+    mean(x)
+  } )
+  return(res)
+})
+
+ns2
+
+# JACCARD
+jac <- lapply(ngenes ,function(n) {
+  res2 <- subset(res1,Var3==n)
+  ns <- matrix( res2$jaccard , ncol=9 )
+  rownames(ns) <- names(d)
+  colnames(ns) <- names(gs)
+  return(ns)
+} )
+names(jac) <- ngenes
+
+jac
+
+jac <- lapply(jac,function(y) {
+  res <- apply(y,2,function(x) {
+    x <- x[!is.na(x)]
+    x <- x[is.finite(x)]
+    mean(x)
+  } )
+  return(res)
+})
+
+jac
+
+jac <- do.call(rbind,jac)
+
+plot(jac[,"c5.go"],type="b", pch=19,cex=1.5,lwd=2,
+  xaxt = "n",ylim=c(0,1),ylab="jaccard",xlab="no. genes selected")
+axis(1, at = seq(1, 5, by = 1),
+     labels = ngenes)
+points(jac[,"c2.cp.reactome"],type="b",col="red",pch=19,cex=1.5,lwd=2)
+points(jac[,"c8.all"],type="b",col="blue",pch=19,cex=1.5,lwd=2)
+points(jac[,"c3.tft.gtrd"],type="b",col="gray50",pch=19,cex=1.5,lwd=2)
+mtext("impact of gene list length on severity of FDR problem")
+
+legend("bottomright", legend=c("GO","Reactome", "Markers", "TFT"),
+       col=c("black", "red", "blue", "gray50"), lty=1, cex=1.2,lwd=3,pch=19)
+
+```
+
 ## Make charts showing changes in pvalues
@@
 lapply(1:length(gs), function(i) {
-  bad <- orabad(d[[1]],gs[[i]])
-  good <- oragood(d[[1]],gs[[i]])
+  bad <- orabad(d[[1]],gs[[i]],ngenes=125)
+  good <- oragood(d[[1]],gs[[i]],ngenes=125)
+  compare_ora_plots(bad,good)
+} )
+
+lapply(1:length(gs), function(i) {
+  bad <- orabad(d[[1]],gs[[i]],ngenes=250)
+  good <- oragood(d[[1]],gs[[i]],ngenes=250)
+  compare_ora_plots(bad,good)
+} )
+
+lapply(1:length(gs), function(i) {
+  bad <- orabad(d[[1]],gs[[i]],ngenes=500)
+  good <- oragood(d[[1]],gs[[i]],ngenes=500)
+  compare_ora_plots(bad,good)
+} )
+
+lapply(1:length(gs), function(i) {
+  bad <- orabad(d[[1]],gs[[i]],ngenes=1000)
+  good <- oragood(d[[1]],gs[[i]],ngenes=1000)
+  compare_ora_plots(bad,good)
+} )
+
+lapply(1:length(gs), function(i) {
+  bad <- orabad(d[[1]],gs[[i]],ngenes=2000)
+  good <- oragood(d[[1]],gs[[i]],ngenes=2000)
   compare_ora_plots(bad,good)

### 2024-05-31 old/analysis/example_proteomics.Rmd
move

--- 2023-08-30 analysis/example_proteomics.Rmd
@@
 | P-value correction for multiple comparisons | FDR method |
-| Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| Background list | Robust detection according to the study authors |
+| Gene Selection Criteria | limma FDR<0.05 |
 | ORA directionality | Separate tests for up- and down-regulation |
@@
 
-png("prot_kegg_euler.png")
+svg("prot_kegg_euler.svg")
 plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
@@
 
-png("prot_reactome_euler.png")
+svg("prot_reactome_euler.svg")
 plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")

### 2024-05-31 old/analysis/example_rnaseq.Rmd
move

--- 2023-08-30 analysis/example_rnaseq.Rmd
@@
 
-png("rna_kegg_euler")
+svg("rna_kegg_euler.svg")
 plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
@@
 
-png("rna_reactome_euler")
+svg("rna_reactome_euler.svg")
 plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")

### 2024-05-31 old/analysis/example_scrnaseq.Rmd
move

--- 2023-08-02 analysis/example_scrnaseq.Rmd
@@
 | P-value correction for multiple comparisons | FDR method |
-| Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | Muscat FDR<0.05 CD4 T-cells |
+| Background list | As defined by Seurat |
+| Gene Selection Criteria | Muscat FDR<0.05 |
 | ORA directionality | Separate tests for up- and down-regulation |
@@
 | P-value correction for multiple comparisons | FDR method |
-| Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| Background list | As defined by Seurat |
+| Gene Selection Criteria | Muscat FDR<0.05 |
 | ORA directionality | Separate tests for up- and down-regulation |
@@
 
+svg("scrna_kegg_euler.svg")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
+dev.off()
+
 ```
@@
 
+svg("scrna_reactome_euler.svg")
+plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
+dev.off()
+
 ```

### 2024-05-31 old/analysis/muscat_vignette.Rmd
move

---
title: "Differential state analysis with `muscat`"
author:
- name: Helena L Crowell
  affiliation:
  - &IMLS Institute for Molecular Life Sciences, University of Zurich, Zurich, Switzerland
  - &SIB Swiss Institute of Bioinformatics (SIB), Zurich, Switzerland
  email: helena.crowell@uzh.ch
- name: Charlotte Soneson
  affiliation:
  - *IMLS
  - 'Present address: Friedrich Miescher Institute Basel, Switzerland <br> & Swiss Institute of Bioinformatics (SIB), Basel, Switzerland'
- name: Pierre-Luc Germain
  affiliation:
  - *IMLS
  - *SIB
- name: Mark D Robinson
  affiliation:
  - *IMLS
  - *SIB
package: "`r BiocStyle::pkg_ver('muscat')`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{"1. DS analysis"}
  %\VignettePackage{muscat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "`r file.path(system.file('extdata', package = 'muscat'), 'refs.bib')`"
abstract: > 
  <p> Single-cell RNA sequencing (scRNA-seq) has quickly become an empowering technology to profile the transcriptomes of individual cells on a large scale. Many early analyses of differential expression have aimed at identifying differences between subpopulations, and thus are focused on finding subpopulation markers either in a single sample or across multiple samples. More generally, such methods can compare expression levels in multiple sets of cells, thus leading to cross-condition analyses.
  <p> However, given the emergence of replicated multi-condition scRNA-seq  datasets,  an  area  of  increasing  focus  is making  sample-level inferences, termed here as *differential state* (DS) analysis. For example, one could investigate the condition-specific responses of cell subpopulations measured from patients in each condition.
  <p> `muscat`: **mu**lti-sample **mu**lti-group **sc**RNA-seq **a**nalysis **t**ools [@Crowell2019-muscat] provides various methods and visualization tools for DS analysis in multi-sample, multi-group, multi-(cell-)subpopulation scRNA-seq data, including cell-level mixed models and methods based on aggregated “pseudobulk” data, as well as a flexible simulation platform that mimics both single and multi-sample scRNA-seq data.
---

<style type="text/css">
.smaller {
  font-size: 10px
}
</style>

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(BiocStyle)
library(cowplot)
```

***

For details on the concept and technicalities of DS analysis, and the methods presented here, consider having a look at our publication:

> Crowell HL, Soneson C\*, Germain P-L\*, Calini D,  
Collin L, Raposo C, Malhotra D, and Robinson MD:  
*muscat* detects subpopulation-specific state transitions from   
multi-sample multi-condition single-cell transcriptomics data.  
Nature Communications **11**, 6077 (2020).  
[DOI: 10.1038/s41467-020-19894-4](https://doi.org/10.1038/s41467-020-19894-4)

# Load packages {-}

```{r load-libs, message = FALSE,  warning = FALSE}
library(dplyr)
library(ggplot2)
library(limma)
library(muscat)
library(purrr)
```

# Introduction

## What is DS analysis?

A fundamental task in the analysis of single-cell RNA-sequencing (scRNA-seq) data is the identification of systematic transcriptional changes [@Stegle2015]. Such analyses are a critical step in the understanding of molecular responses, and have applications in development, in perturbation studies or in disease.  
Most of the current scRNA-seq differential expression (DE) analysis methods are designed to test one set of cells against another (or more generally, multiple sets together), and can be used to compare cell clusters (e.g., for identifying marker genes) or across conditions (cells from one condition versus another) [@Soneson2018]. In such statistical models, the cells are the experimental units and thus represent the population that inferences will extrapolate to.

Using established terminology, we refer to cell *identity* as the combination of cell *type*, a stable molecular signature, and cell *state*, a transient snapshot of a cell's molecular events [@Wagner2016; @Trapnell2015]. This classification is inherently arbitrary, but still provides a basis for biological interpretation and a framework for discovering interesting expression patterns from scRNA-seq datasets. For example, T cells could be defined as a single (albeit diverse) cell type or could be divided into discrete subtypes, if relevant information to categorize each cell at this level were available. In either case, the framework presented here would be able to focus on the cell type of interest and look for changes (in expression) across samples.  
Given the emergence of multi-sample multi-group scRNA-seq datasets, the goal becomes making sample-level inferences (i.e., experimental units are samples). Thus, differential state (DS) analysis is defined as following a given cell type across a set of samples (e.g., individuals) and experimental conditions (e.g., treatments), in order to identify cell-type-specific responses, i.e., changes in cell state. DS analysis: i) should be able to detect diluted changes that only affect a single cell type, a subset of cell types or even a subset of a single subpopulation; and, ii) is intended to be orthogonal to clustering or cell type assignment.

## Starting point

The starting point for a DS analysis is a (sparse) matrix of gene expression, either as counts or some kind of normalized data, where rows = genes and columns = cells. Each cell additionally has a cluster (subpopulation) label as well as a sample label; metadata should accompany the list of samples, such that they can be organized into comparable groups with sample-level replicates (e.g., via a design matrix).

The approach presented here is modular and thus subpopulation labels could originate from an earlier step in the analysis, such as clustering [@Duo2018; @Freytag2018-clustering], perhaps after integration  [@Butler2018-Seurat; @Stuart2019] or after labeling of clusters [@Diaz-Mejia2019] or after cell-level type assignment [@Zhang2019].

# Getting started

## Data description

For this vignette, we will use a `r Biocpkg("SingleCellExperiment")` (SCE) containing 10x droplet-based scRNA-seq PBCM data from 8 Lupus patients obtained before and after 6h-treatment with IFN-$\beta$ [@Kang2018-demuxlet]. The complete raw data, as well as gene and cell metadata is available through the NCBI GEO, accession number [GSE96583](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE96583).

## Loading the data

The @Kang2018-demuxlet dataset has been made available through Bioconductor's `ExperimentHub` and can be loaded into R as follows: We first initialize a Hub instance to search for and load available data with the `ExperimentHub` function, and store the complete list of records in the variable `eh`. Using `query`, we then retrieve any records that match our keyword(s) of interest, as well as their corresponding accession ID (EH1234).

```{r eh, message = FALSE}
library(ExperimentHub)
eh <- ExperimentHub()
query(eh, "Kang")
```

Finally, we load the data of interest into R via `[[` and the corresponding accession ID. The dataset contains >35,000 genes and ~29,000 cells:

```{r load-data, message = FALSE}
(sce <- eh[["EH2259"]])
```

## Preprocessing

The `r Biocpkg("scater")` package [@McCarthy2017-scater] provides a variety of tools for preprocessing and quality control of single-cell transcriptomic data. For completeness, we will apply some minimal filtering steps to

- remove undetected genes
- remove cells with very few or many detected genes
- remove very lowly expressed genes
- compute normalized expression values for visualization

For more thorough preprocessing, we refer to the [Quality control with *scater*](https://bioconductor.org/packages/release/bioc/vignettes/scater/inst/doc/vignette-qc.html) vignette.

```{r fil}
# remove undetected genes
sce <- sce[rowSums(counts(sce) > 0) > 0, ]
dim(sce)
```

We use `perCellQCMetrics` to compute various per-cell quality control metrics, and proceed with filtering cells and genes as noted above:

```{r qc, message = FALSE}
# calculate per-cell quality control (QC) metrics
library(scater)
qc <- perCellQCMetrics(sce)

# remove cells with few or many detected genes
ol <- isOutlier(metric = qc$detected, nmads = 2, log = TRUE)
sce <- sce[, !ol]
dim(sce)

# remove lowly expressed genes
sce <- sce[rowSums(counts(sce) > 1) >= 10, ]
dim(sce)
```

Finally, we use `logNormCounts` to calculate log$_2$-transformed normalized expression values by dividing each count by its size factor, adding a pseudo-count of 1, and log-transforming[^1].

[^1]: Note that, in this workflow, expression values are used for visualization only, and that differential analyses are performed on pseudobulks (section \@ref(sec-pbDS)) or the count data directly (section \@ref(sec-mmDS)).

```{r norm}
# compute sum-factors & normalize
sce <- computeLibraryFactors(sce)
sce <- logNormCounts(sce)
```

Alternatively, expression values could be obtained via `vst` (variance stabilizing transformation) from the `r CRANpkg("sctransform")` package [@Hafemeister2019-sctransform], which returns Pearson residuals from a regularized negative binomial regression model that can be interpreted as normalized expression values:

```{r vst, eval = FALSE}
library(sctransform)
assays(sce)$vstresiduals <- vst(counts(sce), verbosity = FALSE)$y
```

By default, `r BiocStyle::Biocpkg("scater")`'s functions will try to access the assay data specified via argument `exprs_values` (default `logcounts`) for e.g. visualization and dimension reduction. When an alternative assay such as the `vstresiduals` above should be used, it is thus necessary to explicitly specify this, for example, via `runUMAP(sce, exprs_values = "vstresiduals")` to compute UMAP cell embeddings on the assay data compute above.

## Data preparation

`muscat` expects a certain format of the input SCE. Specifically, the following cell metadata (`colData`) columns have to be provided:

- `"sample_id"`: unique sample identifiers (e.g., PeterPan_ref1, Nautilus_trt3, ...)
- `"cluster_id"`: subpopulation (cluster) assignments (e.g., T cells, monocytes, ...)
- `"group_id"`: experimental group/condition (e.g., control/treatment, healthy/diseased, ...)

```{r prep-sce} 
sce$id <- paste0(sce$stim, sce$ind)
(sce <- prepSCE(sce, 
    kid = "cell", # subpopulation assignments
    gid = "stim",  # group IDs (ctrl/stim)
    sid = "id",   # sample IDs (ctrl/stim.1234)
    drop = TRUE))  # drop all other colData columns
```

For consistency and easy accession throughout this vignette, we will store cluster and sample IDs, as well as the number of clusters and samples into the following *simple* variables:

```{r ids}
nk <- length(kids <- levels(sce$cluster_id))
ns <- length(sids <- levels(sce$sample_id))
names(kids) <- kids; names(sids) <- sids
```

## Data overview

### Cluster-sample sizes

As we will be aggregating measurements at the cluster-sample level, it is of particular importance to check the number of cells captured for each such instance. While `aggregateData` (see Section \@ref(sec-agg)) allows excluding cluster-sample combinations with less than a threshold number of cells, clusters or samples with overall very low cell-counts may be excluded from further analysis at this point already.

For the @Kang2018-demuxlet dataset, for example, one might consider removing the *Dendritic cells* and *Megakaryocytes* clusters, as these containg less than 50 cells across all samples.

```{r ncells, size = "small"}
# nb. of cells per cluster-sample
t(table(sce$cluster_id, sce$sample_id))
```

### Dimension reduction {.tabset}

The dimension reductions (DR) available within the SCE can be accessed via `reducedDims` from the `r Biocpkg("scater")` package. The data provided by @Kang2018-demuxlet already contains t-SNE coordinates; however, we can of course compute additional dimension reductions using one of `r Biocpkg("scater")`'s `runX` functions:

```{r umap}
# compute UMAP using 1st 20 PCs
sce <- runUMAP(sce, pca = 20)
```

Using `r Biocpkg("scater")`'s `plotReducedDim` function, we can plot t-SNE and UMAP representations colored by cluster and group IDs, respectively. We additionaly create a small wrapper function, `.plot_dr()`, to improve the readability of color legends and simplify the plotting theme:

```{r}
# wrapper to prettify reduced dimension plots
.plot_dr <- function(sce, dr, col)
  plotReducedDim(sce, dimred = dr, colour_by = col) +
    guides(fill = guide_legend(override.aes = list(alpha = 1, size = 3))) +
    theme_minimal() + theme(aspect.ratio = 1)
```

For our dataset, the t-SNE and UMAP colored by `cluster_id`s show that cell-populations are well-separated from one another. IFN-$\beta$ stimulation manifests as a severe shift in the low-dimensional projection of cells when coloring by `group_id`s, indicating widespread, genome-scale transcriptional changes.

```{r eval = FALSE}
# downsample to max. 100 cells per cluster
cs_by_k <- split(colnames(sce), sce$cluster_id)
cs100 <- unlist(sapply(cs_by_k, function(u) 
  sample(u, min(length(u), 100))))

# plot t-SNE & UMAP colored by cluster & group ID
for (dr in c("TSNE", "UMAP"))
  for (col in c("cluster_id", "group_id"))
    .plot_dr(sce[, cs100], dr, col)
```

```{r dr-ids, echo = FALSE, results = "asis", fig.height = 4, fig.width = 12, fig.cap = "Dimension reduction plots. Cells are colored by cluster ID (A) and group ID (B), respectively. For each cluster, at most 100 cells were sampled for plotting."}
cs_by_k <- split(colnames(sce), sce$cluster_id)
cs100 <- unlist(sapply(cs_by_k, function(u) 
  sample(u, min(length(u), 100))))

for (dr in c("TSNE", "UMAP")) {
  cat("#### ", dr, "{-}\n")
  ps <- lapply(c("cluster_id", "group_id"), 
    function(col) .plot_dr(sce[, cs100], dr, col = col))
  assign(paste0("ps_", tolower(dr)), ps)
  print(plot_grid(plotlist = ps, align = "vh", labels = c("A", "B")))
  cat("\n\n")
}
```

# Differential State (DS) analysis

To test for state changes across conditions, we will consider two types of approaches: i) mixed models that act directly on cell-level measurements; and ii) aggregation-based methods that act on *pseudobulk* data. For both approaches, each gene is tested for state changes in each cluster. Thus, a total of $\#(genes) \times \#(clusters)$ tests will be performed per comparison of interest. The following schematic summarizes the data representation considered by cell- and sample-level approaches, respectively:

```{r echo = FALSE, out.height = 4, fig.cap = "Schematic overview of cell- and sample-level approaches for DS analysis. Top panels show a schematic of the data distributions or aggregates across samples (each violin is a group or sample; each dot is a sample) and conditions (blue or orange). The bottom panels highlight the data organization in sub-matrix slices of the original count table."}
knitr::include_graphics(system.file('extdata', '1d.png', package = 'muscat'))
```

## Aggregation of single-cell to pseudobulk data {#sec-agg}

In order to leverage existing robust bulk RNA-seq DE frameworks, such as `r Biocpkg("edgeR")` [@Robinson2010-edgeR], `r Biocpkg("DESeq2")` [@Love2014-DESeq2], and `r Biocpkg("limma")` [@Ritchie2015-limma], we first aggregate measurements for each sample (in each cluster) to obtain pseudobulk data.

In general, `aggregateData()` will aggregate the data by the `colData` variables specified with argument `by`, and return a `SingleCellExperiment` containing pseudobulk data.  
For DS analysis, measurements must be aggregated at the cluster-sample level (default `by = c("cluster_id", "sample_id"`). In this case, the returned `SingleCellExperiment` will contain one assay per cluster, where rows = genes and columns = samples. Arguments `assay` and `fun` specify the input data and summary statistic, respectively, to use for aggregation.  
While, in principle, various combinations of input data (raw/(log-)normalized counts, CPM ect.) and summary statistics (sum, mean, median) could be applied, we here default to the sum of raw counts:

```{r agg}
pb <- aggregateData(sce,
    assay = "counts", fun = "sum",
    by = c("cluster_id", "sample_id"))
# one sheet per subpopulation
assayNames(pb)
# pseudobulks for 1st subpopulation
t(head(assay(pb)))
```

## Pseudobulk-level MDS plot

Prior to conducting any formal testing, we can compute a multi-dimensional scaling (MDS) plot of aggregated signal to explore overall sample similarities.

`pbMDS` takes as input any SCE containg PB data as returned by `aggregateData`, and computes MDS dimensions using `r Biocpkg("edgeR")`. Ideally, such a representation of the data should separate both clusters and groups from one another. Vice versa, samples from the same cluster or group should cluster together.

In our MDS plot on pseudo-bulk counts (Fig. \@ref(fig:pb-mds)), we can observe that the first dimension (MDS1) clearly separates cell populations (clusters), while the second (MDS2) separates control and stimulated samples (groups). Furthermore, the two T-cell clusters fall close to each other.

```{r pb-mds, fig.height = 4, fig.cap = "Pseudobulk-level multidimensional scaling (MDS) plot. Each point represents a cluster-sample instance; points are colored by cluster ID and shaped by group ID."}
(pb_mds <- pbMDS(pb))
```

If you're not satisfied with how the plot looks, here's an example of how to modify the `ggplot`-object from above in various ways:

```{r message = FALSE, fig.height = 4, fig.cap = "Pseudobulk-level MDS plot v2. Default plotting aesthetics were modified to change shaping of groups, coloring of clusters, as well as point size and transparency."}
# use very distinctive shaping of groups & change cluster colors
pb_mds <- pb_mds + 
  scale_shape_manual(values = c(17, 4)) +
  scale_color_manual(values = RColorBrewer::brewer.pal(8, "Set2"))
# change point size & alpha
pb_mds$layers[[1]]$aes_params$size <- 5
pb_mds$layers[[1]]$aes_params$alpha <- 0.6
pb_mds
```

## Sample-level analysis: Pseudobulk methods {#sec-pbDS}

Once we have assembled the pseudobulk data, we can test for DS using `pbDS`. By default, a $\sim group\_id$ model is fit, and the last coefficient of the linear model is tested to be equal to zero.

```{r}
# run DS analysis
res <- pbDS(pb, verbose = FALSE)
# access results table for 1st comparison
tbl <- res$table[[1]]
# one data.frame per cluster
names(tbl)
# view results for 1st cluster
k1 <- tbl[[1]]
head(format(k1[, -ncol(k1)], digits = 2))
```

Depening on the complexity of the experimental design (e.g., when there are more than two groups present), comparison(s) of interest may need to be specified explicitly. We can provide `pbDS` with a design matrix capturing the experimental design using `model.matrix` (package `r Rpackage("stats")`), and a contrast matrix that specifies our comparison of interesting using `makeContrasts` from the `r Biocpkg("limma")` package. Alternatively, the comparison(s) of interest (or a list thereof) can be specified with via `coefs` (see `?glmQLFTest` for details). For the @Kang2018-demuxlet dataset, we want to carry out a single comparison of stimulated against control samples, thus placing `"ctrl"` on the right-hand side as the reference condition:

```{r eval = FALSE}
# construct design & contrast matrix
ei <- metadata(sce)$experiment_info
mm <- model.matrix(~ 0 + ei$group_id)
dimnames(mm) <- list(ei$sample_id, levels(ei$group_id))
contrast <- makeContrasts("stim-ctrl", levels = mm)

# run DS analysis
pbDS(pb, design = mm, contrast = contrast)
```

## Cell-level analysis: Mixed models {#sec-mmDS}

Alternative to the above sample-level approach, we fit (for each gene) a mixed model (MM) to the cell-level measurement data. `muscat` provides implementations of MM that use 3 main approaches:

1. fitting linear mixed models (LMMs) on log-normalized data with observational weights, 
2. fitting LMMs on variance-stabilized data; and,
3. fitting generalized linear mixed models (GLMMs) directly on counts

In each case, a $\sim 1 + \text{group_id} + (1\,|\,\text{sample_id})$ model is fit for each gene, optimizing the log-likelihood (i.e., `REML = FALSE`). P-values are calculated using the estimates of degrees of freedom specifying by argument `df` (default `"Satterthwaite"`). Fitting, testing and moderation are applied subpopulation-wise. For differential testing, `mmDS` will only consider:

- subpopulations with at least `n_cells` cells (default 10) in at least `n_samples` samples (default 2)
- genes with a count >= `min_count` (default 1) in at least `min_cells` (default 20)

Mixed model based approaches can be run directly on cell-level measurements, and do not require prior aggregation:

```{r mm, eval = FALSE}
# 1st approach
mm <- mmDS(sce, method = "dream",
  n_cells = 10, n_samples = 2,
  min_counts = 1, min_cells = 20)

# 2nd & 3rd approach
mm <- mmDS(sce, method = "vst", vst = "sctransform")
mm <- mmDS(sce, method = "nbinom")
```

# Handling results

## Results filtering & overview

To get a general overview of the differential testing results, we first filter them to retain hits FDR < 5\% and abs(logFC) > 1, and count the number and frequency of differential findings by cluster. Finally, we can view the top hits (lowest adj. p-value) in each cluster.

```{r}
# filter FDR < 5%, abs(logFC) > 1 & sort by adj. p-value
tbl_fil <- lapply(tbl, function(u) {
  u <- dplyr::filter(u, p_adj.loc < 0.05, abs(logFC) > 1)
  dplyr::arrange(u, p_adj.loc)
})

# nb. of DS genes & % of total by cluster
n_de <- vapply(tbl_fil, nrow, numeric(1))
p_de <- format(n_de / nrow(sce) * 100, digits = 3)
data.frame("#DS" = n_de, "%DS" = p_de, check.names = FALSE)

# view top 2 hits in each cluster
top2 <- bind_rows(lapply(tbl_fil, top_n, 2, p_adj.loc))
format(top2[, -ncol(top2)], digits = 2)
```

## Calculating expression frequencies

Besides filter DS results based on magnitude (logFCs) and significance (FDR), it is often worthwhile to also consider the expression frequencies of each gene, i.e., the fraction of cells that express a given gene in each sample and/or group.  
`muscat` provides wrapper, `calcExprFreqs` to compute cluster-sample/-group wise expression frequencies. Here, a gene is considered to be expressed when the specified measurement value (argument `assay`) falls above a certain threshold (argument `th`). Note that, `assay = "counts"` and `th = 0` (default) amounts to the fraction of cells for which a respective gene has been detected.  
`calcExprFreqs` will return a `r Biocpkg("SingleCellExperiment")` object, where sheets (assays) = clusters, rows = genes, and columns = samples (and groups, if `group_id`s are present in the `colData` of the input SCE).

```{r frq}
frq <- calcExprFreqs(sce, assay = "counts", th = 0)
# one sheet per cluster
assayNames(frq)
# expression frequencies in each
# sample & group; 1st cluster
t(head(assay(frq), 5))
```

We can use the obtained frequencies to, for instance, only retain genes that are expressed in an average of 10\% of cells in at least 1 group:

```{r}
gids <- levels(sce$group_id)
frq10 <- vapply(as.list(assays(frq)), 
  function(u) apply(u[, gids] > 0.1, 1, any), 
  logical(nrow(sce)))
t(head(frq10))

tbl_fil2 <- lapply(kids, function(k)
  dplyr::filter(tbl_fil[[k]], 
    gene %in% names(which(frq10[, k]))))

# nb. of DS genes & % of total by cluster
n_de <- vapply(tbl_fil2, nrow, numeric(1))
p_de <- format(n_de / nrow(sce) * 100, digits = 3)
data.frame("#DS" = n_de, "%DS" = p_de, check.names = FALSE)
```

## Formatting results

Especially when testing multiple contrasts or coefficients, the results returned by `runDS` may become very complex and unhandy for exploration or exporting. Results can be formatted using `resDS`, which provides two alternative modes for formatting: `bind = "row"/"col"`.

When `bind = "row"`, results from all comparisons will be merged vertically (analogous to `do.call("rbind", ...)`) into a tidy format table, with column `contrast/coef` specifying the comparison.

Otherwise, `bind = "col"`, results will be merged horizontally into a single wide table where all results for a given gene and cluster are kept in one row. An identifier of the respective contrast of coefficient is then appended to the column names. This format is useful when wanting to view a specific gene's behavior across, for example, multiple treatments, but will become *messy* when many comparisons are included.

Expression frequencies computed with `calcExprFreqs`, as well as cluster-sample level avg. CPM, can be included in the results by setting `frq/cpm = TRUE`. Alternatively, if the former have been pre-computed, they can be supplied directly as an input to `resDS` (see example below).

```{r eval = FALSE}
# tidy format; attach pre-computed expression frequencies
resDS(sce, res, bind = "row", frq = frq)

# big-table (wide) format; attach CPMs
resDS(sce, res, bind = "col", cpm = TRUE)
```

Alternatively, if expression frequencies have not been pre-computed with `calcExprFreqs`, they may be added to the results table directly by specifying `frq = TRUE`:

```{r eval = FALSE}
# compute expression frequencies on the fly
resDS(sce, res, frq = TRUE)
```

# Visualizing results

## Between-cluster concordance

DS analysis aims at identifying population-specific changes in state (or expression) across conditions. In this setting, key questions of interest arise, e.g., which genes are DE in only a single (or very few) clusters? How many DE genes are shared between clusters? In summary, what is the general concordance in differential findings between clusters?

To gain an impression of the between-cluster (dis-)agreement on DE genes, we generate an UpSet-plot that visualizes the number of DE genes that are shared across or unique to certain clusters:

```{r upset, fig.width = 10, fig.cap = "Upset plot. Included are DS findings (FDR < 0.05, |logFC| > 1) across all clusters; shown are the 50 most frequent interactions."}
library(UpSetR)
de_gs_by_k <- map(tbl_fil, "gene")
upset(fromList(de_gs_by_k))
```

An UpSet plot as the one above tells us, for instance, that 185 genes are differential for all subpopulations; 387 across both *Monocytes* clusters; and 159 only in the *B cells* cluster.

## DR colored by expression

The code chunk generates a set of t-SNEs colored by gene expression for the top-8 DS genes. To match the affected cells to their cluster and experimental group, see the t-SNEs colored by cluster and group ID from above.

```{r fig.width = 14, fig.height = 8, fig.cap = "t-SNE colored by gene expression. Show are t-SNE projections with cells colored by the expression of the top-8 DS genes. For each cluster, at most 100 cells were sampled for plotting."}
# pull top-8 DS genes across all clusters
top8 <- bind_rows(tbl_fil) %>% 
  top_n(8, dplyr::desc(p_adj.loc)) %>% 
  pull("gene")

# for ea. gene in 'top8', plot t-SNE colored by its expression 
ps <- lapply(top8, function(g)
  .plot_dr(sce[, cs100], "TSNE", g) + 
    ggtitle(g) + theme(legend.position = "none"))

# arrange plots
plot_grid(plotlist = ps, ncol = 4, align = "vh")
```

## Cell-level viz.: Violin plots

For changes of high interest, we can view the cell-level expression profiles of a specific gene across samples or groups using `plotExpression` (`r Biocpkg("scater")` package). Here, we generate violin plots for the top-6 DS genes (lowest adj. p-value) in the *B cells* cluster[^2].

[^2]: Note that, as DS testing is done at the cluster-level, we need to subset the cells that have been assigned to the corresponding cluster for plotting.

```{r violins, fig.width = 10, fig.height = 5, fig.cap = "Violin plots. Show are the top 6 hits (lowest adj. p-value) for the B cells cluster. Each violin is a sample; points are colored by group ID."}
plotExpression(sce[, sce$cluster_id == "B cells"],
  features = tbl_fil$`B cells`$gene[seq_len(6)],
  x = "sample_id", colour_by = "group_id", ncol = 3) +
  guides(fill = guide_legend(override.aes = list(size = 5, alpha = 1))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Sample-level viz.: Pseudobulk heatmaps

Especially when wanting to gain an overview of numerous DE testing results for many clusters, both dimension reduction and cell-level visualisations require a lot of space can become cumbersome to interpret. In this setting, it is thus recommended to visualise aggregated measures, e.g., mean expressions by cluster sample.

```{r pb-hm-1}
# top-5 DS genes per cluster
pbHeatmap(sce, res, top_n = 5)
```

Alternatively, `pbHeatmap` provides a set of options regarding which cluster(s), gene(s), and comparison to include (arguments `k`, `g` and `c`, respectively). For example, the following options render a heatmap visualizing the top 20 DS genes for the *B cells* cluster:

```{r pb-hm-2}
# top-20 DS genes for single cluster
pbHeatmap(sce, res, k = "B cells")
```

Similarly, we can visualize the cluster-sample means of a single gene of interest across all clusters in order to identify cell-types that are affected similarly by different experimental conditions:

```{r pb-hm-3}
# single gene across all clusters
pbHeatmap(sce, res, g = "ISG20")
```

# Session info {- .smaller}

```{r session-info}
sessionInfo()
```

# Save image

```{r save}
save.image("muscat_vignette.RData")
```

# References

### 2024-05-31 old/manuscript/manuscript.Rmd
move

--- 2023-07-09 manuscript/manuscript.Rmd
@@
 
+## Methods
+
+### Demonstrating the impact of the problem
+
+To demonstrate the impact of this problem on real omics data, we conducted enrichment analysis of
+four different types of data.
+We obtained differential proteomics data from Stukalov et al (https://pubmed.ncbi.nlm.nih.gov/33845483/),
+who profiled the effect of SARS-COV2 infection on A549 cells.
+At 24 hrs infection with SARS-CoV2, Stukalov et al reported the detection of 5825 proteins, including
+165 downregulated and 44 upregulated (FDR<0.05).
+For bulk RNA-seq, we obtained count data from DEE2.io (Ziemann et al, 2019) for control and azacitidine
+exposed AML3 cells (SRA accession SRP038101) (Lund et al, 2014, PMID:25315154).
+Genes with fewer than 10 reads per sample on average were considered not detected and were omitted from
+downstream analysis.
+DESeq2 was used to compare treatment groups and genes with FDR<0.05 were selected (Love 2014).
+For single cell RNA-seq 10X Genomics droplet-based, scRNA-seq PBMC data from 8 Lupus patients obtained
+before and after 6h-treatment with IFN-β (Kang et al. 2018).
+The complete raw data, as well as gene and cell metadata is available through the NCBI GEO,
+accession number GSE96583.
+Analysis of this data set  was done using a pseudobulk approach according to the Muscat package vignette
+(Crowell et al, 2020), yielding lists of detected genes for each cell state as well as up- and
+down-regulated gene lists (FDR<0.05).
+Infinium 450K BeadChip based Epigenenome Wise Association Study (EWAS) data was obtained from
+Joehanes et al (PMID:27651444), a meta-analysis investigating the DNA methylation differences between
+smokers and non-smokers.
+Data was downloaded from EWAS Catalog at the link (http://ewascatalog.org/?study=27651444_smoking_current_vs_never_smoking).
+As a background gene list, all genes represented by probes on the 450K array were listed.
+For all of the above analyses, up- and down-regulated genes were partitioned into separate lists,
+for separate enrichment analysis.
+These gene lists were loaded into R and were subject to ClusterProfiler analysis along with the
+respective background gene set.
+KEGG (REF) or Reactome (REF) pathway annotations were used (GMT format).
+KEGG pathway annotations v2023.1 were obtained from MSigDB (REF), while Reactome pathways were downloaded
+from the Reactome website (2023-03-06).
+Pathways (gene sets) with fewer than 10 detected genes were not included in the analysis.
+In addition to the default ClusterProfiler analysis, a modification was made to circumvent the
+background filtering problem.
+To do this, the list of detected genes was appended as a new "pathway" to the KEGG or Reactome annotation
+set.
+The results of this "fixed" analysis was compared those using the default algorithm using an Euler
+diagram.
+
+### Examining the breadth of the problem
+
+Next, we investigated the behaviour of many different ORA tools to find out whether they suffered from
+the same problem or not.
+We analysed the 20 most popular ORA tools as described by Xie et al, 2021 (PMID:33858350).
+To understand whether the background list filtering problem was apparent, we used the gene lists from
+the bulk RNA-seq dataset, with various pathway /gene set databases.
+If the tool reported the number of genes in the background list, we used this as evidence to classify
+the tool's behaviour.
+If not, then we undertook forensic investigation of the enrichment results to infer the program's
+behaviour.
+
+## Results
+
+### Background list filtering biases ORA results
+
+### Inappropriate background list filtering is widespread
+
 ## Acknowledgements

### 2024-06-01 analysis/sim/run_simpw.Rmd
try dose
sims

@@
 
-library("tidyverse")
+#library("tidyverse")
 library("parallel")
@@
 library("kableExtra")
+library("fgsea")
+library("clusterProfiler")
 source("simpw_func.R")
@@
 
+xxx object slots
+
+1. expression counts
+
+2. ground truth up genes
+
+3. ground truth down genes
+
+4. ground truth up gene sets
+
+5. ground truth down gene sets
+
+6. DE result (DESeq2)
+
+7. DE genes up observed
+
+8. DE genes down observed
+
+9. fgsea up gene sets
+
+10. fgsea down gene sets
+
+11. phyper up gene sets
+
+12. phyper down gene sets
+
+13. clusterprofiler_default up gene sets
+
+14. clusterprofiler_default down gene sets
+
+15. clusterprofiler_fixed up gene sets
+
+16. clusterprofiler_fixed down gene sets
+
+17. clusterprofiler_nobg up gene sets
+
+18. clusterprofiler_nobg down gene sets
+
+19. fisher up gene sets
+
+20. fisher down gene sets
+
+21. DOSE up gene sets
+
+22. DOSE down gene sets
+
 # Get count data
 
-```{r,getcounts}
+```{r,getcounts,warning=FALSE}
 
-a<-countData()
+a <- countData()
 
@@
 
-gsets<-randomGeneSets(a,setsize=50,nsets=500)
+gsets <- randomGeneSets(a,setsize=20,nsets=200)
 
@@
 
-SIMS=16
+SIMS=8
 FRAC_DE=0.05
 FC=1
-N_REPS=4
+N_REPS=5
 DGE_FUNC="deseq2"
 SUM_COUNT=3e7
-VARIANCE=c(0.25,0.3,0.35,0.4,0.45)
+VARIANCE=c(0.3,0.35,0.4)
 
@@
 
+res <- do.call(rbind,res)
+
 ```

### 2024-06-03 analysis/sim/run_simpw.Rmd
updating simulation analysis

@@
----
+652 4---
 title: "Examining background gene lists using simulation - comparison of different enrichment techniques"
@@
 
-SIMS=8
+SIMS=16
 FRAC_DE=0.05
@@
 DGE_FUNC="deseq2"
-SUM_COUNT=3e7
-VARIANCE=c(0.3,0.35,0.4)
+SUM_COUNT=2e7
+VARIANCE=c(0.3,0.35,0.4,0.45)
 

### 2024-06-03 analysis/sim/simpw_func.R

@@ -1,4 +1,4 @@
-library("tidyverse")
+#library("tidyverse")
 library("parallel")
 library("edgeR")
 library("DESeq2")
@@ -374,7 +374,10 @@ x
 # clusterprofiler default function
 ##################################
 # Note that clusterprofiler requires different gene set format
-run_clusterprofiler_default <-function(x,DGE_FUNC,gsets2,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS){
+run_clusterprofiler_default <- function(x,DGE_FUNC,gsets,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS){
+
+gsets2 <- stack(gsets)[,c(2,1)]
+colnames(gsets2) <- c("term","gene")
 
 dge <- sapply(x,"[",6)
 
@@ -393,27 +396,35 @@ n_dns=n_ups=p_ups=p_dns=obs_up=obs_dn=list()
 for (d in 1:length(dge)) {
 
 # clusterprofiler UP
-ora_up <- as.data.frame(enricher(gene = ups[[d]] ,
-  universe = bgs[[d]],  minGSSize=2, maxGSSize = 500000, TERM2GENE = gsets2,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+if ( length(ups[[d]]) > 2 ) {
+  ora_up <- as.data.frame(enricher(gene = ups[[d]] ,
+    universe = bgs[[d]],  minGSSize=2, maxGSSize = 500000, TERM2GENE = gsets2,
+    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
 
-ora_up$geneID <- NULL
-ora_up <- subset(ora_up,p.adjust<0.05 )
-ora_ups <- rownames(ora_up)
-obs_up[[d]] <- ora_ups
+  ora_up$geneID <- NULL
+  ora_up <- subset(ora_up,p.adjust<0.05 )
+  ora_ups <- rownames(ora_up)
+  obs_up[[d]] <- ora_ups
+} else {
+  ora_ups = NULL
+}
 
 # clusterprofiler DOWN
-ora_dn <- as.data.frame(enricher(gene = dns[[d]] ,
-  universe = bgs[[d]],  minGSSize=2, maxGSSize = 500000, TERM2GENE = gsets2,
[…]

### 2024-06-03 analysis/sim/run_simpw.Rmd
update sims

@@
 
-SIMS=16
+SIMS=100
 FRAC_DE=0.05
 FC=1
-N_REPS=5
+N_REPS=6
 DGE_FUNC="deseq2"
-SUM_COUNT=2e7
+SUM_COUNT=3e7
 VARIANCE=c(0.3,0.35,0.4,0.45)
@@
   VARIANCE=mygrid[i,"VARIANCE"]
-  x <- agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets)
+  x <- agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets,cores=32)
   as.data.frame(do.call(rbind, x))

### 2024-06-03 analysis/sim/simpw_func.R

@@ -451,7 +451,7 @@ p<-true_pos/(true_pos+false_pos)
 r<-true_pos/(true_pos+false_neg)
 f<-2*p*r/(p+r)
 
-attr(x,'clusterprofiler_default') <-data.frame(N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,true_pos,false_pos,true_neg,false_neg,p,r,f)
+attr(x,'clusterprofiler') <-data.frame(N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,true_pos,false_pos,true_neg,false_neg,p,r,f)
 x
 
 }
@@ -808,10 +808,10 @@ xxx <- mclapply(xxx , DGE_FUNC , mc.cores = cores )
 xxx <- run_clusterprofiler_default(x=xxx,DGE_FUNC,gsets,N_REPS=N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS)
 
 # run clusterprofiler default pos 11,12
-xxx <- run_clusterprofiler_fixed(x=xxx,DGE_FUNC,gsets,N_REPS=N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS)
+#xxx <- run_clusterprofiler_fixed(x=xxx,DGE_FUNC,gsets,N_REPS=N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS)
 
 # run clusterprofiler nobg pos 13,14
-xxx <- run_clusterprofiler_nobg(x=xxx,DGE_FUNC,gsets,N_REPS=N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS)
+#xxx <- run_clusterprofiler_nobg(x=xxx,DGE_FUNC,gsets,N_REPS=N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS)
 
 # run fora pos 15,16
 xxx <- run_fora(xxx,DGE_FUNC,gsets,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS)

### 2024-06-03 analysis/sim/run_simpw.Rmd
update sims

@@
 
-# Intro
+## Intro
 
@@
 
-9. fgsea up gene sets
+9. clusterprofiler_default up gene sets
 
-10. fgsea down gene sets
+10. clusterprofiler_default down gene sets
 
-11. phyper up gene sets
+11. clusterprofiler_fixed up gene sets
 
-12. phyper down gene sets
+12. clusterprofiler_fixed down gene sets
 
-13. clusterprofiler_default up gene sets
+13. clusterprofiler_nobg up gene sets
 
-14. clusterprofiler_default down gene sets
+14. clusterprofiler_nobg down gene sets
 
-15. clusterprofiler_fixed up gene sets
+15. fora up gene sets
 
-16. clusterprofiler_fixed down gene sets
+16. fora down gene sets
 
-17. clusterprofiler_nobg up gene sets
+17. fgsea up gene sets
 
-18. clusterprofiler_nobg down gene sets
-
-19. fisher up gene sets
-
-20. fisher down gene sets
-
-21. DOSE up gene sets
-
-22. DOSE down gene sets
+18. fgsea down gene sets
 
@@
 
-# Session information
+## Plot
+
+```{r,plot1}
+
+cp <- subset(res,PWAY_FUNC == "clusterprofiler")
+fo <- subset(res,PWAY_FUNC == "fora")
+fg <- subset(res,PWAY_FUNC == "fgsea")
+
+par(mfrow=c(1,3))
+
+plot(cp$VARIANCE,cp$p,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="precision")
+points(fo$VARIANCE,fo$p,type="b",pch=19,col="red")
+points(fg$VARIANCE,fg$p,type="b",pch=19,col="blue")
+
+legend("bottomleft", inset=.02, title="tool",
+   c("clusterprofiler","fora","fgsea"), col=c("black","red","blue"), horiz=FALSE, cex=1.2, pch=19,lwd=2)
+
+plot(cp$VARIANCE,cp$r,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="recall")
+points(fo$VARIANCE,fo$r,type="b",pch=19,col="red")
+points(fg$VARIANCE,fg$r,type="b",pch=19,col="blue")
+
+plot(cp$VARIANCE,cp$f,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="f1")
+points(fo$VARIANCE,fo$f,type="b",pch=19,col="red")
+points(fg$VARIANCE,fg$f,type="b",pch=19,col="blue")
+
+par(mfrow=c(1,1))
+
+```
+
+## Session information
 

### 2024-06-03 analysis/sim/simpw_func.R

@@ -221,7 +221,7 @@ p<-true_pos/(true_pos+false_pos)
 r<-true_pos/(true_pos+false_neg)
 f<-2*p*r/(p+r)
 
-attr(y,'mitch_res') <-data.frame(N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,true_pos,false_pos,true_neg,false_neg,p,r,f)
+attr(y,'mitch') <-data.frame(N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,true_pos,false_pos,true_neg,false_neg,p,r,f)
 y
 }
 
@@ -280,7 +280,7 @@ p<-true_pos/(true_pos+false_pos)
 r<-true_pos/(true_pos+false_neg)
 f<-2*p*r/(p+r)
 
-attr(x,'phyper_res') <-data.frame(N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,true_pos,false_pos,true_neg,false_neg,p,r,f)
+attr(x,'phyper') <-data.frame(N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,true_pos,false_pos,true_neg,false_neg,p,r,f)
 x
 
 }
@@ -365,7 +365,7 @@ p<-true_pos/(true_pos+false_pos)
 r<-true_pos/(true_pos+false_neg)
 f<-2*p*r/(p+r)
 
-attr(x,'fisher_res') <-data.frame(N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,true_pos,false_pos,true_neg,false_neg,p,r,f)
+attr(x,'fisher') <-data.frame(N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,true_pos,false_pos,true_neg,false_neg,p,r,f)
 x
 
 }
@@ -682,7 +682,7 @@ p<-true_pos/(true_pos+false_pos)
 r<-true_pos/(true_pos+false_neg)
 f<-2*p*r/(p+r)
 
[…]

### 2024-06-03 analysis/sim/run_simpw.Rmd
sims dome

@@
-652 4---
+---
 title: "Examining background gene lists using simulation - comparison of different enrichment techniques"
@@
 
-#library("tidyverse")
 library("parallel")
@@
 
+19. mitch up
+
+20. mitch dn
+
 # Get count data
@@
 
-gsets <- randomGeneSets(a,setsize=20,nsets=200)
+gsets <- randomGeneSets(a,setsize=30,nsets=200)
 
@@
 
-SIMS=100
+SIMS=500
 FRAC_DE=0.05
 FC=1
-N_REPS=6
+N_REPS=5
 DGE_FUNC="deseq2"
-SUM_COUNT=3e7
-VARIANCE=c(0.3,0.35,0.4,0.45)
+SUM_COUNT=2e7
+VARIANCE=c(0.3,0.325,0.35,0.375,0.4,0.425,0.45)
 
@@
 
-```{r,results}
+```{r,rqesults}
 
@@
 fg <- subset(res,PWAY_FUNC == "fgsea")
+mi <- subset(res,PWAY_FUNC == "mitch")
 
@@
 legend("bottomleft", inset=.02, title="tool",
-   c("clusterprofiler","fora","fgsea"), col=c("black","red","blue"), horiz=FALSE, cex=1.2, pch=19,lwd=2)
+   c("clusterprofiler","fora","fgsea"), col=c("black","red","blue"),
+   horiz=FALSE, cex=1.2, pch=19,lwd=2)
 

### 2024-06-03 analysis/sim/simpw_func.R

@@ -161,68 +161,74 @@ RepParallel <- function(n, expr, simplify = "array",...) {
 # define DESeq2 function
 ##################################################
 deseq2<-function(x) {
-library("DESeq2")
-label="simulate"
-y<-x[[1]]
-samplesheet<-as.data.frame(colnames(y))
-colnames(samplesheet)="sample"
-samplesheet$trt<-factor(as.numeric(grepl("trt",colnames(y))))
-dds <- DESeqDataSetFromMatrix(countData = y, colData = samplesheet, design = ~ trt )
-res <- DESeq(dds)
-z<- DESeq2::results(res)
-vsd <- vst(dds, blind=FALSE)
-zz<-cbind(z,assay(vsd))
-x[[6]]<-as.data.frame(zz[order(zz$padj),])
-sig<-subset(zz,padj<0.05)
-x[[7]]<-rownames(sig[which(sig$log2FoldChange>0),])
-x[[8]]<-rownames(sig[which(sig$log2FoldChange<0),])
-x
+  library("DESeq2")
+  label = "simulate"
+  y <- x[[1]]
+  samplesheet <- as.data.frame(colnames(y))
+  colnames(samplesheet) = "sample"
+  samplesheet$trt <- factor(as.numeric(grepl("trt",colnames(y))))
+  dds <- DESeqDataSetFromMatrix(countData = y, colData = samplesheet, design = ~ trt )
+  res <- DESeq(dds)
+  z <- DESeq2::results(res)
+  vsd <- vst(dds, blind=FALSE)
+  zz <- cbind(z,assay(vsd))
+  x[[6]] <- as.data.frame(zz[order(zz$padj),])
+  sig <- subset(zz,padj<0.05)
+  x[[7]] <- rownames(sig[which(sig$log2FoldChange>0),])
+  x[[8]] <- rownames(sig[which(sig$log2FoldChange<0),])
+  x
 }
 
 
 #################################################
 # define mitch function
 ##################################################
-run_mitch<-function(y,DGE_FUNC,gsets, N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS) {
-library("mitch")
-dge<-sapply(y,"[",6)
[…]

### 2024-06-03 analysis/sim/run_simpw.Rmd
fig

@@
     code_folding: hide
-    fig_width: 5
+    fig_width: 7
     fig_height: 5
@@
 
+saveRDS(res,file="res.Rds")
+
 ```
@@
 
-```{r,rqesults}
+```{r,results}
 
@@
 fg <- subset(res,PWAY_FUNC == "fgsea")
-mi <- subset(res,PWAY_FUNC == "mitch")
+
+cp %>% kbl(caption="clusterprofiler") %>% kable_paper("hover", full_width = F)
+fo %>% kbl(caption="fora") %>% kable_paper("hover", full_width = F)
+fg %>% kbl(caption="fg") %>% kable_paper("hover", full_width = F)
 

### 2024-06-05 analysis/analyse_rnaseq1.Rmd
improve charts

@@
 
-names(gs) <- gsub(".v2023.2.Hs.symbols.gmt","",names(gs))
+names(gs)
+names(gs) <- c("KEGG", "Reactome", "Wikipathways", "miR targets", "TFT GTRD",
+  "GO", "HPO", "Cellmarkers", "Hallmark" )
 
@@
 # good=same as bad, except fixed the background problem
-compare_ora_plots <- function(bad,good){
+compare_ora_plots <- function(bad,good,libname=NULL){
   badup <- bad$orabad_up
@@
   MAX=max(c(mup$es.x,mup$es.y))
+  UPHEADER=paste(libname,"up")
+  DNHEADER=paste(libname,"dn")
+
+  par(mfrow=c(2,2))
 
@@
     xlim=c(0,MAX), ylim=c(0,MAX),
-    xlab="original", ylab="corrected", main="Enrichment score")
+    xlab="original ES", ylab="corrected ES", main=UPHEADER)
   grid()
@@
     xlim=c(0,MAX), ylim=c(0,MAX),
-    xlab="original", ylab="corrected", main="-log10 p-value")
+    xlab="original -log10 p-value", ylab="corrected -log10 p-value", main=UPHEADER)
   grid()
@@
     xlim=c(0,MAX), ylim=c(0,MAX),
-    xlab="original", ylab="corrected", main="Enrichment score")
+    xlab="original ES", ylab="corrected ES", main=DNHEADER)
   grid()
@@
     xlim=c(0,MAX), ylim=c(0,MAX),
-    xlab="original", ylab="corrected", main="-log10 p-value")
+    xlab="original -log10 p-value", ylab="corrected -log10 p-value", main=DNHEADER)
   grid()
@@
 par(mar=c(c(5.1, 7.1, 4.1, 2.1) ))
-barplot(t(df),beside=TRUE,horiz=TRUE,las=1,
-  xlab="no. significant sets", main="Comparison of original and corrected data",
+barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,1250),
+  xlab="no. significant sets", main="Effect of correcting the background bug",
   col=c("gray30","gray80"))
+text(df$ns1+50,((1:nrow(df))*3)-1.5,labels=signif(df[,1],3))
+text(df$ns2+50,((1:nrow(df))*3)-0.5,labels=signif(df[,2],3))
+#abline(v=seq(100,1000,100),lty=2,lwd=0.5,col="gray")
+legend("bottomright", inset=.02, legend=c("corrected","original"),
+  fill=c("gray80","gray30"), horiz=FALSE, cex=1.1)
 
-legend("bottomright", inset=.02, legend=c("original","corrected"),
-  fill=c("gray30","gray80"), horiz=TRUE, cex=1)
-
-# RATIO
-rat <- matrix( (res1$nsig_good+0) / (res1$nsig_bad+0) , ncol=9 )
-rownames(rat) <- names(d)
-colnames(rat) <- names(gs)
-
-signif(rat,3) %>%
-  kbl(row.names = TRUE, caption="Ratio of significant genesets Corrected:Original") %>%
-  kable_paper("hover", full_width = F)
-
-rat2 <- apply(rat,2,function(x) { x <- x[!is.na(x)] ; x <- x[is.finite(x)] ;  mean(x) } )
-rat2
-
-barplot(sort(rat2),horiz=TRUE,las=1,main="increase in significant sets after correction",xlab="fold increase")
+mylabels <- gsub("$","%",gsub("^","+",signif(((df$ns2/df$ns1)-1)*100,3)))
+text(df$ns2+160,((1:nrow(df))*3)-1,labels=mylabels,cex=1.1)
 
@@
   good <- oragood(d[[1]],gs[[i]])
-  compare_ora_plots(bad,good)
+  compare_ora_plots(bad,good,libname=names(gs)[i])
 } )

### 2024-06-05 analysis/analyse_rnaseq1.Rmd
fig size

@@
     code_folding: hide
-    fig_width: 5
-    fig_height: 5
+    fig_width: 7
+    fig_height: 7
 theme: cosmo

### 2024-06-05 analysis/analyse_rnaseq1.Rmd
update figs

@@
   jac <- length(intersect(sig_bad,sig_good)) / length(union(sig_bad,sig_good))
-  result=c("nsig_bad"=nsig_bad,"nsig_good"=nsig_good,"jaccard"=jac)
+  result=c("nsig_bad"=nsig_bad,"nsig_good"=nsig_good,"Jaccard"=jac)
   return(result)

### 2024-06-05 analysis/analyse_rnaseq2.Rmd
update figs

@@
 
-The goal of this work is to understand how ClusterProfiler manages the background list,
+The goal of this work is to understand the behaviour of ClusterProfiler's p-value adjustment method
 as we have observed some weird behaviour.
-In particular we see that when we provide a custom background, those genes do not appear in the
-final analysis unless they are also included in the gene list.
+In particular we see that clusterprofiler's criteria for inclusion/exclusion of gene sets is irregular.
+For example clusterprofiler seems to include gene sets as detected if there is just 1 member detected
+in the foreground list.
+If a gene set has 10 in the background and  0 in the foreground it looks like it gets excluded.
+This is a bit strange and contrary to best practice, where the detection threshold should be set by
+the overlap size in the background list, rather than the foreground list.
 
-In the code chunk below called `libs`, you can add and remove required R library dependancies.
-Check that the libraries listed here match the Dockerfile, otherwise you might get errors.
+Theoretically, this behaviour could have a negative impact on performance, as FDR is applied to the
+wrong pathways.
 
@@
 # good=same as bad, except fixed the background problem
-compare_ora_plots <- function(bad,good){
+compare_ora_plots <- function(bad,good,libname=NULL,setsize=NULL){
   badup <- bad$orabad_up
@@
   MAX=max(-log10(c(mup$p.adjust.x,mup$p.adjust.y)))
+  UPHEADER=paste(libname,"up",setsize,"genes")
+  DNHEADER=paste(libname,"dn",setsize,"genes")
+
+  par(mfrow=c(1,2))
 
@@
     xlim=c(0,MAX), ylim=c(0,MAX),
-    xlab="original", ylab="corrected", main="-log10 FDR")
+    xlab="original -log10 FDR", ylab="corrected -log10 FDR", main=UPHEADER)
   grid()
@@
     xlim=c(0,MAX), ylim=c(0,MAX),
-    xlab="original", ylab="corrected", main="-log10 FDR")
+    xlab="original -log10 FDR", ylab="corrected -log10 FDR", main=DNHEADER)
   grid()
@@
 summary(res1$nsig_good)
-summary(res1$jaccard)
+summary(res1$Jaccard)
 
@@
 par(mar=c(c(5.1, 7.1, 4.1, 2.1) ))
-barplot(t(df),beside=TRUE,horiz=TRUE,las=1,
-  xlab="no. significant sets", main="Comparison of original and corrected data",
+barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,330),
+  xlab="no. significant sets", main="Effect of correcting the FDR bug",
   col=c("gray30","gray80"))
-
-legend("bottomright", inset=.02, legend=c("original","corrected"),
-  fill=c("gray30","gray80"), horiz=TRUE, cex=1)
+text(df$ns1+20,((1:nrow(df))*3)-1.5,labels=signif(df[,1],3))
+text(df$ns2+20,((1:nrow(df))*3)-0.5,labels=signif(df[,2],3))
+legend("bottomright", inset=.02, legend=c("corrected","original"),
+  fill=c("gray80","gray30"), horiz=FALSE, cex=1)
 
@@
 
-barplot(sort(rat2),horiz=TRUE,las=1,main="increase in significant sets after correction",xlab="fold increase")
-
 # JACCARD
-jac <- matrix( res1$jaccard , ncol=9 )
+jac <- matrix( res1$Jaccard , ncol=9 )
 rownames(jac) <- names(d)
@@
 lapply(ngenes, function(n) {
-  res2 <- subset(res1,Var3==n) ; summary(res2$jaccard)
+  res2 <- subset(res1,Var3==n) ; summary(res2$Jaccard)
 } )
@@
   res2 <- subset(res1,Var3==n)
-  ns <- matrix( res2$jaccard , ncol=9 )
+  ns <- matrix( res2$Jaccard , ncol=9 )
   rownames(ns) <- names(d)
@@
 
-plot(jac[,"c5.go"],type="b", pch=19,cex=1.5,lwd=2,
-  xaxt = "n",ylim=c(0,1),ylab="jaccard",xlab="no. genes selected")
+plot(jac[,"GO"],type="b", pch=19,cex=1.5,lwd=2,
+  xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected")
 axis(1, at = seq(1, 5, by = 1),
      labels = ngenes)
-points(jac[,"c2.cp.reactome"],type="b",col="red",pch=19,cex=1.5,lwd=2)
-points(jac[,"c8.all"],type="b",col="blue",pch=19,cex=1.5,lwd=2)
-points(jac[,"c3.tft.gtrd"],type="b",col="gray50",pch=19,cex=1.5,lwd=2)
-mtext("impact of gene list length on severity of FDR problem")
+points(jac[,"Reactome"],type="b",col="red",pch=19,cex=1.5,lwd=2)
+points(jac[,"Cellmarkers"],type="b",col="blue",pch=19,cex=1.5,lwd=2)
+points(jac[,"TFT GTRD"],type="b",col="gray50",pch=19,cex=1.5,lwd=2)
+mtext("impact of gene list length on FDR problem severity")
 
-legend("bottomright", legend=c("GO","Reactome", "Markers", "TFT"),
+legend("bottomright", legend=c("GO","Reactome", "Cellmarkers", "TFT"),
        col=c("black", "red", "blue", "gray50"), lty=1, cex=1.2,lwd=3,pch=19)
@@
 
-```{r,charts1}
+```{r,charts1,fig.height=4}
 
@@
   good <- oragood(d[[1]],gs[[i]],ngenes=125)
-  compare_ora_plots(bad,good)
+  compare_ora_plots(bad,good,libname=names(gs)[i],setsize=125)
 } )
@@
   good <- oragood(d[[1]],gs[[i]],ngenes=250)
-  compare_ora_plots(bad,good)
+  compare_ora_plots(bad,good,libname=names(gs)[i],setsize=250)
 } )
@@
   good <- oragood(d[[1]],gs[[i]],ngenes=500)
-  compare_ora_plots(bad,good)
+  compare_ora_plots(bad,good,libname=names(gs)[i],setsize=500)
 } )
@@
   good <- oragood(d[[1]],gs[[i]],ngenes=1000)
-  compare_ora_plots(bad,good)
+  compare_ora_plots(bad,good,libname=names(gs)[i],setsize=1000)
 } )
@@
   good <- oragood(d[[1]],gs[[i]],ngenes=2000)
-  compare_ora_plots(bad,good)
+  compare_ora_plots(bad,good,libname=names(gs)[i],setsize=2000)
 } )

### 2024-06-05 analysis/analyse_rnaseq1.Rmd
titles

@@
 ---
-title: "Examining background gene lists in clusterProfiler - RNA-seq example"
+title: "Examining background gene list issue in clusterProfiler - RNA-seq example"
 author: "Mark Ziemann & Anusuiya Bora"

### 2024-06-05 analysis/analyse_rnaseq2.Rmd
update figs
titles

@@
 ---
-title: "Examining background gene lists in clusterProfiler - RNA-seq example"
+title: "Examining FDR correction issue in clusterProfiler - RNA-seq example"
 author: "Mark Ziemann & Anusuiya Bora"
@@
 # overall
-oa <- matrix(paste(res1$nsig_bad,res1$nsig_good,signif(res1$jaccard,2)),ncol=9)
+oa <- matrix(paste(res1$nsig_bad,res1$nsig_good,signif(res1$Jaccard,2)),ncol=9)
 rownames(oa) <- names(d)

### 2024-06-06 manuscript/manuscript.Rmd
ms

--- 2023-07-08 manuscript/manuscript.Rmd
@@
 ---
-title: "Importance of using the correct background list for over-representation analysis"
-author: "Mark Ziemann and Anusuiya Bora"
+title: "Two subtle problems with over-representation analysis"
 date: "`r Sys.Date()`"
@@
 
-Alternate title: *Problems in over-representation analysis software*
-
-[Mark Ziemann<sup>1</sup>*](https://orcid.org/0000-0002-7688-6974),
-[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352)
+[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352),
+[Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974)
 
@@
 
-1. Deakin University, Geelong, Australia, School of Life and Environmental Sciences.
+1. Burnet Institute, Melbourne, Australia.
 
-2. Vellore Institute of Technology, Vellore, India.
+2. Deakin University, School of Life and Environmental Sciences, Geelong, Australia.
 
-(*) Corresponding author: m.ziemann@deakin.edu.au
+(*) Corresponding author: mark.ziemann@burnet.edu.au
 
-## Abstract
+# Abstract
 
-* Over-representation analysis (ORA) is the most common form of functional enrichment and one of the
-most used techniques in computational biology, yet there are common methodological issues that
-undermine its validity.
+TODO.
 
-* These problems are typically ascribed to errors in analysis and reporting by the end-users,
-however software design is another contributor.
+# Keywords
 
-* Here, we identify and characterise a potential bug popular ORA software
-package and web tool.
+* Bioinformatics
 
-* We show that this bug deleteriously affects the results of ORA tests, and we present a simple
-work-around to circumvent it.
+* Reproducibility
+
+* Enrichment analysis
+
+* Pathway analysis
+
+* Gene set enrichment analysis
+
+* Docker
+
+# Main Body
 
@@
 library("DiagrammeR")
+library("RColorBrewer")
+library("kableExtra")
+#library("gt")
 
@@
 
-Functional enrichment analysis is a family of computational approaches used to interpret large omics
-data.
-The most commonly used method is over-representation analysis (ORA) which involves the selection of
-genes of interest, followed by a statistical test to evaluate whether genes of certain annotated
-categories are over-represented in the set of interest.
-The set of interest may be defined in different ways, such as genes that meet a differential expression
-significance threshold, or those that have been clustered into a group based on co-expression patterns.
-There are various sources of annotated categories, including gene ontologies, protein-coding features,
-gene co-expression and signalling or metabolic pathways.
-Providers of such annotations include but are not limited to the gene ontology consortium, Reactome,
-KEGG, WikiPathways and MSigDB.
+Functional enrichment analysis involves the summarisation of omics data to reflect
+biological changes, and has become one of the most popular methods in bioinformatics.
+Enrichment analysis software packages have been collectively cited 131,332 times as of
+late 2019 [@Xie2021-ch].
+The algorithms used by these tools can largely be divided into two groups, functional
+class scoring (FCS) and over-representation analysis (ORA).
+In FCS, all detected genes are ranked by their degree of differential expression followed
+by some statistical test for enrichment of gene categories at either extreme of the
+ranked list.
+ORA involdes the selection of genes of interest, followed by a test to ascertain whether
+certain functional categories are over-represented in the selected genes.
 
-Central to ORA tests is the selection of a list of background genes.
-This is also sometimes called the "universe" or the "reference" list.
-These are defined as any gene that is detected robustly, and has a chance at making the set of interest.
-For RNA-seq studies, this could be a threshold detection limit based either on raw counts or counts per
-million.
+Of these two methods, FCS is reported to be the most accurate, while ORA seems to be the
+most popular.
 
-The first use of ORA-based functional enrichment that we are aware of is a 1999 report that used the
-hypergeometric test to determine the chance probability of observing over-representation of certain
-annotated functions in co-regulated genes in yeast [@Tavazoie1999-kf].
-This general approach is the basis of most ORA-based significance testing and can be represented by the
-following equation:
+According to a study that used simulated differential expression data, FCS accuracy was
+superior to ORA over a range of experimental designs [@Kaspi2020-rn].
 
-$$
-P=1-\sum_{i=0}^{k-1}\frac{ {f \choose i} {g-f \choose n-1} }{g \choose n}
-$$
+A survey of articles published in 2019 showed that seven of the top eight most popular
+methods were based on ORA.
 
-Where (P) is the probability of observing at least (k) genes from a functional category within the
-list of selected genes with size (n).
-Here, (g) is the total number of detected genes, and (f) is the number of detected genes belonging to a
-functional category.
+The general approach for over-representation analysis was developed in 1999 and
 
-In a hypothetical example, there were 134 genes classified as detected.
-Of these, 38 were defined as upregulated in response to a stimulus based on a differential expression
-test.
-From the 38, 8 genes belonged to the "cytokine signaling pathway" category.
-This category contains 26 genes in total, but only 18 were detected in this study.
+According to a survey of related publications in 2022,
 
-So the background can be defined as 18/134 and the foreground as 8/38 and the enrichment score
-expressed as (8/38)/(18/135)=1.58
+is one of the most frequently used methods in
+computational biology and involves the summarisation of omics data to reflect biological changes,
+such as the detection of pathway activation in development or disease
+[@Slonim2002-sf; @Khatri2012-do].
+As of August 2020, the various types of enrichment analysis were collectively cited 131,332 times [@Xie2021-ch].
 
-In this example:
+The most popular tool for enrichment analysis appears to be DAVID [@Sherman2022-hf].
+Its 11 related journal articles have received a total of 37,035 citations from 27,129
+separate PubMed articles (as of 29th June 2023).
+Examination of related keywords in PubMed suggests the number of articles describing
+"enrichment analysis" is doubling every two years, as omics analysis becomes accessible to
+more researchers around the world.
 
-*k*=8
+These observations highlight that pathway enrichment tools are essential in the interpretation of omics data.
+However, there are concerns that inadvertent misuse of these methods leads to invalid results
+[@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
+According to a recent survey of 186 articles, incorrect methods are very common [@Wijesooriya2022-li].
+For example, a suitable background list for over-representation tests is reported in only 4.1% of studies.
+Correction of p-values to account for multiple testing was reported in only 54% of studies.
+This survey also revealed a general lack of reporting detail around other aspects of enrichment analysis,
+such as the tool version used.
 
-*n*=26
+We hypothesise that the quality of enrichment analysis studies could be dramatically improved by adopting
+established simple rules for reproducible computing [@Sandve2013-lk], and specific guidelines for
+reporting enrichment analysis [@Zhao2023-ok].
 
-*g*=134
+Our aim is to determine how reproducible published enrichment analyses are.
+To achieve this, we systematically examined the reproducibility of 20 published articles that used a web-based
+enrichment tool.
+We investigate root causes of reproduction failure and describe some potential solutions.
 
-*f*=18
+## A systematic assessment of reproducibility in enrichment analysis
 
-We can calculate the probability of getting exactly 7 members of the set.
+### Methods
 
-$$
-P=\frac{ {18 \choose 7} {134-18 \choose 26-1} }{134 \choose 26}
-$$
+From a list of 186 PubMed central articles using enrichment analysis that we examined previously
+@Wijesooriya2022-li, we selected 20 articles that fulfilled the following criteria; (1) the
+organism under study is human; (2) the tool used was DAVID (Database for Annotation,
+Visualization, and Integrated Discovery) @Huang2007-yw ; (3) the omics type was gene
+expression (array or RNA-seq); and (4) that the gene list was provided in the supplementary data.
+The basis for these selection criteria is that this combination represents the most common application
+of enrichment analysis according to our previous study, which could be reproduced without the need to
+reanalyse raw data. These articles are listed in the bibliography
+[@Liu2019-nm; @Zhang2019-bq; @Zhang2019-zs; @Wang2019-il; @Balusamy2019-ue; @Ahn2019-wl; @Chai2019-cn;
+@Guo2019-fx; @Alebrahim2019-zc; @Hong2019-ym; @Han2019-ix; @McSweeney2019-ij; @Wang2019-yw; @Pan2019-iv;
+@Wang2019-jy; @Weigt2019-mk; @Najafi2019-ce; @Feltrin2019-to; @Wang2019-xx; @Sun2019-xr].
 
-$$
-P=\frac{ 480700 * 1.71e29 }{ 3.84e+33 } = 1/21.4 = 0.047
-$$
+We screened reporting criteria of these articles, including whether the version of the application
+is given, whether p-values underwent correction for multiple testing, whether a background list was
+described and whether gene selection was described unambiguously.
 
-This needs to be repeated for other values of *i*, namely 6 to 0 and then sum all of these.
+The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
+special attention to preventing gene name errors from occurring [@Zeeberg2004-na,Ziemann2016-pv, Abeysooriya2021-uz].
+The reproducibility analysis was performed using the respective methodologies stated in the methods
+sections of the 20 articles using DAVID v6.8, as at this time v6.7 was not available
+(conducted March-May 2022) [@Sherman2022-hf].
 
-```{r,hyperdist}
-res<-unlist(lapply(0:38, function(i) {  (choose(18,i) * choose((134-18),(38-1)) / choose(134,38)) }))
-plot(res,type="b")
+Briefly, we used the DAVID v6.8 "Analysis Wizard" and pasted in the provided gene list into the text box,
+selecting this as "gene list", together with the appropriate gene identifier type.
+Typically, this was the official gene symbol.
+The species was specified as *Homo sapiens*.
+Next, we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results"
+page.
+We followed the steps indicated in the respective papers to select appropriate gene sets investigated
+in the articles such as Gene Ontology (GO; biological process, molecular function), Kyoto Encyclopedia
+of Genes and Genomes (KEGG) pathways, etc.
+Next, the "Functional Annotation Chart" was clicked on, opening a new window with a table of
+enrichment results, which were saved for future reference.
+
+The pathway enrichment results presented in the paper were compared to those we obtained.
+It would be ideal to compare standardised effect sizes (fold enrichment scores) between original and
+reproduction results, however these were reported in the main text of only 3/20 articles.
+Therefore we adopted a binary classification system, based on head-counting pathways that were presented as
+statistically significant in the original article.
+We defined a pathway as confirmed when it reached a similar level of statistical significance
+upon reproduction.
+We define "similar significance level" to be a value within 50% of the value presented in the original article.
+For example if an article shows a pathway with FDR=0.04, then an acceptable FDR in reproduction could be in
+the range of 0.02 to 0.06.
+Additionally, statements made in the abstract, results, and discussion sections based on specific pathway
+enrichment results were carefully examined and cross-checked with our results to understand whether those
+assertions survived reproduction or not.
+
+Each article was critically evaluated and given a score of one (1; low) to three (3; high).
+Articles were rated 1 when <50% of pathways shown in the results were confirmed
+upon reproduction, and most conclusions derived from enrichment analysis were not supported by reproduction.
+Articles were rated 2 when ≥50% of pathways shown in the results were confirmed
+upon reproduction, and most conclusions derived from the pathway enrichment analysis were supported.
+Articles were rated 3 when ≥80% of terms shown in the original article were observed again in
+reproduction with a similar significance value, and all conclusions derived from the pathway analysis were
+supported in reproduction.
+
+### Results
+
+As reproducibility is closely associated with the quality of documentation, we collected key methodological
+information about how the enrichment analysis was done and reported (Table 1).
+Only seven articles reported the version of DAVID that was used.
+Version 6.7 was reportedly used in two articles while version 6.8 was specified in five.
+Correction of p-values for multiple testing was specified or inferred in 11 articles, while the results of
+the remaining nine articles did not take into consideration the importance of p-value correction for
+minimising false positives.
+Of the 20 articles, none mentioned explicitly that a custom background list was used for enrichment analysis.
+We also examined whether the gene selection criteria were unambiguously clear, which was true in 16 articles.
+
+```{r,table1,echo=FALSE}
+
+x <- read.table("../data/checklist.tsv",header=TRUE,row.names=1)
+
+colnames(x) <- gsub("_"," ",colnames(x))
+x[,3]=NULL
+x |> gt(rownames_to_stub = TRUE) |> tab_header(title = "Table 1. Article reporting checklist results.")
+#x %>% kbl(caption = "Table 1. Article reporting checklist results.") %>% kable_paper("hover", full_width = F)
+
 ```
+
+We attempted reproduction according to the methods described in the respective articles, with the limitation
+that we had to use DAVID v6.8 as v6.7 was no longer available.
+The 20 articles were classified based on the degree of reproducibility (**Figure 1**).
+Seven studies were classified as low reproducibility, while nine were classified as medium and just four
+were classified as high.
+Justifications for these classifications are provided in Table S1.
+
+```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 1.
+Examining reproducibility of enrichment analysis in a sample of 20 journal articles.
+Beside the PubMed Central identifier, the reported version of DAVID used in the original publication (blank if none given)."}
+
+score1 <- c("PMC6405693 (v6.7)","PMC6425008 (v6.8)","PMC6535219","PMC6539328 (v6.8)",
+  "PMC6542760","PMC6561911 (v6.8)","PMC6663624")
+score2 <- c("PMC6381667","PMC6463127","PMC6557785","PMC6580941","PMC6591946",
+  "PMC6582306","PMC6333352 (v6.8)","PMC6526186","PMC6607402 (v6.7)")
+score3 <- c("PMC6349697 (v6.8)","PMC6444048","PMC6587650", "PMC6368841")
+l <- list("1-low"=score1,"2-medium"=score2,"3-high"=score3)
+
+barplot(unlist(lapply(l,length)),col="lightgray",
+  xlab="Reproducibility score",ylab="no. articles",
+  ylim=c(0,11))
+
+#grid()
+text(0.7,1:7-0.5,labels=score1,cex=0.85)
+text(1.9,1:9-0.5,labels=score2,cex=0.85)
+text(3.1,1:4-0.5,labels=score3,cex=0.85)
+
+pc <- unlist(lapply(l,length)) / sum(unlist(lapply(l,length))) * 100
+pc2 <- paste(pc,"%",sep="")
+text(c(0.7,1.9,3.1),unlist(lapply(l,length))+0.4,labels=pc2)
+
+pdf("fig1.pdf",width=6,height=5)
+barplot(unlist(lapply(l,length)),col="lightgray",
+  xlab="reproducibility score",ylab="no. articles",
+  ylim=c(0,11))
+grid()
+text(0.7,1:7-0.5,labels=score1)
+text(1.9,1:9-0.5,labels=score2)
+text(3.1,1:4-0.5,labels=score3)
+pc <- unlist(lapply(l,length)) / sum(unlist(lapply(l,length))) * 100
+pc2 <- paste(pc,"%",sep="")
+text(c(0.7,1.9,3.1),unlist(lapply(l,length))+0.4,labels=pc2)
+dev.off()
+
+```
+
+| S.No.
+Pubmed ID: Replication of (Table/Figure) | A.
+Was Replication possible?
+B.
+Were there other significant terms found which were not mentioned in the article?
+| Replication score and justification |
+| --- | --- | --- |
+| 1.PMC6349697: Figure 2C and 2D | A.
+Yes, however, corrected p values are different after replication.
+B.
+Yes, only BP terms were taken into account so several CC and MF significant terms were not mentioned in the paper.
+|  **3 (high).** Considering top 6 BP Terms and KEGG pathways.
+The basis of the selection of just 6 gene ontologies and 4 KEGG pathways shown in the paper needs to be addressed, when there is a higher amount of significant GOs and pathways that are observed after replication.
+It was impossible to determine if the article utilised a background gene list or not.
+The steps to perform DAVID analysis could have been elaborated in the paper.
+The results shown in the paper are observed after replication but only 6 BP GO significant terms are shown in the paper.
+|
+| 2.PMC6368841: Figure 3 A - D | A.
+Only GO Terms were the similar as the paper.
+The FDR values are inconsistent.
+However, the paper should have clearly mentioned the different tools under figure captions to bring more clarity of the terms obtained by respective tools like DAVID and ClusterProfiler.
+The KEGG terms do not come up in DAVID at all.
+B.
+No | **3 (high).** The article receives a score of 3 solely based on GO results.
+GO results from the original manuscript are somewhat similar to reproduced analysis in DAVID.
+The FDR values however are inconsistent.
+The list of genes as input into DAVID analysis is unclear.
+The paper should have mentioned that both up and down-regulated proteins were considered for DAVID analysis.
+Another important question is why were two different tools used for GO and KEGG when both can be performed in DAVID or clusterprofiler?
+|
+| 3.PMC6381667: Figure 1 and Figure 2 | A.
+Results are not reproducible and are highly inconsistent B.
+N.A.
+| **2 (medium).** GO terms obtained after replication showed inconsistent results compared to the results in the original article.
+KEGG results were similar.
+Better naming convention for supplementary file for 117 genes for DAVID input is necessary.
+GO Results are not reproducible and are highly inconsistent.
+It was not made clear which results were obtained with DAVID or IPA.
+|
+| 4.PMC6405693: Table 1 and Table 2; Texts under “Protein-Protein Interaction Network Building and Interrelation Analysis Between Pathways” | A.
+Results do not match.
+B.
+N.A.
+| **1 (low).** GO analysis from up-regulated and down-regulated proteins are less similar to the results given in the paper.
+The FDR values for some pathways are highly inconsistent.
+KEGG Analysis of all DEGs do not match at all.
+|
+| 5.PMC6425008: Most of the tables and figures of the paper (Figure 5, 6, 7), Table 2 | A.
+No.
+B.
+N.A.
+| **1 (low).** This paper could have done proper labelling of supplementary files as it was difficult to find the gene list for DAVID analysis.
+Lacks information on when control genes are used and when citral-affected gene lists are used.
+Replication of GO and KEGG analysis is not consistent with results from the paper.
+How were Figures 5-7 derived?
+The overall protocol is not so coherent.
+|
+| 6.PMC6444048: The text under “Pathway analysis of SUVmax related genes” And Table 2 | A.
+Terms are the same but gene counts have been different.
+The DAVID results are somewhat reproducible.
+B.
+No | **3 (high).** Gene counts in the terms significantly enriched in the “process of cell division” and “nucleoplasm” have changed.
+Only cell cycle is highly enriched in the replicated results, the claims of metabolic pathways and hypoxia signalling (through GSEA analysis) to be enriched are not accurate.
+The results claimed are somewhat similar.
+All the processes have different FDR values.
+|
+| 7.PMC6463127: Table 4, Table 5, Table 6 | A.
+Yes but the significance differed.
+B.
+Yes, several terms from replication were not mentioned in the paper for GO Analysis.
+| **2 (medium).** Does not consider corrected p values.
+Overall, significant terms from GO analysis of up and down-regulated DEGs are missing upon replication.
+|
+| 8.PMC6535219: Figure 2 | A.
+Somewhat similar.
+B.
+There were many top significant terms that were not mentioned.
+| **1 (low).** Narrowing/Filtering the given gene list is a required step that could have been mentioned.
+Many terms were not so significant anymore.
+There were new terms that are of top positions after replication.
+|
+| 9.PMC6539328: Figure 1A | A.
+Mostly not.
+B.
+Yes, many | **1 (low).** The significant pathways stated in the paper are not showing up once replicated.
+There is no clarity of the up-regulated genes used for GO and KEGG analysis.
+Only 21 up-regulated genes are listed whereas the paper states 4831.
+Also, why are only up-regulated genes used?
+|
+| 10.PMC6542760: Figure 4 | A. No. B. Yes, many significant terms were observed. | **1 (low).** Replication results of GO terms are not at all similar with the results shown in the paper. |
+| 11.PMC6557785: Figure 5 A-B | A.
+Yes, but results in paper are not significant upon replication.
+B.
+Yes | **2 (medium).** For GO terms, there are other terms that have higher numbers of genes, making them more enriched than immune-related pathways as mentioned in the paper.
+For KEGG Pathways, there are other pathways that are more significant and contain a higher number of genes enriched for example “Salmonella infection” that has a higher count of genes but is not mentioned in the paper.
+|
+| 12.PMC6561911: Figure 5 a and b | A. No. B. N.A. | **1 (low).** The enrichment analysis results are inconsistent with the results displayed in the paper. There was no KEGG analysis done |
+| 13.PMC6580941: Table 1 and Table 2 | A.
+Yes.
+B.
+Yes!
+Significant terms like “nucleoplasm”, “RNA Binding”, “protein binding” are missing from the paper.
+| **2 (medium).** For GO analysis, the replicated results are not as significant as the GO terms shown in Tables 1 and 2 of the paper.
+The Gene list for input is not so clearly mentioned, it should have been labelled properly in the name of the supplemental file.
+P-values are completely different.
+|
+| 14.PMC6587650: Figure 2 | A.
+Yes.
+B.
+Yes, some.
+| **3 (high).** (Best) The GO terms results of BP, MF, CC and KEGG pathway results from the paper are compliant with replicated results.
+They could have mentioned why they chose the type of sorting shown in the paper to display their results.
+Overall, all results have been replicated.
+|
+| 15.PMC6591946: Figure 5 c and Figure 6 a | A.
+Yes for GO analysis only.
+B.
+Yes for KEGG analysis | **2 (medium).** The GO term results of replication are compliant with Figure 5 c of the paper.
+KEGG pathway results are semi-consistent.
+|
+| 16.PMC6663624: Table 3 and Table 4 | A.
+Yes.
+B.
+Yes | **1 (low).** The top 2 significant GO terms in the paper are not significant after replication analysis is done.
+Other GO terms are the same as replicated results.
+For KEGG Analysis, some terms are more significant after replication.
+Eg: PD-L1 expression and PD-1 checkpoint pathway in cancer, Th1 and Th2 cell differentiation, Th17 cell differentiation were more significant after “natural killer (NK) cell-mediated cytotoxicity” KEGG Pathway.
+The paper is somewhat replicable.
+|
+| 17.PMC6582306: Figure 5 | A.
+Partly.
+B.
+Yes, many.
+| **2 (medium).** The paper uses 44 sets of DEGs grouped in 4 major types.
+For replication, OFC Set 9 was used, DAVID GO analysis was done twice where a discrepancy of results was observed.
+Compared to the paper, the most significant term was missing after replication analysis was conducted.
+For OFC set 9, several pathways are seen after replication but only one pathway is shown in paper.
+This should be addressed.
+|
+| 18.PMC6333352: Figure 7 | A.
+Yes.
+B.
+Yes.
+There was a difference in gene counts as well.
+| **2 (medium).** DAVID was used for KEGG Analysis: From WGCNA - Bahn - KEGG Analysis: Epstein-Barr virus infection, the pathway with the highest gene count is missing in the results shown in the paper.
+From NERI - BAHN - KEGG Analysis: “Pathways of neurodegeneration - multiple diseases”, “Pathways in cancer” - pathways with the highest gene count (30 and 28 respectively) are not mentioned in the paper.
+Other pathways are present and the paper is somewhat reproducible.
+|
+| 19.PMC6526186: Table 1 | A.
+Yes.
+B.
+Yes.
+| **2 (medium).** The top BP GO terms from the paper are no longer significantly higher in the replicated GO results.
+The top 2 KEGG Pathways after replication are “Human cytomegalovirus infection” followed by “Pathways in cancer”.
+The p-values have changed after the results were replicated, compared to the values given in the paper.
+The rest of the pathways are similar to the results given in the paper.
+|
+| 20.PMC6607402: Figure 1 A-D | A.
+Yes.
+B.
+Yes, many!
+| **2 (medium).** For GO analysis and their replication, top enriched BP, CC and MF GO terms are not mentioned.
+Most of the terms are mentioned in the replicated results, but different levels of enrichment.
+|
+
+: Table S1. Classification and justification of reproducibility scores for each article.
+
+The ability to reproduce DAVID enrichment analysis using the supplementary gene
+lists was highly variable.
+The version of DAVID used is likely to be a big contributor to irreproducibility as we know that there are major
+differences in the underlying annotation set (knowledgebase) used by version 6.7 and 6.8.
+This is supported by the fact that none of the v6.7 studies received a high reproduicibility score.
+On the other hand these versions did not receive regular updates, so the knowledgebases were static for relatively
+long times; v6.7 (Jan. 2010 - Oct 2020) and v6.8 (Oct 2016 - Jun 2022).
+We can then deduce that knowledgebase version is likely not the cause of irreproducibility for PMC6561911,
+PMC6539328 and PMC6425008; studies that reportedly used v6.8 but received low reproducibility scores.
+
+We note that DAVID version 6.8 was retired in June 2022, so these 20 studies are currently no longer reproducible
+with the original tools.
+The implications of this are outlined in the Discussion section below.
+
+## A five-pillars approach to addressing reproducibility of enrichment analysis
+
+Our results emphasise the need for better methodological reporting to reduce ambiguity in methods sections
+and highlight that webtools can suddenly become unavailable.
+Although there are various solutions to these issues (refer to the Discussion section), here we will outline
+a potential solution whose design is guided by current best practices
+[@Tan2010-is; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg; @Ziemann2023-dg].
+Our aim is to provide a beginner-friendly guide to enrichment analysis that will be
+reproducible for many years to come.
+
+### Method overview
+
+We have developed a step-by-step protocol for how to conduct a highly reproducible
+enrichment analysis, which is attached as a supplementary file (Suppl Info 1) and published to
+protocols.io with a permissive license [@Ziemann2023-kz].
+This protocol is prepared in accordance with the *five pillars* principles [@Ziemann2023-dg], and includes
+version control with git and GitHub, compute environment control using Docker, literate programming
+with R Markdown, persistent code and data sharing with Zenodo and documentation of all of these to
+empower others to reproduce the workflow.
+Central to this protocol is the provision of template Dockerfile and an R Markdown script, which
+are intended to be remixed and altered by end users to suit their project's needs.
+The general steps involved are shown in **Figure 2**.
+
+```{r,method_diagram,echo=FALSE,fig.height=6,fig.cap = "Figure 2. Flow chart of example workflow. An estimated amount of time to conduct each step is given."}
+
+library("RColorBrewer")
+library("shape")
+
+cols <- brewer.pal(5, "Set2")
+cols2 <- adjustcolor( cols, alpha.f = 0.7)
+plot(0,type='n',axes=FALSE,ann=FALSE, xaxs="i", yaxs="i", xlim=c(0, 11), ylim=c(-3, 8.5))
+
+polygon(c(0,6,6,0),c(8.5,8.5,7.5,7.5), col=cols2[1], border = NA)
+text( 3, 8.0, lab="1. Install Docker: 7 mins",cex=1.5)
+
+polygon(c(0.5,7.5,7.5,0.5),c(7.0,7.0,6.0,6.0), col=cols2[2], border = NA)
+text( 3.9, 6.5, lab="2. Validate example: 14 mins",cex=1.5)
+
+polygon(c(1,8.5,8.5,1),c(5.5,5.5,4.5,4.5), col=cols2[3], border = NA)
+text( 4.8, 5.0, lab="3. Create GitHub repo: 29 mins",cex=1.5)
+
+polygon(c(1.5,10.5,10.5,1.5),c(4,4,3,3), col=cols2[4], border = NA)
+text( 6.1, 3.5, lab="4. Build custom Docker image: 45 mins",cex=1.5)
+
+polygon(c(2,10.5,10.5,2),c(2.5,2.5,1.5,1.5), col=cols2[5], border = NA)
+text( 6.3, 2, lab="5. Customise R Markdown: 3-12 hrs",cex=1.5)
+
+polygon(c(2.5,9.5,9.5,2.5),c(1,1,0,0),col="lightgray", border = NA)
+text( 6.1, 0.5, lab="6. Update repository: 70 mins",cex=1.5)
+
+polygon(c(3,11,11,3),c(-0.5,-0.5,-1.5,-1.5), col=cols2[1], border = NA)
+text( 7.0, -1, lab="7. Update Docker image: 40 mins",cex=1.5)
+
+polygon(c(3.5,12,12,3.5),c(-2,-2,-3,-3), col=cols2[2], border = NA)
+text( 7.6, -2.5, lab="8. Long Term Archiving: 180 mins",cex=1.5)
+
+#Arrows(2, 7.5, 2.5, 7.0, lwd=2, arr.type="triangle")
+#Arrows(2.5, 6, 3, 5.5, lwd=2, arr.type="triangle")
+#Arrows(3, 4.5, 3.5, 4, lwd=2, arr.type="triangle")
+#Arrows(3.5, 3, 4, 2.5, lwd=2, arr.type="triangle")
+#Arrows(4, 1.5, 4.5, 1, lwd=2, arr.type="triangle")
+
+pdf("fig2.pdf",width=8,height=6)
+cols <- brewer.pal(5, "Set2")
+cols2 <- adjustcolor( cols, alpha.f = 0.7)
+plot(0,type='n',axes=FALSE,ann=FALSE, xaxs="i", yaxs="i", xlim=c(0, 11), ylim=c(-3, 8.5))
+
+polygon(c(0,6,6,0),c(8.5,8.5,7.5,7.5), col=cols2[1], border = NA)
+text( 3, 8.0, lab="1. Install Docker: 7 mins",cex=1.5)
+
+polygon(c(0.5,7.5,7.5,0.5),c(7.0,7.0,6.0,6.0), col=cols2[2], border = NA)
+text( 3.9, 6.5, lab="2. Validate example: 14 mins",cex=1.5)
+
+polygon(c(1,8.5,8.5,1),c(5.5,5.5,4.5,4.5), col=cols2[3], border = NA)
+text( 4.8, 5.0, lab="3. Create GitHub repo: 29 mins",cex=1.5)
+
+polygon(c(1.5,10.5,10.5,1.5),c(4,4,3,3), col=cols2[4], border = NA)
+text( 6.1, 3.5, lab="4. Build custom Docker image: 45 mins",cex=1.5)
+
+polygon(c(2,10.5,10.5,2),c(2.5,2.5,1.5,1.5), col=cols2[5], border = NA)
+text( 6.3, 2, lab="5. Customise R Markdown: 3-12 hrs",cex=1.5)
+
+polygon(c(2.5,9.5,9.5,2.5),c(1,1,0,0),col="lightgray", border = NA)
+text( 6.1, 0.5, lab="6. Update repository: 70 mins",cex=1.5)
+
+polygon(c(3,11,11,3),c(-0.5,-0.5,-1.5,-1.5), col=cols2[1], border = NA)
+text( 7.0, -1, lab="7. Update Docker image: 40 mins",cex=1.5)
+
+polygon(c(3.5,12,12,3.5),c(-2,-2,-3,-3), col=cols2[2], border = NA)
+text( 7.6, -2.5, lab="8. Long Term Archiving: 180 mins",cex=1.5)
+dev.off()
+
+```
+
+The Dockerfile uses a base image from Bioconductor that corresponds to the latest available
+stable release version, on the Ubuntu operating system.
+The image contains instructions for installing a few useful utilities including `nano` and `git` for
+modifying scripts, and `magic-wormhole` for transferring data between computers.
+It installs R packages from CRAN and Bioconductor repositories to fulfill common tasks in transcriptome
+analysis such as fetching RNA-seq counts from DEE2 with `getDEE2` @Ziemann2019-zo,
+differential expression with `DESeq2` @Love2014-hd,
+enrichment analysis with `clusterProfiler`, `fgsea` and `mitch`
+[@Yu2012-ts; @Korotkevich2016-ls; @Kaspi2020-dq].
+It also contains instructions to clone a GitHub repository, which contains the scripts that run the
+analytical workflow.
+Lastly, it copies a gene set library from Reactome @Gillespie2022-gz with a clear version
+history to guarantee provenance of the annotation data.
+
+The R Markdown script called `example.Rmd` undertakes the transcriptome and enrichment analyses.
+R Markdown allows the unification of extended descriptions, code and results into a single document,
+which is becoming ubiquitous in bioinformatics and data science more generally
+@Grolemund2017-zl.
+R Markdown supports languages other than R, including Python, Shell, Julia and others.
+The ability to provide extended descriptions is helpful for providing contextualising information about
+the purpose of the work, detailed methods and interpretations derived from the analysis.
+The R Markdown script contains code chunks that perform specific tasks including fetching an example
+data set from DEE2, inspecting data quality, differential expression and two types of enrichment
+analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 3**).
+
+```{r,diagram2,echo=FALSE,fig.cap = "Figure 3. Flow chart of example workflow."}
+
+library("DiagrammeR")
+grViz("digraph flowchart {
+      node [fontname = Helvetica, shape = rectangle]
+      tab1 [label = '@@1']
+      tab2 [label = '@@2']
+      tab3 [label = '@@3']
+      tab4 [label = '@@4']
+      tab5 [label = '@@5']
+      tab6 [label = '@@6']
+      tab1 -> tab2 -> tab3 -> tab4 -> tab6;
+      tab3 -> tab5 -> tab6 }
+      [1]: 'Fetch RNA-seq from DEE2'
+      [2]: 'Quality control'
+      [3]: 'Differential expression'
+      [4]: 'Over-representation analysis'
+      [5]: 'Functional class scoring'
+      [6]: 'Euler diagram'
+")
+
+grViz("digraph flowchart {
+      node [fontname = Helvetica, shape = rectangle]
+      tab1 [label = '@@1']
+      tab2 [label = '@@2']
+      tab3 [label = '@@3']
+      tab4 [label = '@@4']
+      tab5 [label = '@@5']
+      tab6 [label = '@@6']
+      tab1 -> tab2 -> tab3 -> tab4 -> tab6;
+      tab3 -> tab5 -> tab6 }
+      [1]: 'Fetch RNA-seq from DEE2'
+      [2]: 'Quality control'
+      [3]: 'Differential expression'
+      [4]: 'Over-representation analysis'
+      [5]: 'Functional class scoring'
+      [6]: 'Euler diagram'
+") %>%
+  DiagrammeRsvg::export_svg() %>%
+  charToRaw() %>%
+  rsvg::rsvg_pdf("fig3.pdf")
+
+```
+
+For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
+[@Wijesooriya2021-lb; @Zhao2023-ok].
+An example checklist is shown in **Table 2**.
+
+| Reporting criteria | Method or resource used |
+| -- | -- |
+| Origin of gene sets | Reactome (2023-03-06) |
+| Tool used | `clusterProfiler` |
+| Statistical test used | hypergeometric test |
+| P-value correction for multiple comparisons | FDR method |
+| Background list | Genes with >=10 reads per sample on average across all samples |
+| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| ORA directionality | Separate tests for up- and down-regulation |
+| Data availability | via dee2.io at accession SRP038101 |
+| Other parameters | Minimum gene set size of 10 |
+
+: Table 2. An example reporting checklist for enrichment analysis.
+
+The software versions used in this example analysis are provided in Table 3.
+
+| Software | Version | Purpose |
+| --- | --- | --- |
+| Ubuntu | 22.04 | Operating System |
+| R | 4.3.0 | Statistical computing language |
+| nano | 6.2 | text editing |
+| git | 2.34.1 | source control |
+| R/kableExtra | 1.3.4 | render formatted tables |
+| R/vioplot | 0.4.0 | violin charts |
+| R/gplots | 3.1.3 | heatmaps |
+| R/eulerr | 7.0.0 | Venn diagrams |
+| BioC/getDEE2 | 1.6.0 | fetch example RNA-seq data |
+| BioC/DESeq2 | 1.36.0 | differential expression |
+| BioC/fgsea | 1.22.0 | functional class scoring |
+| BioC/clusterProfiler | 4.4.4 | over-representation analysis |
+| BioC/mitch | 1.10.0 | multi-contrast enrichment |
+
+: Table 3. The software bundled in the example Docker image.
+
+Reproducing the example workflow is relatively quick and simple, requiring six commands and approximately 4-7
+minutes starting from a base Linux install.
+The steps include installing Docker, pulling and running the container, running the R Markdown script,
+exiting the container and copy the newly created report from the container to the current working
+directory (Box 1).
+
+```bash
+sudo apt update && sudo apt install docker.io -y  # 30 sec
+sudo docker run --platform linux/amd64 mziemann/enrichment_recipe
+Rscript -e 'rmarkdown::render("example.Rmd")' # 150 sec
+docker cp $(docker ps -aql):/enrichment_recipe/example.html . # instant
+firefox example.html
+```
+<p style="text-align: center;">
+*Box 1. Linux shell commands to reproduce a transcriptome and enrichment analysis using a Docker image.
+Ubuntu 22.04 was used.*
+</p>
+
+### Demonstration
+
+To demonstrate the utility of this workflow, we present the results of this example data analysis;
+it is a comparison of control and 5-azacitidine-treated AML3 cell transcriptomes with RNA-seq
+in triplicate @Lund2014-et.
+We use this analysis to better understand the effect of DNA methyltransferase inhibitors on
+gene expression and elaborate upon previous findings [@Lund2014-et], which mainly focused on
+the relationship between RNA expression and DNA methylation but did not show results of a
+transcriptome-only enrichment analysis.
+Furthermore, we provide a comparison of the two most common approaches for enrichment analysis; ORA and
+FCS, and give guidance for interpreting results of such tests.
+
+For quality control the number of reads per sample was calculated and found to be in the range of
+11-16M.
+Genes with fewer than 10 reads per sample on average were removed from the analysis, leaving 13,168
+genes.
+A multi-dimensional scaling plot showed clustering of samples into their treatment
+groups, suggesting a robust effect of the chemical exposure (**Figure 4A**).
+Differential expression analysis revealed 3,598 differentially expressed genes with FDR<0.05, with
+1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells
+(**Figure 4B**).
+Up and down-regulated gene sets were subjected to ORA separately, using the list of
+13,168 genes as the background, and gene sets from REACTOME.
+
+![Figure 4. Differential expression and enrichment analysis with the example workflow.
+(A) An MDS plot of gene expression data for the largest two principal components.
+Blue and pink points represent control and azacitidine-treated
+samples respectively.
+(B) An MA plot.
+The x-axis represents the log<sub>10</sub> base mean expression, and the y-axis represents the
+log<sub>2</sub> fold change.
+Points in red indicate genes with statistically significant differential expression (FDR<0.05), while
+those in grey did not meet this threshold.
+(C) Bar plot of ORA-based enrichment test results.
+(D) Bar plot of FCS-based enrichment test results.
+(C,D) REACTOME gene sets with FDR>0.05 were discarded and the remaining gene sets were
+ranked by enrichment score (ES)
+The most extreme up- and down-regulated genes sets are shown (based on ES).
+Red bars indicate up-regulation and blue bars indicate down-regulation.
+(E) Euler diagrams of differentially regulated gene sets identified with ORA and FCS approaches.
+](../img/fig_example.png "Example enrichment analysis on")
+
+With ORA, we observed 201 up-regulated and 99 down-regulated gene sets (FDR<0.05).
+We prioritised enrichment results by enrichment score, a surrogate measure for effect size,
+which tends to highlight smaller and more specific gene sets undergoing more drastic differential
+expression, rather than large gene sets that undergo small expression changes.
+This led to the identification of up-regulated pathways related to nuclear RNA import and
+glucokinase regulation; while down-regulated pathways
+were associated with interferon signaling, interleukin signaling and NADPH oxidase
+(host defence) (**Figure 4C**).
+
+After FCS, we identified 239 up-regulated gene sets and 158 down-regulated gene sets (FDR<0.05).
+FCS up-regulated pathways were associated with prometaphase chromosome condensation, nuclear RNA
+transport and defective homologous recombination repair (**Figure 4D**).
+This finding is consistent with changes to the cell cycle observed in 5-azacytidine treated cells,
+where the proportion of cells in G2/M phase is reduced, and proxy measures of apoptosis appear
+slightly higher @Lund2014-et.
+
+Down-regulated pathways related to endosomal/vacuolar (antigen presentation) pathway, interferon
+signaling and TLR/TNF signaling, which was not described in the original publication @Lund2014-et.
+Interestingly, these pathways are reported to be *upregulated* in other cancer cell studies of DNA
+hypomethylating agents for similar durations [@Li2014-op; @Chiappinelli2015-tv; @Okoye-Okafor2019-iq].
+
+Regarding up-regulated pathways, the similarity between ORA and FCS was relatively high, exhibiting a
+Jaccard index of 0.58, while this agreement was lower for down-regulated pathways, with a Jaccard
+index of 0.46 (**Figure 4E**).
+
+## Discussion
+
+There are some notable limitations of this reproduction work.
+The sample size of 20 is relatively small.
+The articles were selected randomly from a list of articles that cited DAVID, and they are mostly from
+low and mid-tier journals (mean Scimago Journal Rank for 2022: 1.2, SD=0.68) and the results may not be representative of
+higher-tier journal articles.
+This is not a complete reproduction of all work in those papers; we only reproduced the DAVID functional enrichment analysis,
+which typically is just one component of an article that can have several other parts, including
+experimental validation of key findings.
+Therefore, if we identified that the enrichment analysis was not reproducible, it does not automatically mean that
+all results and conclusions are flawed.
+
+Furthermore, our reproducibility assessment is based on "head-counting" pathways that meet an arbitrary significance range.
+This binary classification of reproduction results is less powerful as opposed to comparison of standardised effect sizes [@Lakens2013-fx; @Goodman2016-ie].
+For DAVID analysis, the fold enrichment score could be reported as a type of effect size.
+Unfortunately, fold enrichment scores are infrequently reported, so it was not a feasible approach for a systematic and representative
+reproduction study like this.
+In future, we recommend researchers use both significance values and enrichment scores to prioritise and report
+enrichment results.
+
+Despite these limitations, there are a few key observations.
+We observe that only 4/20(20%) of studies achieved a high degree of reproducibility.
+In a way, this is unexpected, given the simplicity of using DAVID.
+Users provide a list of gene identifiers in a text box and select some options.
+There are several possible explanations for this inconsistency:
+
+* The methods stated in the respective articles are inconsistent with what was actually done.
+
+* The list of genes in the supplement is not the same as what was used for enrichment analysis.
+
+* Ambiguity around whether up- and down-regulated genes were considered
+separately or together.
+
+* Ambiguity around whether all GO classes were considered or just one, such as "biological process".
+
+* Ambiguity around whether a custom background gene set was used.
+
+* Differences between DAVID version 6.7 and 6.8.
+
+* Ambiguity around which tool was used, as some articles described using other tools in addition to DAVID
+but did not label figures and tables with the tool used.
+
+This irreproducibility might be expected given that substantial methodological flaws and reporting
+deficiencies were present in 85% of enrichment analyses from which this set of 20 studies were
+drawn @Wijesooriya2022-li.
+Authors can prevent these reporting deficiencies providing a checklist of methodological details similar
+to ours (**Table 2**).
+Checklists like these are a useful resource as they are simple for journals to enforce and reviewers to check.
+
+While researching the literature on computational reproducibility in bioinformatics, we identified
+many case studies, but only two systematic examinations of multiple journal articles.
+Ioannidis et al @Ioannidis2009-hv examined a set of 10 gene expression studies with available data
+published in 2005-2006, and found only two could be reproduced to a satisfactory level.
+A recent preprint reports that only 245 out of 9,625 Jupyter notebooks (2.5%) in biomedical journals
+successfully reproduced results identical to the respective journal article [@Samuel2022-dy].
+More work in this direction is urgently needed to understand the causes of irreproducible bioinformatics
+research so it can be remedied, as reproducibility is considered only the first step towards
+reliable research [@Nosek2022-hw].
+
+This article is an attempt to foster increased reproducibility in genomics by demonstrating the
+implementation of the five pillars principles into a commonly used routine, namely differential
+expression and pathway enrichment analysis.
+However, these concepts may be difficult for novice analysts, and so we provide a step-by-step
+protocol to guide them on how to customise template Dockerfile and R Markdown scripts to suit their
+project's needs [@Ziemann2023-kz].
+Although we aimed to make this process as easy as possible, it requires a substantial time outlay, which
+we estimate at 15-20 hours.
+This is significantly longer than using a web-based tool that provides results in just a few
+minutes, but we consider the time investment is well spent.
+It ensures analytical reproducibility, assists with auditing and makes the process of reproduction
+simple and fast (Box 1), which should aid the early identification of errors.
+Using a literate scripting approach such as R Markdown, Jupyter or Quarto notebook, enables very high
+levels of workflow transparency including inspection of intermediate data objects, and evidence that
+the published methods were followed.
+
+With the retirement of DAVID 6.8 and earlier versions, which occurred in June 2022, all DAVID analyses
+prior to 2021 have become irreproducible, including the sample of 20 studies featured here.
+This is concerning, as these studies were only three years old at the time of the tool's retirement.
+Moreover, according to PubMed citations, the DAVID project accrued 27,572 citations before 2021 from 20,878
+PubMed articles, suggesting that ~20,000 omics studies can no longer be reproducible with the original tools.
+
+The problem of bioinformatic resources including web services becoming unavailable is not a new one, being
+raised over a decade ago [@Hennessey2013-tj] and revisited in 2020 [@Kern2020-iz].
+We are worried that a reliance on web tools that lack open source code or containers can leave
+researchers in violation of their data management responsibilities, as preservation mandates
+include software as well as original research data
+[@NHMRC2019-wu ; @Harvard_2023-cl].
+
+With this, we must reconsider whether web-based tools are valid for scientific publications if they are at
+risk of becoming unavailable without notice due to link decay.
+Some progress is being made to make web-based point-and-click tools more reproducible.
+The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
+workflows that enable reproducibility [@Kuehn2008-su; @Goecks2010-lq; @Teichman2023-fx].
+The Degust differential expression suite is a graphical interface to popular
+expression tools like edgeR and limma and allows users to download the computer script to facilitate
+future non-interactive reproducibility [@Powell2019-lv].
+The ShinyGO web-based enrichment tool is a satisfactory GUI-based alternative to DAVID as it provides Docker image
+downloads for all past versions [@Ge2020-qe], facilitating preservation and aiding future reproducibility.
+Still, such tools leave documentation and reproducibility gaps which are resolved with the highly reproducible
+approach detailed here.
+
+Although container-based workflows are frequently described as being a complete solution to computational
+reproducibility, they have some limitations.
+For example, building a Docker image relies on a network of servers providing software packages which may work now
+but not in a decade's time.
+Therefore building such images has a short reproducibility horizon.
+The Guix package and container manager, together with the Software Heritage code repository enables reproducible
+image builds, which enhances the transparency of compute environments, and lengthening the reproducibility horizon
+[@Vallet2022-gw].
+It is also not guaranteed that Docker images published now will be compatible with future Docker versions,
+whereas Guix can build an older version of itself.
+
+Although the recipe described here is designed for gene expression analysis, it is modifiable to accommodate
+other omics, for example genomic, epigenomic and proteomic data, given the names of analytes measured are
+consistent with the annotation databases.
+
+Although we hope that highly reproducible enrichment analysis will become more widespread, we do not
+anticipate any dramatic changes in researcher behaviour unless the motivating factors are addressed.
+Currently, reward structures in academia favour career progression for individuals who publish many
+articles as opposed to individuals who publish fewer but more rigorously conducted works.
+Solutions to this could include mandates by funding agencies, institutions or journals [@Garijo2013-zb],
+badging of articles that comply with best practices [@Kidwell2016-pm], and recognition of research rigour
+in track record evaluations.
+
+# Data and Software Availability
+
+## Underlying data
+
+Publicly available data were obtained from Digital Expression Explorer 2 (dee2.io).
+Expression data from untreated and Aza treated AML3 cells.
+NCBI Sequence Read Archive accession Number SRP038101.
+
+## Software and Code
+
+* The code repository including template Dockerfile and R Markdown script are available on GitHub
+(https://github.com/markziemann/enrichment_recipe).
+
+* The example Docker image is available on DockerHub (https://hub.docker.com/r/mziemann/enrichment_recipe).
+
+* The code repository and Docker image have been uploaded to Zenodo for long-term archiving
+(https://zenodo.org/record/8170984).
+
+## Availability of other materials
+
+* The full protocol can be found at protocols.io
+(https://www.protocols.io/view/a-recipe-for-extremely-reproducible-enrichment-ana-j8nlkwpdxl5r/v2) @Ziemann2023-kz.
+
+* Instructional video guides are available on YouTube
+(https://www.youtube.com/playlist?list=PLAAydBPtqFMXDpLa796q7f7W1HK4t_6Db).
+
+## Author Contributions
+
+Conceptualization: MZ, AB.
+Data Curation: MZ.
+Formal Analysis: MZ.
+Funding Acquisition: N/A.
+Investigation: MZ, AB.
+Methodology: MZ, AB.
+Project Administration: MZ.
+Resources: MZ.
+Software: MZ, AB.
+Supervision: MZ.
+Validation: AB.
+Visualization: MZ, AB.
+Writing – Original Draft Preparation: MZ, AB.
+Writing – Review & Editing: MZ, AB.
+
+## Competing Interests
+
+No competing interests were disclosed.
+
+## Grant Information
+
+The authors declared that no grants were involved in supporting this work.
 
@@
 
+We thank Ms Megan Soria and Ms Kaumadi Wijesooriya (Deakin University) for discussions on the concept and manuscript.
 This research was supported by use of the Nectar Research Cloud, a collaborative Australian research
 platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
+The authors gratefully acknowledge the contribution to this work of the
+Victorian Operational Infrastructure Support Program received by the Burnet Institute.
 
-## Availability of materials
-
-* Code repository:
-
-* Example Docker image:
-
-## Bibliography
+# Bibliography

### 2024-06-06 manuscript/manuscript.Rmd
ms

@@
 
-* Reproducibility
-
-* Enrichment analysis
-
 * Pathway analysis
@@
 
-* Docker
+* Research integrity
 
@@
 Functional enrichment analysis involves the summarisation of omics data to reflect
-biological changes, and has become one of the most popular methods in bioinformatics.
+biological differences, and has become one of the most popular methods in bioinformatics.
 Enrichment analysis software packages have been collectively cited 131,332 times as of
@@
 In FCS, all detected genes are ranked by their degree of differential expression followed
-by some statistical test for enrichment of gene categories at either extreme of the
+by a statistical test for enrichment of gene categories at either extreme of the
 ranked list.
@@
 
-Of these two methods, FCS is reported to be the most accurate, while ORA seems to be the
-most popular.
-
 According to a study that used simulated differential expression data, FCS accuracy was
 superior to ORA over a range of experimental designs [@Kaspi2020-rn].
+Despite this performance difference, a survey of articles published in 2019 showed that
+seven of the top eight most popular tools were based on ORA [@Wijesooriya2022-fw].
+The discrepancy between performance and popularity likely arose due to the relative ease of
+conducting ORA as compared to FCS.
+A minimalist ORA might involve pasting a single list of genes into a text box on a website,
+where the results appear nearly instantaneously.
+On the other hand, FCS typically involves downloading specific softwares and dealing with a
+dataset that represents every detected gene, typically ~20,000 rows and ensuring that the
+format from upstream tools can be shoehorned to meet the requirements of FCS packages.
 
-A survey of articles published in 2019 showed that seven of the top eight most popular
-methods were based on ORA.
+Since it's initial development in 1999 [@Tavazoie1999-bu], ORA has proliferated widely,
+becoming part of many websites and software packages.
+The most popular ORA website is DAVID [@Dennis2003-ar;@Huang2009-gi;@Sherman2022-jn],
+followed by Ingenuity Pathway Analysis, a commercial software package provided by QIAGEN Inc.
+ORA has been implemented into popular R/Bioconductor packages including `ClusterProfiler`
+[@Yu2012-jj],`limma`[@Ritchie2015-oz] and `GOseq`[@Young2010-iw].
+
+Despite the popularity of the technique, there are concerns that
 

### 2024-06-06 manuscript/manuscript.Rmd
ms

--- 2023-07-09 manuscript/manuscript.Rmd
@@
 ---
-title: "Importance of using the correct background list for over-representation analysis"
-author: "Mark Ziemann, Barry Schroeter, Anusuiya Bora"
+title: "Two subtle problems with over-representation analysis"
 date: "`r Sys.Date()`"
@@
 
-Alternate title: *Problems in over-representation analysis software*
-
-[Mark Ziemann<sup>1</sup>*](https://orcid.org/0000-0002-7688-6974),
-[Barry Schroeter<sup>1</sup>]()
-[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352)
+[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352),
+[Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974)
 
@@
 
-1. Deakin University, Geelong, Australia, School of Life and Environmental Sciences.
+1. Burnet Institute, Melbourne, Australia.
 
-2. Vellore Institute of Technology, Vellore, India.
+2. Deakin University, School of Life and Environmental Sciences, Geelong, Australia.
 
-(*) Corresponding author: m.ziemann@deakin.edu.au
+(*) Corresponding author: mark.ziemann@burnet.edu.au
 
-## Abstract
+# Abstract
 
-* Over-representation analysis (ORA) is the most common form of functional enrichment and one of the
-most used techniques in computational biology, yet there are common methodological issues that
-undermine its validity.
+TODO.
 
-* These problems are typically ascribed to errors in analysis and reporting by the end-users,
-however software design is another contributor.
+# Keywords
 
-* Here, we identify and characterise a potential bug popular ORA software
-package and web tool.
+* Bioinformatics
 
-* We show that this bug deleteriously affects the results of ORA tests, and we present a simple
-work-around to circumvent it.
+* Pathway analysis
+
+* Gene set enrichment analysis
+
+* Research integrity
+
+# Main Body
 
@@
 
-Functional enrichment analysis is a family of computational approaches used to interpret large omics
+Functional enrichment analysis involves the summarisation of omics data to reflect
+biological differences, and has become one of the most popular methods in bioinformatics.
+Enrichment analysis software packages have been collectively cited 131,332 times as of
+late 2019 [@Xie2021-ch].
+The algorithms used by these tools can largely be divided into two groups, functional
+class scoring (FCS) and over-representation analysis (ORA).
+In FCS, all detected genes are ranked by their degree of differential expression followed
+by a statistical test for enrichment of gene categories at either extreme of the
+ranked list.
+ORA involdes the selection of genes of interest, followed by a test to ascertain whether
+certain functional categories are over-represented in the selected genes.
+
+According to a study that used simulated differential expression data, FCS accuracy was
+superior to ORA over a range of experimental designs [@Kaspi2020-rn].
+Despite this performance difference, a survey of articles published in 2019 showed that
+seven of the top eight most popular tools were based on ORA [@Wijesooriya2022-fw].
+The discrepancy between performance and popularity likely arose due to the relative ease of
+conducting ORA as compared to FCS.
+A minimalist ORA might involve pasting a single list of genes into a text box on a website,
+where the results appear nearly instantaneously.
+On the other hand, FCS typically involves downloading specific softwares and dealing with a
+dataset that represents every detected gene, typically ~20,000 rows and ensuring that the
+format from upstream tools can be shoehorned to meet the requirements of FCS packages.
+
+Since it's initial development in 1999 [@Tavazoie1999-bu], ORA has proliferated widely,
+becoming part of many websites and software packages.
+The most popular ORA website is DAVID [@Dennis2003-ar;@Huang2009-gi;@Sherman2022-jn],
+followed by Ingenuity Pathway Analysis, a commercial software package provided by QIAGEN Inc.
+ORA has been implemented into popular R/Bioconductor packages including `ClusterProfiler`
+[@Yu2012-jj],`limma` [@Ritchie2015-oz] and `GOseq` [@Young2010-iw].
+
+Despite the popularity of ORA, there are concerns that this technique is being misused.
+Due to technological and biological reasons, not all genes are detected in transcriptomic
+studies, meaning that some genes will more readily appear as differentially expressed.
+To address this bias, a background list of detected genes is required for comparison to the
+list of selected genes (foreground list) [@Tilford2009-vx].
+Failure to use a background gene list can lead to dramatic changes in enrichment results,
+rendering them unreliable [@Timmons2015-ki].
+Unfortunately, use of an appropriate background list is reported in only a small fraction
+(~4%) of peer-reviewed articles describing enrichment analysis [@Wijesooriya2022-fw].
+In addition, correction of p-values for multiple testing is crucial in controlling the
+false positive rate, but this is reported in only ~54% of studies [@Wijesooriya2022-fw].
+These discoveries have resulted in the development of best practices for end users
+to improve awareness of the issues and help enforce reporting standards [@Zhao2023-mx].
+
+The focus of this work is to raise awareness to two existing problems in ORA as implemented
+by some popular enrichment tools.
+The first problem is that genes belonging to the background list are removed from the analysis
+entirely if they are not annotated as belonging to any functional categories.
+This results in the background list appearing smaller than it should, leading to an
+underestimation of fold enrichment scores and significance values.
+In principle, this problem would differentially affect analyses involving different gene set
+libraries, with smaller libraries such as Kyoto Encyclopedia of Genes and Genomes (KEGG)
+[@Kanehisa2023-ni], which describes 2,727 genes affected more severely as compared to larger
+libraries like Gene Ontology [@The_Gene_Ontology_Consortium2023-xp] which describes 19,428
+genes (figures obtained from MSigDB [@Liberzon2015-cs]).
+
+The second problem is that adjusting p-values for multiple testing, also known as
+false-discovery correction is implemented improperly.
+ORA tools should implement a threshold by which a pathway could be considered as detected.
+Ideally, a pathway should be considered detected based on the presence of a predetermined
+number of member genes in the background set.
+However some tools use the foreground set to define a pathway as detected.
+This is problematic, as the act of trying to find an intersect between a foreground gene list
+and a pathway is in itself a test, even if there are no common genes.
+This results in pathways being discarded from the analysis if no common genes are found in
+the foreground, despite a large number of those genes being present in the background list.
+This leads to FDR values appearing lower than they should.
+We expect this problem to have a more severe effect when the foreground list is fairly small.
+
+Here, we show the effect of these two problems on real RNA-seq based enrichment analysis
+results and demonstrate these two issues lead to poor performance using simulated expression
 data.
-The most commonly used method is over-representation analysis (ORA) which involves the selection of
-genes of interest, followed by a statistical test to evaluate whether genes of certain annotated
-categories are over-represented in the set of interest.
-The set of interest may be defined in different ways, such as genes that meet a differential expression
-significance threshold, or those that have been clustered into a group based on co-expression patterns.
-There are various sources of annotated categories, including gene ontologies, protein-coding features,
-gene co-expression and signalling or metabolic pathways.
-Providers of such annotations include but are not limited to the gene ontology consortium, Reactome,
-KEGG, WikiPathways and MSigDB.
+Finally we classify popular tools regarding these two problems and provide general
+recommendations for end users.
 
-Central to ORA tests is the selection of a list of background genes.
-This is also sometimes called the "universe" or the "reference" list.
-These are defined as any gene that is detected robustly, and has a chance at making the set of interest.
-For RNA-seq studies, this could be a threshold detection limit based either on raw counts or counts per
-million.
+## A systematic assessment of reproducibility in enrichment analysis
 
-The first use of ORA-based functional enrichment that we are aware of is a 1999 report that used the
-hypergeometric test to determine the chance probability of observing over-representation of certain
-annotated functions in co-regulated genes in yeast [@Tavazoie1999-kf].
-This general approach is the basis of most ORA-based significance testing and can be represented by the
-following equation:
+### Methods
 
-$$
-P=1-\sum_{i=0}^{k-1}\frac{ {f \choose i} {g-f \choose n-1} }{g \choose n}
-$$
+### Results
 
-Where (P) is the probability of observing at least (k) genes from a functional category within the
-list of selected genes with size (n).
-Here, (g) is the total number of detected genes, and (f) is the number of detected genes belonging to a
-functional category.
+## Discussion
 
-The fold enrichment score (ES) is commonly used to express the degree of enrichment and is a ratio-of-ratios.
-The foreground ratio is the number of selected genes belonging a functional class (A) divided by the
-number of detected genes in the functional class (B).
-The background ratio is the number of selected genes not belonging to the functional class (C) divided
-by the number of detected genes not in the functional class (D).
+# Data and Software Availability
 
-$$
-ES=\frac{ A / B}{ C / D}
-$$
+## Underlying data
 
-Both equations illustrate that correct definition of the number of background genes is important to
-obtain accurate enrichment scores and significance values.
-A common error in ORA of differential gene expression is to use all genes in the genome as the
-background @Wijesooriya2022-li.
-This leads to a severe over-estimation of the number of background genes as there are ~45k annotated
-genes in the human genome (according to Ensembl v109), but RNA-seq robustly detects only ~15k genes.
-This leads to inflation of p-values and results that share only ~40% similarity with properly conducted
-ORA analysis.
-This issue is even more severe for other omics, where the number of detected genes is lower, for example
-proteomics and single cell RNA-seq where there are only a few thousand detected analytes.
-To address this issue, most major ORA tools allow for a user-provided background list of detected genes.
+Publicly available data were obtained from Digital Expression Explorer 2 (dee2.io).
 
-While testing some of these tools, we observed an unusual behaviour whereby the number of background
-genes reported by the tool varied significantly from the length of the gene list we provided.
-Moreover, this number also varied when we used a different library of gene function annotations.
-Puzzled by this behaviour, we browsed the computational biology forums and were met with other confused
-users wondering why the number of genes in the background list would vary like this.
-The tool developers kindly clarified the situation, by stating that the tool removed genes from the
-entire analysis if they were not members of a functional category.
-We thought this behaviour odd, because it has the potential to significantly alter the test being
-conducted, away from it's original application.
+## Software and Code
 
-In this paper, we examine the effect of this filtering step on ORA of real RNA-seq data, use simulated
-RNA-seq data to determine the impact upon precision and recall, and provide practical recommendations
-to end users to obtain the most accurate results.
+* The code repository including template Dockerfile and R Markdown script are available on GitHub
+(https://github.com/markziemann/enrichment_recipe).
 
-In a hypothetical example, there were 134 genes classified as detected.
-Of these, 38 were defined as upregulated in response to a stimulus based on a differential expression
-test.
-From the 38, 8 genes belonged to the "cytokine signaling pathway" category.
-This category contains 26 genes in total, but only 18 were detected in this study.
+* The example Docker image is available on DockerHub (https://hub.docker.com/r/mziemann/enrichment_recipe).
 
-So the background can be defined as 18/134 and the foreground as 8/38 and the enrichment score
-expressed as (8/38)/(18/135)=1.58
+* The code repository and Docker image have been uploaded to Zenodo for long-term archiving
+(https://zenodo.org/record/8170984).
 
-In this example:
+## Availability of other materials
 
-*k*=8
+* The full protocol can be found at protocols.io
+(https://www.protocols.io/view/a-recipe-for-extremely-reproducible-enrichment-ana-j8nlkwpdxl5r/v2) @Ziemann2023-kz.
 
-*n*=26
+* Instructional video guides are available on YouTube
+(https://www.youtube.com/playlist?list=PLAAydBPtqFMXDpLa796q7f7W1HK4t_6Db).
 
-*g*=134
+## Author Contributions
 
-*f*=18
+Conceptualization: MZ, AB.
+Data Curation: MZ.
+Formal Analysis: MZ.
+Funding Acquisition: N/A.
+Investigation: MZ, AB.
+Methodology: MZ, AB.
+Project Administration: MZ.
+Resources: MZ.
+Software: MZ, AB.
+Supervision: MZ.
+Validation: AB.
+Visualization: MZ, AB.
+Writing – Original Draft Preparation: MZ, AB.
+Writing – Review & Editing: MZ, AB.
 
-We can calculate the probability of getting exactly 7 members of the set.
+## Competing Interests
 
-$$
-P=\frac{ {18 \choose 7} {134-18 \choose 26-1} }{134 \choose 26}
-$$
+No competing interests were disclosed.
 
-$$
-P=\frac{ 480700 * 1.71e29 }{ 3.84e+33 } = 1/21.4 = 0.047
-$$
+## Grant Information
 
-This needs to be repeated for other values of *i*, namely 6 to 0 and then sum all of these.
-
-```{r,hyperdist}
-res<-unlist(lapply(0:38, function(i) {  (choose(18,i) * choose((134-18),(38-1)) / choose(134,38)) }))
-plot(res,type="b")
-```
+The authors declared that no grants were involved in supporting this work.
 

### 2024-06-07 analysis/analyse_rnaseq1.Rmd
fig1

@@
 # number of sets
-lapply(gs,length)
+num <- unlist(lapply(gs,length))
+num
 
 # set size summary
-lapply(gs,function(x) {
+sz <- unlist(lapply(gs,function(x) {
   y <- unlist(lapply(x, length ))
   summary(y)
-} )
+} ))
+sz <- t(matrix(sz,ncol=9))
+rownames(sz) <- names(gs)
+colnames(sz) <- names(summary(c(1,2,3,4)))
+sz
 
 # gene coverage
-lapply(gs,function(x) {
+cov <- unlist(lapply(gs,function(x) {
   y <- unique(unlist(x))
   length(y)
-} )
+} ))
+cov
 
 # total number of annotations
-lapply(gs, function(x) {
+tot <- unlist(lapply(gs, function(x) {
   sum(unlist(lapply(x,length)))
-} )
+} ))
+tot
+
+```
+
+## First test
+
+```{r,initialtest1}
+
+x1 <- readRDS("../dataprep/bulkrna1.Rds")
+head(x1)
+
+# remove genes without symbols
+x1f <- x1[lapply(strsplit(rownames(x1) , " "),length)>1,]
+
+dim(x1)
+dim(x1f)
+
+# about 1000 genes are lost after removing redundant gene names
+bg <- unique(sapply(strsplit(rownames(x1f) , " "),"[[",2))
+length(bg)
+
+up <-rownames(subset(x1f, log2FoldChange > 0 & padj < 0.05 ))
+up <- unique(sapply(strsplit(up," "),"[[",2))
+length(up)
+
+dn <-rownames(subset(x1f, log2FoldChange < 0 & padj < 0.05 ))
+dn <- unique(sapply(strsplit(dn," "),"[[",2))
+length(dn)
+
+# kegg gene set
+kegg <- gs[[1]]
+kegg <- stack(kegg)
+colnames(kegg) <- c("gene","term")
+kegg <- kegg[,c(2,1)]
+kegg$term <- gsub("KEGG_MEDICUS_","",kegg$term)
+kegg$term <- gsub("_"," ",kegg$term)
+
+ora_up <- as.data.frame(enricher(gene = up ,
+    universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = kegg,
+    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))
+
+  br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))
+
+  ora_up$es <- gr/br
+
+ora_up$geneID = ora_up$ID = ora_up$Description = NULL
+
+ora_up_kegg <- ora_up
+
+head(ora_up,20) %>%
+  kbl(row.names = TRUE, caption="Upregulated KEGGs in SRP128998") %>%
+  kable_paper("hover", full_width = F)
+
+# GO gene set
+go <- gs[[6]]
+go <- stack(go)
+colnames(go) <- c("gene","term")
+go <- go[,c(2,1)]
+go$term <- gsub("^GO","",go$term)
+go$term <- gsub("_"," ",go$term)
+
+ora_up <- as.data.frame(enricher(gene = up ,
+    universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = go,
+    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))
+
+  br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))
+
+  ora_up$es <- gr/br
+
+ora_up$geneID = ora_up$ID = ora_up$Description = NULL
+
+ora_up_go <- ora_up
+
+head(ora_up,20) %>%
+  kbl(row.names = TRUE, caption="Upregulated GOs in SRP128998") %>%
+  kable_paper("hover", full_width = F)
+
+```
+
+Transfer a gene set from GO to KEGG analysis.
+
+```{r,intitialtest2}
+
+pw <- gs[[6]][grep("GOBP_COMPLEMENT_ACTIVATION_CLASSICAL_PATHWAY", names(gs[[6]]))]
+
+pw
+
+pw <- stack(pw)[,c(2,1)]
+colnames(pw) <- c("term","gene")
+kegg2 <- rbind(kegg,pw)
+
+ora_up <- as.data.frame(enricher(gene = up ,
+    universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = kegg2,
+    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))
+
+  br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))
+
+  ora_up$es <- gr/br
+
+ora_up$geneID = ora_up$ID = ora_up$Description = NULL
+
+ora_up_kegg2 <- ora_up
+
+head(ora_up,20) %>%
+  kbl(row.names = TRUE, caption="Upregulated KEGGs in SRP128998 with Complement") %>%
+  kable_paper("hover", full_width = F)
+
+message("GO")
+ora_up_go[rownames(ora_up_go)=="BP COMPLEMENT ACTIVATION CLASSICAL PATHWAY",]
+
+message("KEGG + complement")
+ora_up_kegg2[rownames(ora_up_kegg2)=="GOBP_COMPLEMENT_ACTIVATION_CLASSICAL_PATHWAY",]
+
+```
+
+Fix KEGG analysis.
+
+```{r,keggfix1}
+
+bgdf <- data.frame("background",bg)
+colnames(bgdf) <- c("term","gene")
+keggfix <- rbind(kegg,bgdf)
+
+ora_up <- as.data.frame(enricher(gene = up ,
+    universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = keggfix,
+    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))
+
+  br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))
+
+  ora_up$es <- gr/br
+
+ora_up$geneID = ora_up$ID = ora_up$Description = NULL
+
+ora_up_keggfix <- ora_up
+
+head(ora_up,20) %>%
+  kbl(row.names = TRUE, caption="Upregulated KEGG fixed in SRP128998 with Complement") %>%
+  kable_paper("hover", full_width = F)
+
+message("Original")
+ora_up_kegg[which(rownames(ora_up_kegg)=="REFERENCE CLASSICAL PATHWAY OF COMPLEMENT CASCADE C4 C2 TO C3 CONVERTASE FORMATION"),]
+
+message("Fixed")
+ora_up_keggfix[which(rownames(ora_up_keggfix)=="REFERENCE CLASSICAL PATHWAY OF COMPLEMENT CASCADE C4 C2 TO C3 CONVERTASE FORMATION"),]
 
@@
 barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,1250),
-  xlab="no. significant sets", main="Effect of correcting the background bug",
+  xlab="no. significant sets",
   col=c("gray30","gray80"))
@@
 
+png("fig1_ORAsig.png")
+par(mar=c(c(5.1, 7.1, 2.1, 2.1) ))
+barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,1250),
+  xlab="no. significant sets",
+  col=c("gray30","gray80"))
+text(df$ns1+50,((1:nrow(df))*3)-1.5,labels=signif(df[,1],3))
+text(df$ns2+50,((1:nrow(df))*3)-0.5,labels=signif(df[,2],3))
+#abline(v=seq(100,1000,100),lty=2,lwd=0.5,col="gray")
+legend("bottomright", inset=.02, legend=c("corrected","original"),
+  fill=c("gray80","gray30"), horiz=FALSE, cex=1.1)
+
+mylabels <- gsub("$","%",gsub("^","+",signif(((df$ns2/df$ns1)-1)*100,3)))
+text(df$ns2+160,((1:nrow(df))*3)-1,labels=mylabels,cex=1.1)
+dev.off()
+
 # JACCARD
-jac <- matrix( res1$jaccard , ncol=9 )
+jac <- matrix( res1$Jaccard , ncol=9 )
 rownames(jac) <- names(d)
@@
 
-barplot(sort(jac2),horiz=TRUE,las=1,xlim=c(0,1),
-  main="Jaccard similarity Corrected:Original",xlab="Jaccard index")
+par(mar=c(c(5.1, 7.1, 2.1, 2.1) ))
+barplot(sort(jac2),horiz=TRUE,las=1,xlim=c(0,1), xlab="Jaccard index")
+
+par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
+plot(cov,jac2,xlim=c(0,30000),ylim=c(0,1),pch=19,col="lightgray",cex=3,
+  xlab="gene coverage",ylab="Jaccard index")
+text(cov,jac2,labels=names(cov))
+mylm <- lm(jac2~cov)
+abline(mylm,lty=2,lwd=2)
+SLOPE=unname(signif(mylm$coefficients[2],3))
+INT=unname(signif(mylm$coefficients[1],3))
+PVAL=signif(summary(mylm)$coefficients[4],3)
+RSQ=signif(summary(mylm)$r.squared,3)
+HEADER=paste("Rsq =",RSQ,"; p-val =",PVAL)
+mtext(HEADER)
+
+png("fig1_ORAjac.png")
+par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
+plot(cov,jac2,xlim=c(0,30000),ylim=c(0,1),pch=19,col="lightgray",cex=3,
+  xlab="gene coverage",ylab="Jaccard index")
+text(cov,jac2,labels=names(cov))
+mylm <- lm(jac2~cov)
+abline(mylm,lty=2,lwd=2)
+SLOPE=unname(signif(mylm$coefficients[2],3))
+INT=unname(signif(mylm$coefficients[1],3))
+PVAL=signif(summary(mylm)$coefficients[4],3)
+RSQ=signif(summary(mylm)$r.squared,3)
+HEADER=paste("Rsq =",RSQ,"; p-val =",PVAL)
+mtext(HEADER)
+dev.off()
 

### 2024-06-07 manuscript/manuscript.Rmd
fig1

@@
 
-[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352),
-[Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974)
+[Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974),
+[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352)
 
@@
 
-Functional enrichment analysis involves the summarisation of omics data to reflect
-biological differences, and has become one of the most popular methods in bioinformatics.
-Enrichment analysis software packages have been collectively cited 131,332 times as of
-late 2019 [@Xie2021-ch].
-The algorithms used by these tools can largely be divided into two groups, functional
-class scoring (FCS) and over-representation analysis (ORA).
+Over-representation analysis (ORA) is a type of functional enrichment analysis (FEA) that
+involves the summarisation of omics data to reflect biological differences.
+Generally speaking, FEA has become one of the most popular methods in bioinformatics,
+collectively accumulating 131,332 citations as of late 2019 [@Xie2021-nv].
+
+ORA involdes the selection of genes of interest, followed by a test to ascertain whether
+certain functional categories are over-represented in the selected genes.
+Since it's initial development in 1999 [@Tavazoie1999-bu], ORA has proliferated widely,
+becoming part of many websites and software packages.
+The most popular ORA website is DAVID [@Dennis2003-ar;@Huang2009-gi;@Sherman2022-jn],
+followed by Ingenuity Pathway Analysis, a commercial software package provided by QIAGEN Inc.
+ORA has been implemented into popular R/Bioconductor packages including `ClusterProfiler`
+[@Yu2012-jj;@Wu2021-wy], `limma` [@Ritchie2015-oz] and `GOseq` [@Young2010-iw].
+
+Another popular approach to FEA is functional class scoring (FCS).
 In FCS, all detected genes are ranked by their degree of differential expression followed
 by a statistical test for enrichment of gene categories at either extreme of the
-ranked list.
-ORA involdes the selection of genes of interest, followed by a test to ascertain whether
-certain functional categories are over-represented in the selected genes.
+ranked list [@Subramanian2005-no].
 
@@
 seven of the top eight most popular tools were based on ORA [@Wijesooriya2022-fw].
-The discrepancy between performance and popularity likely arose due to the relative ease of
+The discrepancy between performance and popularity is likely due to the relative ease of
 conducting ORA as compared to FCS.
-A minimalist ORA might involve pasting a single list of genes into a text box on a website,
+A minimal ORA might involve pasting a single list of genes into a text box on a website,
 where the results appear nearly instantaneously.
-On the other hand, FCS typically involves downloading specific softwares and dealing with a
-dataset that represents every detected gene, typically ~20,000 rows and ensuring that the
-format from upstream tools can be shoehorned to meet the requirements of FCS packages.
-
-Since it's initial development in 1999 [@Tavazoie1999-bu], ORA has proliferated widely,
-becoming part of many websites and software packages.
-The most popular ORA website is DAVID [@Dennis2003-ar;@Huang2009-gi;@Sherman2022-jn],
-followed by Ingenuity Pathway Analysis, a commercial software package provided by QIAGEN Inc.
-ORA has been implemented into popular R/Bioconductor packages including `ClusterProfiler`
-[@Yu2012-jj],`limma` [@Ritchie2015-oz] and `GOseq` [@Young2010-iw].
+On the other hand, FCS typically involves installing specific softwares and dealing with a
+dataset representing every detected gene (~20,000 rows) and ensuring that the file
+format from upstream tools can be compatible with FCS packages.
 
@@
 
-The focus of this work is to raise awareness to two existing problems in ORA as implemented
-by some popular enrichment tools.
+The focus of this work is to raise awareness to two existing problems in some implementations
+of ORA used by popular enrichment tools.
 The first problem is that genes belonging to the background list are removed from the analysis
@@
 the foreground, despite a large number of those genes being present in the background list.
-This leads to FDR values appearing lower than they should.
+This artificially reduces the number of tests conducted and makes FDR values appear lower
+than they should, potentially raising the rate of false positives.
 We expect this problem to have a more severe effect when the foreground list is fairly small.
 
-Here, we show the effect of these two problems on real RNA-seq based enrichment analysis
-results and demonstrate these two issues lead to poor performance using simulated expression
+Here, we define the effect of these two problems on real RNA-seq based enrichment analysis
+results and demonstrate how these two issues impact performance using simulated expression
 data.
@@
 
-## A systematic assessment of reproducibility in enrichment analysis
-
 ### Methods
 
+### Expression data preparation
+
+To test the effect of these two problems on RNA-seq based enrichment analysis, data from seven
+transcriptomic studies were downloaded from DEE2 using the getDEE2 R/Bioconductor package
+[@Ziemann2019-ux].
+Raw data for these seven studies are available from NCBI Sequence Read Archive under the
+accession numbers SRP128998, SRP038101, SRP037718, SRP096177, SRP097759, SRP253951 and
+SRP068733 [@Felisbino2021-hc; @Lund2014-ds; @Rafehi2014-rv; @Keating2014-ab;
+@Blanco-Melo2020-hh; @Rafehi2017-qn].
+For each dataset, kallisto counts were aggregated to the gene level.
+Genes with fewer than 10 reads per sample on average were removed from downstream analysis.
+Remaining genes passing this selection were included in the background gene set.
+Differential expression analysis was conducted with DESeq2 v1.42.0 [@Love2014-wg], with
+genes identified by their Ensembl identifiers.
+Gene symbols were then fetched using biomaRt v2.58.0, based on Ensembl version 112 [@Durinck2009-dr].
+
+### Examining the background problem using real transcriptome data
+
+Human gene sets were obtained from MSigDB v2023.2 [@Liberzon2015-cs], and included
+KEGG Medicus, Reactome, Wikipathways, MicroRNA targets from miRdb, transcription factor
+targets from GTRD, Gene Ontology (GO), Human Phenotype Ontology (HPO), Cell markers and
+Hallmark pathways [REFS].
+For an initial demonstration of the background problem, differentially expressed genes with
+FDR<0.05 were subset, with separate lists for up and downregulated genes.
+These gene lists were subjected to ORA using the `enricher` function of clusterProfiler
+v4.10.0, using a minimum set size of 5 and a maximum set size of 500,000.
+
+For one test, a GO gene set called "Complement activation classical pathway" was added
+to the KEGG database to test whether fold enrichment and p-values were affected.
+
+To systematically assess the impact of the background problem, we conducted ORA with
+`enricher` using the different gene set libraries described above for the seven RNA-seq
+datasets and nine gene set libraries.
+`enricher` was used with the same parameters as above, and we devised a simple workaround
+to the background problem, which is to append the entire background list as a gene set to the
+library.
+This forces `enricher` to retain all background genes.
+The ORA analyses were then filtered for significant sets using a FDR threshold of 0.05.
+Jaccard index was used to compare original and corrected analyses.
+
+### Examining the FDR problem
+
 ### Results
+
+### Understanding the background problem
+
+In the analysis of dataset 1, the background list contained 14,144 genes and the foreground
+list consisted of 1,469 genes.
+After ORA with KEGG sets, the background size reported by clusterProfiler was only 2,120.
+With GO sets, the background was reported to be 12,365.
+These numbers correspond to the number of genes in the background list that are annotated with
+one or more functions in the respective gene set libraries.
+
+We used a workaround to eliminate the background problem and compared the results to standard
+analysis.
+This was repeated for seven datasets and the nine gene set libraries.
+The mean number of significant results was calculated for each of the gene set libraries
+before and after correcting the problem (**Figure 1A**).
+Each of the gene set libraries showed an uplift when correcting the background problem.
+The smallest increase was observed in Cellmarkers gene library (11.7%), while the largest
+increase was observed for HPO (229%).
+The similarity between results, as quantified with the Jaccard similarity was highest for
+Cellmarkers and GTRD at 0.88 and 0.76 respectively, and the lowest similarity was recorded for
+HPO and KEGG with 0.20 and 0.48 respectively.
+Observed Jaccard scores correlated with the number of genes annotated to one or more functional
+categories in the gene set library (**Figure 1B**).
+
+![**Figure 1A. Effect of correcting the background problem.** Figures represent the mean number
+of statistically significant sets over seven independent datasets detected with clusterProfiler
+with and without correcting the background problem.
+The percent increase in significant sets are shown.](../analysis/fig1_ORAsig.png "Figure 1A
+shows that correcting the background bug improves the number of significant sets between 12 and
+229%.")
+
+![**Figure 1B. Impact of background problem is worse for smaller gene set libraries.** Mean
+Jaccard similarity index was calculated for original and corrected ORA for seven independent
+datasets. The number of genes represented in one or more functional sets is shown on the x-axis,
+and the mean Jaccard index is shown on the y-axis.](../analysis/fig1_ORAjac.png "Figure 1B
+shows that the background bug impacts smaller gene sets libraries like KEGG and HPO more
+severely.")
+
+### Understanding the FDR problem
 
@@
 ## Availability of other materials
-
-* The full protocol can be found at protocols.io
-(https://www.protocols.io/view/a-recipe-for-extremely-reproducible-enrichment-ana-j8nlkwpdxl5r/v2) @Ziemann2023-kz.
-
-* Instructional video guides are available on YouTube
-(https://www.youtube.com/playlist?list=PLAAydBPtqFMXDpLa796q7f7W1HK4t_6Db).
 

### 2024-06-07 analysis/analyse_rnaseq2.Rmd
ms

@@
 
-par(mar=c(c(5.1, 7.1, 4.1, 2.1) ))
+par(mar=c(c(5.1, 7.1, 2.1, 2.1) ))
 barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,330),
-  xlab="no. significant sets", main="Effect of correcting the FDR bug",
+  xlab="no. significant sets",
   col=c("gray30","gray80"))
@@
   fill=c("gray80","gray30"), horiz=FALSE, cex=1)
+
+png("fig2_ORAfdr.png")
+par(mar=c(c(5.1, 7.1, 2.1, 2.1) ))
+barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,330),
+  xlab="no. significant sets",
+  col=c("gray30","gray80"))
+text(df$ns1+20,((1:nrow(df))*3)-1.5,labels=signif(df[,1],3))
+text(df$ns2+20,((1:nrow(df))*3)-0.5,labels=signif(df[,2],3))
+legend("bottomright", inset=.02, legend=c("corrected","original"),
+  fill=c("gray80","gray30"), horiz=FALSE, cex=1)
+dev.off()
 
@@
 points(jac[,"TFT GTRD"],type="b",col="gray50",pch=19,cex=1.5,lwd=2)
-mtext("impact of gene list length on FDR problem severity")
 
@@
        col=c("black", "red", "blue", "gray50"), lty=1, cex=1.2,lwd=3,pch=19)
+
+png("fig2_ORAsz.png")
+
+plot(jac[,"GO"],type="b", pch=19,cex=1.5,lwd=2,
+  xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected")
+axis(1, at = seq(1, 5, by = 1),
+     labels = ngenes)
+points(jac[,"Reactome"],type="b",col="red",pch=19,cex=1.5,lwd=2)
+points(jac[,"Cellmarkers"],type="b",col="blue",pch=19,cex=1.5,lwd=2)
+points(jac[,"TFT GTRD"],type="b",col="gray50",pch=19,cex=1.5,lwd=2)
+
+legend("bottomright", legend=c("GO","Reactome", "Cellmarkers", "TFT"),
+       col=c("black", "red", "blue", "gray50"), lty=1, cex=1.2,lwd=3,pch=19)
+
+dev.off()
 

### 2024-06-07 manuscript/manuscript.Rmd
ms

@@
 
+A workaround for the FDR problem was devised by filtering the gene sets by the presence of
+at least two genes in the background list prior to running `enricher`.
+After `enricher` using the parameters above, the number of gene sets in the results was
+quantified, and compared to the number that should have been reported using the two gene limit.
+The difference between these numbers (*n*) represents the number of missing gene sets.
+To account for these tests, *n* 1 values were appended to the p-values obtained by enricher,
+followed by FDR correction in R, to obtain the corrected FDR values.
+As above, analysis of seven datasets was done with nine gene set libraries, the significant
+sets were selected and the Jaccard statistics were collected for each run to compare original
+and corrected analysis.
+
+As we predicted that this effect could be more severe for smaller foreground lists, we performed
+parallel analysis with foreground gene lists with sizes varying between 125 and 2,000 genes,
+as ranked by FDR value.
+
 ### Results
@@
 ![**Figure 1A. Effect of correcting the background problem.** Figures represent the mean number
-of statistically significant sets over seven independent datasets detected with clusterProfiler
-with and without correcting the background problem.
+of statistically significant sets over seven independent datasets detected with clusterProfiler's
+`enricher` function with and without correcting the background problem.
 The percent increase in significant sets are shown.](../analysis/fig1_ORAsig.png "Figure 1A
@@
 
+A workaround for the FDR problem was developed and as above, and original and corrected
+analyses were conducted for the seven datasets and nine gene set libraries.
+The results obtained showed no major differences in the number of significant sets between
+original and corrected analyses (**Figure 2A**).
+
+We posited that the foreground gene list length may have exacerbate the problem, so we performed
+parallel analysis with foreground lists of length between 125 and 2,000 genes (**Figure 2B**).
+A clear trend was observed for GO, Reactome, Cellmarkers and TFT GTRD, with Jaccard similarity
+being significantly reduced when foreground gene lists were shorter than 500 genes.
+Interestingly, the drop in Jaccard similarity was more severe for GO and Reactome sets as
+compared to Cellmarkers and TFT GTRD sets, and may be associated with the size of the gene
+sets in the library.
+Median gene set size for these are GO:18, Reactome:11, TFT GTRD:287 and Cellmarkers:115.
+
+![**Figure 2A. Effect of correcting the FDR problem.** Figures represent the mean number of
+significant sets with and without correction of the FDR problem over seven independent RNA-seq
+datasets.](../analysis/fig2_ORAfdr.png "Figure 2A shows that correcting the FDR bug slightly
+reduces the number of significant sets.")
+
+![**Figure 2B. Impact of FDR problem is worse for shorter foreground gene lists.** Mean Jaccard
+index between original and corrected analysis were calculated for seven RNA-seq datasets and nine
+gene set libraries. Only GO, Reactome, Cellmarkers and TFT GTRD are shown.](
+../analysis/fig2_ORAsz.png "Figure 2B shows that the impact of the FDR bug is worse when
+foreground gene lists are shorter than 500.")
+
 ## Discussion
+
+In addition to the subtle problems with ORA, there are some major ones too:
+
+* Rarely used or reported properly (BG,FDR)
+
+* Results vary dramatically depending on the number of genes selected in the foreground
+
+* Less accurate as compared to FCS
 

### 2024-06-07 manuscript/manuscript.Rmd
ms updates

@@
 
+### Quantifying the effect of these problems on precision and recall using simulated expression data
+
+RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
+This dataset was selected due to its high number of assigned reads (367M).
+A library of 200 random gene sets were generated, each containing 30 genes.
+To generate differential expression profiles, 5% of gene sets were selected to be differentially
+expressed, with equal numbers of up- and down-regulated pathways.
+Pseudosamples were generated by first downsampling the counts using the `thincounts` function of
+the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the addition of some extra variation drawn from a
+normal distribution with mean of 1 and standard deviation varying between 0.3 and 0.45, this
+value selected randomly for each gene.
+Five "control" and five "case" pseudosamples were generated.
+For the genes selected to be differentially expressed, expression values were multiplied by a
+fold change in the case samples of 2.0 for upregulated genes and 0.5 for downregulated genes.
+Genes that were selected to be both up and downregulated were left unchanged.
+
+With these expression profiles, DESeq2 was used for differential analysis, and the results were
+subjected to `enricher` which suffers from both problems, `fora`, an ORA function of the `fgsea`
+bioconductor package v1.30.0 which does not suffer these problems, and `fgsea` a fast
+implementation of the preranked GSEA algorithm [@Korotkevich2016-gd].
+For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
+selected for the foreground gene set.
+If fewer than 250 genes are detected as significant, then 250 genes with the smallest p-values
+were used for ORA, similar to a previous report [@Tarca2013-xh].
+For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
+
+After enrichment analysis with these approaches, the significant sets up and down regulated sets
+were selected and compared to the ground truth in order to calculate precision, recall and F1
+score.
+
+At each value of added variance, 500 replications were conducted, each with as set seed for
+reproducibility.
+
 ### Results

### 2024-06-07 analysis/sim/simpw_func.R
sim update minimum 200 genes for ORA

@@ -170,12 +170,13 @@ deseq2<-function(x) {
   dds <- DESeqDataSetFromMatrix(countData = y, colData = samplesheet, design = ~ trt )
   res <- DESeq(dds)
   z <- DESeq2::results(res)
-  vsd <- vst(dds, blind=FALSE)
-  zz <- cbind(z,assay(vsd))
-  x[[6]] <- as.data.frame(zz[order(zz$padj),])
-  sig <- subset(zz,padj<0.05)
-  x[[7]] <- rownames(sig[which(sig$log2FoldChange>0),])
-  x[[8]] <- rownames(sig[which(sig$log2FoldChange<0),])
+  x[[6]] <- as.data.frame(z[order(z$pvalue),])
+  up <- rownames(subset(z,log2FoldChange>0 & padj<0.05))
+  if (length(up)<250) { up <- rownames(head(subset(z,log2FoldChange>0 ),250)) }
+  dn <- rownames(subset(z,log2FoldChange<0 & padj<0.05))
+  if (length(dn)<250) { dn <- rownames(head(subset(z,log2FoldChange<0 ),250)) }
+  x[[7]] <- up
+  x[[8]] <- dn
   x
 }
 
@@ -231,150 +232,6 @@ run_mitch <- function(y,DGE_FUNC,gsets, N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIM
   y
 }
 
-##################################
-# hypergeometric test function (limited to deseq2)
-##################################
-run_hypergeometric<-function(x,DGE_FUNC,gsets,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS){
-
-dge<-sapply(x,"[",6)
-
-ups<-lapply(dge, function(x) { rownames(subset(x, padj<0.05 & log2FoldChange > 0)) } )
-dns<-lapply(dge, function(x) { rownames(subset(x, padj<0.05 & log2FoldChange < 0)) } )
-
-l_ups<-sapply(ups,length)
-l_dns<-sapply(dns,length)
-geneset_sizes<-sapply( gsets , length )
-
-# calculate number of genes in sets that are up and downregulated
-n_dns=n_ups=p_ups=p_dns=obs_up=obs_dn=NULL
-n_dns=n_ups=p_ups=p_dns=obs_up=obs_dn=list()
-
[…]

### 2024-06-07 analysis/sim/simpw_func.R
update

@@ -170,11 +170,12 @@ deseq2<-function(x) {
   dds <- DESeqDataSetFromMatrix(countData = y, colData = samplesheet, design = ~ trt )
   res <- DESeq(dds)
   z <- DESeq2::results(res)
-  x[[6]] <- as.data.frame(z[order(z$pvalue),])
+  z <- as.data.frame(z[order(z$pvalue),])
   up <- rownames(subset(z,log2FoldChange>0 & padj<0.05))
   if (length(up)<250) { up <- rownames(head(subset(z,log2FoldChange>0 ),250)) }
   dn <- rownames(subset(z,log2FoldChange<0 & padj<0.05))
   if (length(dn)<250) { dn <- rownames(head(subset(z,log2FoldChange<0 ),250)) }
+  x[[6]] <- z
   x[[7]] <- up
   x[[8]] <- dn
   x
@@ -243,8 +244,8 @@ gsets2 <- stack(gsets)[,c(2,1)]
 colnames(gsets2) <- c("term","gene")
 
 dge <- sapply(x,"[",6)
-up <- sapply(x,"[",7)
-dn <- sapply(x,"[",8)
+ups <- sapply(x,"[",7)
+dns <- sapply(x,"[",8)
 bgs <- lapply(dge, function(x) { rownames(x) } )
 
 l_ups <- sapply(ups,length)

### 2024-06-08 analysis/sim/run_simpw.Rmd
update
update sim

@@
 SIMS=500
+#run 500 sims
 FRAC_DE=0.05
 FC=1
-N_REPS=5
+N_REPS=3
 DGE_FUNC="deseq2"
 SUM_COUNT=2e7
-VARIANCE=c(0.3,0.325,0.35,0.375,0.4,0.425,0.45)
+VARIANCE=c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1,1.1,1.2,1.3,1.4,1.5)
 
@@
 
-## Plot
+## Plotfig1_ORAjac.png
 
@@
 
+par(mfrow=c(1,3))
+par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
+
+plot(cp$VARIANCE,cp$p,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="precision")
+points(fo$VARIANCE,fo$p,type="b",pch=19,col="red")
+points(fg$VARIANCE,fg$p,type="b",pch=19,col="blue")
+
+legend("bottomleft", inset=.02, title="tool",
+   c("clusterprofiler","fora","fgsea"), col=c("black","red","blue"),
+   horiz=FALSE, cex=1.2, pch=19,lwd=2)
+
+plot(cp$VARIANCE,cp$r,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="recall")
+points(fo$VARIANCE,fo$r,type="b",pch=19,col="red")
+points(fg$VARIANCE,fg$r,type="b",pch=19,col="blue")
+
+plot(cp$VARIANCE,cp$f,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="f1")
+points(fo$VARIANCE,fo$f,type="b",pch=19,col="red")
+points(fg$VARIANCE,fg$f,type="b",pch=19,col="blue")
+
+png("fig3_sim.png",width=720, height=480)
+par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
 par(mfrow=c(1,3))
@@
 
-par(mfrow=c(1,1))
+dev.off()
 

### 2024-06-08 analysis/sim/simpw_func.R

@@ -172,9 +172,9 @@ deseq2<-function(x) {
   z <- DESeq2::results(res)
   z <- as.data.frame(z[order(z$pvalue),])
   up <- rownames(subset(z,log2FoldChange>0 & padj<0.05))
-  if (length(up)<250) { up <- rownames(head(subset(z,log2FoldChange>0 ),250)) }
+  if (length(up)<200) { up <- rownames(head(subset(z,log2FoldChange>0 ),200)) }
   dn <- rownames(subset(z,log2FoldChange<0 & padj<0.05))
-  if (length(dn)<250) { dn <- rownames(head(subset(z,log2FoldChange<0 ),250)) }
+  if (length(dn)<200) { dn <- rownames(head(subset(z,log2FoldChange<0 ),200)) }
   x[[6]] <- z
   x[[7]] <- up
   x[[8]] <- dn

### 2024-06-08 analysis/analyse_rnaseq1.Rmd
ms work

@@
 
-  if ( length(defup)<250 ) {
-    defup <- head(rownames(subset(def,log2FoldChange>0)),250)
+  if ( length(defup)<200 ) {
+    defup <- head(rownames(subset(def,log2FoldChange>0)),200)
     defup <- defup[lapply(strsplit(defup," "),length)==2]
@@
 
-  if ( length(defdn)<250 ) {
-    defdn <- head(rownames(subset(def,log2FoldChange<0)),250)
+  if ( length(defdn)<200 ) {
+    defdn <- head(rownames(subset(def,log2FoldChange<0)),200)
     defdn <- defdn[lapply(strsplit(defdn," "),length)==2]

### 2024-06-08 manuscript/manuscript.Rmd
ms work

@@
 
-ORA involdes the selection of genes of interest, followed by a test to ascertain whether
+ORA involves the selection of genes of interest, followed by a test to ascertain whether
 certain functional categories are over-represented in the selected genes.
 Since it's initial development in 1999 [@Tavazoie1999-bu], ORA has proliferated widely,
-becoming part of many websites and software packages.
+becoming part of many bioinformatics websites and software packages.
 The most popular ORA website is DAVID [@Dennis2003-ar;@Huang2009-gi;@Sherman2022-jn],
 followed by Ingenuity Pathway Analysis, a commercial software package provided by QIAGEN Inc.
-ORA has been implemented into popular R/Bioconductor packages including `ClusterProfiler`
+ORA has been implemented into R/Bioconductor packages including `clusterProfiler`
 [@Yu2012-jj;@Wu2021-wy], `limma` [@Ritchie2015-oz] and `GOseq` [@Young2010-iw].
@@
 dataset representing every detected gene (~20,000 rows) and ensuring that the file
-format from upstream tools can be compatible with FCS packages.
+format from upstream tools is compatible with FCS packages.
 
@@
 list of selected genes (foreground list) [@Tilford2009-vx].
-Failure to use a background gene list can lead to dramatic changes in enrichment results,
+Failure to use a background gene list leads to dramatic changes in enrichment results,
 rendering them unreliable [@Timmons2015-ki].
@@
 (~4%) of peer-reviewed articles describing enrichment analysis [@Wijesooriya2022-fw].
-In addition, correction of p-values for multiple testing is crucial in controlling the
-false positive rate, but this is reported in only ~54% of studies [@Wijesooriya2022-fw].
-These discoveries have resulted in the development of best practices for end users
-to improve awareness of the issues and help enforce reporting standards [@Zhao2023-mx].
+Correction of p-values for multiple testing is crucial in controlling the false positive rate
+as enrichment analysis typically involves throusands of parallel tests.
+But appropriate p-value correction is reported in only ~54% of studies [@Wijesooriya2022-fw].
+These issues have resulted in the development of best practices for end users and
+stronger reporting standards [@Zhao2023-mx].
 
-The focus of this work is to raise awareness to two existing problems in some implementations
-of ORA used by popular enrichment tools.
+The focus of this work is to raise awareness to two existing problems in ORA implementations
+of some popular enrichment tools.
 The first problem is that genes belonging to the background list are removed from the analysis
@@
 libraries like Gene Ontology [@The_Gene_Ontology_Consortium2023-xp] which describes 19,428
-genes (figures obtained from MSigDB [@Liberzon2015-cs]).
+genes (figures obtained from MSigDB June 2024 [@Liberzon2015-cs]).
 
 The second problem is that adjusting p-values for multiple testing, also known as
-false-discovery correction is implemented improperly.
-ORA tools should implement a threshold by which a pathway could be considered as detected.
+false-discovery rate correction is implemented improperly.
+Determining which pathways should be subject to false discovery rate correction depends on
+how a detection threshold is set.
 Ideally, a pathway should be considered detected based on the presence of a predetermined
@@
 @Blanco-Melo2020-hh; @Rafehi2017-qn].
-For each dataset, kallisto counts were aggregated to the gene level.
+For each dataset, kallisto (v0.43.1) [@Bray2016-lu] transcript counts were aggregated to the
+gene level.
 Genes with fewer than 10 reads per sample on average were removed from downstream analysis.
@@
 Human gene sets were obtained from MSigDB v2023.2 [@Liberzon2015-cs], and included
-KEGG Medicus, Reactome, Wikipathways, MicroRNA targets from miRdb, transcription factor
-targets from GTRD, Gene Ontology (GO), Human Phenotype Ontology (HPO), Cell markers and
-Hallmark pathways [REFS].
+KEGG Medicus [@Kanehisa2023-ni],
+Reactome [@Milacic2024-iv],
+Wikipathways [@Agrawal2024-lr],
+MicroRNA targets from miRdb [@Chen2020-wo],
+transcription factor targets from GTRD [@Kolmykov2021-mn],
+Gene Ontology (GO) [@The_Gene_Ontology_Consortium2023-xp],
+Human Phenotype Ontology (HPO) [@Kohler2021-ha],
+Cell markers and
+Hallmark pathways [@Liberzon2015-cs].
 For an initial demonstration of the background problem, differentially expressed genes with
 FDR<0.05 were subset, with separate lists for up and downregulated genes.
+If fewer than 200 genes met this criteria, then the 200 genes with the smallest p-values
+were selected.
 These gene lists were subjected to ORA using the `enricher` function of clusterProfiler
-v4.10.0, using a minimum set size of 5 and a maximum set size of 500,000.
+v4.10.0, using a minimum set size of 5 and a no maximum set size.
 
@@
 `enricher` using the different gene set libraries described above for the seven RNA-seq
-datasets and nine gene set libraries.
+datasets.
 `enricher` was used with the same parameters as above, and we devised a simple workaround
@@
 
+For this test, 2000 of the top up and down regulated genes were selected from each
+differential expression dataset.
 A workaround for the FDR problem was devised by filtering the gene sets by the presence of
@@
 After `enricher` using the parameters above, the number of gene sets in the results was
-quantified, and compared to the number that should have been reported using the two gene limit.
+quantified, and compared to the number that should have been reported using the two gene
+limit.
 The difference between these numbers (*n*) represents the number of missing gene sets.
@@
 selected for the foreground gene set.
-If fewer than 250 genes are detected as significant, then 250 genes with the smallest p-values
+If fewer than 200 genes are detected as significant, then 200 genes with the smallest p-values
 were used for ORA, similar to a previous report [@Tarca2013-xh].
@@
 
+### Impact of these two problems on accuracy determined with simulated data
+
+![**Figure 3. Comparison of clusterProfiler, fora and fgsea accuracy with simulated data.**
+Precision, recall and F1 score are shown with the addition of different amounts of random
+noise to expression levels.](../analysis/sim/fig3_sim.png "Figure 3 shows that fora is more
+precise than clusterProfiler, and fgsea has higher recall than either ORA methods.")
+
 ## Discussion

### 2024-06-08 analysis/sim/run_simpw.Rmd
more sims

@@
 
-SIMS=500
+SIMS=1000
 #run 500 sims
@@
 points(fg$VARIANCE,fg$p,type="b",pch=19,col="blue")
-
+grid()
 legend("bottomleft", inset=.02, title="tool",
@@
 points(fg$VARIANCE,fg$r,type="b",pch=19,col="blue")
-
+grid()
 plot(cp$VARIANCE,cp$f,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="f1")
@@
 points(fg$VARIANCE,fg$f,type="b",pch=19,col="blue")
-
+grid()
 dev.off()

### 2024-06-09 analysis/sim/simpw_func.R
update sim

@@ -87,10 +87,12 @@ if (NDIF>0) {
 
   # select the ones that are also in the profile
   UP_DE<-UP_DE[which(UP_DE %in% row.names(df))]
+  #UP_DE<-sample(UP_DE,length(UP_DE)/2)
 
   # same for down genes
   DN_DE<-unique(unlist(unname(DN_LIST)))
   DN_DE<-DN_DE[which(DN_DE %in% row.names(df))]
+  #DN_DE<-sample(DN_DE,length(DN_DE)/2)
 
   ITX<-intersect(UP_DE,DN_DE)
   # need to eliminate the overlapping ones for simplicity

### 2024-06-09 manuscript/manuscript.Rmd
ns

@@
 The difference between these numbers (*n*) represents the number of missing gene sets.
-To account for these tests, *n* 1 values were appended to the p-values obtained by enricher,
-followed by FDR correction in R, to obtain the corrected FDR values.
+To account for these tests, *n* values of 1 were appended to the p-values obtained by
+enricher, followed by FDR correction in R, to obtain the properly corrected FDR values.
 As above, analysis of seven datasets was done with nine gene set libraries, the significant
@@
 
-As we predicted that this effect could be more severe for smaller foreground lists, we performed
-parallel analysis with foreground gene lists with sizes varying between 125 and 2,000 genes,
-as ranked by FDR value.
+As we predicted that this effect could be more severe for smaller foreground lists, we
+performed parallel analysis with foreground gene lists with sizes varying between 125 and
+2,000 genes, as ranked by FDR value.
 
@@
 A library of 200 random gene sets were generated, each containing 30 genes.
-To generate differential expression profiles, 5% of gene sets were selected to be differentially
-expressed, with equal numbers of up- and down-regulated pathways.
-Pseudosamples were generated by first downsampling the counts using the `thincounts` function of
-the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the addition of some extra variation drawn from a
-normal distribution with mean of 1 and standard deviation varying between 0.3 and 0.45, this
-value selected randomly for each gene.
-Five "control" and five "case" pseudosamples were generated.
+To generate differential expression profiles, 5% of gene sets were selected to be
+differentially expressed, with equal numbers of up- and down-regulated pathways.
+Pseudosamples were generated by first downsampling the counts using the `thincounts`
+function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
+addition of some extra variation drawn from a normal distribution with mean of 1 and
+standard deviation varying between 0.1 and 1.0, this value selected randomly for each gene.
+Three "control" and five "case" pseudosamples were generated.
 For the genes selected to be differentially expressed, expression values were multiplied by a
@@
 Genes that were selected to be both up and downregulated were left unchanged.
-
-With these expression profiles, DESeq2 was used for differential analysis, and the results were
-subjected to `enricher` which suffers from both problems, `fora`, an ORA function of the `fgsea`
-bioconductor package v1.30.0 which does not suffer these problems, and `fgsea` a fast
-implementation of the preranked GSEA algorithm [@Korotkevich2016-gd].
+With these expression profiles, DESeq2 was used for differential analysis, and the results
+were subjected to `enricher` which suffers from both problems, `fora`, an ORA function of
+the `fgsea` bioconductor package v1.30.0 which does not suffer these problems, and `fgsea` a
+fast implementation of the preranked GSEA algorithm [@Korotkevich2016-gd].
 For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
@@
 For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
-
-After enrichment analysis with these approaches, the significant sets up and down regulated sets
-were selected and compared to the ground truth in order to calculate precision, recall and F1
-score.
-
-At each value of added variance, 500 replications were conducted, each with as set seed for
-reproducibility.
+After enrichment analysis with these approaches, the significant sets up and down regulated
+sets were selected and compared to the ground truth in order to calculate precision, recall
+and F1 score.
+At each value of added variance, 1000 replications were conducted.
+No set seed was used, but results were similar when run on different computer systems.
 
@@
 
-In the analysis of dataset 1, the background list contained 14,144 genes and the foreground
-list consisted of 1,469 genes.
-After ORA with KEGG sets, the background size reported by clusterProfiler was only 2,120.
-With GO sets, the background was reported to be 12,365.
-These numbers correspond to the number of genes in the background list that are annotated with
-one or more functions in the respective gene set libraries.
-
-We used a workaround to eliminate the background problem and compared the results to standard
-analysis.
-This was repeated for seven datasets and the nine gene set libraries.
+A workaround to eliminate the background problem was developed and we used it to compare the
+results of ORA of seven datasets with the original and corrected algorithm for nine different
+gene set libraries.
 The mean number of significant results was calculated for each of the gene set libraries

### 2024-06-10 manuscript/manuscript.Rmd
ms

@@
 selected for the foreground gene set.
-If fewer than 200 genes are detected as significant, then 200 genes with the smallest p-values
+If fewer than 200 genes were detected as significant, then 200 genes with the smallest p-values
 were used for ORA, similar to a previous report [@Tarca2013-xh].
 For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
-After enrichment analysis with these approaches, the significant sets up and down regulated
+After enrichment analysis with these approaches, the significant up and down regulated
 sets were selected and compared to the ground truth in order to calculate precision, recall
 and F1 score.
-At each value of added variance, 1000 replications were conducted.
+At each value of added variance, 500 replications were conducted.
 No set seed was used, but results were similar when run on different computer systems.
 
-### Results
+## Results
 
@@
 
+Simulated differential expression profiles with apriori changes to predetermined pathway genes were
+analysed in parallel by clusterProfilers enricher function, which suffers from the two aforementioned
+problems and fora which does not.
+In parallel, fgsea was used to compare the performance of these two ORA methods against a prototypical
+FCS method.
+ClusterProfiler scored very high precision when added variance was less than 0.4, but showed a
+progressive worsening in precision as more noise was added to the dataset, dropping to 0.68 with added
+variance of 1.0.
+In comparison, the precision of fora was relatively stable around 0.97 until added variance reached 0.9.
+The recall of ORA methods dropped suddenly when added variance was 0.5 or larger.
+fgsea showed stable precision of 0.94 across all levels of added variance, and superior recall overall.
+
 ![**Figure 3. Comparison of clusterProfiler, fora and fgsea accuracy with simulated data.**
@@
 ## Discussion
+
+Here we have described the effect of two subtle problems with common implementations of ORA.
+This work was inspired by a doctoral student new to bioinformatics who was puzzled by confusing results
+obtained from clusterProfiler when using a custom set of pathways.
+Our hunch that clusterProfiler was behaving in an undesired way was confirmed by several posts to
+online bioinformatics forums dating to 2020 where confused users were wondering why the number of
+background genes changed so much.
+ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated genes from the
+background `options(enrichment_force_universe = TRUE)`, however this feature isn't yet described in the
+official documentation.
+
+The second problem is potentially more serious in that it could lead to lots of false positives,
+as clusterProfiler underestimates the actual number of tests that were conducted.
+As shown in Figure 2B, this affects results differently depending on the gene set library used and also
+the number of genes selected in the foreground list.
+
+ClusterProfiler is not the only tool to suffer from these problems, they are widespread (**Table 1**).
+
+| Tool | Has problems | Citation |
+| --- | --- | --- |
+| ClusterProfiler | Background and FDR | [@Yu2012-jj;@Wu2021-wy] |
+| DAVID | Background and FDR | [@Dennis2003-ar;@Huang2009-gi;@Sherman2022-jn] |
+| fora (fgsea) | Neither | [@Korotkevich2016-gd] |
+
+: Table 1. Classification of several popular ORA tools for Background and FDR problems.
 
@@
 * Less accurate as compared to FCS
+
+The reason why such problems proliferated for so long is that using a custom background for ORA seems
+to be the exception, rather than the rule.
+Several tools do not even allow for the use of a custom background, suggesting that some tool developers
+were unaware of the seriousness of this error.
+This is one reason why ORA tool documentation also seems to lack any explanation of the importance of
+using a custom background set.
 

### 2024-06-10 analysis/sim/run_simpw.Rmd
refine sims
update sim
fig

@@
 SUM_COUNT=2e7
-VARIANCE=c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1,1.1,1.2,1.3,1.4,1.5)
+VARIANCE=c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)
 
@@
 
-png("fig3_sim.png",width=720, height=480)
+png("fig3_sim.png", width=7,height=5,units="in",res=150,pointsize=12)
 par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))

### 2024-06-10 manuscript/manuscript.Rmd
ms

@@
 conducting ORA as compared to FCS.
-A minimal ORA might involve pasting a single list of genes into a text box on a website,
+A minimal ORA might involve pasting a list of gene symbols into a text box on a website,
 where the results appear nearly instantaneously.
@@
 (~4%) of peer-reviewed articles describing enrichment analysis [@Wijesooriya2022-fw].
-Correction of p-values for multiple testing is crucial in controlling the false positive rate
+Correction of p-values for multiple testing is also crucial in controlling the false positive rate
 as enrichment analysis typically involves throusands of parallel tests.
@@
 than they should, potentially raising the rate of false positives.
-We expect this problem to have a more severe effect when the foreground list is fairly small.
+We expect this problem to have a more severe effect when the foreground list is small.
 
@@
 Hallmark pathways [@Liberzon2015-cs].
-For an initial demonstration of the background problem, differentially expressed genes with
+To demonstrate the background problem, differentially expressed genes with
 FDR<0.05 were subset, with separate lists for up and downregulated genes.
 If fewer than 200 genes met this criteria, then the 200 genes with the smallest p-values
-were selected.
+were selected, as suggested by a previous report [@Tarca2013-xh].
 These gene lists were subjected to ORA using the `enricher` function of clusterProfiler
 v4.10.0, using a minimum set size of 5 and a no maximum set size.
-
-For one test, a GO gene set called "Complement activation classical pathway" was added
-to the KEGG database to test whether fold enrichment and p-values were affected.
 
@@
 For this test, 2000 of the top up and down regulated genes were selected from each
-differential expression dataset.
+differential expression dataset based on p-value.
 A workaround for the FDR problem was devised by filtering the gene sets by the presence of

### 2024-06-11 manuscript/manuscript.Rmd
mswork

@@
 addition of some extra variation drawn from a normal distribution with mean of 1 and
-standard deviation varying between 0.1 and 1.0, this value selected randomly for each gene.
+standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
 Three "control" and five "case" pseudosamples were generated.
 For the genes selected to be differentially expressed, expression values were multiplied by a
-fold change in the case samples of 2.0 for upregulated genes and 0.5 for downregulated genes.
+log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated genes.
 Genes that were selected to be both up and downregulated were left unchanged.
@@
 and F1 score.
-At each value of added variance, 500 replications were conducted.
+At each value of added variance, 1000 replications were conducted.
 No set seed was used, but results were similar when run on different computer systems.
@@
 A workaround to eliminate the background problem was developed and we used it to compare the
-results of ORA of seven datasets with the original and corrected algorithm for nine different
+results of ORA of seven datasets with the original and corrected methods for nine different
 gene set libraries.
@@
 ![**Figure 1A. Effect of correcting the background problem.** Figures represent the mean number
-of statistically significant sets over seven independent datasets detected with clusterProfiler's
-`enricher` function with and without correcting the background problem.
+of statistically significant sets (FDR<0.05) over seven independent datasets detected with
+clusterProfiler's `enricher` function with and without correcting the background problem.
+Separate tests for up- and down-regulated gene lists were conducted.
 The percent increase in significant sets are shown.](../analysis/fig1_ORAsig.png "Figure 1A
@@
 
-A workaround for the FDR problem was developed and as above, and original and corrected
+A workaround for the FDR problem was developed, then original and corrected
 analyses were conducted for the seven datasets and nine gene set libraries.
-The results obtained showed no major differences in the number of significant sets between
+Results showed no major differences in the number of significant sets between
 original and corrected analyses (**Figure 2A**).
 
-We posited that the foreground gene list length may have exacerbate the problem, so we performed
+We posited that the foreground gene list length might exacerbate the problem, so we performed
 parallel analysis with foreground lists of length between 125 and 2,000 genes (**Figure 2B**).
@@
 significant sets with and without correction of the FDR problem over seven independent RNA-seq
-datasets.](../analysis/fig2_ORAfdr.png "Figure 2A shows that correcting the FDR bug slightly
-reduces the number of significant sets.")
+datasets. Foreground list consisted of 2000 genes with the smallest p-values](../analysis/fig2_ORAfdr.png
+"Figure 2A shows that correcting the FDR bug slightly reduces the number of significant sets.")
 
@@
 Simulated differential expression profiles with apriori changes to predetermined pathway genes were
-analysed in parallel by clusterProfilers enricher function, which suffers from the two aforementioned
-problems and fora which does not.
-In parallel, fgsea was used to compare the performance of these two ORA methods against a prototypical
-FCS method.
+analysed in parallel by clusterProfiler's `enricher` function, which suffers from the two
+aforementioned problems and `fora` which does not.
+In parallel, `fgsea` was used to compare the performance of these two ORA methods against a
+prototypical FCS method.
 ClusterProfiler scored very high precision when added variance was less than 0.4, but showed a
-progressive worsening in precision as more noise was added to the dataset, dropping to 0.68 with added
-variance of 1.0.
-In comparison, the precision of fora was relatively stable around 0.97 until added variance reached 0.9.
-The recall of ORA methods dropped suddenly when added variance was 0.5 or larger.
-fgsea showed stable precision of 0.94 across all levels of added variance, and superior recall overall.
+progressive worsening in precision as more noise was added to the dataset, dropping to 0.61 with added
+variance of 0.6.
+In comparison, the precision of fora was relatively stable around 0.97 until added variance reached 0.5.
+The recall of ORA methods dropped suddenly when added variance was 0.2 or larger.
+fgsea showed stable precision of 0.94 when added variance was 0.4 or lower, and dropped slightly to
+0.91 with added variance of 0.6.
+fgsea recall was better than either ORA tool across all levels of added variance.
+This result shows that these two subtle problems result in poorer precision for clusterProfiler,
+which becomes more severe with higher levels of noise.
+Overall, fgsea had similar precision to fora (mean=0.94), but the higher recall of fgsea meant it had
+a higher overall accuracy measured by F1 score (0.83) as compared to fora and clusterProfiler
+(0.61 and 0.64 respectively).
 
@@
 obtained from clusterProfiler when using a custom set of pathways.
-Our hunch that clusterProfiler was behaving in an undesired way was confirmed by several posts to
-online bioinformatics forums dating to 2020 where confused users were wondering why the number of
-background genes changed so much.
-ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated genes from the
-background `options(enrichment_force_universe = TRUE)`, however this feature isn't yet described in the
-official documentation.
+Our hunch that clusterProfiler was behaving in an undesired way was confirmed by the discovery of
+several posts to online bioinformatics forums dating to ~2020 where confused users were wondering why
+the number of background genes changed so much.
+ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated genes from
+the background `options(enrichment_force_universe = TRUE)`, however this feature isn't yet described
+in the official documentation.
 
-The second problem is potentially more serious in that it could lead to lots of false positives,
-as clusterProfiler underestimates the actual number of tests that were conducted.
-As shown in Figure 2B, this affects results differently depending on the gene set library used and also
-the number of genes selected in the foreground list.
-
+The second problem is potentially more serious in that it could lead large numbers false positives
+under specific conditions; when the foreground list is short and the median pathway size is relatively
+small.
+These false positives arise due to an underestimate of the actual number of tests conducted.
 ClusterProfiler is not the only tool to suffer from these problems, they are widespread (**Table 1**).
@@
 
-In addition to the subtle problems with ORA, there are some major ones too:
+Although the purpose of this work was to describe two subtle problems with ORA, it has some major
+drawbacks which limit its utility.
 
-* Rarely used or reported properly (BG,FDR)
+* The method is rarely reported comprehensively, especially crucial details about background gene list
+and FDR correction.
 
-* Results vary dramatically depending on the number of genes selected in the foreground
+* Results vary dramatically depending on the number of genes selected in the foreground.
 
-* Less accurate as compared to FCS
+* Overall ORA is less accurate as compared to FCS
 
-The reason why such problems proliferated for so long is that using a custom background for ORA seems
-to be the exception, rather than the rule.
-Several tools do not even allow for the use of a custom background, suggesting that some tool developers
-were unaware of the seriousness of this error.
-This is one reason why ORA tool documentation also seems to lack any explanation of the importance of
-using a custom background set.
+These bigger problems suggest FCS should replace ORA in most cases.
+Where it makes sense to use ORA would be when FCS is not applicable, for example when investigating
+the over-represented categories in gene expression modules defined by packages like WGCNA (Weighted
+Gene Correlation Network Analysis)[@Langfelder2008-hu].
 

### 2024-06-13 manuscript/manuscript.Rmd
ms

@@
 
-TODO.
+Over-representation analysis (ORA) is used widely to asses for the enrichment of functional
+categories in a gene list compared to a background list.
+ORA is therefore a critical method in the interpretation of 'omics data, relating gene lists
+to biological functions and themes.
+Although ORA is hugely popular, we and others have noticed two potentially undesired
+behaviours of some ORA tools.
+The first one we call the "background problem", because it involves the software eliminating
+large numbers of genes from the background list if they are not annotated as belonging to any
+category.
+The second one we call the "false discovery rate problem", because some tools underestimate
+the true number of parallel tests conducted.
+Here we demonstrate the impact of these issues on several real RNA-seq datasets and use
+simulated RNA-seq data to quantify the impact of these problems.
+We show that the severity of these problems depends on the gene set library, the number of
+genes in the list, and the degree of noise in the dataset.
+These problem can be mitigated by changing packages/websites for ORA or by changing to another
+approach such as functional class scoring.
 
@@
 
-### Methods
+## Methods
 
@@
 For the genes selected to be differentially expressed, expression values were multiplied by a
-log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated genes.
+log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
+genes.
 Genes that were selected to be both up and downregulated were left unchanged.
@@
 selected for the foreground gene set.
-If fewer than 200 genes were detected as significant, then 200 genes with the smallest p-values
-were used for ORA, similar to a previous report [@Tarca2013-xh].
+If fewer than 200 genes were detected as significant, then 200 genes with the smallest
+p-values were used for ORA, similar to a previous report [@Tarca2013-xh].
 For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
@@
 categories in the gene set library (**Figure 1B**).
+This result suggests that ORA with smaller gene set libraries like HPO and KEGG are impacted
+more severely.
 
@@
 Median gene set size for these are GO:18, Reactome:11, TFT GTRD:287 and Cellmarkers:115.
+These results demonstrate that the FDR problem is more severe when dealing with shorter
+foreground lists, and this is exacerbated when functional categories in the gene set library
+are also small.
 
@@
 significant sets with and without correction of the FDR problem over seven independent RNA-seq
-datasets. Foreground list consisted of 2000 genes with the smallest p-values](../analysis/fig2_ORAfdr.png
-"Figure 2A shows that correcting the FDR bug slightly reduces the number of significant sets.")
+datasets.
+Foreground list consisted of 2000 genes with the smallest p-values](../analysis/fig2_ORAfdr.png "Figure 2A shows that correcting the FDR bug slightly reduces the number of significant sets.")
 
@@
 
-Simulated differential expression profiles with apriori changes to predetermined pathway genes were
-analysed in parallel by clusterProfiler's `enricher` function, which suffers from the two
-aforementioned problems and `fora` which does not.
+Simulated differential expression profiles with *a priori* changes to predetermined pathway
+genes were analysed in parallel by clusterProfiler's `enricher` function, which suffers from
+the two aforementioned problems and `fora` which does not.
 In parallel, `fgsea` was used to compare the performance of these two ORA methods against a
 prototypical FCS method.
-ClusterProfiler scored very high precision when added variance was less than 0.4, but showed a
-progressive worsening in precision as more noise was added to the dataset, dropping to 0.61 with added
-variance of 0.6.
-In comparison, the precision of fora was relatively stable around 0.97 until added variance reached 0.5.
-The recall of ORA methods dropped suddenly when added variance was 0.2 or larger.
-fgsea showed stable precision of 0.94 when added variance was 0.4 or lower, and dropped slightly to
-0.91 with added variance of 0.6.
+ClusterProfiler scored very high precision when added variance was less than 0.2, but showed a
+progressive worsening in precision as more noise was added to the dataset, dropping to 0.69
+with added variance of 0.5.
+In comparison, the precision of fora was relatively stable around 0.97 until added variance
+reached 0.5, where it dropped to 0.88.
+The recall of ORA methods dropped suddenly when added variance was 0.2 or greater.
+fgsea showed stable precision of 0.93 or greater when added variance was 0.5 or lower, which
+dropped slightly to 0.91 with added variance of 0.6.
 fgsea recall was better than either ORA tool across all levels of added variance.
-This result shows that these two subtle problems result in poorer precision for clusterProfiler,
-which becomes more severe with higher levels of noise.
-Overall, fgsea had similar precision to fora (mean=0.94), but the higher recall of fgsea meant it had
-a higher overall accuracy measured by F1 score (0.83) as compared to fora and clusterProfiler
-(0.61 and 0.64 respectively).
+This result shows that these two subtle problems result in poorer precision for
+clusterProfiler, which becomes more severe with higher levels of noise.
+Overall, fgsea had similar mean precision to fora (0.94), but the higher recall of fgsea
+meant it had a higher overall accuracy measured by F1 score (0.83) as compared to fora and
+clusterProfiler (0.61 and 0.64 respectively).
 
@@
 
-Here we have described the effect of two subtle problems with common implementations of ORA.
-This work was inspired by a doctoral student new to bioinformatics who was puzzled by confusing results
-obtained from clusterProfiler when using a custom set of pathways.
-Our hunch that clusterProfiler was behaving in an undesired way was confirmed by the discovery of
-several posts to online bioinformatics forums dating to ~2020 where confused users were wondering why
-the number of background genes changed so much.
-ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated genes from
-the background `options(enrichment_force_universe = TRUE)`, however this feature isn't yet described
-in the official documentation.
+The background problem came to our attention when a doctoral student in our department
+was puzzled by results obtained from clusterProfiler when using a custom set of pathways.
+This exact issue was the subject of several posts to online bioinformatics forums dating to
+~2020 where confused users were wondering why the number of background genes changed so much.
+The results shown here show that the background problem effects smaller gene set libraries
+like KEGG and HPC more severely as compared to Cellmarkers and GO libraries which describe
+a larger proportion of all genes.
+ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated
+genes from the background `options(enrichment_force_universe = TRUE)`, however this feature
+isn't yet described in the official documentation.
 
-The second problem is potentially more serious in that it could lead large numbers false positives
-under specific conditions; when the foreground list is short and the median pathway size is relatively
-small.
+The second problem is potentially more serious in that it could lead to increasing false
+positives under specific conditions; when the foreground list is short and the median pathway
+size is relatively small.
 These false positives arise due to an underestimate of the actual number of tests conducted.
-ClusterProfiler is not the only tool to suffer from these problems, they are widespread (**Table 1**).
+In extreme cases, correction of this problem can lead to around half of the significant
+results changing.
+
+ClusterProfiler is not the only tool to suffer from these problems, they are widespread
+(**Table 1**), but there are also several tools free of these problems.
 
@@
 
-Although the purpose of this work was to describe two subtle problems with ORA, it has some major
-drawbacks which limit its utility.
-
-* The method is rarely reported comprehensively, especially crucial details about background gene list
-and FDR correction.
-
-* Results vary dramatically depending on the number of genes selected in the foreground.
-
-* Overall ORA is less accurate as compared to FCS
-
-These bigger problems suggest FCS should replace ORA in most cases.
-Where it makes sense to use ORA would be when FCS is not applicable, for example when investigating
-the over-represented categories in gene expression modules defined by packages like WGCNA (Weighted
-Gene Correlation Network Analysis)[@Langfelder2008-hu].
+Although the purpose of this work was to describe two subtle problems with ORA, this work
+highlights some major drawbacks.
+Firstly, ORA was less accurate as compared to FCS in this simulation work, confirming a
+previous report [@Kaspi2020-rn].
+Secondly ORA results can vary dramatically depending on the number of genes selected in the
+foreground, a mostly arbitrary choice.
+These issues together with previously mentioned pitfalls such as lack of background
+correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
+that ORA should be discouraged in most situations, and replaced by other methods such as FCS.
+Where it makes sense to use ORA would be when FCS is not applicable, for example when
+investigating the over-represented categories in gene expression modules defined by packages
+like WGCNA (Weighted Gene Correlation Network Analysis)[@Langfelder2008-hu], or when
+investigating over-represented functional categories among genes with variants associated
+with a disease.
 

### 2024-06-13 manuscript/manuscript.Rmd
ms

@@
 A minimal ORA might involve pasting a list of gene symbols into a text box on a website,
-where the results appear nearly instantaneously.
+with results appearing nearly instantly.
 On the other hand, FCS typically involves installing specific softwares and dealing with a
@@
 Failure to use a background gene list leads to dramatic changes in enrichment results,
-rendering them unreliable [@Timmons2015-ki].
+rendering them unreliable [@Timmons2015-ki;@Wijesooriya2022-fw].
 Unfortunately, use of an appropriate background list is reported in only a small fraction
 (~4%) of peer-reviewed articles describing enrichment analysis [@Wijesooriya2022-fw].
-Correction of p-values for multiple testing is also crucial in controlling the false positive rate
-as enrichment analysis typically involves throusands of parallel tests.
+Correction of p-values for multiple testing is also crucial in controlling the false positive
+rate as enrichment analysis typically involves throusands of parallel tests [@Tilford2009-vx].
 But appropriate p-value correction is reported in only ~54% of studies [@Wijesooriya2022-fw].

### 2024-06-13 analysis/sim/run_simpw.Rmd
fig3
sims
new sims

--- 2024-06-08 analysis/sim/run_simpw.Rmd
@@
 
-11. clusterprofiler_fixed up gene sets
+11. clusterprofiler bg fix up gene sets
 
-12. clusterprofiler_fixed down gene sets
+12. clusterprofiler bg fix down gene sets
 
-13. clusterprofiler_nobg up gene sets
+13. clusterprofiler FDR fix up gene sets
 
-14. clusterprofiler_nobg down gene sets
+14. clusterprofiler FDR fix down gene sets
 
-15. fora up gene sets
+15. clusterprofiler BG and FDR fix up gene sets
 
-16. fora down gene sets
+16. clusterprofiler BG and FDR fix down gene sets
 
-17. fgsea up gene sets
+17. fora up gene sets
 
-18. fgsea down gene sets
+18. fora down gene sets
 
-19. mitch up
+19. fgsea up gene sets
 
-20. mitch dn
+20. fgsea down gene sets
 
@@
 
-SIMS=500
+SIMS=1000
 #run 500 sims
 FRAC_DE=0.05
-FC=1
+FC=0.5
 N_REPS=3
@@
 SUM_COUNT=2e7
-VARIANCE=c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1,1.1,1.2,1.3,1.4,1.5)
+VARIANCE=c(0,0.1,0.2,0.3,0.4,0.5,0.6)
 
@@
 
-## Plotfig1_ORAjac.png
+## Plot
+
+fig1_ORAjac.png
 
@@
 
-cp <- subset(res,PWAY_FUNC == "clusterprofiler")
+cp <- subset(res,PWAY_FUNC == "clusterProfiler default")
+cpbg <- subset(res,PWAY_FUNC == "clusterProfiler BG fix")
+cpfdr <- subset(res,PWAY_FUNC == "clusterProfiler FDR fix")
+cpbgfdr <- subset(res,PWAY_FUNC == "clusterProfiler BG and FDR fix")
 fo <- subset(res,PWAY_FUNC == "fora")
@@
 
-cp %>% kbl(caption="clusterprofiler") %>% kable_paper("hover", full_width = F)
+cp %>% kbl(caption="clusterProfiler") %>% kable_paper("hover", full_width = F)
+cpbg %>% kbl(caption="clusterProfiler with background bug fix") %>% kable_paper("hover", full_width = F)
+cpfdr %>% kbl(caption="clusterProfiler with FDR bug fix") %>% kable_paper("hover", full_width = F)
+cpbgfdr %>% kbl(caption="clusterProfiler with background and FDR fixes") %>% kable_paper("hover", full_width = F)
 fo %>% kbl(caption="fora") %>% kable_paper("hover", full_width = F)
@@
 plot(cp$VARIANCE,cp$p,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="precision")
+points(cpbg$VARIANCE,cpbg$p,type="b",pch=19,col="orange")
+points(cpfdr$VARIANCE,cpfdr$p,type="b",pch=19,col="darkgreen")
+points(cpbgfdr$VARIANCE,cpbgfdr$p,type="b",pch=19,col="purple")
 points(fo$VARIANCE,fo$p,type="b",pch=19,col="red")
@@
 legend("bottomleft", inset=.02, title="tool",
-   c("clusterprofiler","fora","fgsea"), col=c("black","red","blue"),
-   horiz=FALSE, cex=1.2, pch=19,lwd=2)
+   c("CP default","CP BG fix","CP FDR fix","CP BG and FDR fix","fora","fgsea"),
+   col=c("black","orange","darkgreen","purple","red","blue"),horiz=FALSE, cex=1.2, pch=19,lwd=2)
 
 plot(cp$VARIANCE,cp$r,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="recall")
+points(cpbg$VARIANCE,cpbg$r,type="b",pch=19,col="orange")
+points(cpfdr$VARIANCE,cpfdr$r,type="b",pch=19,col="darkgreen")
+points(cpbgfdr$VARIANCE,cpbgfdr$r,type="b",pch=19,col="purple")
 points(fo$VARIANCE,fo$r,type="b",pch=19,col="red")
@@
 plot(cp$VARIANCE,cp$f,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="f1")
+points(cpbg$VARIANCE,cpbg$f,type="b",pch=19,col="orange")
+points(cpfdr$VARIANCE,cpfdr$f,type="b",pch=19,col="darkgreen")
+points(cpbgfdr$VARIANCE,cpbgfdr$f,type="b",pch=19,col="purple")
 points(fo$VARIANCE,fo$f,type="b",pch=19,col="red")
@@
 plot(cp$VARIANCE,cp$p,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="precision")
+points(cpbg$VARIANCE,cpbg$p,type="b",pch=19,col="orange")
+points(cpfdr$VARIANCE,cpfdr$p,type="b",pch=19,col="darkgreen")
+points(cpbgfdr$VARIANCE,cpbgfdr$p,type="b",pch=19,col="purple")
 points(fo$VARIANCE,fo$p,type="b",pch=19,col="red")
 points(fg$VARIANCE,fg$p,type="b",pch=19,col="blue")
-
+grid()
 legend("bottomleft", inset=.02, title="tool",
-   c("clusterprofiler","fora","fgsea"), col=c("black","red","blue"),
+   c("CP default","CP BG fix","CP FDR fix","CP BG and FDR fix","fora","fgsea"),
+   col=c("black","orange","darkgreen","purple","red","blue"),
    horiz=FALSE, cex=1.2, pch=19,lwd=2)
@@
 plot(cp$VARIANCE,cp$r,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="recall")
+points(cpbg$VARIANCE,cpbg$r,type="b",pch=19,col="orange")
+points(cpfdr$VARIANCE,cpfdr$r,type="b",pch=19,col="darkgreen")
+points(cpbgfdr$VARIANCE,cpbgfdr$r,type="b",pch=19,col="purple")
 points(fo$VARIANCE,fo$r,type="b",pch=19,col="red")
 points(fg$VARIANCE,fg$r,type="b",pch=19,col="blue")
+grid()
 
 plot(cp$VARIANCE,cp$f,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="f1")
+points(cpbg$VARIANCE,cpbg$f,type="b",pch=19,col="orange")
+points(cpfdr$VARIANCE,cpfdr$f,type="b",pch=19,col="darkgreen")
+points(cpbgfdr$VARIANCE,cpbgfdr$f,type="b",pch=19,col="purple")
 points(fo$VARIANCE,fo$f,type="b",pch=19,col="red")
 points(fg$VARIANCE,fg$f,type="b",pch=19,col="blue")
+grid()
+dev.off()
 
-dev.off()
+```
+
+Barplot is simpler.
+
+```{r,bars}
+
+prec <- c("CP default"=mean(cp$p),"CP BG fix"=mean(cpbg$p),"CP FDR fix"=mean(cpfdr$p),"CP BG and FDR fix"=mean(cpbgfdr$p),"fora"=mean(fo$p),"fgsea"=mean(fg$p))
+rec <- c("CP default"=mean(cp$r),"CP BG fix"=mean(cpbg$r),"CP FDR fix"=mean(cpfdr$r),"CP BG and FDR fix"=mean(cpbgfdr$r),"fora"=mean(fo$r),"fgsea"=mean(fg$r))
+f1 <- sapply(1:length(prec), function(i) { 2/(1/rec[i]+1/prec[i]) })
+
+par(mfrow=c(1,3))
+par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
+barplot(prec,ylim=c(0,1.1),las=2,main="precision",ylab="index")
+text((1:6*1.18)-0.38,prec+0.02,labels=signif(prec,3))
+barplot(rec,ylim=c(0,1.1),las=2,main="recall",ylab="index")
+text((1:6*1.18)-0.38,rec+0.02,labels=signif(rec,3))
+
+barplot(f1,ylim=c(0,1.1),las=2,main="F1",ylab="index")
+text((1:6*1.18)-0.38,f1+0.02,labels=signif(f1,3))
 

### 2024-06-13 analysis/sim/simpw_func.R

@@ -183,59 +183,6 @@ deseq2<-function(x) {
   x
 }
 
-
-#################################################
-# define mitch function
-##################################################
-run_mitch <- function(y,DGE_FUNC,gsets, N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,cores=16) {
-  library("mitch")
-  dge <- sapply(y,"[",6)
-  names(dge) <- paste0("x",1:length(dge),sep="")
-
-  mres <- mclapply(1:length(dge),function(d) {
-    w <- mitch_import(dge[[d]], DGE_FUNC)
-    mres <- mitch_calc(w,gsets,priority="significance",cores=1,minsetsize=5)
-  } , mc.cores=cores)
-
-  obs_up <- lapply(1:length(mres), function(d) {
-        up <- subset(mres[[d]]$enrichment_result, p.adjustANOVA<0.05 & s.dist>0)[,1]
-        y[[d]][[19]] <- up
-        up
-  } )
-
-  obs_dn <- lapply(1:length(mres), function(d) {
-        dn <- subset(mres[[d]]$enrichment_result, p.adjustANOVA<0.05 & s.dist<0)[,1]
-        y[[d]][[20]] <- dn
-        dn
-  } )
-
-  gt_up <- sapply(y,"[",4)
-  gt_up <- lapply( gt_up , names)
-  gt_dn <- sapply(y,"[",5)
-  gt_dn <- lapply( gt_dn , names)
-
-  true_pos_up <- as.numeric(mapply( function(x,y) length(intersect(x,y)) , obs_up ,  gt_up ))
-  true_pos_dn <- as.numeric(mapply( function(x,y) length(intersect(x,y)) , obs_dn ,  gt_dn ))
-  false_pos_up <- as.numeric(mapply( function(x,y) length(setdiff(x,y)) , obs_up ,  gt_up ))
-  false_pos_dn <- as.numeric(mapply( function(x,y) length(setdiff(x,y)) , obs_dn , gt_dn ))
-  false_neg_up <- as.numeric(mapply( function(x,y) length(setdiff(x,y)) , gt_up ,  obs_up ))
-  false_neg_dn <- as.numeric(mapply( function(x,y) length(setdiff(x,y)) , gt_dn ,  obs_dn ))
-
[…]

### 2024-06-14 analysis/sim/run_simpw.Rmd
png

@@
 par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
-barplot(prec,ylim=c(0,1.1),las=2,main="precision",ylab="index")
-text((1:6*1.18)-0.38,prec+0.02,labels=signif(prec,3))
-barplot(rec,ylim=c(0,1.1),las=2,main="recall",ylab="index")
-text((1:6*1.18)-0.38,rec+0.02,labels=signif(rec,3))
+barplot(prec,ylim=c(0,1.1),las=2,main="precision",ylab="")
+text((1:6*1.18)-0.45,prec+0.02,labels=signif(prec,3))
+barplot(rec,ylim=c(0,1.1),las=2,main="recall",ylab="")
+text((1:6*1.18)-0.45,rec+0.02,labels=signif(rec,3))
+barplot(f1,ylim=c(0,1.1),las=2,main="F1",ylab="")
+text((1:6*1.18)-0.45,f1+0.02,labels=signif(f1,3))
 
-barplot(f1,ylim=c(0,1.1),las=2,main="F1",ylab="index")
-text((1:6*1.18)-0.38,f1+0.02,labels=signif(f1,3))
+png("fig3_bars.png", width=7,height=5,units="in",res=150,pointsize=12)
+par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
+par(mfrow=c(1,3))
+barplot(prec,ylim=c(0,1.1),las=2,main="precision",ylab="")
+text((1:6*1.18)-0.45,prec+0.02,labels=signif(prec,3))
+barplot(rec,ylim=c(0,1.1),las=2,main="recall",ylab="")
+text((1:6*1.18)-0.45,rec+0.02,labels=signif(rec,3))
+barplot(f1,ylim=c(0,1.1),las=2,main="F1",ylab="")
+text((1:6*1.18)-0.45,f1+0.02,labels=signif(f1,3))
+dev.off()
 

### 2024-06-14 analysis/sim/run_simpw.Rmd
fig3 legend

@@
    c("CP default","CP BG fix","CP FDR fix","CP BG and FDR fix","fora","fgsea"),
-   col=c("black","orange","darkgreen","purple","red","blue"),horiz=FALSE, cex=1.2, pch=19,lwd=2)
+   col=c("black","orange","darkgreen","purple","red","blue"),horiz=FALSE, cex=1.0, pch=19,lwd=2)
 
@@
    col=c("black","orange","darkgreen","purple","red","blue"),
-   horiz=FALSE, cex=1.2, pch=19,lwd=2)
+   horiz=FALSE, cex=1.0, pch=19,lwd=2)
 

### 2024-06-14 manuscript/manuscript.Rmd
reproducibility with master script and dockerfile

@@
 
-library("DiagrammeR")
-library("RColorBrewer")
 library("kableExtra")
-#library("gt")
 
@@
 were selected, as suggested by a previous report [@Tarca2013-xh].
-These gene lists were subjected to ORA using the `enricher` function of clusterProfiler
+These gene lists were subjected to ORA using clusterProfiler's `enricher` function
 v4.10.0, using a minimum set size of 5 and a no maximum set size.
@@
 To systematically assess the impact of the background problem, we conducted ORA with
-`enricher` using the different gene set libraries described above for the seven RNA-seq
+clusterProfiler using the different gene set libraries described above for the seven RNA-seq
 datasets.
-`enricher` was used with the same parameters as above, and we devised a simple workaround
+ClusterProfiler was used with the same parameters as above, and we devised a simple workaround
 to the background problem, which is to append the entire background list as a gene set to the
 library.
-This forces `enricher` to retain all background genes.
+This forces clusterProfiler to retain all background genes.
 The ORA analyses were then filtered for significant sets using a FDR threshold of 0.05.
@@
 A workaround for the FDR problem was devised by filtering the gene sets by the presence of
-at least two genes in the background list prior to running `enricher`.
-After `enricher` using the parameters above, the number of gene sets in the results was
+at least two genes in the background list prior to running clusterProfiler.
+After clusterProfiler using the parameters above, the number of gene sets in the results was
 quantified, and compared to the number that should have been reported using the two gene
@@
 To account for these tests, *n* values of 1 were appended to the p-values obtained by
-enricher, followed by FDR correction in R, to obtain the properly corrected FDR values.
+clusterProfiler, followed by FDR correction in R, to obtain the properly corrected FDR values.
 As above, analysis of seven datasets was done with nine gene set libraries, the significant
@@
 With these expression profiles, DESeq2 was used for differential analysis, and the results
-were subjected to `enricher` which suffers from both problems, `fora`, an ORA function of
-the `fgsea` bioconductor package v1.30.0 which does not suffer these problems, and `fgsea` a
-fast implementation of the preranked GSEA algorithm [@Korotkevich2016-gd].
+were subjected to clusterProfiler which suffers from both problems,
+clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
+clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
+from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
+fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
+[@Korotkevich2016-gd].
 For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
@@
 No set seed was used, but results were similar when run on different computer systems.
+All analyses were undertaken with R v4.4.0.
 
@@
 of statistically significant sets (FDR<0.05) over seven independent datasets detected with
-clusterProfiler's `enricher` function with and without correcting the background problem.
+clusterProfiler with and without correcting the background problem.
 Separate tests for up- and down-regulated gene lists were conducted.
@@
 Simulated differential expression profiles with *a priori* changes to predetermined pathway
-genes were analysed in parallel by clusterProfiler's `enricher` function, which suffers from
-the two aforementioned problems and `fora` which does not.
-In parallel, `fgsea` was used to compare the performance of these two ORA methods against a
+genes were analysed in parallel by clusterProfiler default and with mitigations for
+problems described above.
+In addition, these data underwent ORA with fora which does not suffer from these two
+problems.
+For comparison fgsea was used to compare the performance of these two ORA methods against a
 prototypical FCS method.
-ClusterProfiler scored very high precision when added variance was less than 0.2, but showed a
-progressive worsening in precision as more noise was added to the dataset, dropping to 0.69
-with added variance of 0.5.
-In comparison, the precision of fora was relatively stable around 0.97 until added variance
-reached 0.5, where it dropped to 0.88.
-The recall of ORA methods dropped suddenly when added variance was 0.2 or greater.
-fgsea showed stable precision of 0.93 or greater when added variance was 0.5 or lower, which
-dropped slightly to 0.91 with added variance of 0.6.
-fgsea recall was better than either ORA tool across all levels of added variance.
-This result shows that these two subtle problems result in poorer precision for
-clusterProfiler, which becomes more severe with higher levels of noise.
-Overall, fgsea had similar mean precision to fora (0.94), but the higher recall of fgsea
-meant it had a higher overall accuracy measured by F1 score (0.83) as compared to fora and
-clusterProfiler (0.61 and 0.64 respectively).
+These methods were tested with different levels of introduced variation and the mean
+precision, recall and F1 scores were recorded (**Figure 3A**).
+As anticipated, mitigating the background problem improved recall (by 1.6%), however it caused
+a similar-sized decrease in precision.
+Addressing the FDR problem improved precison by 9.6%, but the recall dropped by 8.5%.
+Addressing both problems improved precision by 8.2% but decreased recall by 5.6%.
+Under these conditions, the FDR problem affected results more severely than the background
+problem.
+Results obtained for clusterProfiler after implementing the two mitigations were identical to
+fora to three significant figures.
+fgsea recorded precision of 0.937, similar to the ORA methods, but recall was far superior,
+with fgsea scoring 22.4% higher than default clusterProfiler and 28% higher than fora.
+As a result, fgsea's overall accuracy as defined by F1 index was 21.2% higher than fora and
+18.4% higher than default clusterProfiler.
 
-![**Figure 3. Comparison of clusterProfiler, fora and fgsea accuracy with simulated data.**
+When looking at the accuracy profile at different levels of added noise (**Figure 3B**), we
+see that ORA techniques have superior precision when the added noise is low (<0.2), but the
+precision of ORA techniques drops at higher levels of added noise.
+The rate of drop in precision was more severe for approaches that suffer from the FDR problem.
+The precision of fgsea was relatively stable despite high levels of added noise, ranging from
+0.95 to 0.91.
+The recall of fgsea was superior to all ORA approaches across all levels of added noise.
+
+![**Figure 3A. Impact of correcting these two problems on accuracy of simulated expression
+data.**
+Precision, recall and F1 scores are the mean of trials at a range of added noise.
+CP; clusterProfiler, CP BG fix; clusterProfiler with a workaround to the background problem,
+CP FDR fix; clusterProfiler with a workaround to the FDR problem,
+CD BG and FDR fix; clusterProfiler with mitigations for both background and FDR problems,
+fora; ORA function of the fgsea package,
+fgsea; an FCS method similar to GSEA.
+](../analysis/sim/fig3_bars.png "Figure 3A shows that fora is more
+precise than clusterProfiler, and fgsea has higher recall than ORA methods.")
+
+![**Figure 3B. Accuracy of clusterProfiler, fora and fgsea with simulated data with various
+levels of added noise.**
 Precision, recall and F1 score are shown with the addition of different amounts of random
-noise to expression levels.](../analysis/sim/fig3_sim.png "Figure 3 shows that fora is more
-precise than clusterProfiler, and fgsea has higher recall than either ORA methods.")
+noise to expression levels.
+The purple line for clusterProfiler with mitigations for both problems is not visible as the
+performance was identical to fora.
+](../analysis/sim/fig3_sim.png "Figure 3B shows that ORA methods have deteriorating accuracy
+with noisy data as compared to fgsea, a FCS method.")
 
@@
 
-Although the purpose of this work was to describe two subtle problems with ORA, this work
-highlights some major drawbacks.
+The purpose of this work was to describe two subtle problems with ORA, but there are some
+bigger problems noted.
 Firstly, ORA was less accurate as compared to FCS in this simulation work, confirming a
 previous report [@Kaspi2020-rn].
+Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
+was relatively stable.
+This means researchers can be more confident with FCS results, even if the data are
+noisy.
+Higher recall of FCS also means that researchers will have more power to identify pathways
+relevant to the biological processes being studied.
+
 Secondly ORA results can vary dramatically depending on the number of genes selected in the

### 2024-06-14 manuscript/manuscript.Rmd
resources

@@
 
-Secondly ORA results can vary dramatically depending on the number of genes selected in the
+The lack of enrichment scores hampers interpretation.
+
+Lastly, ORA results can vary dramatically depending on the number of genes selected in the
 foreground, a mostly arbitrary choice.
@@
 
-* The code repository including template Dockerfile and R Markdown script are available on GitHub
-(https://github.com/markziemann/enrichment_recipe).
+* The code repository including template Dockerfile and R Markdown script are available on
+GitHub (https://github.com/markziemann/background).
 
-* The example Docker image is available on DockerHub (https://hub.docker.com/r/mziemann/enrichment_recipe).
+* The example Docker image is available on DockerHub
+(https://hub.docker.com/r/mziemann/background).
 

### 2024-06-14 manuscript/manuscript.Rmd
mswork

@@
 Remaining genes passing this selection were included in the background gene set.
-Differential expression analysis was conducted with DESeq2 v1.42.0 [@Love2014-wg], with
+Differential expression analysis was conducted with DESeq2 v1.44.0 [@Love2014-wg], with
 genes identified by their Ensembl identifiers.
-Gene symbols were then fetched using biomaRt v2.58.0, based on Ensembl version 112 [@Durinck2009-dr].
+Gene symbols were then fetched using biomaRt v2.60.0, based on Ensembl version 112 [@Durinck2009-dr].
 
@@
 These gene lists were subjected to ORA using clusterProfiler's `enricher` function
-v4.10.0, using a minimum set size of 5 and a no maximum set size.
+v4.12.0, using a minimum set size of 5 and a no maximum set size.
 
@@
 No set seed was used, but results were similar when run on different computer systems.
-All analyses were undertaken with R v4.4.0.
+All analyses were undertaken with R v4.4.0 in a Docker container.
 
@@
 Each of the gene set libraries showed an uplift when correcting the background problem.
-The smallest increase was observed in Cellmarkers gene library (11.7%), while the largest
-increase was observed for HPO (229%).
+The smallest increase was observed in Cellmarkers gene library (11.8%), while the largest
+increase was observed for HPO (224%).
 The similarity between results, as quantified with the Jaccard similarity was highest for
-Cellmarkers and GTRD at 0.88 and 0.76 respectively, and the lowest similarity was recorded for
+Cellmarkers and GTRD at 0.76 and 0.77 respectively, and the lowest similarity was recorded for
 HPO and KEGG with 0.20 and 0.48 respectively.
@@
 precision, recall and F1 scores were recorded (**Figure 3A**).
-As anticipated, mitigating the background problem improved recall (by 1.6%), however it caused
-a similar-sized decrease in precision.
-Addressing the FDR problem improved precison by 9.6%, but the recall dropped by 8.5%.
-Addressing both problems improved precision by 8.2% but decreased recall by 5.6%.
+Mitigating the background problem improved recall (by 1.8%), however it caused
+a similar-sized decrease in precision (1.6% lower).
+Addressing the FDR problem improved precison by 9.9%, but the recall dropped by 7.7%.
+Addressing both problems improved precision by 8.2% but decreased recall by 5.4%.
 Under these conditions, the FDR problem affected results more severely than the background
@@
 fora to three significant figures.
-fgsea recorded precision of 0.937, similar to the ORA methods, but recall was far superior,
-with fgsea scoring 22.4% higher than default clusterProfiler and 28% higher than fora.
-As a result, fgsea's overall accuracy as defined by F1 index was 21.2% higher than fora and
-18.4% higher than default clusterProfiler.
+fgsea recorded precision of 0.940, similar to the ORA methods, but recall was far superior,
+with fgsea scoring 22.9% higher than default clusterProfiler and 28.3% higher than fora.
+As a result, fgsea's overall accuracy as defined by F1 index was 21.0% higher than fora and
+18.6% higher than default clusterProfiler.
 
-When looking at the accuracy profile at different levels of added noise (**Figure 3B**), we
-see that ORA techniques have superior precision when the added noise is low (<0.2), but the
-precision of ORA techniques drops at higher levels of added noise.
+When looking at the accuracy profile at different levels of added noise (**Figure 3B**),
+ORA techniques show high precision when the added noise is low (<0.2), but
+precision drops at higher levels of added noise.
 The rate of drop in precision was more severe for approaches that suffer from the FDR problem.
 The precision of fgsea was relatively stable despite high levels of added noise, ranging from
-0.95 to 0.91.
+0.95 to 0.92.
 The recall of fgsea was superior to all ORA approaches across all levels of added noise.
@@
 The results shown here show that the background problem effects smaller gene set libraries
-like KEGG and HPC more severely as compared to Cellmarkers and GO libraries which describe
+like KEGG and HPO more severely as compared to Cellmarkers and GO libraries which describe
 a larger proportion of all genes.

### 2024-06-15 manuscript/manuscript.Rmd
spellcheck

@@
 
-Over-representation analysis (ORA) is used widely to asses for the enrichment of functional
+Over-representation analysis (ORA) is used widely to assess the enrichment of functional
 categories in a gene list compared to a background list.
@@
 genes in the list, and the degree of noise in the dataset.
-These problem can be mitigated by changing packages/websites for ORA or by changing to another
+These problems can be mitigated by changing packages/websites for ORA or by changing to another
 approach such as functional class scoring.
@@
 involves the summarisation of omics data to reflect biological differences.
-Generally speaking, FEA has become one of the most popular methods in bioinformatics,
+FEA has become one of the most popular methods in bioinformatics,
 collectively accumulating 131,332 citations as of late 2019 [@Xie2021-nv].
@@
 certain functional categories are over-represented in the selected genes.
-Since it's initial development in 1999 [@Tavazoie1999-bu], ORA has proliferated widely,
+Since its initial development in 1999 [@Tavazoie1999-bu], ORA has proliferated widely,
 becoming part of many bioinformatics websites and software packages.
@@
 with results appearing nearly instantly.
-On the other hand, FCS typically involves installing specific softwares and dealing with a
+On the other hand, FCS typically involves installing specific software and dealing with a
 dataset representing every detected gene (~20,000 rows) and ensuring that the file
@@
 Correction of p-values for multiple testing is also crucial in controlling the false positive
-rate as enrichment analysis typically involves throusands of parallel tests [@Tilford2009-vx].
+rate as enrichment analysis typically involves thousands of parallel tests [@Tilford2009-vx].
 But appropriate p-value correction is reported in only ~54% of studies [@Wijesooriya2022-fw].
@@
 The second problem is that adjusting p-values for multiple testing, also known as
-false-discovery rate correction is implemented improperly.
+false-discovery rate correction, is implemented improperly.
 Determining which pathways should be subject to false discovery rate correction depends on
@@
 number of member genes in the background set.
-However some tools use the foreground set to define a pathway as detected.
-This is problematic, as the act of trying to find an intersect between a foreground gene list
-and a pathway is in itself a test, even if there are no common genes.
+However, some tools use the foreground set to define a pathway as detected.
+This is problematic, as the act of trying to find an intersection between a foreground gene list
+and a pathway is a test, even if there are no common genes.
 This results in pathways being discarded from the analysis if no common genes are found in
-the foreground, despite a large number of those genes being present in the background list.
+the foreground, despite many genes being present in the background list.
 This artificially reduces the number of tests conducted and makes FDR values appear lower
@@
 data.
-Finally we classify popular tools regarding these two problems and provide general
+Finally, we classify popular tools regarding these two problems and provide general
 recommendations for end users.
@@
 FDR<0.05 were subset, with separate lists for up and downregulated genes.
-If fewer than 200 genes met this criteria, then the 200 genes with the smallest p-values
+If fewer than 200 genes met this criterion, then the 200 genes with the smallest p-values
 were selected, as suggested by a previous report [@Tarca2013-xh].
@@
 This forces clusterProfiler to retain all background genes.
-The ORA analyses were then filtered for significant sets using a FDR threshold of 0.05.
+The ORA analyses were then filtered for significant sets using an FDR threshold of 0.05.
 Jaccard index was used to compare original and corrected analyses.
@@
 If fewer than 200 genes were detected as significant, then 200 genes with the smallest
-p-values were used for ORA, similar to a previous report [@Tarca2013-xh].
+p-values were used for ORA, like a previous report [@Tarca2013-xh].
 For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
 After enrichment analysis with these approaches, the significant up and down regulated
-sets were selected and compared to the ground truth in order to calculate precision, recall
+sets were selected and compared to the ground truth to calculate precision, recall
 and F1 score.
@@
 gene set libraries.
-The mean number of significant results was calculated for each of the gene set libraries
-before and after correcting the problem (**Figure 1A**).
+The mean number of statistically significant results was calculated for each of the gene set
+libraries before and after correcting the problem (**Figure 1A**).
 Each of the gene set libraries showed an uplift when correcting the background problem.
@@
 increase was observed for HPO (224%).
-The similarity between results, as quantified with the Jaccard similarity was highest for
+The similarity between results, as quantified with the Jaccard similarity, was highest for
 Cellmarkers and GTRD at 0.76 and 0.77 respectively, and the lowest similarity was recorded for
@@
 Interestingly, the drop in Jaccard similarity was more severe for GO and Reactome sets as
-compared to Cellmarkers and TFT GTRD sets, and may be associated with the size of the gene
+compared to Cellmarkers and TFT GTRD sets and may be associated with the size of the gene
 sets in the library.
@@
 problems.
-For comparison fgsea was used to compare the performance of these two ORA methods against a
+For comparison, fgsea was used to compare the performance of these two ORA methods against a
 prototypical FCS method.
@@
 a similar-sized decrease in precision (1.6% lower).
-Addressing the FDR problem improved precison by 9.9%, but the recall dropped by 7.7%.
+Addressing the FDR problem improved precision by 9.9%, but the recall dropped by 7.7%.
 Addressing both problems improved precision by 8.2% but decreased recall by 5.4%.
@@
 ~2020 where confused users were wondering why the number of background genes changed so much.
-The results shown here show that the background problem effects smaller gene set libraries
+The results shown here show that the background problem affects smaller gene set libraries
 like KEGG and HPO more severely as compared to Cellmarkers and GO libraries which describe
@@
 The second problem is potentially more serious in that it could lead to increasing false
-positives under specific conditions; when the foreground list is short and the median pathway
+positives under specific conditions; when the foreground list is short, and the median pathway
 size is relatively small.
 These false positives arise due to an underestimate of the actual number of tests conducted.
-In extreme cases, correction of this problem can lead to around half of the significant
-results changing.
+In extreme cases, correction of this problem can lead to around half of the statistically
+significant results changing.
 
@@
 investigating the over-represented categories in gene expression modules defined by packages
-like WGCNA (Weighted Gene Correlation Network Analysis)[@Langfelder2008-hu], or when
+like WGCNA (Weighted Gene Correlation Network Analysis) [@Langfelder2008-hu], or when
 investigating over-represented functional categories among genes with variants associated

### 2024-06-15 manuscript/manuscript.Rmd
mswork

@@
 relevant to the biological processes being studied.
-
-The lack of enrichment scores hampers interpretation.
-
+Secondly, many ORA tools do not provide enrichment scores in their default output tables.
+Enrichment scores are a surrogate for effect size, and without them, users may over-rely
+on significance scores when interpreting enrichment results [@Ziemann2024-ts].
 Lastly, ORA results can vary dramatically depending on the number of genes selected in the
 foreground, a mostly arbitrary choice.
+
 These issues together with previously mentioned pitfalls such as lack of background

### 2024-06-29 app/app.R
app idea

@@ -1,0 +1,158 @@
+library(tidyverse)
+library(ggplot2)
+library(rmarkdown)
+library(broom)
+
+# Define UI
+ui <- fluidPage(
+  titlePanel("Two subtle issues with over-representation analysis: Demonstration"),
+  sidebarLayout(
+    sidebarPanel(
+      fileInput("fg", "Choose foreground gene list",
+                accept = c("text/csv",
+                           "text/comma-separated-values,text/plain",
+                           ".txt")),
+      fileInput("bg", "Choose background gene list",
+                accept = c("text/csv",
+                           "text/comma-separated-values,text/plain",
+                           ".txt")),
+      uiOutput("comparison"),
+      uiOutput("genesetlibrary"),
+      downloadButton("download_report", "Download Report")
+    ),
+    mainPanel(
+      tabsetPanel(
+        tabPanel("Data",
+                 "Foreground", verbatimTextOutput("contents1"),
+                 "Background",verbatimTextOutput("contents2"),
+                 "No. genes in foregorund", verbatimTextOutput("summary1"),
+                 "No. genes in background", verbatimTextOutput("summary2")),
+        tabPanel("Scatterplot", plotOutput("scatterplot")),
+        tabPanel("Regression Summary", verbatimTextOutput("regression_summary"))
+      )
+    )
+  )
+)
+
+# Define server logic
+server <- function(input, output, session) {
+  
+  fg <- reactive({
+    req(input$fg)
+    fg <- readLines(input$fg$datapath)
+    return(fg)
+  })
+  
+  bg <- reactive({
+    req(input$bg)
+    bg <- readLines(input$bg$datapath)
+    return(bg)
+  })
+  
+  output$contents1 <- renderPrint({
+    req(fg())
+    head(fg())
+  })
+  
+  output$contents2 <- renderPrint({
+    req(bg())
+    head(bg())
+  })
+  
+  output$summary1 <- renderPrint({
+    req(fg())
+    length(fg())
+  })
+  
+  output$summary2 <- renderPrint({
+    req(bg())
+    length(bg())
+  })
+  
+  output$comparison <- renderUI({
+    selectInput("comparison", "Error Comparison", choices = c("Background error","FDR error","Both errors"), selected = "both")
+  })
+  
+  output$genesetlibrary <- renderUI({
[…]

### 2024-06-29 app/app.R
progress

@@ -1,7 +1,9 @@
-library(tidyverse)
-library(ggplot2)
-library(rmarkdown)
-library(broom)
+library("clusterProfiler")
+library("markdown")
+library("rmarkdown")
+library("shiny")
+library("broom")
+library("RhpcBLASctl")
 
 # Define UI
 ui <- fluidPage(
@@ -22,12 +24,13 @@ ui <- fluidPage(
     ),
     mainPanel(
       tabsetPanel(
+        tabPanel("Instructions", includeMarkdown("intro.md")),
         tabPanel("Data",
                  "Foreground", verbatimTextOutput("contents1"),
                  "Background",verbatimTextOutput("contents2"),
                  "No. genes in foregorund", verbatimTextOutput("summary1"),
                  "No. genes in background", verbatimTextOutput("summary2")),
-        tabPanel("Scatterplot", plotOutput("scatterplot")),
+        tabPanel("Comparative Analysis", tableOutput("comparativeanalysis")),
         tabPanel("Regression Summary", verbatimTextOutput("regression_summary"))
       )
     )
@@ -74,8 +77,68 @@ server <- function(input, output, session) {
   })
   
   output$genesetlibrary <- renderUI({
-    selectInput("genesetlibrary", "Gene set library", choices = c("Gene Ontology","Reactome Pathways","KEGG Pathways"), selected = names(data())[2])
+    selectInput("genesetlibrary", "Gene set library", 
+                choices = c("CellMarkers", "Gene Ontology","GTRD TF Targets",
+                            "Hallmark","Human Phenotype Ontology",
+                            "KEGG Pathways","miR targets",
+                            "Reactome Pathways","WikiPathways"))
   })
+  
+  
+  mygs <- reactive({
+    gs <- readRDS("genesets/gs.Rds")
+    if ( input$genesetlibrary == "CellMarkers") {
+      mygs <- gs[["CellMarkers"]]
+    }
+    if ( input$genesetlibrary == "Gene Ontology") {
+      mygs <- gs[["Gene Ontology"]]
+    }
+    if ( input$genesetlibrary == "GTRD TF Targets") {
+      mygs <- gs[["GTRD TF Targets"]]
+    }
+    if ( input$genesetlibrary == "Hallmark") {
[…]

### 2024-06-29 app/genesets/genesets.R

@@ -1,0 +1,21 @@
+# Convert GMT to RDS
+library("clusterProfiler")
+l <- list.files(".",pattern = "gmt")
+l <- l[order(l)]
+gs <- lapply(l,function(x) {
+  y <- read.gmt(x)
+  y$term <- gsub("_"," ",y$term)
+  return(y)
+})
+str(gs)
+names(gs) <- c("KEGG Pathways",
+               "Reactome Pathways",
+               "WikiPathways",
+               "miR targets",
+               "GTRD TF Targets", 
+               "Gene Ontology",
+               "Human Phenotype Ontology",
+               "CellMarkers", 
+               "Hallmark")
+                            
+saveRDS(gs,"gs.Rds")

### 2024-06-29 app/app.R
app progress

@@ -4,6 +4,7 @@ library("rmarkdown")
 library("shiny")
 library("broom")
 library("RhpcBLASctl")
+library("eulerr")
 
 # Define UI
 ui <- fluidPage(
@@ -30,8 +31,8 @@ ui <- fluidPage(
                  "Background",verbatimTextOutput("contents2"),
                  "No. genes in foregorund", verbatimTextOutput("summary1"),
                  "No. genes in background", verbatimTextOutput("summary2")),
-        tabPanel("Comparative Analysis", tableOutput("comparativeanalysis")),
-        tabPanel("Regression Summary", verbatimTextOutput("regression_summary"))
+        tabPanel("Comparative Analysis", tableOutput("tbl1")),
+        tabPanel("Charts", textOutput("counts"),plotOutput("euler"))
       )
     )
   )
@@ -84,7 +85,6 @@ server <- function(input, output, session) {
                             "Reactome Pathways","WikiPathways"))
   })
   
-  
   mygs <- reactive({
     gs <- readRDS("genesets/gs.Rds")
     if ( input$genesetlibrary == "CellMarkers") {
@@ -117,11 +117,11 @@ server <- function(input, output, session) {
     return(mygs)
   })
     
-  output$comparativeanalysis <- renderTable({
+  original <- reactive({
     req(fg())
     req(bg())
     req(mygs())
-    #length(unique(mygs()[,1]))
+    options(enrichment_force_universe = FALSE)
     ora <- as.data.frame(enricher(gene = fg() ,
                                      universe = bg(), minGSSize = 5, maxGSSize = 500000, TERM2GENE = mygs(),
                                      pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
@@ -135,11 +135,67 @@ server <- function(input, output, session) {
     ora$ES <- gr/br
     ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
     colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
-    ora <- ora[,c("Description","GeneRatio","BgRatio","ES","pvalue","FDR")]
+    colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
[…]

### 2024-06-30 app/app.R
progress charts

@@ -31,8 +31,13 @@ ui <- fluidPage(
                  "Background",verbatimTextOutput("contents2"),
                  "No. genes in foregorund", verbatimTextOutput("summary1"),
                  "No. genes in background", verbatimTextOutput("summary2")),
-        tabPanel("Comparative Analysis", tableOutput("tbl1")),
-        tabPanel("Charts", textOutput("counts"),plotOutput("euler"))
+        tabPanel("Comparative Analysis", 
+                 textOutput("Comparison of original and BG fixed analysis - top 20 pathways with divergent FDR values"),
+                 tableOutput("tbl1")),
+        tabPanel("Charts", textOutput("counts"),
+                 plotOutput("euler"),
+                 plotOutput("scatter_es",height=550),
+                 plotOutput("scatter_fdr",height=550)),
       )
     )
   )
@@ -195,7 +200,32 @@ server <- function(input, output, session) {
     v1 <- list("Original"=orig_sets, "BG fix"=bgfix_sets)
     plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))
   })
+  
+  output$scatter_es <- renderPlot({
+    orig_df <- original()
+    bgfix_df <- bgfix()
+    m <- merge(orig_df,bgfix_df,by="Description")
+    m <- m[,c("ES.x","ES.y")]
+    MAX=max(c(m$ES.x,m$ES.y))
+    plot(m$ES.x,m$ES.y,xlim=c(0,MAX),ylim=c(0,MAX),
+         xlab="Original",ylab="BG Fix")
+    mtext("Fold Enrichment Scores")
+    abline(a = 0, b = 1,lwd=2,lty=2,col="red")
+  })
 
+  output$scatter_fdr <- renderPlot({
+    orig_df <- original()
+    bgfix_df <- bgfix()
+    m <- merge(orig_df,bgfix_df,by="Description")
+    m <- m[,c("FDR.x","FDR.y")]
+    MAX=max(c(-log10(m$FDR.x),-log10(m$FDR.y)))
+    plot(-log10(m$FDR.x),-log10(m$FDR.y),
+         xlim=c(0,MAX),ylim=c(0,MAX),
+         xlab="Original",ylab="BG Fix")
+    mtext("FDR Values")
+    abline(a = 0, b = 1,lwd=2,lty=2,col="red")
+  })
+  
   plot_data <- reactive({
     req(data(), input$x_axis, input$y_axis)
[…]

### 2024-06-30 app/app.R
plotly

@@ -5,6 +5,7 @@ library("shiny")
 library("broom")
 library("RhpcBLASctl")
 library("eulerr")
+library("plotly")
 
 # Define UI
 ui <- fluidPage(
@@ -37,7 +38,8 @@ ui <- fluidPage(
         tabPanel("Charts", textOutput("counts"),
                  plotOutput("euler"),
                  plotOutput("scatter_es",height=550),
-                 plotOutput("scatter_fdr",height=550)),
+                 plotOutput("scatter_fdr",height=550),
+                 plotlyOutput("scatter_es2")),
       )
     )
   )
@@ -121,15 +123,15 @@ server <- function(input, output, session) {
     }
     return(mygs)
   })
-    
+  
   original <- reactive({
     req(fg())
     req(bg())
     req(mygs())
     options(enrichment_force_universe = FALSE)
     ora <- as.data.frame(enricher(gene = fg() ,
-                                     universe = bg(), minGSSize = 5, maxGSSize = 500000, TERM2GENE = mygs(),
-                                     pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+                                  universe = bg(), minGSSize = 5, maxGSSize = 500000, TERM2GENE = mygs(),
+                                  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
     
     gr <- as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",1)) /
       as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",2))
@@ -167,7 +169,7 @@ server <- function(input, output, session) {
     ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
     return(ora)
   })
-
+  
   bgfixtbl <- reactive({
     orig_df <- original()
     bgfix_df <- bgfix()
@@ -193,6 +195,12 @@ server <- function(input, output, session) {
   })
   
   output$euler <- renderPlot({
+    if (input$comparison == "Background error") {
+      euler_bgfix()
+    }
+  })
+  
+  euler_bgfix <- reactive({
     orig_df <- original()
     bgfix_df <- bgfix()
     orig_sets <- subset(orig_df,FDR < 0.05)$Description
@@ -202,6 +210,12 @@ server <- function(input, output, session) {
   })
   
[…]

### 2024-07-01 app/app.R
interactive charts

@@ -32,14 +32,16 @@ ui <- fluidPage(
                  "Background",verbatimTextOutput("contents2"),
                  "No. genes in foregorund", verbatimTextOutput("summary1"),
                  "No. genes in background", verbatimTextOutput("summary2")),
-        tabPanel("Comparative Analysis", 
-                 textOutput("Comparison of original and BG fixed analysis - top 20 pathways with divergent FDR values"),
+        tabPanel("Comparative Analysis",
+                 "Comparison of original and BG fixed analysis - top 20 pathways with divergent FDR values.",
+                 "'x' is original, and 'y' is fixed analysis,",
                  tableOutput("tbl1")),
         tabPanel("Charts", textOutput("counts"),
                  plotOutput("euler"),
-                 plotOutput("scatter_es",height=550),
-                 plotOutput("scatter_fdr",height=550),
-                 plotlyOutput("scatter_es2")),
+                 #plotOutput("scatter_es",height=550),
+                 #plotOutput("scatter_fdr",height=550),
+                 plotlyOutput("scatter_es2",height=550),
+                 plotlyOutput("scatter_fdr2",height=550)),
       )
     )
   )
@@ -170,27 +172,37 @@ server <- function(input, output, session) {
     return(ora)
   })
   
-  bgfixtbl <- reactive({
+  tbl1_bgfix <- reactive({
     orig_df <- original()
     bgfix_df <- bgfix()
     m <- merge(orig_df,bgfix_df,by="Description")
     m <- m[,c("Description","FgRatio.x","FgRatio.y","BgRatio.x","BgRatio.y","ES.x","ES.y","FDR.x","FDR.y")]
-    diff <- abs(m$FDR.x - m$FDR.y)
+    diff <- abs(-log10(m$FDR.x) - -log10(m$FDR.y))
     m <- m[order(-diff),]
     return(m)
   })
   
   output$tbl1 <- renderTable({
     if ( input$comparison == "Background error" ) {
-      tbl <- head(bgfixtbl(),20)
+      tbl <- head(tbl1_bgfix(),100)
     }
     return(tbl)
   })  
   
-  output$counts <- renderText({
+  counts_bgfix <- reactive({
[…]

### 2024-07-01 app/report.Rmd
adding report

---
title: "Two subtle issues with over-representation analysis: Demonstration"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
params:
  fg: NA
  bg: NA
  comparison: NA
  genesetlibrary: NA
---

Source: https://github.com/markziemann/background

## Background

Over-representation analysis is frequently used to determine the functionally
important themes from gene lists.
These gene lists are classified in different ways, but are commonly defined
by differential regulation in omics experiments like RNA-seq, proteomics or
micro-array.
Many different tools are available to conduct this type of analysis including
[DAVID](https://david.ncifcrf.gov/) and [Enricher](https://maayanlab.cloud/Enrichr/).

For the results to be reliable, the gene list needs to be provided together
with a background list.
The background list is the full list of genes that were reliably detected by
the assay.
This is important, because these assays typically cannot detect all genes
reliably due to technical limitations.
As different tissues express genes very differently, the right background list
is critical to obtain the functional categories that are specific to the
experiment conducted.
Using a generic background list of all annotated genes could lead to a major
bias.

Additionally, as such analyses involve the analysis of hundreds of
functional categories in parallel, the resulting p-values cannot be taken at
face value due to the high rate of false positives.
To mitigate this, false discovery rate correction is applied to adjust the
p-values.

In this work we highlight two subtle issues with some popular implementations 
of over-representation analysis.

## Genes without annotated categories are discarded

The first issue related to how the background list is defined.
If a user does the right thing and provides a background list along with the
foreground list, then some tools will exclude any genes from the background
that do not have any annotated functional categories.
This results in a large fraction of genes in the background being excluded,
and the enrichment ratio being unreliable.
This issue is more severe for analyses that involve smaller libraries of
functional annotations.
For example KEGG has annotations for a few thousand genes, meaning most of
the background genes get discarded.

## p-value adjustment is applied incorrectly

The second issue is that the false discovery rate is implemented incorrectly.
ORA enrichment tools intersect the gene list with each functional category
and return results for each functional category with at least one common
gene.
But there are some functional categories with zero genes in common with the
gene list.
These are excluded from the end results.
The process of attempting to intersect these categories with the submitted
gene list is itself a test, and so statistically speaking, the result of
that failed test should be reported in the final results with no overlaps and
a p-value of 1.
This may not be a problem when running enrichment analysis on a gene list 
with 3000 members, but it might be serious when it is small, like <200
members.
This results in adjusted p-values being slightly smaller than they should be.

This software was developed by Mark Ziemann, report issues at the [GitHub repository](https://github.com/markziemann/background).

Some information about the parameters used:

```{r, start}

fg <- params$fg
bg <- params$bg
comparison <- params$comparison
genesetlibrary <- params$genesetlibrary
minsetsize <- 5

message(paste("Minimum gene set size:",minsetsize))

```

Minimum set size is defined as 5.
Sets with fewer than this number of genes are discarded.

## Input gene lists

Here is the first few lines of the input foreground and background lists.

```{r,checklibraries}

suppressPackageStartupMessages({
  library("clusterProfiler")
  library("markdown")
  library("rmarkdown")
  library("shiny")
  library("broom")
  library("RhpcBLASctl")
  library("eulerr")
  library("plotly")
  library("DT")
})

```

```{r,peek}

message("First 10 genes in the foreground")
head(fg,10)

message("First 10 genes in the background")
head(bg,10)

message("Num genes in foreground")
length(fg)

message("Num genes in background")
length(bg)

```

Here is a look at the gene set library.

```{r, gsl1}

genesetlibrary

```

Overlap of genes in the gene set library.

```{r,gsl2}

print("abc")

```

## Default ORA

```{r,default }

print("run clusterprofiler default")

print("allow download?")

```

Make bar chart.

```{r,default_bars}

barplot(c(1,2,3))

```

## Fixed analysis

There are two issues with ORA.
The one being characterised here is shown in the text below:

```{r,fixed1 }

mymessage <- paste("Here we are characterising","comparison")
message(mymessage)

print("run clusterprofiler fixed")
# actually there will be three different types of analysis possible here
# depending on the selection

print("allow download?")

```

Make bar chart.

```{r,fixed_bars}

barplot(c(1,2,3))

```

## Comparison of default and fixed analysis

* Data table

* Euler diagram

* Scatterplot of ES

* Scatterplot of logFDR

## Session information

Here is the session info with all the versions of packages used.

```{r session_info}

sessionInfo()

```

END of report

### 2024-07-01 app/app.R

@@ -6,6 +6,8 @@ library("broom")
 library("RhpcBLASctl")
 library("eulerr")
 library("plotly")
+library("DT")
+
 
 # Define UI
 ui <- fluidPage(
@@ -35,7 +37,7 @@ ui <- fluidPage(
         tabPanel("Comparative Analysis",
                  "Comparison of original and BG fixed analysis - top 20 pathways with divergent FDR values.",
                  "'x' is original, and 'y' is fixed analysis,",
-                 tableOutput("tbl1")),
+                 DT::dataTableOutput("tbl1")),
         tabPanel("Charts", textOutput("counts"),
                  plotOutput("euler"),
                  #plotOutput("scatter_es",height=550),
@@ -179,15 +181,19 @@ server <- function(input, output, session) {
     m <- m[,c("Description","FgRatio.x","FgRatio.y","BgRatio.x","BgRatio.y","ES.x","ES.y","FDR.x","FDR.y")]
     diff <- abs(-log10(m$FDR.x) - -log10(m$FDR.y))
     m <- m[order(-diff),]
+    m$ES.x <- signif(m$ES.x,3)
+    m$ES.y <- signif(m$ES.y,3)
+    m$FDR.x <- signif(m$FDR.x,3)
+    m$FDR.y <- signif(m$FDR.y,3)
     return(m)
   })
   
-  output$tbl1 <- renderTable({
+  output$tbl1 <- DT::renderDataTable({
     if ( input$comparison == "Background error" ) {
-      tbl <- head(tbl1_bgfix(),100)
+      tbl <- tbl1_bgfix()
     }
     return(tbl)
-  })  
+  }, rownames= FALSE)  
   
   counts_bgfix <- reactive({
     oricnt <- nrow(subset(original(),FDR<0.05))
@@ -377,11 +383,10 @@ server <- function(input, output, session) {
       tempReport <- file.path(tempdir(), "report.Rmd")
       file.copy("report.Rmd", tempReport, overwrite = TRUE)
       
-      params <- list(plot = plot_data(),
-                     stats = regression_stats(),
-                     x_var = input$x_axis,
-                     y_var = input$y_axis,
-                     coefficients = summary(regression_model())$coefficients)
+      params <- list(fg = fg(),
+                     bg = bg(),
+                     comparison = input$comparison,
+                     genesetlibrary = input$genesetlibrary)
       
       rmarkdown::render(tempReport, output_file = file,
                         params = params,

### 2024-07-01 app/report.Rmd
progress on report

@@
   genesetlibrary: NA
+  mygs: NA
 ---
@@
 
-## Genes without annotated categories are discarded
+### Genes without annotated categories are discarded
 
@@
 
-## p-value adjustment is applied incorrectly
+### p-value adjustment is applied incorrectly
 
@@
 
-This software was developed by Mark Ziemann, report issues at the [GitHub repository](https://github.com/markziemann/background).
-
-Some information about the parameters used:
-
 ```{r, start}
@@
 genesetlibrary <- params$genesetlibrary
-minsetsize <- 5
-
-message(paste("Minimum gene set size:",minsetsize))
+mygs <- params$mygs
+minsetsize = 5
 
 ```
-
-Minimum set size is defined as 5.
-Sets with fewer than this number of genes are discarded.
-
-## Input gene lists
-
-Here is the first few lines of the input foreground and background lists.
 
@@
 
-```{r,peek}
+### Notes
 
-message("First 10 genes in the foreground")
+Minimum set size is defined as 5.
+Sets with fewer than this number of genes are discarded.
+
+This software was developed by Mark Ziemann, report issues at the [GitHub repository](https://github.com/markziemann/background).
+
+## Input gene lists
+
+First 10 genes in the foreground.
+
+```{r,peek1}
 head(fg,10)
+```
 
-message("First 10 genes in the background")
+First 10 genes in the background.
+
+```{r,peek2}
 head(bg,10)
+```
 
-message("Num genes in foreground")
+Number of genes in the foreground list.
+
+```{r,peek3}
 length(fg)
+```
 
-message("Num genes in background")
+Number of genes in the background list.
+
+```{r,peek4}
 length(bg)
+```
 
-```
+## Gene set library
 
@@
 
+nsets <- length(unique(mygs$term))
+message(paste("Number of sets:",nsets))
 ```
 
-Overlap of genes in the gene set library.
+Summary of gene set length.
 
 ```{r,gsl2}
+summary(table(mygs$term))
+```
 
-print("abc")
+Total number of genes with one or more annotations in the library.
+
+```{r,gsl3}
+length(unique(mygs$gene))
+```
+
+Total number of annotations in the gene set library.
+
+```{r,gsl4}
+nrow(mygs)
+```
+
+Overlap of foreground genes in the gene set library.
+
+```{r,gsl5}
+
+mygs_genes <- unique(mygs$gene)
+
+I <- length(intersect(fg,mygs_genes))
+
+message(paste(I,"from",length(fg)))
+
+```
+
+Overlap of background genes in the gene set library.
+
+```{r,gsl6}
+
+I <- length(intersect(bg,mygs_genes))
+
+message(paste(I,"from",length(bg)))
 
@@
 
-print("run clusterprofiler default")
+options(enrichment_force_universe = FALSE)
+ora <- as.data.frame(enricher(gene = fg ,
+                              universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = mygs,
+                              pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
 
-print("allow download?")
+gr <- as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",2))
+
+ora$ES <- gr/br
+ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
+colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
+colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
+ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
+ora$ES <- signif(ora$ES,3)
+ora$pvalue <- signif(ora$pvalue,3)
+ora$FDR <- signif(ora$FDR,3)
+ora1 <- ora
 
@@
 
-Make bar chart.
+```{r,defaulttable}
+
+DT::datatable(
+  { ora1 },
+
+  extensions = 'Buttons',
+  filter = list(position = 'top', clear = FALSE),
+
+  options = list(
+    paging = TRUE,
+    searching = TRUE,
+    fixedColumns = TRUE,
+    autoWidth = TRUE,
+    ordering = TRUE,
+    dom = 'tB',
+    buttons = c('copy', 'csv', 'excel'),
+    pageLength = 50,
+    search = list(regex = TRUE, caseInsensitive = TRUE)),
+
+  rownames= FALSE)
+
+```
+
+Bar chart.
 
@@
 
-barplot(c(1,2,3))
+edata <- head(subset(ora1,FDR<0.05),20)
+if (nrow(edata)>1) {
+  edata <- edata[order(edata$ES),]
+  vec <- edata$ES
+  names(vec) <- edata$Description
+  names(vec) <- substr(names(vec), start = 1, stop = 60)
+  names(vec) <- gsub("_"," ",names(vec))
+  par(mar=c(4,27,1,1))
+  barplot(vec,horiz=TRUE,las=1,cex.names=0.8,
+          xlab="Enrichment Score")
+  mtext("Default ORA",side=3,adj=1)
+} else {
+  message("No gene sets with FDR<0.05 found")
+}
 
@@
 
-mymessage <- paste("Here we are characterising","comparison")
+mymessage <- paste("Here we are characterising",comparison)
 message(mymessage)
 
-print("run clusterprofiler fixed")
-# actually there will be three different types of analysis possible here
-# depending on the selection
+#Background error
 
-print("allow download?")
+if (comparison == "Background error") {
+  options(enrichment_force_universe = TRUE)
+  ora <- as.data.frame(enricher(gene = fg ,
+                                universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = mygs,
+                                pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+  gr <- as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",2))
+
+  br <- as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",2))
+
+  ora$ES <- gr/br
+  ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
+  colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
+  colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
+  ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
+  ora$ES <- signif(ora$ES,3)
+  ora$pvalue <- signif(ora$pvalue,3)
+  ora$FDR <- signif(ora$FDR,3)
+}
+
+ora2 <- ora
+```
+
+```{r,fixed2}
+
+DT::datatable(
+  { ora2 },
+
+  extensions = 'Buttons',
+  filter = list(position = 'top', clear = FALSE),
+
+  options = list(
+    paging = TRUE,
+    searching = TRUE,
+    fixedColumns = TRUE,
+    autoWidth = TRUE,
+    ordering = TRUE,
+    dom = 'tB',
+    buttons = c('copy', 'csv', 'excel'),
+    pageLength = 50,
+    search = list(regex = TRUE, caseInsensitive = TRUE)),
+
+  rownames= FALSE)
 
@@
 
-Make bar chart.
+Fixed bar chart.
 
@@
 
-barplot(c(1,2,3))
+edata <- head(subset(ora2,FDR<0.05),20)
+if (nrow(edata)>1) {
+  edata <- edata[order(edata$ES),]
+  vec <- edata$ES
+  names(vec) <- edata$Description
+  names(vec) <- substr(names(vec), start = 1, stop = 60)
+  names(vec) <- gsub("_"," ",names(vec))
+  par(mar=c(4,27,1,1))
+  barplot(vec,horiz=TRUE,las=1,cex.names=0.8,
+          xlab="Enrichment Score")
+  mtext("Fixed ORA",side=3,adj=1)
+} else {
+  message("No gene sets with FDR<0.05 found")
+}
 
@@
 
-* Data table
+Side by side data table.
+Columns labeled 'x' are original algorithm, and 'y' are fixed.
 
-* Euler diagram
+```{r,comparison1}
+m <- merge(ora1,ora2,by="Description")
+m <- m[,c("Description","FgRatio.x","FgRatio.y","BgRatio.x","BgRatio.y","ES.x","ES.y","FDR.x","FDR.y")]
+diff <- abs(-log10(m$FDR.x) - -log10(m$FDR.y))
+m <- m[order(-diff),]
+m$ES.x <- signif(m$ES.x,3)
+m$ES.y <- signif(m$ES.y,3)
+m$FDR.x <- signif(m$FDR.x,3)
+m$FDR.y <- signif(m$FDR.y,3)
 
-* Scatterplot of ES
+DT::datatable( { m },
+  extensions = 'Buttons',
+  filter = list(position = 'top', clear = FALSE),
+  options = list(
+    paging = TRUE,
+    searching = TRUE,
+    fixedColumns = TRUE,
+    autoWidth = TRUE,
+    ordering = TRUE,
+    dom = 'tB',
+    buttons = c('copy', 'csv', 'excel'),
+    pageLength = 50,
+    search = list(regex = TRUE, caseInsensitive = TRUE)),
+  rownames= FALSE)
 
-* Scatterplot of logFDR
+```
+
+Euler diagram.
+
+```{r,comparison2}
+
+ora1_sets <- subset(ora1,FDR < 0.05)$Description
+ora2_sets <- subset(ora2,FDR < 0.05)$Description
+
+if ( comparison == "Background error") {
+  v1 <- list("Original"=ora1_sets, "BG fix"=ora2_sets)
+}
+plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))
+
+```
+
+Scatterplot of ES.
+
+```{r,comparison3}
+
+orig_df <- ora1
+bgfix_df <- ora2
+m <- merge(orig_df,bgfix_df,by="Description")
+m <- m[,c("Description","ES.x","ES.y")]
+MAX <- max(c(m$ES.x,m$ES.y))
+INCREMENT <- MAX/nrow(m)
+VEC <- seq(0,MAX,INCREMENT)
+
+if ( comparison == "Background error") {
+
+m$Original <- m$BGcorrected <- round(VEC[1:nrow(m)],0.1)
+m$ES.x <- signif(m$ES.x,3)
+m$ES.y <- signif(m$ES.y,3)
+
+plot_ly(
+  m, x = ~ES.x, y = ~ES.y
+) %>%
+  add_trace(m, x = ~Original, y = ~BGcorrected, type = "scatter",
+            mode="lines", line=list(color='grey')) %>%
+  add_trace(m, x = ~ES.x, y = ~ES.y, type = "scatter", mode = "markers",
+            showlegend = FALSE, text = m$Description,
+            hoverinfo = 'text') %>%
+  layout(title="Enrichment score comparison")
+}
+
+```
+
+Scatterplot of logFDR values.
+
+```{r,comparison4}
+
+m <- merge(ora1,ora2,by="Description")
+m <- m[,c("Description","FDR.x","FDR.y")]
+m$logFDR.x <- -log10(m$FDR.x)
+m$logFDR.y <- -log10(m$FDR.y)
+MAX <- max(c(m$logFDR.x,m$logFDR.y))
+INCREMENT <- MAX/nrow(m)
+VEC <- seq(0,MAX,INCREMENT)
+
+if ( comparison == "Background error") {
+
+  m$Original <- m$BGcorrected <- round(VEC[1:nrow(m)],0.1)
+  m$logFDR.x <- signif(m$logFDR.x,3)
+  m$logFDR.y <- signif(m$logFDR.y,3)
+
+  plot_ly(
+    m, x = ~logFDR.x, y = ~logFDR.y
+  ) %>%
+    add_trace(m, x = ~Original, y = ~BGcorrected, type = "scatter",
+              mode="lines", line=list(color='grey')) %>%
+    add_trace(m, x = ~logFDR.x, y = ~logFDR.y, type = "scatter", mode = "markers",
+              showlegend = FALSE, text = m$Description,
+              hoverinfo = 'text') %>%
+    layout(title="log FDR comparison")
+
+}
+
+```
 

### 2024-07-01 app/app.R

@@ -314,7 +314,7 @@ server <- function(input, output, session) {
     m$logFDR.y <- signif(m$logFDR.y,3)
     
     fig <- plot_ly(
-      m, x = ~ES.x, y = ~ES.y
+      m, x = ~logFDR.x, y = ~logFDR.y
     ) %>%
       add_trace(m, x = ~Original, y = ~BGcorrected, type = "scatter",
                 mode="lines", line=list(color='grey')) %>%
@@ -386,7 +386,8 @@ server <- function(input, output, session) {
       params <- list(fg = fg(),
                      bg = bg(),
                      comparison = input$comparison,
-                     genesetlibrary = input$genesetlibrary)
+                     genesetlibrary = input$genesetlibrary,
+                     mygs = mygs())
       
       rmarkdown::render(tempReport, output_file = file,
                         params = params,

### 2024-07-01 app/report.Rmd
FDR fix

@@
 
-Source: https://github.com/markziemann/background
+This software was developed by Mark Ziemann under an MIT licence.
+Report issues at the [GitHub repository](https://github.com/markziemann/background).
 
@@
 
-### Genes without annotated categories are discarded
+### Background problem: Genes without annotated categories are discarded
 
@@
 
-### p-value adjustment is applied incorrectly
+### FDR problem: p-value adjustment is applied incorrectly
 
@@
 
-```
-
-```{r,checklibraries}
-
 suppressPackageStartupMessages({
@@
 
-This software was developed by Mark Ziemann, report issues at the [GitHub repository](https://github.com/markziemann/background).
+The problem we are characterising is:
+
+```{r,notes}
+
+mymessage <- paste("Here we are characterising",comparison)
+message(mymessage)
+
+```
 
@@
 ```{r, gsl1}
+genesetlibrary
+```
 
-genesetlibrary
+Number of sets.
 
-nsets <- length(unique(mygs$term))
-message(paste("Number of sets:",nsets))
+```{r,gsl1a}
+length(unique(mygs$term))
 ```
@@
 ```{r,gsl2}
-summary(table(mygs$term))
+summary(as.numeric(unname(table(mygs$term))))
 ```
@@
   ora$FDR <- signif(ora$FDR,3)
+} else if (comparison == "FDR error") {
+  gs1 <- mygs
+  gs1 <- gs1[which(gs1$gene %in% bg),]
+  terms <- names(which(table(gs1$term)>5))
+  gs1 <- gs1[gs1$term %in% terms,]
+  options(enrichment_force_universe = FALSE)
+  ora <- as.data.frame(enricher(gene = fg ,
+                                universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = gs1,
+                                pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+  gr <- as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",2))
+
+  br <- as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",2))
+
+  ora$ES <- gr/br
+  ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
+  colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
+  colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
+  ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
+
+  # proper FDR correction
+  nsets <- length(which(table(gs1$term)>5))
+  nres <- nrow(ora)
+  diff <- nsets - nres
+  pvals <- c(ora$pvalue,rep(1,diff))
+  ora$FDR <- p.adjust(pvals,method="fdr")[1:nrow(ora)]
 }
@@
 
-Euler diagram.
+Euler diagram of pathways with FDR<0.05.
 
@@
   v1 <- list("Original"=ora1_sets, "BG fix"=ora2_sets)
+} else if (comparison == "FDR error") {
+  v1 <- list("Original"=ora1_sets, "FDR fix"=ora2_sets)
 }
 plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))
+
+```
+
+Jaccard similarity index of pathways with FDR<0.05.
+
+```{r,jac1}
+
+jac <- length(intersect(ora1_sets,ora2_sets)) / length(union(ora1_sets,ora2_sets))
+
+message(paste("Jaccard index:",signif(jac,3)))
 
@@
     add_trace(m, x = ~Original, y = ~BGcorrected, type = "scatter",
+              mode="lines", line=list(color='grey')) %>%
+    add_trace(m, x = ~ES.x, y = ~ES.y, type = "scatter", mode = "markers",
+              showlegend = FALSE, text = m$Description,
+              hoverinfo = 'text') %>%
+    layout(title="Enrichment score comparison")
+} else if (comparison == "FDR error") {
+  m$Original <- m$FDRcorrected <- round(VEC[1:nrow(m)],0.1)
+  m$ES.x <- signif(m$ES.x,3)
+  m$ES.y <- signif(m$ES.y,3)
+
+  plot_ly(
+    m, x = ~ES.x, y = ~ES.y
+  ) %>%
+    add_trace(m, x = ~Original, y = ~FDRcorrected, type = "scatter",
               mode="lines", line=list(color='grey')) %>%
@@
 
+} else if (comparison == "FDR error") {
+  m$Original <- m$FDRcorrected <- round(VEC[1:nrow(m)],0.1)
+  m$logFDR.x <- signif(m$logFDR.x,3)
+  m$logFDR.y <- signif(m$logFDR.y,3)
+
+  plot_ly(
+    m, x = ~logFDR.x, y = ~logFDR.y
+  ) %>%
+    add_trace(m, x = ~Original, y = ~FDRcorrected, type = "scatter",
+              mode="lines", line=list(color='grey')) %>%
+    add_trace(m, x = ~logFDR.x, y = ~logFDR.y, type = "scatter", mode = "markers",
+              showlegend = FALSE, text = m$Description,
+              hoverinfo = 'text') %>%
+    layout(title="log FDR comparison")
 }

### 2024-07-01 app/app.R

@@ -35,8 +35,7 @@ ui <- fluidPage(
                  "No. genes in foregorund", verbatimTextOutput("summary1"),
                  "No. genes in background", verbatimTextOutput("summary2")),
         tabPanel("Comparative Analysis",
-                 "Comparison of original and BG fixed analysis - top 20 pathways with divergent FDR values.",
-                 "'x' is original, and 'y' is fixed analysis,",
+                 "Comparison of original (x) and fixed (y) analysis - top 20 pathways with divergent FDR values.",
                  DT::dataTableOutput("tbl1")),
         tabPanel("Charts", textOutput("counts"),
                  plotOutput("euler"),
@@ -174,6 +173,43 @@ server <- function(input, output, session) {
     return(ora)
   })
   
+  fdrfix <- reactive({
+    req(fg())
+    req(bg())
+    req(mygs())
+    options(enrichment_force_universe = FALSE)
+    
+    # select sets with 5 in bg
+    gs1 <- mygs()
+    gs1 <- gs1[which(gs1$gene %in% bg()),]
+    terms <- names(which(table(gs1$term)>5))
+    gs1 <- gs1[gs1$term %in% terms,]
+    
+    ora <- as.data.frame(enricher(gene = fg() ,
+                                  universe = bg(), minGSSize = 5, maxGSSize = 500000, TERM2GENE = gs1,
+                                  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+    
+    gr <- as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",1)) /
+      as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",2))
+    
+    br <- as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",1)) /
+      as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",2))
+    
+    ora$ES <- gr/br
+    ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
+    colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
+    colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
+    ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
+    
+    # proper FDR correction
[…]

### 2024-07-01 app/report.Rmd
working

@@
   library("DT")
+  library("kableExtra")
 })
@@
 ora$ES <- gr/br
+ora1_full <- ora
 ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
@@
 DT::datatable(
-  { ora1 },
+  { head(ora1,1000) },
 
@@
 
+```
+
+```{r,defaultfull,results="asis",warning=FALSE}
+
+HEADER=paste("<br><details><summary><b>","Click HERE to show results for top 1000 sets","</b></summary><br><p>",sep=" " )
+
+cat(HEADER)
+
+ora1_full$ID = ora1_full$p.adjust = NULL
+colnames(ora1_full) <- gsub("qvalue","FDR",colnames(ora1_full))
+ora1_full$geneID <- gsub("/"," ",ora1_full$geneID)
+ora1_full <- ora1_full[,c("Description","GeneRatio","BgRatio","Count","ES","pvalue","FDR","geneID")]
+kbl(head(ora1_full,1000), format="html", row.names=FALSE, caption = "Complete results",digits=100) %>%
+kable_styling("hover", full_width = FALSE)
+
+cat("<br></p></details>")
+cat("<hr><br>")
 ```
@@
   ora$ES <- gr/br
+  ora2_full <- ora
   ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
@@
   ora$FDR <- p.adjust(pvals,method="fdr")[1:nrow(ora)]
+  ora$ES <- signif(ora$ES,3)
+  ora$pvalue <- signif(ora$pvalue,3)
+  ora$FDR <- signif(ora$FDR,3)
+} else if (comparison == "Both errors") {
+  gs1 <- mygs
+  gs1 <- gs1[which(gs1$gene %in% bg),]
+  terms <- names(which(table(gs1$term)>5))
+  gs1 <- gs1[gs1$term %in% terms,]
+  options(enrichment_force_universe = TRUE)
+  ora <- as.data.frame(enricher(gene = fg ,
+                                universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = gs1,
+                                pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+  gr <- as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",2))
+
+  br <- as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",1)) /
+    as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",2))
+
+  ora$ES <- gr/br
+  ora2_full <- ora
+  ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
+  colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
+  colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
+  ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
+
+  # proper FDR correction
+  nsets <- length(which(table(gs1$term)>5))
+  nres <- nrow(ora)
+  diff <- nsets - nres
+  pvals <- c(ora$pvalue,rep(1,diff))
+  ora$FDR <- p.adjust(pvals,method="fdr")[1:nrow(ora)]
+  ora$ES <- signif(ora$ES,3)
+  ora$pvalue <- signif(ora$pvalue,3)
+  ora$FDR <- signif(ora$FDR,3)
 }
-
 ora2 <- ora
@@
 DT::datatable(
-  { ora2 },
+  { head(ora2,1000) },
 
@@
   rownames= FALSE)
+
+```
+
+```{r,fixedfull,results="asis"}
+
+HEADER=paste("<br><details><summary><b>","Click HERE to show results for top 1000 sets","</b></summary><br><p>",sep=" " )
+
+cat(HEADER)
+
+ora2_full$ID = ora2_full$p.adjust = NULL
+colnames(ora2_full) <- gsub("qvalue","FDR",colnames(ora2_full))
+ora2_full$geneID <- gsub("/"," ",ora2_full$geneID)
+ora2_full <- ora2_full[,c("Description","GeneRatio","BgRatio","Count","ES","pvalue","FDR","geneID")]
+
+kbl(head(ora2_full,1000), format="html", row.names=FALSE, caption = "Complete results",digits=100) %>%
+kable_styling("hover", full_width = FALSE)
+
+cat("<br></p></details>")
+cat("<hr><br>")
 
@@
 
-```{r,comparison2}
+```{r,comparison2,fig.height=5}
 
@@
   v1 <- list("Original"=ora1_sets, "FDR fix"=ora2_sets)
+} else if (comparison == "Both errors") {
+  v1 <- list("Original"=ora1_sets, "Both fix"=ora2_sets)
 }
@@
     layout(title="Enrichment score comparison")
+} else if (comparison == "Both errors") {
+  m$Original <- m$BOTHcorrected <- round(VEC[1:nrow(m)],0.1)
+  m$ES.x <- signif(m$ES.x,3)
+  m$ES.y <- signif(m$ES.y,3)
+
+  plot_ly(
+    m, x = ~ES.x, y = ~ES.y
+  ) %>%
+    add_trace(m, x = ~Original, y = ~BOTHcorrected, type = "scatter",
+              mode="lines", line=list(color='grey')) %>%
+    add_trace(m, x = ~ES.x, y = ~ES.y, type = "scatter", mode = "markers",
+              showlegend = FALSE, text = m$Description,
+              hoverinfo = 'text') %>%
+    layout(title="Enrichment score comparison")
 }
@@
     layout(title="log FDR comparison")
+} else if (comparison == "Both errors") {
+  m$Original <- m$BOTHcorrected <- round(VEC[1:nrow(m)],0.1)
+  m$logFDR.x <- signif(m$logFDR.x,3)
+  m$logFDR.y <- signif(m$logFDR.y,3)
+
+  plot_ly(
+    m, x = ~logFDR.x, y = ~logFDR.y
+  ) %>%
+    add_trace(m, x = ~Original, y = ~BOTHcorrected, type = "scatter",
+              mode="lines", line=list(color='grey')) %>%
+    add_trace(m, x = ~logFDR.x, y = ~logFDR.y, type = "scatter", mode = "markers",
+              showlegend = FALSE, text = m$Description,
+              hoverinfo = 'text') %>%
+    layout(title="log FDR comparison")
 }

### 2024-07-01 app/app.R

@@ -22,13 +22,13 @@ ui <- fluidPage(
                 accept = c("text/csv",
                            "text/comma-separated-values,text/plain",
                            ".txt")),
-      uiOutput("comparison"),
       uiOutput("genesetlibrary"),
+      uiOutput("comparison"),
       downloadButton("download_report", "Download Report")
     ),
     mainPanel(
       tabsetPanel(
-        tabPanel("Instructions", includeMarkdown("intro.md")),
+        tabPanel("Get started", includeMarkdown("intro.md")),
         tabPanel("Data",
                  "Foreground", verbatimTextOutput("contents1"),
                  "Background",verbatimTextOutput("contents2"),
@@ -177,7 +177,6 @@ server <- function(input, output, session) {
     req(fg())
     req(bg())
     req(mygs())
-    options(enrichment_force_universe = FALSE)
     
     # select sets with 5 in bg
     gs1 <- mygs()
@@ -185,6 +184,45 @@ server <- function(input, output, session) {
     terms <- names(which(table(gs1$term)>5))
     gs1 <- gs1[gs1$term %in% terms,]
     
+
+    options(enrichment_force_universe = FALSE)
+    ora <- as.data.frame(enricher(gene = fg() ,
+                                  universe = bg(), minGSSize = 5, maxGSSize = 500000, TERM2GENE = gs1,
+                                  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+    
+    gr <- as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",1)) /
+      as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",2))
+    
+    br <- as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",1)) /
+      as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",2))
+    
+    ora$ES <- gr/br
+    ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
+    colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
+    colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
+    ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
+    
+    # proper FDR correction
[…]

### 2024-07-02 manuscript/manuscript.Rmd
update

@@
 genes in the list, and the degree of noise in the dataset.
-These problems can be mitigated by changing packages/websites for ORA or by changing to another
-approach such as functional class scoring.
+These problems can be mitigated by changing packages/websites for ORA or by changing to
+another approach such as functional class scoring.
 
@@
 
-* Research integrity
+* Research rigour
 
@@
 at least two genes in the background list prior to running clusterProfiler.
-After clusterProfiler using the parameters above, the number of gene sets in the results was
-quantified, and compared to the number that should have been reported using the two gene
-limit.
+After running clusterProfiler using the parameters above, the number of gene sets in the
+results was quantified, and compared to the number that should have been reported using the
+two gene limit.
 The difference between these numbers (*n*) represents the number of missing gene sets.
@@
 correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
-that ORA should be discouraged in most situations, and replaced by other methods such as FCS.
-Where it makes sense to use ORA would be when FCS is not applicable, for example when
-investigating the over-represented categories in gene expression modules defined by packages
-like WGCNA (Weighted Gene Correlation Network Analysis) [@Langfelder2008-hu], or when
-investigating over-represented functional categories among genes with variants associated
-with a disease.
+that if the data can be scored/ranked, methods like FCS are recommended over ORA.
+ORA is better suited to other cases where genes are classified into binary groups.
+For example when investigating the over-represented categories in gene expression modules
+[@Langfelder2008-hu], or when investigating over-represented functional categories among
+genes with variants associated with a disease [@White2019-ps].
 

### 2024-07-02 app/app.R
fix

@@ -2,7 +2,6 @@ library("clusterProfiler")
 library("markdown")
 library("rmarkdown")
 library("shiny")
-library("broom")
 library("RhpcBLASctl")
 library("eulerr")
 library("plotly")

### 2024-07-02 app/report.Rmd
no broom

@@
   library("shiny")
-  library("broom")
   library("RhpcBLASctl")

### 2024-07-08 manuscript/manuscript.Rmd
bmc bioinfo

@@
 bibliography: references.bib
-csl: plos-computational-biology.csl
+csl: bmc-bioinformatics.csl
 ---
@@
 
-# Abstract
+## Abstract
 
-Over-representation analysis (ORA) is used widely to assess the enrichment of functional
+**Background**: Over-representation analysis (ORA) is used widely to assess the enrichment of functional
 categories in a gene list compared to a background list.
@@
 the true number of parallel tests conducted.
-Here we demonstrate the impact of these issues on several real RNA-seq datasets and use
+
+**Results**: Here we demonstrate the impact of these issues on several real RNA-seq datasets and use
 simulated RNA-seq data to quantify the impact of these problems.
@@
 genes in the list, and the degree of noise in the dataset.
-These problems can be mitigated by changing packages/websites for ORA or by changing to
+
+**Conclusion**: These problems can be mitigated by changing packages/websites for ORA or by changing to
 another approach such as functional class scoring.
 
-# Keywords
+## Keywords
 
@@
 
-# Main Body
-
-## Introduction
+## Background
 
@@
 
-To test the effect of these two problems on RNA-seq based enrichment analysis, data from seven
+To quantify the effect of these two problems on RNA-seq based enrichment analysis, data from seven
 transcriptomic studies were downloaded from DEE2 using the getDEE2 R/Bioconductor package
@@
 
-The second problem is potentially more serious in that it could lead to increasing false
+The FDR problem is potentially more serious in that it could lead to increasing false
 positives under specific conditions; when the foreground list is short, and the median pathway
@@
 ClusterProfiler is not the only tool to suffer from these problems, they are widespread
-(**Table 1**), but there are also several tools free of these problems.
+(**Table 1**). Still, there are several tools free of these problems.
 
@@
 bigger problems noted.
-Firstly, ORA was less accurate as compared to FCS in this simulation work, confirming a
+Firstly, all variations of ORA were less accurate as compared to FCS in this simulation work, confirming a
 previous report [@Kaspi2020-rn].
@@
 
-# Data and Software Availability
+## Declarations
 
-## Underlying data
+### Ethics approval and consent to participate
 
-Publicly available data were obtained from Digital Expression Explorer 2 (dee2.io).
+Not applicable.
 
-## Software and Code
+### Consent for publication
 
-* The code repository including template Dockerfile and R Markdown script are available on
+Not applicable.
+
+### Availability of data and materials
+
+Publicly available data were obtained from Digital Expression Explorer 2 (dee2.io),
+using the accession numbers mentioned in the Methods section.
+The code repository including template Dockerfile and R Markdown script are available on
 GitHub (https://github.com/markziemann/background).
+The Docker image is available on DockerHub (https://hub.docker.com/r/mziemann/background).
+The code repository and Docker image have been uploaded to Zenodo for long-term archiving
+(https://zenodo.org/record/TODOXXX).
 
-* The example Docker image is available on DockerHub
-(https://hub.docker.com/r/mziemann/background).
+## Competing Interests
 
-* The code repository and Docker image have been uploaded to Zenodo for long-term archiving
-(https://zenodo.org/record/8170984).
+The authors declare that they have no competing interests.
 
-## Availability of other materials
+## Funding
+
+This work was supported by Burnet Institute and Deakin University.
+Authors declare that no grants were involved in supporting this work.
 
@@
 
-## Competing Interests
-
-No competing interests were disclosed.
-
-## Grant Information
-
-The authors declared that no grants were involved in supporting this work.
-
 ## Acknowledgements
@@
 
-# Bibliography
+## Bibliography

### 2024-07-09 analysis/analyse_rnaseq1.Rmd
pdf1

@@
 
+pdf("fig1_ORAsig.pdf",width=7,height=7)
+par(mar=c(c(5.1, 7.1, 2.1, 2.1) ))
+barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,1250),
+  xlab="no. significant sets",
+  col=c("gray30","gray80"))
+text(df$ns1+50,((1:nrow(df))*3)-1.5,labels=signif(df[,1],3))
+text(df$ns2+50,((1:nrow(df))*3)-0.5,labels=signif(df[,2],3))
+#abline(v=seq(100,1000,100),lty=2,lwd=0.5,col="gray")
+legend("bottomright", inset=.02, legend=c("corrected","original"),
+  fill=c("gray80","gray30"), horiz=FALSE, cex=1.1)
+
+mylabels <- gsub("$","%",gsub("^","+",signif(((df$ns2/df$ns1)-1)*100,3)))
+text(df$ns2+160,((1:nrow(df))*3)-1,labels=mylabels,cex=1.1)
+dev.off()
+
 # JACCARD
@@
 
+pdf("fig1_ORAjac.pdf",width=7, height=7)
+par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
+plot(cov,jac2,xlim=c(0,30000),ylim=c(0,1),pch=19,col="lightgray",cex=3,
+  xlab="gene coverage",ylab="Jaccard index")
+text(cov,jac2,labels=names(cov))
+mylm <- lm(jac2~cov)
+abline(mylm,lty=2,lwd=2)
+SLOPE=unname(signif(mylm$coefficients[2],3))
+INT=unname(signif(mylm$coefficients[1],3))
+PVAL=signif(summary(mylm)$coefficients[4],3)
+RSQ=signif(summary(mylm)$r.squared,3)
+HEADER=paste("Rsq =",RSQ,"; p-val =",PVAL)
+mtext(HEADER)
+dev.off()
 ```

### 2024-07-09 analysis/analyse_rnaseq2.Rmd
pdf2

@@
 
+pdf("fig2_ORAfdr.pdf",width=7,height=7)
+par(mar=c(c(5.1, 7.1, 2.1, 2.1) ))
+barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,330),
+  xlab="no. significant sets",
+  col=c("gray30","gray80"))
+text(df$ns1+20,((1:nrow(df))*3)-1.5,labels=signif(df[,1],3))
+text(df$ns2+20,((1:nrow(df))*3)-0.5,labels=signif(df[,2],3))
+legend("bottomright", inset=.02, legend=c("corrected","original"),
+  fill=c("gray80","gray30"), horiz=FALSE, cex=1)
+dev.off()
+
 # RATIO
@@
 png("fig2_ORAsz.png")
-
 plot(jac[,"GO"],type="b", pch=19,cex=1.5,lwd=2,
@@
        col=c("black", "red", "blue", "gray50"), lty=1, cex=1.2,lwd=3,pch=19)
+dev.off()
 
+pdf("fig2_ORAsz.pdf",width=7, height=7)
+plot(jac[,"GO"],type="b", pch=19,cex=1.5,lwd=2,
+  xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected")
+axis(1, at = seq(1, 5, by = 1),
+     labels = ngenes)
+points(jac[,"Reactome"],type="b",col="red",pch=19,cex=1.5,lwd=2)
+points(jac[,"Cellmarkers"],type="b",col="blue",pch=19,cex=1.5,lwd=2)
+points(jac[,"TFT GTRD"],type="b",col="gray50",pch=19,cex=1.5,lwd=2)
+
+legend("bottomright", legend=c("GO","Reactome", "Cellmarkers", "TFT"),
+       col=c("black", "red", "blue", "gray50"), lty=1, cex=1.2,lwd=3,pch=19)
 dev.off()

### 2024-07-09 analysis/sim/run_simpw.Rmd
pdf3

@@
 
+#################################
+
+pdf("fig3_sim.pdf", width=7,height=5,pointsize=12)
+par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
+par(mfrow=c(1,3))
+
+plot(cp$VARIANCE,cp$p,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="precision")
+points(cpbg$VARIANCE,cpbg$p,type="b",pch=19,col="orange")
+points(cpfdr$VARIANCE,cpfdr$p,type="b",pch=19,col="darkgreen")
+points(cpbgfdr$VARIANCE,cpbgfdr$p,type="b",pch=19,col="purple")
+points(fo$VARIANCE,fo$p,type="b",pch=19,col="red")
+points(fg$VARIANCE,fg$p,type="b",pch=19,col="blue")
+grid()
+legend("bottomleft", inset=.02, title="tool",
+   c("CP default","CP BG fix","CP FDR fix","CP BG and FDR fix","fora","fgsea"),
+   col=c("black","orange","darkgreen","purple","red","blue"),
+   horiz=FALSE, cex=1.0, pch=19,lwd=2)
+
+plot(cp$VARIANCE,cp$r,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="recall")
+points(cpbg$VARIANCE,cpbg$r,type="b",pch=19,col="orange")
+points(cpfdr$VARIANCE,cpfdr$r,type="b",pch=19,col="darkgreen")
+points(cpbgfdr$VARIANCE,cpbgfdr$r,type="b",pch=19,col="purple")
+points(fo$VARIANCE,fo$r,type="b",pch=19,col="red")
+points(fg$VARIANCE,fg$r,type="b",pch=19,col="blue")
+grid()
+
+plot(cp$VARIANCE,cp$f,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="f1")
+points(cpbg$VARIANCE,cpbg$f,type="b",pch=19,col="orange")
+points(cpfdr$VARIANCE,cpfdr$f,type="b",pch=19,col="darkgreen")
+points(cpbgfdr$VARIANCE,cpbgfdr$f,type="b",pch=19,col="purple")
+points(fo$VARIANCE,fo$f,type="b",pch=19,col="red")
+points(fg$VARIANCE,fg$f,type="b",pch=19,col="blue")
+grid()
+dev.off()
+
 ```
@@
 
+#################################
+pdf("fig3_bars.pdf", width=7,height=5,pointsize=12)
+par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
+par(mfrow=c(1,3))
+barplot(prec,ylim=c(0,1.1),las=2,main="precision",ylab="")
+text((1:6*1.18)-0.45,prec+0.02,labels=signif(prec,3))
+barplot(rec,ylim=c(0,1.1),las=2,main="recall",ylab="")
+text((1:6*1.18)-0.45,rec+0.02,labels=signif(rec,3))
+barplot(f1,ylim=c(0,1.1),las=2,main="F1",ylab="")
+text((1:6*1.18)-0.45,f1+0.02,labels=signif(f1,3))
+dev.off()
+
 ```

### 2024-07-09 manuscript/manuscript.Rmd
ms

@@
 
-(*) Corresponding author: mark.ziemann@burnet.edu.au
+(*) Corresponding author: mark.ziemann@burnet.edu.au Burnet Institute, 85 Commercial Rd Melbourne VIC 3004, Australia.
 
@@
 ~2020 where confused users were wondering why the number of background genes changed so much.
-The results shown here show that the background problem affects smaller gene set libraries
+The results shown here indicate the background problem affects smaller gene set libraries
 like KEGG and HPO more severely as compared to Cellmarkers and GO libraries which describe
@@
 
+So end-users can explore the effect of these two problems on their work, we have developed
+an R/Shiny web application.
+Users can upload their foreground and background lists, select their preferred gene set
+library and compare results from default and patched analyses.
+The resource is currently available at https://oratool.ziemann-lab.net/ and available as a
+docker container (https://hub.docker.com/repository/docker/mziemann/background_app/general).
+
 ClusterProfiler is not the only tool to suffer from these problems, they are widespread
@@
 
-The purpose of this work was to describe two subtle problems with ORA, but there are some
+The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there are some
 bigger problems noted.
@@
 
-## Competing Interests
+### Competing Interests
 
@@
 
-## Funding
+### Funding
 
@@
 
-## Author Contributions
+### Author Contributions
 
@@
 
-## Acknowledgements
+### Acknowledgements
 

### 2024-07-10 manuscript/manuscript.Rmd
ms work

@@
 In extreme cases, correction of this problem can lead to around half of the statistically
-significant results changing.
+significant results changing (see Figure 2B).
 
@@
 ClusterProfiler is not the only tool to suffer from these problems, they are widespread
-(**Table 1**). Still, there are several tools free of these problems.
+(**Table 1**). Still, there are some tools free of such problems.
+For example, fora is a good option for R-based workflows, however the function does not calculate the fold enrichment score,
+so users will need to do it themselves [@Korotkevich2016-gd].
+ShinyGO is an excellent alternative for web based analysis [@Ge2020-lu].
+It doesn't suffer from the two issues described here, it is easy to use,
+and has a high degree of reproducibility thanks to its stand-alone versioned docker
+container option for local execution.
+One problem that all of these tools listed in Table 1 have is that a background list is optional.
+Sometimes this feature is hidden under "advanced options".
+Given the strong biological and technical biases that are present in contemporary omics such as including single cell and spatial transcriptomics,
+it should be mandatory step for users to provide a custom background list.
 
-| Tool | Has problems | Citation |
-| --- | --- | --- |
-| ClusterProfiler | Background and FDR | [@Yu2012-jj;@Wu2021-wy] |
-| DAVID | Background and FDR | [@Dennis2003-ar;@Huang2009-gi;@Sherman2022-jn] |
-| fora (fgsea) | Neither | [@Korotkevich2016-gd] |
+| Tool | Version | Type | Provides FDR values | Provides enrichment score | Proper background handling | Proper FDR | Reference |
+| --- | --- | --- | --- | --- | --- | --- | --- |
+| DAVID | v2023q4 | Web | Yes | Yes | No | Yes | [@Sherman2022-jn] |
+| Panther | v19 | Web | Yes | Yes | Yes | Yes | [@Mi2019-ak] |
+| Enrichr | June 2023 | Web | Yes | Yes | Yes | No | [@Kuleshov2016-og] |
+| KOBAS-i | KOBAS 3.0 | Web | Yes | No | Yes | No | [@Bu2021-ak] |
+| WebGestalt | 2024 | Web | Yes | Yes | No | Yes | [@Elizarraras2024-nr] |
+| g:Profiler | Feb 2024 | Web | Yes | No | No | Yes | [@Kolberg2023-nr] |
+| STRING-DB | v12.0 | Web | Yes | Yes | ? | ? | [@Szklarczyk2023-an] |
+| ShinyGO | v0.80 | Web | Yes | Yes | Yes | Yes | [@Ge2020-lu] |
+| Ingenuity Pathway Analysis | ? | ? | ? | ? | ? | ? | [@Kramer2014-kd] |
+| BinGO | v3.0.5 | Cytoscape | Yes | No | No | No | [@Maere2005-gq] |
+| ClueGO | v2.5.10 | Cytoscape | Yes | No | No | No | [@Bindea2009-jn] |
+| clusterProfiler | v4.12.0 | R package | Yes | No | No | No | [@Wu2021-wy] |
+| topGO | v2.56.0 | R package | No | No | No | Yes+ | [@Adrian_Alexa2017-nf] |
+| GOseq | v1.56.0 | R package | No | No | No* | Yes+ | [@Young2010-iw] |
+| goana/kegga | limma v3.60.3 | R package | No | No | No** | Yes+ | [@Ritchie2015-oz] |
+| fora | fgsea v1.30.0 | R package | Yes | No | Yes | Yes | [@Korotkevich2016-gd] |
 
-: Table 1. Classification of several popular ORA tools for Background and FDR problems.
+: Table 1. Classification of several popular ORA tools regarding the provision of information such as
+enrichment scores and FDR values, as well as behaviours regarding the two algorithmic issues described above.
+(*) Denotes functions where options are provided to adjust this behaviour and are documented in the user manual.
+(**) Documentation states unannotated genes are discarded with default settings using whole genome
+background, but not when a custom background is used.
+(+) Denotes that although no FDR values are provided, the results include sets where no overlaps were found, so users can adjust raw p-values themselves.
 
@@
 For example when investigating the over-represented categories in gene expression modules
-[@Langfelder2008-hu], or when investigating over-represented functional categories among
-genes with variants associated with a disease [@White2019-ps].
+[@Langfelder2008-hu], or with variant-harbouring genes associated with a disease [@White2019-ps].
 

### 2024-07-10 manuscript/manuscript.Rmd
works
refine ms

@@
 The second problem is that adjusting p-values for multiple testing, also known as
-false-discovery rate correction, is implemented improperly.
+false-discovery rate correction, is sometimes implemented improperly.
 Determining which pathways should be subject to false discovery rate correction depends on
@@
 
-![**Figure 1B. Impact of background problem is worse for smaller gene set libraries.** Mean
+![**Figure 1B. Impact of the background problem is worse for smaller gene set libraries.** Mean
 Jaccard similarity index was calculated for original and corrected ORA for seven independent
@@
 
-![**Figure 2B. Impact of FDR problem is worse for shorter foreground gene lists.** Mean Jaccard
+![**Figure 2B. Impact of the FDR problem is worse for shorter foreground gene lists.** Mean Jaccard
 index between original and corrected analysis were calculated for seven RNA-seq datasets and nine
@@
 Given the strong biological and technical biases that are present in contemporary omics such as including single cell and spatial transcriptomics,
-it should be mandatory step for users to provide a custom background list.
+it should be a mandatory step for users to provide a custom background list.
 
@@
 Enrichment scores are a surrogate for effect size, and without them, users may over-rely
-on significance scores when interpreting enrichment results [@Ziemann2024-ts].
+on significance scores when interpreting enrichment results.
 Lastly, ORA results can vary dramatically depending on the number of genes selected in the

### 2024-07-11 manuscript/manuscript.Rmd
string db

@@
 | g:Profiler | Feb 2024 | Web | Yes | No | No | Yes | [@Kolberg2023-nr] |
-| STRING-DB | v12.0 | Web | Yes | Yes | ? | ? | [@Szklarczyk2023-an] |
+| STRING-DB | v12.0 | Web | Yes | Yes | Yes | Yes | [@Szklarczyk2023-an] |
 | ShinyGO | v0.80 | Web | Yes | Yes | Yes | Yes | [@Ge2020-lu] |

### 2024-07-12 manuscript/manuscript.Rmd
omit IPA

--- 2024-07-10 manuscript/manuscript.Rmd
@@
 | g:Profiler | Feb 2024 | Web | Yes | No | No | Yes | [@Kolberg2023-nr] |
-| STRING-DB | v12.0 | Web | Yes | Yes | ? | ? | [@Szklarczyk2023-an] |
+| STRING-DB | v12.0 | Web | Yes | Yes | Yes | Yes | [@Szklarczyk2023-an] |
 | ShinyGO | v0.80 | Web | Yes | Yes | Yes | Yes | [@Ge2020-lu] |
-| Ingenuity Pathway Analysis | ? | ? | ? | ? | ? | ? | [@Kramer2014-kd] |
 | BinGO | v3.0.5 | Cytoscape | Yes | No | No | No | [@Maere2005-gq] |
@@
 | goana/kegga | limma v3.60.3 | R package | No | No | No** | Yes+ | [@Ritchie2015-oz] |
-| fora | fgsea v1.30.0 | R package | Yes | No | Yes | Yes | [@Korotkevich2016-gd] |
+| fora | fgsea v1.30.0 | R package | Yes | No^ | Yes | Yes | [@Korotkevich2016-gd] |
 
-: Table 1. Classification of several popular ORA tools regarding the provision of information such as
+: Table 1. Classification of several popular freely available ORA tools regarding the provision of information such as
 enrichment scores and FDR values, as well as behaviours regarding the two algorithmic issues described above.
@@
 (+) Denotes that although no FDR values are provided, the results include sets where no overlaps were found, so users can adjust raw p-values themselves.
+(^) Denotes enrichment scores will be included in the next Bioconductor release.
 
@@
 on significance scores when interpreting enrichment results.
-Lastly, ORA results can vary dramatically depending on the number of genes selected in the
+ORA results can vary dramatically depending on the number of genes selected in the
 foreground, a mostly arbitrary choice.

### 2024-07-17 analysis/GSE158420_rna.Rmd
downsampling for sensitivity

---
title: "GSE158420: RNA expression downsampling"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/background/

## Introduction

Here, we look at the sensitivity of pathway analysis of RNA-seq data.

```{r,packages}

suppressPackageStartupMessages({
  library("DESeq2")
  library("eulerr")
  library("HGNChelper")
  library("mitch")
  library("kableExtra")
  library("beeswarm")
  library("gplots")
  library("gridExtra")
  library("png")
  library("fgsea")
  library("parallel")
})

CORES=16

```

## Load annotations and methylation data

Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.

```{r,gsdl}

gs <- gmt_import("../gmt/c2.cp.reactome.v2023.2.Hs.symbols.gmt")

names(gs) <- gsub("REACTOME_","",names(gs))
names(gs) <- gsub("_"," ",names(gs))

```

## Load RNA expression

From GEO https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE158420
GSE158420_counts.txt.gz

```{r,deseq1}

if (!file.exists("GSE158420_counts.txt.gz")) {
  download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE158420&format=file&file=GSE158420%5Fcounts%2Etxt%2Egz",
    destfile="GSE158420_counts.txt.gz")
}

x <- read.table("GSE158420_counts.txt.gz",row.names=NULL)
dim(x)
x <- aggregate(. ~ row.names,x,sum)
dim(x)

#RUFKM
head(x)

```

There are some gene name errors, so I will use HGNChelper to repair them.

```{r,gene_name_fix}

human=x$row.names
new.hgnc.table <- readRDS("new.hgnc.table.rds")
fix <- checkGeneSymbols(human,map=new.hgnc.table)
x$gene <- fix$Suggested.Symbol
x$row.names=NULL
x <- aggregate(. ~ gene,x,sum)
dim(x)
row.names(x)=x$gene
x$gene=NULL
head(x)

```

Now do DESeq2

```{r,deseq2}

patient <- head(as.character(unlist(lapply(1:ncol(x),function(i) {c(i,i)}))),ncol(x))
tumor <- rep(c(0,1),ncol(x)/2)
ss <- data.frame(patient,tumor)
rownames(ss) <- colnames(x)

dds <- DESeqDataSetFromMatrix(countData = x , colData=ss, design = ~ patient + tumor)
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <-cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])

sig <- subset(dge,padj<0.05)
SIG = nrow(sig)
DN = nrow(subset(sig,log2FoldChange<0))
UP = nrow(subset(sig,log2FoldChange>0))
HEADER = paste("mRNA", SIG , "DEGs,", UP ,"upregulated,", DN, "downregulated")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3,
  xlab="log2 base mean",
  ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 ,xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- rep(c(0,1),ncol(x)/2)
colCols <- gsub("0","gray",colCols)
colCols <- gsub("1","black",colCols)

colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,
  trace="none",margins = c(6,10), cexRow=.6, cexCol=.5, main="Top 50 genes")

```

RNA expression pathway enrichment with mitch

```{r,mitch}

rna <- mitch_import(dge, DEtype="deseq2")

res <- mitch_calc(rna, genesets=gs, priority="effect",cores=16, minsetsize=5)

mres <- res$enrichment_result
msig <- subset(mres,p.adjustANOVA<0.05)
nrow(msig)

head(subset(mres,p.adjustANOVA<0.05 & s.dist>0),20) %>%
  kbl(caption="mRNA upregulated pathways") %>%
  kable_styling("hover",full_width=FALSE)

head(subset(mres,p.adjustANOVA<0.05 & s.dist<0),20) %>%
  kbl(caption="mRNA downregulated pathways") %>%
  kable_styling("hover",full_width=FALSE)

```

FGSEA

```{r,fgsea}

dge$stat[which(is.na(dge$stat))] <- 1

stat <- dge$stat
names(stat) <- rownames(dge)

pll <- lapply(1:32,function(i) {
  tic()
  fres <- fgsea(pathways=gs,stats=stat,minSize=5, nproc=i)
  tm <- toc()
  tm$callback_msg
})

pll <- unlist(pll)
pll <- as.numeric(sapply(strsplit(pll," "),"[[",1))

names(pll) <- 1:32
barplot(pll)
plot(1:32,pll,type="b",pch=19)

fsig <- subset(fres,padj<0.05)
fres <- fres[order(-abs(fres$ES)),]

head(subset(fres,padj<0.05 & ES>0),20) %>%
  kbl(caption="mRNA upregulated pathways") %>%
  kable_styling("hover",full_width=FALSE)

head(subset(fres,padj<0.05 & ES<0),20) %>%
  kbl(caption="mRNA downregulated pathways") %>%
  kable_styling("hover",full_width=FALSE)

fsets <- paste(fsig$pathway,as.character(fsig$ES > 0))

```

FORA

```{r,fora1}

sigup <- rownames(subset(dge,padj<0.05 & log2FoldChange>0))
sigdn <- rownames(subset(dge,padj<0.05 & log2FoldChange<0))
bg <- rownames(dge)

oresup <- as.data.frame(fora(pathways=gs,genes=sigup,universe=bg,minSize=5))
oresup$es <- (oresup$overlap / length(sigup) ) / (oresup$size / length(bg))
head(oresup,10)

oresdn <- as.data.frame(fora(pathways=gs,genes=sigdn,universe=bg,minSize=5))
oresdn$es <- (oresdn$overlap / length(sigdn) ) / (oresdn$size / length(bg))
head(oresup,10)

oresup <- subset(oresup,padj<0.05)
oresup <- paste(oresup$pathway,"TRUE")

oresdn <- subset(oresdn,padj<0.05)
oresdn <- paste(oresdn$pathway,"FALSE")

osets <- c(oresup,oresdn)
str(osets)

```

Assess the effect of gene number on pathway results.

```{r,fora2}

sizes <- c(100,200,500,1000,2000,3000,4000,5000,6000,7000,8000)

fres <- lapply(sizes, function(size) {
  sigup <- head(rownames(subset(dge,padj<0.05 & log2FoldChange>0)),size)
  sigdn <- head(rownames(subset(dge,padj<0.05 & log2FoldChange<0)),size)
  bg <- rownames(dge)

  oresup <- as.data.frame(fora(pathways=gs,genes=sigup,universe=bg,minSize=5))
  oresup$es <- (oresup$overlap / length(sigup) ) / (oresup$size / length(bg))
  nup <- nrow(subset(oresup,padj<0.05))

  oresdn <- as.data.frame(fora(pathways=gs,genes=sigdn,universe=bg,minSize=5))
  oresdn$es <- (oresdn$overlap / length(sigdn) ) / (oresdn$size / length(bg))
  ndn <- nrow(subset(oresdn,padj<0.05))

  return(c(nup,ndn))
})

fres <- do.call(rbind,fres)
rownames(fres) <- sizes
colnames(fres) <- c("up","dn")
fres <- as.data.frame(fres)
fres$tot <- fres$up + fres$dn
tot <- fres$tot
names(tot) <- sizes
par(mar=c(mar = c(5.1, 5.1, 4.1, 2.1)))
barplot(tot,horiz=TRUE,las=1,xlab="no. pathways FDR<0.05",ylab="no. genes selected")

```

## Downsample FGSEA

This will run the LAM analysis at group sizes of 2, 3, 6, 10, 16, 22 and 30,
50 times each, with pseudorandom seeds from 100 to 5000 jumping up by 100.

```{r,fgsea_run,fig.height=10}

RhpcBLASctl::blas_set_num_threads(1) #set automatic cores to 1

SAMPLESIZES=c(2,3,6,10,16,22,30)
#SAMPLESIZES=c(3,6,10,22)
ssres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,5000,100) # 50 repetitions
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    ss2 <- ss[mysample,]
    x2 <- x[,mysample]
    x2 <- x2[rowMeans(x2)>=10,]

    dds <- DESeqDataSetFromMatrix(countData = x2 , colData=ss2, design = ~ patient + tumor)
    res <- DESeq(dds)
    z <- results(res)
    vsd <- vst(dds, blind=FALSE)
    zz <-cbind(as.data.frame(z),assay(vsd))
    dge <- as.data.frame(zz[order(zz$pvalue),])

    dge$stat[which(is.na(dge$stat))] <- 1
    stat <- dge$stat
    names(stat) <- rownames(dge)
    fres <- fgsea(pathways=gs,stats=stat,minSize=5,nproc=3)
    fres <- fres[order(-abs(fres$ES)),]
    fsig <- subset(fres,padj<0.05)
    fsets2 <- paste(fsig$pathway,as.character(fsig$ES > 0))

    TOT=length(fsets2)
    TP=length(which(fsets2 %in% fsets))
    FP=length(which(!fsets2 %in% fsets))
    FN=length(which(!fsets %in% fsets2))
    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
    RES
  },mc.cores=16)
  do.call(rbind,dsres)
})

par(mfrow=c(3,1))

dots <- lapply(ssres,function(x) { x[,"TP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fgsea",ylim=c(0,nrow(fsig))) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(fsig))

dots <- lapply(ssres,function(x) { x[,"FP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
lapply(ratio,summary)
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

pdf("sens_fgsea.pdf",height=7,width=5)

dots <- lapply(ssres,function(x) { x[,"TP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fgsea",ylim=c(0,nrow(fsig))) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(fsig))

dots <- lapply(ssres,function(x) { x[,"FP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
lapply(ratio,summary)
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

dev.off()

```

## FORA Downsample

```{r,fora_run,fig.height=10}

SAMPLESIZES=c(2,3,6,10,16,22,30)
ssres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,5000,100) # 50 repetitions
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    ss2 <- ss[mysample,]
    x2 <- x[,mysample]
    x2 <- x2[rowMeans(x2)>=10,]

    dds <- DESeqDataSetFromMatrix(countData = x2 , colData=ss2, design = ~ patient + tumor)
    res <- DESeq(dds)
    z <- results(res)
    vsd <- vst(dds, blind=FALSE)
    zz <-cbind(as.data.frame(z),assay(vsd))
    dge <- as.data.frame(zz[order(zz$pvalue),])

    sigup <- rownames(subset(dge,padj<0.05 & log2FoldChange>0))
    if (length(sigup) < 250) { sigup <- head(rownames(subset(dge, log2FoldChange>0)),250) }
    sigdn <- rownames(subset(dge,padj<0.05 & log2FoldChange<0))
    if (length(sigdn) < 250) { sigdn <- head(rownames(subset(dge, log2FoldChange<0)),250) }
    bg <- rownames(dge)

    oresup <- as.data.frame(fora(pathways=gs,genes=sigup,universe=bg,minSize=5))
    oresup <- subset(oresup,padj<0.05)
    oresup <- paste(oresup$pathway,"TRUE")

    oresdn <- as.data.frame(fora(pathways=gs,genes=sigdn,universe=bg,minSize=5))
    oresdn <- subset(oresdn,padj<0.05)
    oresdn <- paste(oresdn$pathway,"FALSE")

    osets2 <- c(oresup,oresdn)

    TOT=length(osets2)
    TP=length(which(osets2 %in% osets))
    FP=length(which(!osets2 %in% osets))
    FN=length(which(!osets %in% osets2))
    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
    RES
  },mc.cores=16)
  do.call(rbind,dsres)
})

par(mfrow=c(3,1))

dots <- lapply(ssres,function(x) { x[,"TP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fora",ylim=c(0,nrow(fsig))) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(fsig))

dots <- lapply(ssres,function(x) { x[,"FP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
lapply(ratio,summary)
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

pdf("sens_fora.pdf",height=7,width=5)

dots <- lapply(ssres,function(x) { x[,"TP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fora",ylim=c(0,nrow(fsig))) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(fsig))

dots <- lapply(ssres,function(x) { x[,"FP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
lapply(ratio,summary)
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

dev.off()

## Session information

```{r,save}

save.image("GSE158422_sensitivity.Rdata")

sessionInfo()

```

### 2024-07-17 analysis/GSE158420_rna.Rmd
fix figure
plotting

@@
 
-fres <- lapply(sizes, function(size) {
+ores <- lapply(sizes, function(size) {
   sigup <- head(rownames(subset(dge,padj<0.05 & log2FoldChange>0)),size)
@@
 
-fres <- do.call(rbind,fres)
-rownames(fres) <- sizes
-colnames(fres) <- c("up","dn")
-fres <- as.data.frame(fres)
-fres$tot <- fres$up + fres$dn
+ores <- do.call(rbind,ores)
+rownames(ores) <- sizes
+colnames(ores) <- c("up","dn")
+ores <- as.data.frame(ores)
+ores$tot <- ores$up + ores$dn
 tot <- fres$tot
@@
 
-SAMPLESIZES=c(2,3,6,10,16,22,30)
+SAMPLESIZES=c(2,3,5,10,15,20,25,30)
 #SAMPLESIZES=c(3,6,10,22)
-ssres <- lapply(SAMPLESIZES,function(n) {
+fgres <- lapply(SAMPLESIZES,function(n) {
   SEEDS <- seq(100,5000,100) # 50 repetitions
@@
 
+ssres <- fgres
 par(mfrow=c(3,1))
@@
 lapply(ratio,summary)
-boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0,ylim=c(0,0.5)) ; grid()
 beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
@@
 pdf("sens_fgsea.pdf",height=7,width=5)
+par(mfrow=c(3,1))
 
@@
 dev.off()
+
+ssres <- fgres
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+fgres2 <- do.call(rbind,lapply(dots,summary))
 
@@
 
-SAMPLESIZES=c(2,3,6,10,16,22,30)
-ssres <- lapply(SAMPLESIZES,function(n) {
+SAMPLESIZES=c(2,3,5,10,15,20,25,30)
+fores <- lapply(SAMPLESIZES,function(n) {
   SEEDS <- seq(100,5000,100) # 50 repetitions
@@
 
+ssres <- fores
+
 par(mfrow=c(3,1))
@@
 lapply(dots,summary)
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fora",ylim=c(0,nrow(fsig))) ; grid()
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fora",ylim=c(0,length(osets))) ; grid()
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
-abline(h=nrow(fsig))
+abline(h=length(osets))
 
@@
 pdf("sens_fora.pdf",height=7,width=5)
+par(mfrow=c(3,1))
 
@@
 
+## Summarise downnsampling
+
+```{r,downsample,fig.width=10}
+
+par(mfrow=c(1,3))
+
+dots <- lapply(fgres,function(x) { x[,"TP"] })
+fgres2 <- do.call(rbind,lapply(dots,summary))
+
+dots <- lapply(fores,function(x) { x[,"TP"] })
+fores2 <- do.call(rbind,lapply(dots,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+MAX=length(fsets)
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways", ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+abline(h=length(osets),col="red")
+abline(h=length(fsets),col="blue")
+mtext("no. consistent pathways FDR<0.05")
+grid()
+
+legend("bottomright", legend=c("fgsea","fora"),
+       col=c("blue","red"), lty=1, pch=19,  cex=1)
+
+dots <- lapply(fgres,function(x) { x[,"FP"] })
+fgres2 <- do.call(rbind,lapply(dots,summary))
+
+dots <- lapply(fores,function(x) { x[,"FP"] })
+fores2 <- do.call(rbind,lapply(dots,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+mtext("no. inconsistent pathways FDR<0.05")
+grid()
+
+dots <- lapply(fores,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(fores,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+fores2 <- do.call(rbind,lapply(ratio,summary))
+
+dots <- lapply(fgres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(fgres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+fgres2 <- do.call(rbind,lapply(ratio,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="proportion",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+mtext("Proportion of inconsistent pathways")
+grid()
+
+### PDF
+pdf("sens.pdf",height=4,width=8)
+par(mfrow=c(1,3))
+
+dots <- lapply(fgres,function(x) { x[,"TP"] })
+fgres2 <- do.call(rbind,lapply(dots,summary))
+
+dots <- lapply(fores,function(x) { x[,"TP"] })
+fores2 <- do.call(rbind,lapply(dots,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+MAX=length(fsets)
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways", main="no. consistent pathways FDR<0.05",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+abline(h=length(osets),col="red")
+abline(h=length(fsets),col="blue")
+grid()
+
+legend("bottomright", legend=c("fgsea","fora"),
+       col=c("blue","red"), lty=1, pch=19,  cex=1)
+
+dots <- lapply(fgres,function(x) { x[,"FP"] })
+fgres2 <- do.call(rbind,lapply(dots,summary))
+
+dots <- lapply(fores,function(x) { x[,"FP"] })
+fores2 <- do.call(rbind,lapply(dots,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways",main="no. inconsistent pathways FDR<0.05",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+grid()
+
+dots <- lapply(fores,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(fores,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+fores2 <- do.call(rbind,lapply(ratio,summary))
+
+dots <- lapply(fgres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(fgres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+fgres2 <- do.call(rbind,lapply(ratio,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="proportion",main="Proportion of inconsistent pathways",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+grid()
+dev.off()
+
+```
 ## Session information

### 2024-07-17 analysis/GSE158420_rna.Rmd
cores

@@
   library("parallel")
+  library("RhpcBLASctl")
 })
@@
 ```{r,deseq2}
+
+RhpcBLASctl::blas_set_num_threads(4) #set automatic cores to 1
 

### 2024-07-17 analysis/GSE158420_rna.Rmd
lib

@@
   library("RhpcBLASctl")
+  library("tictoc")
 })

### 2024-07-17 analysis/GSE158420_rna.Rmd
add data
fix

@@
 
+fres <- fgsea(pathways=gs,stats=stat,minSize=5, nproc=8)
+
 pll <- unlist(pll)
@@
 ores$tot <- ores$up + ores$dn
-tot <- fres$tot
+tot <- ores$tot
 names(tot) <- sizes
@@
 ssres <- fgres
-dots <- lapply(ssres,function(x) { x[,"TP"] })
-fgres2 <- do.call(rbind,lapply(dots,summary))
 
@@
 
+```
+
 ## Summarise downnsampling
@@
 
+df
+
 MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))

### 2024-07-17 analysis/GSE158420_rna.Rmd
mds looks okay
show data
add data
data
thincounts

@@
 
-## Load annotations and methylation data
+## Load pathway data
 
@@
 
-Now do DESeq2
+MDS plot
 
-```{r,deseq2}
-
-RhpcBLASctl::blas_set_num_threads(4) #set automatic cores to 1
+```{r,mds}
 
@@
 rownames(ss) <- colnames(x)
+
+mds <- cmdscale(dist(t(scale(x))))
+
+cols <- gsub("1","pink",gsub("0","lightblue",tumor))
+
+plot(mds, xlab="Coordinate 1", ylab="Coordinate 2", pch=16,col=cols,cex=2)
+
+text(mds, labels=colnames(x),cex=0.7)
+
+```
+
+Now do DESeq2
+
+```{r,deseq2}
+RhpcBLASctl::blas_set_num_threads(4) #set automatic cores to 1
 
@@
 
+str(fsets)
+
 ```
@@
 
+# Full counts
+
 ## Downsample FGSEA
 
-This will run the LAM analysis at group sizes of 2, 3, 6, 10, 16, 22 and 30,
+This will run the FGSEA analysis at group sizes of 2, 3, 6, 10, 16, 22 and 30,
 50 times each, with pseudorandom seeds from 100 to 5000 jumping up by 100.
@@
 dots <- lapply(ssres,function(x) { x[,"TP"] })
-lapply(dots,summary)
 boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fgsea",ylim=c(0,nrow(fsig))) ; grid()
@@
 dots <- lapply(ssres,function(x) { x[,"FP"] })
-lapply(dots,summary)
 boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
@@
 ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
-lapply(ratio,summary)
 boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0,ylim=c(0,0.5)) ; grid()
@@
 dots <- lapply(ssres,function(x) { x[,"TP"] })
-lapply(dots,summary)
 boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fora",ylim=c(0,length(osets))) ; grid()
@@
 
+df
+
 MAX=length(fsets)
@@
 ```
+
+# Thin Counts
+
+```{x,thincounts}
+
+x2 <- thinCounts(x, prob=NULL, target.size=10000000)
+colSums(x2)
+x <- x2
+
+```
+
+## Downsample FGSEA
+
+This will run the FGSEA analysis at group sizes of 2, 3, 6, 10, 16, 22 and 30,
+50 times each, with pseudorandom seeds from 100 to 5000 jumping up by 100.
+
+```{r,fgsea_run2,fig.height=10}
+
+RhpcBLASctl::blas_set_num_threads(1) #set automatic cores to 1
+
+SAMPLESIZES=c(2,3,5,10,15,20,25,30)
+#SAMPLESIZES=c(3,6,10,22)
+fgres <- lapply(SAMPLESIZES,function(n) {
+  SEEDS <- seq(100,5000,100) # 50 repetitions
+  dsres <- mclapply(SEEDS, function(SEED) {
+    set.seed(SEED)
+    mysample <- sample(1:37,n,replace=FALSE)*2
+    mysample <- c(mysample,(mysample-1))
+    mysample <- mysample[order(mysample)]
+    ss2 <- ss[mysample,]
+    x2 <- x[,mysample]
+    x2 <- x2[rowMeans(x2)>=10,]
+
+    dds <- DESeqDataSetFromMatrix(countData = x2 , colData=ss2, design = ~ patient + tumor)
+    res <- DESeq(dds)
+    z <- results(res)
+    vsd <- vst(dds, blind=FALSE)
+    zz <-cbind(as.data.frame(z),assay(vsd))
+    dge <- as.data.frame(zz[order(zz$pvalue),])
+
+    dge$stat[which(is.na(dge$stat))] <- 1
+    stat <- dge$stat
+    names(stat) <- rownames(dge)
+    fres <- fgsea(pathways=gs,stats=stat,minSize=5,nproc=3)
+    fres <- fres[order(-abs(fres$ES)),]
+    fsig <- subset(fres,padj<0.05)
+    fsets2 <- paste(fsig$pathway,as.character(fsig$ES > 0))
+
+    TOT=length(fsets2)
+    TP=length(which(fsets2 %in% fsets))
+    FP=length(which(!fsets2 %in% fsets))
+    FN=length(which(!fsets %in% fsets2))
+    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
+    RES
+  },mc.cores=16)
+  do.call(rbind,dsres)
+})
+
+ssres <- fgres
+par(mfrow=c(3,1))
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+lapply(dots,summary)
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fgsea",ylim=c(0,nrow(fsig))) ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+abline(h=nrow(fsig))
+
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+lapply(dots,summary)
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+lapply(ratio,summary)
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0,ylim=c(0,0.5)) ; grid()
+beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
+
+pdf("sens_fgsea_thin.pdf",height=7,width=5)
+par(mfrow=c(3,1))
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fgsea",ylim=c(0,nrow(fsig))) ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+abline(h=nrow(fsig))
+
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0,ylim=c(0,0.5)) ; grid()
+beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
+
+dev.off()
+
+ssres <- fgres
+
+```
+
+## FORA Downsample
+
+```{r,fora_run2,fig.height=10}
+
+SAMPLESIZES=c(2,3,5,10,15,20,25,30)
+fores <- lapply(SAMPLESIZES,function(n) {
+  SEEDS <- seq(100,5000,100) # 50 repetitions
+  dsres <- mclapply(SEEDS, function(SEED) {
+    set.seed(SEED)
+    mysample <- sample(1:37,n,replace=FALSE)*2
+    mysample <- c(mysample,(mysample-1))
+    mysample <- mysample[order(mysample)]
+    ss2 <- ss[mysample,]
+    x2 <- x[,mysample]
+    x2 <- x2[rowMeans(x2)>=10,]
+
+    dds <- DESeqDataSetFromMatrix(countData = x2 , colData=ss2, design = ~ patient + tumor)
+    res <- DESeq(dds)
+    z <- results(res)
+    vsd <- vst(dds, blind=FALSE)
+    zz <-cbind(as.data.frame(z),assay(vsd))
+    dge <- as.data.frame(zz[order(zz$pvalue),])
+
+    sigup <- rownames(subset(dge,padj<0.05 & log2FoldChange>0))
+    if (length(sigup) < 250) { sigup <- head(rownames(subset(dge, log2FoldChange>0)),250) }
+    sigdn <- rownames(subset(dge,padj<0.05 & log2FoldChange<0))
+    if (length(sigdn) < 250) { sigdn <- head(rownames(subset(dge, log2FoldChange<0)),250) }
+    bg <- rownames(dge)
+
+    oresup <- as.data.frame(fora(pathways=gs,genes=sigup,universe=bg,minSize=5))
+    oresup <- subset(oresup,padj<0.05)
+    oresup <- paste(oresup$pathway,"TRUE")
+
+    oresdn <- as.data.frame(fora(pathways=gs,genes=sigdn,universe=bg,minSize=5))
+    oresdn <- subset(oresdn,padj<0.05)
+    oresdn <- paste(oresdn$pathway,"FALSE")
+
+    osets2 <- c(oresup,oresdn)
+
+    TOT=length(osets2)
+    TP=length(which(osets2 %in% osets))
+    FP=length(which(!osets2 %in% osets))
+    FN=length(which(!osets %in% osets2))
+    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
+    RES
+  },mc.cores=16)
+  do.call(rbind,dsres)
+})
+
+ssres <- fores
+
+par(mfrow=c(3,1))
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+lapply(dots,summary)
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fora",ylim=c(0,length(osets))) ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+abline(h=length(osets))
+
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+lapply(dots,summary)
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+lapply(ratio,summary)
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0,ylim=c(0,0.5)) ; grid()
+beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
+
+pdf("sens_fora_thin.pdf",height=7,width=5)
+par(mfrow=c(3,1))
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fora",ylim=c(0,length(osets))) ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+abline(h=length(osets))
+
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0,ylim=c(0,0.5)) ; grid()
+beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
+
+dev.off()
+
+```
+
+## Summarise downnsampling
+
+```{r,downsample,fig.width=10}
+
+par(mfrow=c(1,3))
+
+dots <- lapply(fgres,function(x) { x[,"TP"] })
+fgres2 <- do.call(rbind,lapply(dots,summary))
+
+dots <- lapply(fores,function(x) { x[,"TP"] })
+fores2 <- do.call(rbind,lapply(dots,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+MAX=length(fsets)
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways", ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+abline(h=length(osets),col="red")
+abline(h=length(fsets),col="blue")
+mtext("no. consistent pathways FDR<0.05")
+grid()
+
+legend("bottomright", legend=c("fgsea","fora"),
+       col=c("blue","red"), lty=1, pch=19,  cex=1)
+
+dots <- lapply(fgres,function(x) { x[,"FP"] })
+fgres2 <- do.call(rbind,lapply(dots,summary))
+
+dots <- lapply(fores,function(x) { x[,"FP"] })
+fores2 <- do.call(rbind,lapply(dots,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+mtext("no. inconsistent pathways FDR<0.05")
+grid()
+
+dots <- lapply(fores,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(fores,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+fores2 <- do.call(rbind,lapply(ratio,summary))
+
+dots <- lapply(fgres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(fgres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+fgres2 <- do.call(rbind,lapply(ratio,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="proportion",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+mtext("Proportion of inconsistent pathways")
+grid()
+
+### PDF
+pdf("sens_thin.pdf",height=4,width=8)
+par(mfrow=c(1,3))
+
+dots <- lapply(fgres,function(x) { x[,"TP"] })
+fgres2 <- do.call(rbind,lapply(dots,summary))
+
+dots <- lapply(fores,function(x) { x[,"TP"] })
+fores2 <- do.call(rbind,lapply(dots,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+df
+
+MAX=length(fsets)
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways", main="no. consistent pathways FDR<0.05",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+abline(h=length(osets),col="red")
+abline(h=length(fsets),col="blue")
+grid()
+
+legend("bottomright", legend=c("fgsea","fora"),
+       col=c("blue","red"), lty=1, pch=19,  cex=1)
+
+dots <- lapply(fgres,function(x) { x[,"FP"] })
+fgres2 <- do.call(rbind,lapply(dots,summary))
+
+dots <- lapply(fores,function(x) { x[,"FP"] })
+fores2 <- do.call(rbind,lapply(dots,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+df
+
+MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways",main="no. inconsistent pathways FDR<0.05",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+grid()
+
+dots <- lapply(fores,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(fores,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+fores2 <- do.call(rbind,lapply(ratio,summary))
+
+dots <- lapply(fgres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(fgres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+fgres2 <- do.call(rbind,lapply(ratio,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+df
+
+MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="proportion",main="Proportion of inconsistent pathways",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+grid()
+dev.off()
+
+```
+
 ## Session information

### 2024-07-17 analysis/GSE158420_rna.Rmd
chunkname
fix
library

@@
   library("tictoc")
+  library("edgeR")
 })
@@
 
-```{x,thincounts}
+```{r,thincounts}
 
@@
 
-```{r,downsample,fig.width=10}
+```{r,downsample2,fig.width=10}
 

### 2024-07-18 manuscript/manuscript.Rmd
downsampling added to ms

@@
 For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
-selected for the foreground gene set.
++selected for the foreground gene set.
 If fewer than 200 genes were detected as significant, then 200 genes with the smallest
@@
 
+## Does FCS have better sensitivity than ORA with real RNA-seq data?
+
+To understand whether FCS has better sensitivity than ORA in a real setting, we conducted a
+systematic downsampling of a gene expression dataset cosisting of normal and tumor samples
+of 37 lung cancer patients.
+In the full set of patient samples, fora (ORA) yielded 375 significantly differentially
+regulated Reactomes, while fgsea (FCS) gave 470 (FDR<0.05).
+Fgsea yielded consistently more significant pathways than fora in downsampled data
+(**Figure 4A**).
+Fgsea also identified a larger proportion of pathways in downsampled data compared to fora.
+For example at a sample size of 5, fgsea identified 74.5% of pathways from the full set as
+compared to 41.5% for fora, confirming better recall for FCS over ORA.
+It was noted that fgsea identified more pathways in downsampled data that were
+inconsistent with the full dataset as compared to fora (**Figure 4B**).
+However, inconsistent pathways were a smaller proportion of all findings for fgsea as
+compared to fora (**Figure 4C**), indicating an overall lower false positive rate for FCS
+compared to ORA methods.
+
+![**Figure 4. Downsampling confirms higher sensitivity for FCS over ORA.
+(A) The number of significant pathways identified in downsampled data that are consistent
+with findings in the full dataset, using FCS (fgsea) and ORA (fora) methods.
+(B) The number of significant pathways identified in downsampled data that are not
+consistent the full dataset.
+(C) The proportion of pathways in downsampled data that are inconsistent with the full
+dataset.
+All values are the median of 50 simulations.](../analysis/fig4.png "Figure 4A shows that fgsea is more sensitive than fora.
+Figure 4B shows fgsea has higher absule number of false positives, but Figure 4C shows that as a proportion of all findings fgsea has lower false positive rate than fora.")
+
 ## Discussion
@@
 noisy.
-Higher recall of FCS also means that researchers will have more power to identify pathways
+The higher recall of FCS demonstrated in simulated differential expression profiles and confirmed with
+downsampled real cancer RNA-seq data means that researchers will have more power to identify pathways
 relevant to the biological processes being studied.

### 2024-07-18 manuscript/manuscript.Rmd
ms work

@@
 
-## Does FCS have better sensitivity than ORA with real RNA-seq data?
+### Exploring sensitivity in real RNA-seq data
 
@@
 
-![**Figure 4. Downsampling confirms higher sensitivity for FCS over ORA.
+![**Figure 4. Downsampling confirms higher sensitivity for FCS over ORA.**
 (A) The number of significant pathways identified in downsampled data that are consistent

### 2024-07-18 manuscript/manuscript.Rmd
ms work

@@
 For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
-+selected for the foreground gene set.
+selected for the foreground gene set.
 If fewer than 200 genes were detected as significant, then 200 genes with the smallest
@@
 All analyses were undertaken with R v4.4.0 in a Docker container.
+
+### Comparing ORA and FCS methods by downsampling real RNA-seq data
+
+To understand the sensitivity and false positive rate of ORA and FCS methods, we employed a downsampling
+approach to a previously published dataset derived from 37 lung cancer patients, having tumor and normal
+tissues for each patient [@Cho2022-ri].
+These RNA-seq based gene expression counts were downloaded from NCBI GEO under accession number GSE158420.
+After reading them into R, we noticed some gene names converted to dates [@Abeysooriya2021-qt], so we
+used the HGNChelper package to fix them [@Oh2022-gj].
+DESeq2 was then used for differential expression analysis correcting for patient-of-origin to identify genes
+differentially expressed in tumor compared to normal tissue.
+FCS was conducted using fgsea using the DESeq2 test statistic to score each gene.
+ORA was conducted using fora, using genes with differential expression FDR<0.05.
+Separate tests were conducted for up- and down-regulated genes.
+For the ORA background, genes with a mean of 10 or more reads per sample on average across the comparison were
+included.
+Reactome pathways were used.
+For downsampling, sample size was varied between 2 and 30 patients, selected pseudorandomly with a set seed,
+followed by parallel analysis with fgsea and fora pipelines, using a significance filter of FDR<0.05.
+We then calculated the number of pathways identified in downsampled data that were consistent with the full
+dataset.
+The number of consistent pathways gives an indication of sensitivity of each method.
+Then, we calculated the number of pathways identified in the downsampled data that were inconsistent with the
+full dataset.
+The proportion of significant pathways classified as inconsistent provides an estimate of the real false
+discovery rate.
+This was repeated 50 times at each sample size, and the median values were reported.
 
@@
 with findings in the full dataset, using FCS (fgsea) and ORA (fora) methods.
+The blue and red horizontal lines indicate the number of significant pathways identified
+in the full set of 27 patients using fgsea and fora respectively.
 (B) The number of significant pathways identified in downsampled data that are not
@@
 So end-users can explore the effect of these two problems on their work, we have developed
-an R/Shiny web application.
+an R/Shiny web application (see "Availability of data and materials" below).
 Users can upload their foreground and background lists, select their preferred gene set
 library and compare results from default and patched analyses.
-The resource is currently available at https://oratool.ziemann-lab.net/ and available as a
-docker container (https://hub.docker.com/repository/docker/mziemann/background_app/general).
 
@@
 (**Table 1**). Still, there are some tools free of such problems.
-For example, fora is a good option for R-based workflows, however the function does not calculate the fold enrichment score,
+For example, fora is a good option for R-based workflows, however the function does not calculate
+the fold enrichment score,
 so users will need to do it themselves [@Korotkevich2016-gd].
@@
 
-The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there are some
-bigger problems noted.
-Firstly, all variations of ORA were less accurate as compared to FCS in this simulation work, confirming a
-previous report [@Kaspi2020-rn].
+The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there
+are some bigger problems noted.
+Firstly, all variations of ORA were less accurate (based on F1 score) as compared to FCS in this
+simulation work, confirming a previous report [@Kaspi2020-rn].
 Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
@@
 noisy.
-The higher recall of FCS demonstrated in simulated differential expression profiles and confirmed with
-downsampled real cancer RNA-seq data means that researchers will have more power to identify pathways
-relevant to the biological processes being studied.
-Secondly, many ORA tools do not provide enrichment scores in their default output tables.
+The higher recall of FCS demonstrated in simulated differential expression profiles and confirmed
+with downsampled real cancer RNA-seq data means that researchers will have more power to identify
+pathways relevant to the biological processes being studied.
+Secondly, many popular ORA tools listed in Table 1 do not provide enrichment scores in their
+default output tables.
 Enrichment scores are a surrogate for effect size, and without them, users may over-rely
 on significance scores when interpreting enrichment results.
-ORA results can vary dramatically depending on the number of genes selected in the
+Lastly, ORA results can vary dramatically depending on the number of genes selected in the
 foreground, a mostly arbitrary choice.
@@
 correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
-that if the data can be scored/ranked, methods like FCS are recommended over ORA.
+that if the data can be scored/ranked, methods like FCS are strongly recommended over ORA.
 ORA is better suited to other cases where genes are classified into binary groups.
@@
 
-Publicly available data were obtained from Digital Expression Explorer 2 (dee2.io),
-using the accession numbers mentioned in the Methods section.
-The code repository including template Dockerfile and R Markdown script are available on
-GitHub (https://github.com/markziemann/background).
-The Docker image is available on DockerHub (https://hub.docker.com/r/mziemann/background).
-The code repository and Docker image have been uploaded to Zenodo for long-term archiving
-(https://zenodo.org/record/TODOXXX).
+Publicly available data were obtained from Digital Expression Explorer 2 (dee2.io) and
+NCBI GEO using the accession numbers mentioned in the Methods section.
+
+Reproduction of the results here can be achieved using the GitHub code repository GitHub
+(https://github.com/markziemann/background) and Docker image
+(https://hub.docker.com/r/mziemann/background), which have both been deposited to Zenodo
+for long-term preservation (https://zenodo.org/record/TODOXXX).
+
+The R/Shiny tool for comparing ORA with and without the two problems is currently
+online at https://oratool.ziemann-lab.net/ and available as a
+downloadable docker container for local execution
+(https://hub.docker.com/repository/docker/mziemann/background_app/general).
 

### 2024-07-22 manuscript/manuscript.Rmd
grammar

@@
 To understand the sensitivity and false positive rate of ORA and FCS methods, we employed a downsampling
-approach to a previously published dataset derived from 37 lung cancer patients, having tumor and normal
+approach to a previously published dataset derived from 37 lung cancer patients, having tumour and normal
 tissues for each patient [@Cho2022-ri].
@@
 DESeq2 was then used for differential expression analysis correcting for patient-of-origin to identify genes
-differentially expressed in tumor compared to normal tissue.
+differentially expressed in tumour compared to normal tissue.
 FCS was conducted using fgsea using the DESeq2 test statistic to score each gene.
@@
 To understand whether FCS has better sensitivity than ORA in a real setting, we conducted a
-systematic downsampling of a gene expression dataset cosisting of normal and tumor samples
+systematic downsampling of a gene expression dataset consisting of normal and tumour samples
 of 37 lung cancer patients.
@@
 The blue and red horizontal lines indicate the number of significant pathways identified
-in the full set of 27 patients using fgsea and fora respectively.
+in the full set of 37 patients using fgsea and fora respectively.
 (B) The number of significant pathways identified in downsampled data that are not
-consistent the full dataset.
+consistent with the full dataset.
 (C) The proportion of pathways in downsampled data that are inconsistent with the full
@@
 Sometimes this feature is hidden under "advanced options".
-Given the strong biological and technical biases that are present in contemporary omics such as including single cell and spatial transcriptomics,
+Given the strong biological and technical biases that are present in contemporary omics like single cell and spatial transcriptomics,
 it should be a mandatory step for users to provide a custom background list.

### 2024-07-22 manuscript/manuscript.Rmd
words
grammar

@@
 This forces clusterProfiler to retain all background genes.
-The ORA analyses were then filtered for significant sets using an FDR threshold of 0.05.
+The ORA results were then filtered for significant sets using an FDR threshold of 0.05.
 Jaccard index was used to compare original and corrected analyses.
@@
 These RNA-seq based gene expression counts were downloaded from NCBI GEO under accession number GSE158420.
-After reading them into R, we noticed some gene names converted to dates [@Abeysooriya2021-qt], so we
+After reading them into R, we noticed some gene names were converted to dates [@Abeysooriya2021-qt], so we
 used the HGNChelper package to fix them [@Oh2022-gj].
@@
 libraries before and after correcting the problem (**Figure 1A**).
-Each of the gene set libraries showed an uplift when correcting the background problem.
+Each of the gene set libraries showed an uplift in the number of significant gene sets when correcting the background problem.
 The smallest increase was observed in Cellmarkers gene library (11.8%), while the largest
@@
 
-We posited that the foreground gene list length might exacerbate the problem, so we performed
+We posited that a shorter foreground gene list length might exacerbate the problem, so we performed
 parallel analysis with foreground lists of length between 125 and 2,000 genes (**Figure 2B**).
@@
 Enrichment scores are a surrogate for effect size, and without them, users may over-rely
-on significance scores when interpreting enrichment results.
+on significance values when interpreting enrichment results.
 Lastly, ORA results can vary dramatically depending on the number of genes selected in the

### 2024-09-05 analysis/analyse_rnaseq1.Rmd
save data

@@
 
+## Table for reviewer
+
+```{r,table}
+
+library(reshape2)
+res2a <- as.matrix(acast(res1, Var1~Var2, value.var="nsig_bad"))
+rownames(res2a) <- gsub("bulkrna","d",rownames(res2a))
+rownames(res2a) <- gsub(".Rds"," original",rownames(res2a))
+
+res2b <- as.matrix(acast(res1, Var1~Var2, value.var="nsig_good"))
+rownames(res2b) <- gsub("bulkrna","d",rownames(res2b))
+rownames(res2b) <- gsub(".Rds"," corrected",rownames(res2b))
+
+res2 <- rbind(res2a,res2b)
+res2 <- res2[order(rownames(res2)),]
+res2c <- rowMeans(res2)
+res2 <- cbind(res2,res2c)
+
+res2 %>%
+  kbl(row.names = TRUE, caption="Effect of background problem correction across seven studies and nine gene set libraries") %>%
+  kable_paper("hover", full_width = F)
+
+saveRDS(res2,"bgprob.rds")
+
+```
+
 ## Session information

### 2024-09-05 analysis/analyse_rnaseq2.Rmd
table of results for fdr prob

@@
 
+## Table for reviewer
+
+```{r,table}
+
+library(reshape2)
+res2a <- as.matrix(acast(res1, Var1~Var2, fun.aggregate = sum, value.var="nsig_bad"))
+rownames(res2a) <- gsub("bulkrna","d",rownames(res2a))
+rownames(res2a) <- gsub(".Rds"," original",rownames(res2a))
+
+res2b <- as.matrix(acast(res1, Var1~Var2, fun.aggregate = sum, value.var="nsig_good"))
+rownames(res2b) <- gsub("bulkrna","d",rownames(res2b))
+rownames(res2b) <- gsub(".Rds"," corrected",rownames(res2b))
+
+res2 <- rbind(res2a,res2b)
+res2 <- res2[order(rownames(res2)),]
+res2c <- rowMeans(res2)
+res2 <- cbind(res2,res2c)
+
+res2 %>%
+  kbl(row.names = TRUE, caption="Effect of FDR problem correction across seven studies and nine gene set libraries") %>%
+  kable_paper("hover", full_width = F)
+
+saveRDS(res2,"fdrprob.rds")
+
+```
+
 ## Session information

### 2024-09-05 analysis/analyse_rnaseq1.Rmd
get set len

@@
 
+## Characterise sets specific to original and uncorrected analysis
+
+```{r,spec}
+
+compare_ora2 <- function(bad,good){
+  badup <- bad$orabad_up
+  baddn <- bad$orabad_dn
+  goodup <- good$oragood_up
+  gooddn <- good$oragood_dn
+  mup <- merge(badup,goodup,by=0)
+  mdn <- merge(baddn,gooddn,by=0)
+  bup <- subset(mup,p.adjust.x<0.05)$Row.names
+  gup <- subset(mup,p.adjust.y<0.05)$Row.names
+  bdn <- subset(mdn,p.adjust.x<0.05)$Row.names
+  gdn <- subset(mdn,p.adjust.y<0.05)$Row.names
+  sig_bad <- unique(c(bup,bdn))
+  sig_good <- unique(c(gup,gdn))
+  itx <- intersect(sig_bad,sig_good)
+  spec_good <- setdiff(sig_good,sig_bad)
+  result=list("intersect"=itx,"spec_good"=spec_good)
+  return(result)
+}
+
+params <- expand.grid(1:length(d),1:length(gs))
+params2 <- expand.grid(names(d),names(gs))
+
+colnames(params) <- c("d","gs")
+
+ora_dat3 <- mclapply(1:nrow(params), function(i) {
+  j=params[i,1]
+  k=params[i,2]
+  bad <- orabad(d[[j]],gs[[k]])
+  good <- oragood(d[[j]],gs[[k]])
+  res <- compare_ora2(bad,good)
+  return(res)
+} , mc.cores=8)
+
+gslengths <- unlist(lapply(gs,function(x) { lapply(x,length) } ) )
+names(gslengths) <- sapply((strsplit(names(gslengths),"\\.")),"[[",2)
+
+speclen <- lapply(ora_dat3, function(j) {
+  itxlen <- mean(gslengths[which(names(gslengths) %in% j[[1]])])
+  speclen <- mean(gslengths[which(names(gslengths) %in% j[[2]])])
+  return(c("itxlen"=itxlen,"spec"=speclen))
+})
+
+itxlenmx <- matrix(sapply(speclen,"[[",1),nrow=7)
+speclenmx <- matrix(sapply(speclen,"[[",2),nrow=7)
+
+colMeans(itxlenmx,na.rm=TRUE) / colMeans(speclenmx,na.rm=TRUE)
+
+mean( ( colMeans(speclenmx,na.rm=TRUE) - colMeans(itxlenmx,na.rm=TRUE) ) / colMeans(itxlenmx,na.rm=TRUE) )
+
+```
+
+Overall, the size of gene sets specific to the "fixed" analysis were 9.3% smaller than
+the intersect.
+
 ## Make charts showing changes in pvalues

### 2024-09-06 analysis/analyse_rnaseq2.Rmd
investigating gene sets specific to uncorrected annalysis

@@
 
+## Investigate which gene sets are specific to the ORIGINAL and FIXED analysis
+
+Will focus on the 500 gene analysis if GO in the seven datasets.
+
+j=1:7
+
+k=gs=6
+
+l=ngenes=500
+
+```{r,deepdive}
+
+compare_ora2 <- function(bad,good){
+  badup <- bad$orabad_up
+  baddn <- bad$orabad_dn
+  goodup <- good$oragood_up
+  gooddn <- good$oragood_dn
+  bup <- rownames(subset(badup,p.adjust<0.05))
+  gup <- rownames(subset(goodup,p.adjust<0.05))
+  bdn <- rownames(subset(baddn,p.adjust<0.05))
+  gdn <- rownames(subset(gooddn,p.adjust<0.05))
+  sig_bad <- unique(c(bup,bdn))
+  sig_good <- unique(c(gup,gdn))
+  spec_good <- setdiff(sig_good,sig_bad)
+  spec_bad <- setdiff(sig_bad,sig_good)
+  itx <- intersect(sig_bad,sig_good)
+  result=list("spec_bad"=spec_bad,"itx"=itx)
+  return(result)
+}
+
+ora_dat3 <- mclapply(1:7, function(i) {
+  j=params[i,1]
+  k=6
+  l=500
+  bad <- orabad(d[[j]],gs[[k]],l)
+  good <- oragood(d[[j]],gs[[k]],l)
+  res <- compare_ora2(bad,good)
+  return(res)
+} , mc.cores=8)
+
+gslengths <- unlist(lapply(gs,function(x) { lapply(x,length) } ) )
+names(gslengths) <- sapply((strsplit(names(gslengths),"\\.")),"[[",2)
+
+i=1
+
+spec <- lapply(1:7,function(i) {
+  itx_lens <- gslengths[names(gslengths) %in% ora_dat3[[i]][[2]]]
+  spec_bad_lens <- gslengths[names(gslengths) %in% ora_dat3[[i]][[1]]]
+  if ( length(itx_lens) >= 10 & length(spec_bad_lens) >= 10 ) {
+    return(c("spec_bad"=mean(spec_bad_lens),"itx"=mean(itx_lens)))
+  } else {
+    return(c("spec_bad"=NA,"itx"=NA))
+  }
+})
+
+spec <- as.data.frame(do.call(rbind,spec))
+
+( spec$spec_bad - spec$itx ) / spec$itx
+
+```
+
 ## Make charts showing changes in pvalues

### 2024-09-06 analysis/analyse_rnaseq1.Rmd
add whiskers

@@
 
-colMeans(itxlenmx,na.rm=TRUE) / colMeans(speclenmx,na.rm=TRUE)
+rowMeans(itxlenmx,na.rm=TRUE)
 
-mean( ( colMeans(speclenmx,na.rm=TRUE) - colMeans(itxlenmx,na.rm=TRUE) ) / colMeans(itxlenmx,na.rm=TRUE) )
+rowMeans(speclenmx,na.rm=TRUE)
+
+( rowMeans(speclenmx,na.rm=TRUE) - rowMeans(itxlenmx,na.rm=TRUE) ) / rowMeans(itxlenmx,na.rm=TRUE)
+
+mean( ( rowMeans(speclenmx,na.rm=TRUE) - rowMeans(itxlenmx,na.rm=TRUE) ) / rowMeans(itxlenmx,na.rm=TRUE) )
 
@@
 
+rowMeans( ( res2b - res2a ) / res2a ) * 100
+
 ```

### 2024-09-06 analysis/analyse_rnaseq2.Rmd
add whiskers

@@
 
+jac3 <- do.call(rbind,lapply(jac,function(x) {x[,"GO"] } ))
+
+apply(jac3,1,function(x) { mean(x, na.rm=TRUE) } )
+apply(jac3,1,function(x) { sd(x, na.rm=TRUE) } )
+
+jsd <- lapply(jac,function(y) {
+  res <- apply(y,2,function(x) {
+    x <- x[!is.na(x)]
+    x <- x[is.finite(x)]
+    sd(x)
+  } )
+  return(res)
+})
+
+jsd <- do.call(rbind,jsd)
+
+jsd %>%
+  kbl(row.names = TRUE, caption="Standard deviation in Jaccard stats across gene list length") %>%
+  kable_paper("hover", full_width = F)
+
 jac <- lapply(jac,function(y) {
@@
 jac <- do.call(rbind,jac)
+
+jac %>%
+  kbl(row.names = TRUE, caption="Jaccard stats across gene list length") %>%
+  kable_paper("hover", full_width = F)
+
+# GO with error bars
+plot(jac[,"GO"],type="b", pch=19,cex=1.5,lwd=2,
+  xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected")
+axis(1, at = seq(1, 5, by = 1),
+     labels = ngenes)
+mysd <- jsd[,"GO"]
+arrows(x0=1:5, y0=(jac[,"GO"] - mysd) ,  x1=1:5, y1=(jac[,"GO"] + mysd), code=3, angle=90, length=0.1 )
+
+mysd <- jsd[,"Reactome"]
+points(jac[,"Reactome"],type="b",col="red",pch=19,cex=1.5,lwd=2)
+arrows(x0=1:5, y0=(jac[,"Reactome"] - mysd) ,  x1=1:5, y1=(jac[,"Reactome"] + mysd), code=3, angle=90, length=0.1 ,col="red")
+
+mysd <- jsd[,"Cellmarkers"]
+points(jac[,"Cellmarkers"],type="b",col="blue",pch=19,cex=1.5,lwd=2)
+arrows(x0=1:5, y0=(jac[,"Cellmarkers"] - mysd) ,  x1=1:5, y1=(jac[,"Cellmarkers"] + mysd), code=3, angle=90, length=0.1 , col="blue")
+
+mysd <- jsd[,"TFT GTRD"]
+points(jac[,"TFT GTRD"],type="b",col="gray50",pch=19,cex=1.5,lwd=2)
+arrows(x0=1:5, y0=(jac[,"TFT GTRD"] - mysd) ,  x1=1:5, y1=(jac[,"TFT GTRD"] + mysd), code=3, angle=90, length=0.1 , col="gray50" )
 
@@
 
+Table when using 2000 genes.
+
 ```{r,table}
@@
 library(reshape2)
-res2a <- as.matrix(acast(res1, Var1~Var2, fun.aggregate = sum, value.var="nsig_bad"))
-rownames(res2a) <- gsub("bulkrna","d",rownames(res2a))
-rownames(res2a) <- gsub(".Rds"," original",rownames(res2a))
+res2000 <- subset(res1,Var3==2000)
 
-res2b <- as.matrix(acast(res1, Var1~Var2, fun.aggregate = sum, value.var="nsig_good"))
-rownames(res2b) <- gsub("bulkrna","d",rownames(res2b))
-rownames(res2b) <- gsub(".Rds"," corrected",rownames(res2b))
+res3a <- as.matrix(acast(res2000, Var1~Var2, fun.aggregate = sum, value.var="nsig_bad"))
+rownames(res3a) <- gsub("bulkrna","d",rownames(res3a))
+rownames(res3a) <- gsub(".Rds"," original",rownames(res3a))
 
-res2 <- rbind(res2a,res2b)
-res2 <- res2[order(rownames(res2)),]
-res2c <- rowMeans(res2)
-res2 <- cbind(res2,res2c)
+res3b <- as.matrix(acast(res2000, Var1~Var2, fun.aggregate = sum, value.var="nsig_good"))
+rownames(res3b) <- gsub("bulkrna","d",rownames(res3b))
+rownames(res3b) <- gsub(".Rds"," corrected",rownames(res3b))
 
-res2 %>%
+res3 <- rbind(res3a,res3b)
+res3 <- res3[order(rownames(res3)),]
+res3c <- rowMeans(res3)
+res3 <- cbind(res3,res3c)
+
+res3 %>%
   kbl(row.names = TRUE, caption="Effect of FDR problem correction across seven studies and nine gene set libraries") %>%
@@
 
-saveRDS(res2,"fdrprob.rds")
+saveRDS(res3,"fdrprob.rds")
+
+```
+
+Another table showing downsampling result.
+
+```{r,dstable}
+
+ds <- res1
+
+colnames(ds) <- c("Dataset","GeneSetLibrary","GeneListSize","nSig_original","nSig_corrected","Jaccard")
+
+ds %>%
+  kbl(row.names = TRUE, caption="FDR problem worsens when the input gene list is shorter") %>%
+  kable_paper("hover", full_width = F)
+
+saveRDS(ds,"downsample_fdr.rds")
 

### 2024-09-06 analysis/analyse_rnaseq1.Rmd
plot individual points

@@
 
+## Make beeswarm chart
+
+```{r,beeswarms}
+
+gslibs <- unique(params2$Var2)
+
+nsl <- lapply(gslibs,function(g) {
+  orig <- subset(res1,Var2==g)$nsig_bad
+  corr <- subset(res1,Var2==g)$nsig_good
+  df1 <- data.frame("gsl"=g,"type"="orig","count"=orig)
+  df2 <- data.frame("gsl"=g,"type"="corr","count"=corr)
+  return(rbind(df1,df2))
+} )
+
+nsl <- do.call(rbind,nsl)
+nsl$set <- paste(nsl$gsl,nsl$type)
+
+unsl <- unstack(nsl[,c(3,4)])
+
+par(mar = c(5.1, 8.1, 4.1, 2.1))
+library(beeswarm)
+boxplot(unsl,horizontal=TRUE,las=1,col=c("lightblue","pink"),cex=0,xlab="no. significant sets FDR<0.05")
+beeswarm(unsl,horizontal=TRUE,add=TRUE,col=c("blue","red"),pch=19,cex=0.6)
+
+par(mar = c(5.1, 4.1, 4.1, 2.1))
+
+```
+
 ## Make charts showing changes in pvalues

### 2024-09-09 analysis/analyse_rnaseq1.Rmd
adding error bars

@@
 dev.off()
+
+# with SD
+mysd <- apply(jac,2,function(x) { sd(x,na.rm=TRUE) } )
+par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
+plot(cov,jac2,xlim=c(0,30000),ylim=c(0,1),pch=19,col="lightgray",cex=3,
+  xlab="gene coverage",ylab="Jaccard index")
+text(cov,jac2,labels=names(cov))
+mylm <- lm(jac2~cov)
+abline(mylm,lty=2,lwd=2)
+SLOPE=unname(signif(mylm$coefficients[2],3))
+INT=unname(signif(mylm$coefficients[1],3))
+PVAL=signif(summary(mylm)$coefficients[4],3)
+RSQ=signif(summary(mylm)$r.squared,3)
+HEADER=paste("Rsq =",RSQ,"; p-val =",PVAL)
+mtext(HEADER)
+arrows(x0=cov,x1=cov,y0=jac2+mysd,y1=jac2-mysd,code=3, angle=90, length=0.1 , col="gray40")
+
+png("fig1_ORAjacsd.png")
+par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
+plot(cov,jac2,xlim=c(0,30000),ylim=c(0,1),pch=19,col="lightgray",cex=3,
+  xlab="gene coverage",ylab="Jaccard index")
+text(cov,jac2,labels=names(cov))
+mylm <- lm(jac2~cov)
+abline(mylm,lty=2,lwd=2)
+SLOPE=unname(signif(mylm$coefficients[2],3))
+INT=unname(signif(mylm$coefficients[1],3))
+PVAL=signif(summary(mylm)$coefficients[4],3)
+RSQ=signif(summary(mylm)$r.squared,3)
+HEADER=paste("Rsq =",RSQ,"; p-val =",PVAL)
+mtext(HEADER)
+arrows(x0=cov,x1=cov,y0=jac2+mysd,y1=jac2-mysd,code=3, angle=90, length=0.1 , col="gray40")
+dev.off()
+
+pdf("fig1_ORAjacsd.pdf",width=7, height=7)
+par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
+plot(cov,jac2,xlim=c(0,30000),ylim=c(0,1),pch=19,col="lightgray",cex=3,
+  xlab="gene coverage",ylab="Jaccard index")
+text(cov,jac2,labels=names(cov))
+mylm <- lm(jac2~cov)
+abline(mylm,lty=2,lwd=2)
+SLOPE=unname(signif(mylm$coefficients[2],3))
+INT=unname(signif(mylm$coefficients[1],3))
+PVAL=signif(summary(mylm)$coefficients[4],3)
+RSQ=signif(summary(mylm)$r.squared,3)
+HEADER=paste("Rsq =",RSQ,"; p-val =",PVAL)
+mtext(HEADER)
+arrows(x0=cov,x1=cov,y0=jac2+mysd,y1=jac2-mysd,code=3, angle=90, length=0.1 , col="gray40")
+dev.off()
+
 ```
@@
 
+#rearrange KEGG, Hallmark, HPO, Wikipath, miR, TFT, Reactome, Cellmarkers, GO
+unsl <- unsl[,c(10,9, 6,5, 8,7, 18,17, 12,11, 16,15, 14,13, 2,1, 4,3)]
+
 par(mar = c(5.1, 8.1, 4.1, 2.1))
 library(beeswarm)
-boxplot(unsl,horizontal=TRUE,las=1,col=c("lightblue","pink"),cex=0,xlab="no. significant sets FDR<0.05")
-beeswarm(unsl,horizontal=TRUE,add=TRUE,col=c("blue","red"),pch=19,cex=0.6)
+boxplot(unsl,horizontal=TRUE,las=1,col=c("pink","lightblue"),cex=0,xlab="no. significant sets FDR<0.05")
+beeswarm(unsl,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+
+unsl1 <- unsl+1
+boxplot(unsl1,horizontal=TRUE,log="x",las=1,col=c("pink","lightblue"),cex=0,xlab="no. significant sets FDR<0.05*")
+beeswarm(unsl1,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+
+legend("bottomright", inset=.02, legend=c("corrected","original"),
+  fill=c("lightblue","pink"),col=c("blue","red"), horiz=FALSE, cex=1.1)
+
+png("fig1_ORAsigbox.png")
+boxplot(unsl1,horizontal=TRUE,log="x",las=1,col=c("pink","lightblue"),cex=0,xlab="no. significant sets FDR<0.05*")
+beeswarm(unsl1,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+legend("bottomright", inset=.02, legend=c("corrected","original"),
+  fill=c("lightblue","pink"),col=c("blue","red"), horiz=FALSE, cex=1.1)
+dev.off()
+
+pdf("fig1_ORAsigbox.pdf",width=7, height=7)
+par(mar = c(5.1, 8.1, 4.1, 2.1))
+boxplot(unsl1,horizontal=TRUE,log="x",las=1,col=c("pink","lightblue"),cex=0,xlab="no. significant sets FDR<0.05*")
+beeswarm(unsl1,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+legend("bottomright", inset=.02, legend=c("corrected","original"),
+  fill=c("lightblue","pink"),col=c("blue","red"), horiz=FALSE, cex=1.1)
+dev.off()
 

### 2024-09-09 analysis/analyse_rnaseq2.Rmd
adding error bars

@@
 
-# GO with error bars
-plot(jac[,"GO"],type="b", pch=19,cex=1.5,lwd=2,
-  xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected")
-axis(1, at = seq(1, 5, by = 1),
-     labels = ngenes)
+# with error bars 2x2 charts
+ubtrim <- function(x) { if(is.na(x)) { x} else if(x > 1) { 1 } else { x } }
+lbtrim <- function(x) { if(is.na(x)) { x} else if(x < 0) { 0 } else { x } }
+
+par(mfrow=c(2,2))
+
+# GO
+plot(jac[,"GO"],type="b", pch=19,cex=1.5,lwd=2, xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected",main="GO")
+axis(1, at = seq(1, 5, by = 1), labels = ngenes)
 mysd <- jsd[,"GO"]
-arrows(x0=1:5, y0=(jac[,"GO"] - mysd) ,  x1=1:5, y1=(jac[,"GO"] + mysd), code=3, angle=90, length=0.1 )
+ub <- (jac[,"GO"] + mysd)
+ub <- sapply(ub,ubtrim)
+lb <-(jac[,"GO"] - mysd)
+lb <- sapply(lb,lbtrim)
+arrows(x0=1:5, y0=lb ,  x1=1:5, y1=ub, code=3, angle=90, length=0.1 )
 
+# REACTOME
 mysd <- jsd[,"Reactome"]
-points(jac[,"Reactome"],type="b",col="red",pch=19,cex=1.5,lwd=2)
-arrows(x0=1:5, y0=(jac[,"Reactome"] - mysd) ,  x1=1:5, y1=(jac[,"Reactome"] + mysd), code=3, angle=90, length=0.1 ,col="red")
+plot(jac[,"Reactome"],type="b", col="red", pch=19,cex=1.5,lwd=2, xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected",main="Reactome")
+axis(1, at = seq(1, 5, by = 1), labels = ngenes)
+ub <- (jac[,"Reactome"] + mysd)
+ub <- sapply(ub,ubtrim)
+lb <-(jac[,"Reactome"] - mysd)
+lb <- sapply(lb,lbtrim)
+arrows(x0=1:5, y0=lb ,  x1=1:5, y1=ub, code=3, angle=90, length=0.1 ,col="red")
 
+# Cellmarkers
 mysd <- jsd[,"Cellmarkers"]
-points(jac[,"Cellmarkers"],type="b",col="blue",pch=19,cex=1.5,lwd=2)
-arrows(x0=1:5, y0=(jac[,"Cellmarkers"] - mysd) ,  x1=1:5, y1=(jac[,"Cellmarkers"] + mysd), code=3, angle=90, length=0.1 , col="blue")
+plot(jac[,"Cellmarkers"],type="b", col="blue", pch=19,cex=1.5,lwd=2, xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected",main="Cellmarkers")
+mysd <- jsd[,"GO"]
+ub <- (jac[,"Cellmarkers"] + mysd)
+ub <- sapply(ub,ubtrim)
+lb <-(jac[,"Cellmarkers"] - mysd)
+lb <- sapply(lb,lbtrim)
+axis(1, at = seq(1, 5, by = 1), labels = ngenes)
+arrows(x0=1:5, y0=lb,  x1=1:5, y1=ub, code=3, angle=90, length=0.1 , col="blue")
 
+# TFT
 mysd <- jsd[,"TFT GTRD"]
-points(jac[,"TFT GTRD"],type="b",col="gray50",pch=19,cex=1.5,lwd=2)
-arrows(x0=1:5, y0=(jac[,"TFT GTRD"] - mysd) ,  x1=1:5, y1=(jac[,"TFT GTRD"] + mysd), code=3, angle=90, length=0.1 , col="gray50" )
+plot(jac[,"TFT GTRD"],type="b", col="gray50", pch=19,cex=1.5,lwd=2, xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected",main="TFT GTRD")
+axis(1, at = seq(1, 5, by = 1), labels = ngenes)
+ub <- (jac[,"TFT GTRD"] + mysd)
+ub <- sapply(ub,ubtrim)
+lb <-(jac[,"TFT GTRD"] - mysd)
+lb <- sapply(lb,lbtrim)
+arrows(x0=1:5, y0=lb,  x1=1:5, y1=ub, code=3, angle=90, length=0.1 , col="gray50" )
 
+png("fig2_ORAszsd.png")
+# GO
+plot(jac[,"GO"],type="b", pch=19,cex=1.5,lwd=2, xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected",main="GO")
+axis(1, at = seq(1, 5, by = 1), labels = ngenes)
+mysd <- jsd[,"GO"]
+ub <- (jac[,"GO"] + mysd)
+ub <- sapply(ub,ubtrim)
+lb <-(jac[,"GO"] - mysd)
+lb <- sapply(lb,lbtrim)
+arrows(x0=1:5, y0=lb ,  x1=1:5, y1=ub, code=3, angle=90, length=0.1 )
+
+# REACTOME
+mysd <- jsd[,"Reactome"]
+plot(jac[,"Reactome"],type="b", col="red", pch=19,cex=1.5,lwd=2, xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected",main="Reactome")
+axis(1, at = seq(1, 5, by = 1), labels = ngenes)
+ub <- (jac[,"Reactome"] + mysd)
+ub <- sapply(ub,ubtrim)
+lb <-(jac[,"Reactome"] - mysd)
+lb <- sapply(lb,lbtrim)
+arrows(x0=1:5, y0=lb ,  x1=1:5, y1=ub, code=3, angle=90, length=0.1 ,col="red")
+
+# Cellmarkers
+mysd <- jsd[,"Cellmarkers"]
+plot(jac[,"Cellmarkers"],type="b", col="blue", pch=19,cex=1.5,lwd=2, xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected",main="Cellmarkers")
+mysd <- jsd[,"GO"]
+ub <- (jac[,"Cellmarkers"] + mysd)
+ub <- sapply(ub,ubtrim)
+lb <-(jac[,"Cellmarkers"] - mysd)
+lb <- sapply(lb,lbtrim)
+axis(1, at = seq(1, 5, by = 1), labels = ngenes)
+arrows(x0=1:5, y0=lb,  x1=1:5, y1=ub, code=3, angle=90, length=0.1 , col="blue")
+
+# TFT
+mysd <- jsd[,"TFT GTRD"]
+plot(jac[,"TFT GTRD"],type="b", col="gray50", pch=19,cex=1.5,lwd=2, xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected",main="TFT GTRD")
+axis(1, at = seq(1, 5, by = 1), labels = ngenes)
+ub <- (jac[,"TFT GTRD"] + mysd)
+ub <- sapply(ub,ubtrim)
+lb <-(jac[,"TFT GTRD"] - mysd)
+lb <- sapply(lb,lbtrim)
+arrows(x0=1:5, y0=lb,  x1=1:5, y1=ub, code=3, angle=90, length=0.1 , col="gray50" )
+dev.off()
+
+pdf("fig2_ORAszsd.pdf",width=7, height=7)
+par(mfrow=c(2,2))
+# GO
+plot(jac[,"GO"],type="b", pch=19,cex=1.5,lwd=2, xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected",main="GO")
+axis(1, at = seq(1, 5, by = 1), labels = ngenes)
+mysd <- jsd[,"GO"]
+ub <- (jac[,"GO"] + mysd)
+ub <- sapply(ub,ubtrim)
+lb <-(jac[,"GO"] - mysd)
+lb <- sapply(lb,lbtrim)
+arrows(x0=1:5, y0=lb ,  x1=1:5, y1=ub, code=3, angle=90, length=0.1 )
+
+# REACTOME
+mysd <- jsd[,"Reactome"]
+plot(jac[,"Reactome"],type="b", col="red", pch=19,cex=1.5,lwd=2, xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected",main="Reactome")
+axis(1, at = seq(1, 5, by = 1), labels = ngenes)
+ub <- (jac[,"Reactome"] + mysd)
+ub <- sapply(ub,ubtrim)
+lb <-(jac[,"Reactome"] - mysd)
+lb <- sapply(lb,lbtrim)
+arrows(x0=1:5, y0=lb ,  x1=1:5, y1=ub, code=3, angle=90, length=0.1 ,col="red")
+
+# Cellmarkers
+mysd <- jsd[,"Cellmarkers"]
+plot(jac[,"Cellmarkers"],type="b", col="blue", pch=19,cex=1.5,lwd=2, xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected",main="Cellmarkers")
+mysd <- jsd[,"GO"]
+ub <- (jac[,"Cellmarkers"] + mysd)
+ub <- sapply(ub,ubtrim)
+lb <-(jac[,"Cellmarkers"] - mysd)
+lb <- sapply(lb,lbtrim)
+axis(1, at = seq(1, 5, by = 1), labels = ngenes)
+arrows(x0=1:5, y0=lb,  x1=1:5, y1=ub, code=3, angle=90, length=0.1 , col="blue")
+
+# TFT
+mysd <- jsd[,"TFT GTRD"]
+plot(jac[,"TFT GTRD"],type="b", col="gray50", pch=19,cex=1.5,lwd=2, xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected",main="TFT GTRD")
+axis(1, at = seq(1, 5, by = 1), labels = ngenes)
+ub <- (jac[,"TFT GTRD"] + mysd)
+ub <- sapply(ub,ubtrim)
+lb <-(jac[,"TFT GTRD"] - mysd)
+lb <- sapply(lb,lbtrim)
+arrows(x0=1:5, y0=lb,  x1=1:5, y1=ub, code=3, angle=90, length=0.1 , col="gray50" )
+dev.off()
+
+par(mfrow=c(1,1))
+
+# original
 plot(jac[,"GO"],type="b", pch=19,cex=1.5,lwd=2,
@@
 
+Boxplot.
+
+```{r,boxplot}
+
+colnames(res3a) <- paste(colnames(res3a),"orig")
+colnames(res3b) <- paste(colnames(res3b),"corr")
+res3d <- cbind(res3a,res3b)
+
+res3d <- as.data.frame(res3d[,c(7,16, 1,10, 9,18, 3,12, 4,13, 5,14, 8,17, 2,11, 6,15)])
+
+par(mar = c(5.1, 8.1, 4.1, 2.1))
+library(beeswarm)
+boxplot(res3d,horizontal=TRUE,las=1,col=c("pink","lightblue"),cex=0,xlab="no. significant sets FDR<0.05")
+beeswarm(res3d,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+
+res3d1 <- res3d + 1
+boxplot(res3d1,log="x",horizontal=TRUE,las=1,col=c("pink","lightblue"),cex=0,xlab="no. significant sets FDR<0.05*")
+beeswarm(res3d1,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+
+legend("bottomright", inset=.02, legend=c("corrected","original"),
+  fill=c("lightblue","pink"),col=c("blue","red"), horiz=FALSE, cex=1.1)
+
+png("fig2_ORAfdrbox.png")
+par(mar=c(c(5.1, 7.1, 2.1, 2.1) ))
+
+boxplot(res3d1,log="x",horizontal=TRUE,las=1,col=c("pink","lightblue"),cex=0,xlab="no. significant sets FDR<0.05*")
+beeswarm(res3d1,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+
+legend("bottomright", inset=.02, legend=c("corrected","original"),
+  fill=c("lightblue","pink"),col=c("blue","red"), horiz=FALSE, cex=1.1)
+
+dev.off()
+
+pdf("fig2_ORAfdrbox.pdf",width=7,height=7)
+par(mar=c(c(5.1, 7.1, 2.1, 2.1) ))
+
+boxplot(res3d1,log="x",horizontal=TRUE,las=1,col=c("pink","lightblue"),cex=0,xlab="no. significant sets FDR<0.05*")
+beeswarm(res3d1,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+
+legend("bottomright", inset=.02, legend=c("corrected","original"),
+  fill=c("lightblue","pink"),col=c("blue","red"), horiz=FALSE, cex=1.1)
+
+dev.off()
+
+```
+
 Another table showing downsampling result.

### 2024-09-10 analysis/sim/run_simpw.Rmd
sim update for granular data viz with vioplots boxplots etc

@@
 
-SIMS=1000
-#run 500 sims
+SIMS=50
+#run 1000 sims
 FRAC_DE=0.05
@@
   VARIANCE=mygrid[i,"VARIANCE"]
-  x <- agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets,cores=32)
-  as.data.frame(do.call(rbind, x))
+  agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets,cores=16)
 })
 
-res <- do.call(rbind,res)
+saveRDS(res,file="res.Rds")
 
-saveRDS(res,file="res.Rds")
+```
+
+## Aggregate data for charting as vioplot
+
+```{r,vioagg}
+
+library(vioplot)
+par(mfrow=c(3,3))
+
+## clusterprofiler default
+cp <- lapply(res,"[[",1)
+cp2 <- lapply(cp,function(df) {
+  df$p <- df$true_pos/(df$true_pos+ df$false_pos)
+  df$r <- df$true_pos/(df$true_pos+ df$false_neg)
+  df$f <- 2*df$p*df$r/(df$p+df$r)
+  df
+} )
+
+cp2p <- lapply(cp2,"[[","p")
+cp2p <- lapply(cp2p,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+cp2r <- lapply(cp2,"[[","r")
+cp2r <- lapply(cp2r,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+cp2f <- lapply(cp2,"[[","f")
+cp2f <- lapply(cp2f,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+names(cp2p) <- names(cp2r) <- names(cp2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")
+
+vioplot(cp2p,xlab="introduced noise",ylab="precision",main="clusterProfiler") ; grid()
+vioplot(cp2r,xlab="introduced noise",ylab="recall",main="clusterProfiler") ; grid()
+vioplot(cp2f,xlab="introduced noise",ylab="F1",main="clusterProfiler") ; grid()
+
+## clusterprofiler bgfix
+cpbg <- lapply(res,"[[",2)
+cpbg2 <- lapply(cpbg,function(df) {
+  df$p <- df$true_pos/(df$true_pos+ df$false_pos)
+  df$r <- df$true_pos/(df$true_pos+ df$false_neg)
+  df$f <- 2*df$p*df$r/(df$p+df$r)
+  df
+} )
+
+cpbg2p <- lapply(cpbg2,"[[","p")
+cpbg2p <- lapply(cpbg2p,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+cpbg2r <- lapply(cpbg2,"[[","r")
+cpbg2r <- lapply(cpbg2r,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+cpbg2f <- lapply(cpbg2,"[[","f")
+cpbg2f <- lapply(cpbg2f,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+names(cpbg2p) <- names(cpbg2r) <- names(cpbg2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")
+
+vioplot(cpbg2p,xlab="introduced noise",ylab="precision",main="clusterProfiler BG fix") ; grid()
+vioplot(cpbg2r,xlab="introduced noise",ylab="recall",main="clusterProfiler BG fix") ; grid()
+vioplot(cpbg2f,xlab="introduced noise",ylab="F1",main="clusterProfiler BG fix") ; grid()
+
+## clusterprofiler FDR fix
+cpfdr <- lapply(res,"[[",3)
+cpfdr2 <- lapply(cpfdr,function(df) {
+  df$p <- df$true_pos/(df$true_pos+ df$false_pos)
+  df$r <- df$true_pos/(df$true_pos+ df$false_neg)
+  df$f <- 2*df$p*df$r/(df$p+df$r)
+  df
+} )
+
+cpfdr2p <- lapply(cpfdr2,"[[","p")
+cpfdr2p <- lapply(cpfdr2p,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+cpfdr2r <- lapply(cpfdr2,"[[","r")
+cpfdr2r <- lapply(cpfdr2r,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+cpfdr2f <- lapply(cpfdr2,"[[","f")
+cpfdr2f <- lapply(cpfdr2f,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+names(cpfdr2p) <- names(cpfdr2r) <- names(cpfdr2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")
+
+vioplot(cpfdr2p,xlab="introduced noise",ylab="precision",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpfdr2r,xlab="introduced noise",ylab="recall",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpfdr2f,xlab="introduced noise",ylab="F1",main="clusterProfiler FDR fix") ; grid()
+
+## clusterprofiler BG and FDR fix
+cpbgfdr <- lapply(res,"[[",4)
+cpbgfdr2 <- lapply(cpbgfdr,function(df) {
+  df$p <- df$true_pos/(df$true_pos+ df$false_pos)
+  df$r <- df$true_pos/(df$true_pos+ df$false_neg)
+  df$f <- 2*df$p*df$r/(df$p+df$r)
+  df
+} )
+
+cpbgfdr2p <- lapply(cpbgfdr2,"[[","p")
+cpbgfdr2p <- lapply(cpbgfdr2p,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+cpbgfdr2r <- lapply(cpbgfdr2,"[[","r")
+cpbgfdr2r <- lapply(cpbgfdr2r,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+cpbgfdr2f <- lapply(cpbgfdr2,"[[","f")
+cpbgfdr2f <- lapply(cpbgfdr2f,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+names(cpbgfdr2p) <- names(cpbgfdr2r) <- names(cpbgfdr2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")
+
+vioplot(cpbgfdr2p,xlab="introduced noise",ylab="precision",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(cpbgfdr2r,xlab="introduced noise",ylab="recall",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(cpbgfdr2f,xlab="introduced noise",ylab="F1",main="clusterProfiler BG and FDR fix") ; grid()
+
+## FORA
+fo <- lapply(res,"[[",5)
+fo2 <- lapply(fo,function(df) {
+  df$p <- df$true_pos/(df$true_pos+ df$false_pos)
+  df$r <- df$true_pos/(df$true_pos+ df$false_neg)
+  df$f <- 2*df$p*df$r/(df$p+df$r)
+  df
+} )
+
+fo2p <- lapply(fo2,"[[","p")
+fo2p <- lapply(fo2p,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+fo2r <- lapply(fo2,"[[","r")
+fo2r <- lapply(fo2r,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+fo2f <- lapply(fo2,"[[","f")
+fo2f <- lapply(fo2f,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+names(fo2p) <- names(fo2r) <- names(fo2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")
+
+vioplot(fo2p,xlab="introduced noise",ylab="precision",main="fora") ; grid()
+vioplot(fo2r,xlab="introduced noise",ylab="recall",main="fora") ; grid()
+vioplot(fo2f,xlab="introduced noise",ylab="F1",main="fora") ; grid()
+
+## FGSEA
+fg <- lapply(res,"[[",6)
+fg2 <- lapply(fg,function(df) {
+  df$p <- df$true_pos/(df$true_pos+ df$false_pos)
+  df$r <- df$true_pos/(df$true_pos+ df$false_neg)
+  df$f <- 2*df$p*df$r/(df$p+df$r)
+  df
+} )
+
+fg2p <- lapply(fg2,"[[","p")
+fg2p <- lapply(fg2p,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+fg2r <- lapply(fg2,"[[","r")
+fg2r <- lapply(fg2r,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+fg2f <- lapply(fg2,"[[","f")
+fg2f <- lapply(fg2f,function(x) { x[is.na(x) ]<- 0 ; x } )
+
+names(fg2p) <- names(fg2r) <- names(fg2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")
+
+vioplot(fg2p,xlab="introduced noise",ylab="precision",main="fgsea") ; grid()
+vioplot(fg2r,xlab="introduced noise",ylab="recall",main="fgsea") ; grid()
+vioplot(fg2f,xlab="introduced noise",ylab="F1",main="fgsea") ; grid()
+
+png("fig3_vio.png", width=7,height=7,units="in",res=150,pointsize=12)
+par(mfrow=c(3,3))
+vioplot(cp2p,xlab="introduced noise",ylab="precision",main="clusterProfiler") ; grid()
+vioplot(cp2r,xlab="introduced noise",ylab="recall",main="clusterProfiler") ; grid()
+vioplot(cp2f,xlab="introduced noise",ylab="F1",main="clusterProfiler") ; grid()
+vioplot(cpbg2p,xlab="introduced noise",ylab="precision",main="clusterProfiler BG fix") ; grid()
+vioplot(cpbg2r,xlab="introduced noise",ylab="recall",main="clusterProfiler BG fix") ; grid()
+vioplot(cpbg2f,xlab="introduced noise",ylab="F1",main="clusterProfiler BG fix") ; grid()
+vioplot(cpfdr2p,xlab="introduced noise",ylab="precision",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpfdr2r,xlab="introduced noise",ylab="recall",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpfdr2f,xlab="introduced noise",ylab="F1",main="clusterProfiler FDR fix") ; grid()
+dev.off()
+png("fig3_vio2.png", width=7,height=7,units="in",res=150,pointsize=12)
+par(mfrow=c(3,3))
+vioplot(cpbgfdr2p,xlab="introduced noise",ylab="precision",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(cpbgfdr2r,xlab="introduced noise",ylab="recall",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(cpbgfdr2f,xlab="introduced noise",ylab="F1",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(fo2p,xlab="introduced noise",ylab="precision",main="fora") ; grid()
+vioplot(fo2r,xlab="introduced noise",ylab="recall",main="fora") ; grid()
+vioplot(fo2f,xlab="introduced noise",ylab="F1",main="fora") ; grid()
+vioplot(fo2p,xlab="introduced noise",ylab="precision",main="fora") ; grid()
+vioplot(fo2r,xlab="introduced noise",ylab="recall",main="fora") ; grid()
+vioplot(fo2f,xlab="introduced noise",ylab="F1",main="fora") ; grid()
+dev.off()
+
+pdf("fig3_vio.pdf", width=7,height=7,pointsize=12)
+par(mfrow=c(3,3))
+vioplot(cp2p,xlab="introduced noise",ylab="precision",main="clusterProfiler") ; grid()
+vioplot(cp2r,xlab="introduced noise",ylab="recall",main="clusterProfiler") ; grid()
+vioplot(cp2f,xlab="introduced noise",ylab="F1",main="clusterProfiler") ; grid()
+vioplot(cpbg2p,xlab="introduced noise",ylab="precision",main="clusterProfiler BG fix") ; grid()
+vioplot(cpbg2r,xlab="introduced noise",ylab="recall",main="clusterProfiler BG fix") ; grid()
+vioplot(cpbg2f,xlab="introduced noise",ylab="F1",main="clusterProfiler BG fix") ; grid()
+vioplot(cpfdr2p,xlab="introduced noise",ylab="precision",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpfdr2r,xlab="introduced noise",ylab="recall",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpfdr2f,xlab="introduced noise",ylab="F1",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpbgfdr2p,xlab="introduced noise",ylab="precision",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(cpbgfdr2r,xlab="introduced noise",ylab="recall",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(cpbgfdr2f,xlab="introduced noise",ylab="F1",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(fo2p,xlab="introduced noise",ylab="precision",main="fora") ; grid()
+vioplot(fo2r,xlab="introduced noise",ylab="recall",main="fora") ; grid()
+vioplot(fo2f,xlab="introduced noise",ylab="F1",main="fora") ; grid()
+vioplot(fo2p,xlab="introduced noise",ylab="precision",main="fora") ; grid()
+vioplot(fo2r,xlab="introduced noise",ylab="recall",main="fora") ; grid()
+vioplot(fo2f,xlab="introduced noise",ylab="F1",main="fora") ; grid()
+dev.off()
 
@@
 ```{r,results}
+
+res2 <- lapply(res,function(x) { lapply(x,function(y) { colMeans(y[,1:3]) } ) } )
+res2 <- lapply(res2,function(x) {
+  out <- as.data.frame(do.call(rbind,x))
+  pw <- c("cp","cpbg","cpfdr","cpbgfdr","fora","fgsea")
+  out$pw <- pw
+  out
+} )
+res3 <- do.call(rbind,res2)
+res3$var <- c(rep(0,6),rep(0.1,6),rep(0.2,6),rep(0.3,6),rep(0.4,6),rep(0.5,6),rep(0.6,6))
 
@@
 
+######################################
+# Now as boxplot
+library(beeswarm)
+
+prec <- list("CP default"=cp$p,"CP BG fix"=cpbg$p,"CP FDR fix"=cpfdr$p,"CP BG and FDR fix"=cpbgfdr$p,"fora"=fo$p,"fgsea"=fg$p)
+rec <- list("CP default"=cp$r,"CP BG fix"=cpbg$r,"CP FDR fix"=cpfdr$r,"CP BG and FDR fix"=cpbgfdr$r,"fora"=fo$r,"fgsea"=fg$r)
+f1 <- lapply(1:length(prec), function(i) { 2/(1/rec[[i]]+1/prec[[i]]) })
+names(f1) <- names(prec)
+
+par(mfrow=c(1,3))
+par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
+boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
+beeswarm(prec,add=TRUE,col="black",pch=19)
+boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
+beeswarm(rec,add=TRUE,col="black",pch=19)
+boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
+beeswarm(f1,add=TRUE,col="black",pch=19)
+
+png("fig3_box.png", width=7,height=5,units="in",res=150,pointsize=12)
+par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
+par(mfrow=c(1,3))
+boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
+beeswarm(prec,add=TRUE,col="black",pch=19)
+boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
+beeswarm(rec,add=TRUE,col="black",pch=19)
+boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
+beeswarm(f1,add=TRUE,col="black",pch=19)
+dev.off()
+
+pdf("fig3_bars.pdf", width=7,height=5,pointsize=12)
+par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
+par(mfrow=c(1,3))
+boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
+beeswarm(prec,add=TRUE,col="black",pch=19)
+boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
+beeswarm(rec,add=TRUE,col="black",pch=19)
+boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
+beeswarm(f1,add=TRUE,col="black",pch=19)
+dev.off()
+
 ```

### 2024-09-10 analysis/sim/simpw_func.R

@@ -187,7 +187,7 @@ deseq2<-function(x) {
 # clusterprofiler default function
 ##################################
 # Note that clusterprofiler requires different gene set format
-run_clusterprofiler_default <- function(x,DGE_FUNC,gsets,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,cores=16){
+run_clusterprofiler_default <- function(x,DGE_FUNC,gsets,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,cores=32){
 
 gsets2 <- stack(gsets)[,c(2,1)]
 colnames(gsets2) <- c("term","gene")
@@ -248,22 +248,15 @@ gt_dn <- lapply( gt_dn , names)
 
 true_pos_up <- as.numeric(mapply( function(x,y) length(intersect(x,y)) , obs_up ,  gt_up ))
 true_pos_dn <- as.numeric(mapply( function(x,y) length(intersect(x,y)) , obs_dn ,  gt_dn ))
+true_pos <- true_pos_up + true_pos_dn
 false_pos_up <- as.numeric(mapply( function(x,y) length(setdiff(x,y)) , obs_up ,  gt_up ))
 false_pos_dn <- as.numeric(mapply( function(x,y) length(setdiff(x,y)) , obs_dn , gt_dn ))
+false_pos <- false_pos_up + false_pos_dn
 false_neg_up <- as.numeric(mapply( function(x,y) length(setdiff(x,y)) , gt_up ,  obs_up ))
 false_neg_dn <- as.numeric(mapply( function(x,y) length(setdiff(x,y)) , gt_dn ,  obs_dn ))
+false_neg <- false_neg_up + false_neg_dn
 
-true_pos <- mean(true_pos_up+true_pos_dn)
-false_pos <- mean(false_pos_up+false_pos_dn)
-false_neg <- mean(false_neg_up+false_neg_dn)
-nrows <- as.numeric(lapply( sapply(x,"[",1 ), nrow))
-true_neg <- mean(nrows-(true_pos+false_pos+false_neg))
-
-p <- true_pos/(true_pos+false_pos)
-r <- true_pos/(true_pos+false_neg)
-f <- 2*p*r/(p+r)
-
[…]

### 2024-09-11 dataprep/bulkrna1.Rmd
adding thread control to data prep

@@
   library("biomaRt")
+  library("RhpcBLASctl")
 })
+
+RhpcBLASctl::blas_set_num_threads(1)
 

### 2024-09-11 analysis/analyse_rnaseq1.Rmd
thread control for fig 1 and 2

@@
   library("parallel")
+  library("RhpcBLASctl")
 })
+
+RhpcBLASctl::blas_set_num_threads(1)
 

### 2024-09-11 analysis/sim/run_simpw.Rmd
thread control for sims

@@
 source("simpw_func.R")
+library("RhpcBLASctl")
+
+RhpcBLASctl::blas_set_num_threads(1)
 
@@
 
-SIMS=50
+SIMS=32
 #run 1000 sims
@@
   VARIANCE=mygrid[i,"VARIANCE"]
-  agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets,cores=16)
+  agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets,cores=32)
 })

### 2024-09-11 analysis/sim/simpw_func.R

@@ -7,6 +7,9 @@ library("stringi")
 library("mitch")
 library("fgsea")
 library("clusterProfiler")
+library("RhpcBLASctl")
+
+RhpcBLASctl::blas_set_num_threads(1)
 
 ########################################
 # get some counts

### 2024-09-11 analysis/sim/run_simpw.Rmd
works

--- 2024-07-09 analysis/sim/run_simpw.Rmd
@@
 source("simpw_func.R")
+library("RhpcBLASctl")
+library("vioplot")
+library("beeswarm")
+
+RhpcBLASctl::blas_set_num_threads(1)
 
@@
 
-SIMS=1000
-#run 500 sims
+SIMS=32
+#run 1000 sims
 FRAC_DE=0.05
@@
   VARIANCE=mygrid[i,"VARIANCE"]
-  x <- agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets,cores=32)
-  as.data.frame(do.call(rbind, x))
+  agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets,cores=32)
 })
 
-res <- do.call(rbind,res)
+saveRDS(res,file="res.Rds")
 
-saveRDS(res,file="res.Rds")
+```
+
+## Aggregate data for charting as vioplot
+
+```{r,vioagg}
+
+prf <- function(cp) {
+  cp2 <- lapply(cp,function(df) {
+    df$p <- df$true_pos/(df$true_pos+ df$false_pos)
+    df$r <- df$true_pos/(df$true_pos+ df$false_neg)
+    df$f <- 2*df$p*df$r/(df$p+df$r)
+    df
+  } )
+  cp2p <- lapply(cp2,"[[","p")
+  cp2p <- lapply(cp2p,function(x) { x[is.na(x) ]<- 0 ; x } )
+  cp2r <- lapply(cp2,"[[","r")
+  cp2r <- lapply(cp2r,function(x) { x[is.na(x) ]<- 0 ; x } )
+  cp2f <- lapply(cp2,"[[","f")
+  cp2f <- lapply(cp2f,function(x) { x[is.na(x) ]<- 0 ; x } )
+  names(cp2p) <- names(cp2r) <- names(cp2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")
+  res=list("p"=cp2p,"r"=cp2r,"f"=cp2f)
+  return(res)
+}
+
+par(mfrow=c(3,3))
+
+## clusterprofiler default
+cp <- lapply(res,"[[",1)
+cp2 <- prf(cp)
+vioplot(cp2[["p"]],xlab="introduced noise",ylab="precision",main="clusterProfiler") ; grid()
+vioplot(cp2[["r"]],xlab="introduced noise",ylab="recall",main="clusterProfiler") ; grid()
+vioplot(cp2[["f"]],xlab="introduced noise",ylab="F1",main="clusterProfiler") ; grid()
+
+## clusterprofiler bgfix
+cpbg <- lapply(res,"[[",2)
+cpbg2 <- prf(cpbg)
+vioplot(cpbg2[["p"]],xlab="introduced noise",ylab="precision",main="clusterProfiler BG fix") ; grid()
+vioplot(cpbg2[["r"]],xlab="introduced noise",ylab="recall",main="clusterProfiler BG fix") ; grid()
+vioplot(cpbg2[["f"]],xlab="introduced noise",ylab="F1",main="clusterProfiler BG fix") ; grid()
+
+## clusterprofiler FDR fix
+cpfdr <- lapply(res,"[[",3)
+cpfdr2 <- prf(cpfdr)
+vioplot(cpfdr2[["p"]],xlab="introduced noise",ylab="precision",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpfdr2[["r"]],xlab="introduced noise",ylab="recall",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpfdr2[["f"]],xlab="introduced noise",ylab="F1",main="clusterProfiler FDR fix") ; grid()
+
+## clusterprofiler BG and FDR fix
+cpbgfdr <- lapply(res,"[[",4)
+cpbgfdr2 <- prf(cpbgfdr)
+vioplot(cpbgfdr2[["p"]],xlab="introduced noise",ylab="precision",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(cpbgfdr2[["r"]],xlab="introduced noise",ylab="recall",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(cpbgfdr2[["f"]],xlab="introduced noise",ylab="F1",main="clusterProfiler BG and FDR fix") ; grid()
+
+## FORA
+fo <- lapply(res,"[[",5)
+fo2 <- prf(fo)
+vioplot(fo2[["p"]],xlab="introduced noise",ylab="precision",main="fora") ; grid()
+vioplot(fo2[["r"]],xlab="introduced noise",ylab="recall",main="fora") ; grid()
+vioplot(fo2[["x"]],xlab="introduced noise",ylab="F1",main="fora") ; grid()
+
+## FGSEA
+fg <- lapply(res,"[[",6)
+fg2 <- prf(fg)
+vioplot(fg2[["p"]],xlab="introduced noise",ylab="precision",main="fgsea") ; grid()
+vioplot(fg2[["r"]],xlab="introduced noise",ylab="recall",main="fgsea") ; grid()
+vioplot(fg2[["f"]],xlab="introduced noise",ylab="F1",main="fgsea") ; grid()
+
+png("fig3_vio.png", width=7,height=7,units="in",res=150,pointsize=12)
+par(mfrow=c(3,3))
+vioplot(cp2[["p"]],xlab="introduced noise",ylab="precision",main="clusterProfiler") ; grid()
+vioplot(cp2[["r"]],xlab="introduced noise",ylab="recall",main="clusterProfiler") ; grid()
+vioplot(cp2[["f"]],xlab="introduced noise",ylab="F1",main="clusterProfiler") ; grid()
+vioplot(cpbg2[["p"]],xlab="introduced noise",ylab="precision",main="clusterProfiler BG fix") ; grid()
+vioplot(cpbg2[["r"]],xlab="introduced noise",ylab="recall",main="clusterProfiler BG fix") ; grid()
+vioplot(cpbg2[["f"]],xlab="introduced noise",ylab="F1",main="clusterProfiler BG fix") ; grid()
+vioplot(cpfdr2[["p"]],xlab="introduced noise",ylab="precision",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpfdr2[["r"]],xlab="introduced noise",ylab="recall",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpfdr2[["f"]],xlab="introduced noise",ylab="F1",main="clusterProfiler FDR fix") ; grid()
+dev.off()
+png("fig3_vio2.png", width=7,height=7,units="in",res=150,pointsize=12)
+par(mfrow=c(3,3))
+vioplot(cpbgfdr2[["p"]],xlab="introduced noise",ylab="precision",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(cpbgfdr2[["r"]],xlab="introduced noise",ylab="recall",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(cpbgfdr2[["f"]],xlab="introduced noise",ylab="F1",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(fo2[["p"]],xlab="introduced noise",ylab="precision",main="fora") ; grid()
+vioplot(fo2[["r"]],xlab="introduced noise",ylab="recall",main="fora") ; grid()
+vioplot(fo2[["f"]],xlab="introduced noise",ylab="F1",main="fora") ; grid()
+vioplot(fg2[["p"]],xlab="introduced noise",ylab="precision",main="fgsea") ; grid()
+vioplot(fg2[["r"]],xlab="introduced noise",ylab="recall",main="fgsea") ; grid()
+vioplot(fg2[["f"]],xlab="introduced noise",ylab="F1",main="fgsea") ; grid()
+dev.off()
+
+pdf("fig3_vio.pdf", width=7,height=7,pointsize=12)
+par(mfrow=c(3,3))
+vioplot(cp2[["p"]],xlab="introduced noise",ylab="precision",main="clusterProfiler") ; grid()
+vioplot(cp2[["r"]],xlab="introduced noise",ylab="recall",main="clusterProfiler") ; grid()
+vioplot(cp2[["f"]],xlab="introduced noise",ylab="F1",main="clusterProfiler") ; grid()
+vioplot(cpbg2[["p"]],xlab="introduced noise",ylab="precision",main="clusterProfiler BG fix") ; grid()
+vioplot(cpbg2[["r"]],xlab="introduced noise",ylab="recall",main="clusterProfiler BG fix") ; grid()
+vioplot(cpbg2[["f"]],xlab="introduced noise",ylab="F1",main="clusterProfiler BG fix") ; grid()
+vioplot(cpfdr2[["p"]],xlab="introduced noise",ylab="precision",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpfdr2[["r"]],xlab="introduced noise",ylab="recall",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpfdr2[["f"]],xlab="introduced noise",ylab="F1",main="clusterProfiler FDR fix") ; grid()
+vioplot(cpbgfdr2[["p"]],xlab="introduced noise",ylab="precision",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(cpbgfdr2[["r"]],xlab="introduced noise",ylab="recall",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(cpbgfdr2[["f"]],xlab="introduced noise",ylab="F1",main="clusterProfiler BG and FDR fix") ; grid()
+vioplot(fo2[["p"]],xlab="introduced noise",ylab="precision",main="fora") ; grid()
+vioplot(fo2[["r"]],xlab="introduced noise",ylab="recall",main="fora") ; grid()
+vioplot(fo2[["f"]],xlab="introduced noise",ylab="F1",main="fora") ; grid()
+vioplot(fg2[["p"]],xlab="introduced noise",ylab="precision",main="fgsea") ; grid()
+vioplot(fg2[["r"]],xlab="introduced noise",ylab="recall",main="fgsea") ; grid()
+vioplot(fg2[["f"]],xlab="introduced noise",ylab="F1",main="fgsea") ; grid()
+dev.off()
 
@@
 
+res2 <- lapply(res,function(x) { lapply(x,function(y) { colMeans(y[,1:3]) } ) } )
+res2 <- lapply(res2,function(x) {
+  out <- as.data.frame(do.call(rbind,x))
+  PWAY_FUNC  <- c("clusterProfiler default","clusterProfiler BG fix","clusterProfiler FDR fix",
+    "clusterProfiler BG and FDR fix","fora","fgsea")
+  out$PWAY_FUNC  <- PWAY_FUNC
+  out
+} )
+res3 <- do.call(rbind,res2)
+res3$VARIANCE <- c(rep(0,6),rep(0.1,6),rep(0.2,6),rep(0.3,6),rep(0.4,6),rep(0.5,6),rep(0.6,6))
+
+res3$p <- res3$true_pos/(res3$true_pos+ res3$false_pos)
+res3$r <- res3$true_pos/(res3$true_pos+ res3$false_neg)
+res3$f <- 2*res3$p*res3$r/(res3$p+res3$r)
+
 res %>% kbl(caption="simulation_results") %>% kable_paper("hover", full_width = F)
@@
 
-fig1_ORAjac.png
-
 ```{r,plot1}
 
-cp <- subset(res,PWAY_FUNC == "clusterProfiler default")
-cpbg <- subset(res,PWAY_FUNC == "clusterProfiler BG fix")
-cpfdr <- subset(res,PWAY_FUNC == "clusterProfiler FDR fix")
-cpbgfdr <- subset(res,PWAY_FUNC == "clusterProfiler BG and FDR fix")
-fo <- subset(res,PWAY_FUNC == "fora")
-fg <- subset(res,PWAY_FUNC == "fgsea")
+cp <- subset(res3,PWAY_FUNC == "clusterProfiler default")
+cpbg <- subset(res3,PWAY_FUNC == "clusterProfiler BG fix")
+cpfdr <- subset(res3,PWAY_FUNC == "clusterProfiler FDR fix")
+cpbgfdr <- subset(res3,PWAY_FUNC == "clusterProfiler BG and FDR fix")
+fo <- subset(res3,PWAY_FUNC == "fora")
+fg <- subset(res3,PWAY_FUNC == "fgsea")
 
@@
 
+######################################
+# Now as boxplot
+
+prec <- list("CP default"=cp$p,"CP BG fix"=cpbg$p,"CP FDR fix"=cpfdr$p,"CP BG and FDR fix"=cpbgfdr$p,"fora"=fo$p,"fgsea"=fg$p)
+rec <- list("CP default"=cp$r,"CP BG fix"=cpbg$r,"CP FDR fix"=cpfdr$r,"CP BG and FDR fix"=cpbgfdr$r,"fora"=fo$r,"fgsea"=fg$r)
+f1 <- lapply(1:length(prec), function(i) { 2/(1/rec[[i]]+1/prec[[i]]) })
+names(f1) <- names(prec)
+
+par(mfrow=c(1,3))
+par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
+boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
+beeswarm(prec,add=TRUE,col="black",pch=19)
+boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
+beeswarm(rec,add=TRUE,col="black",pch=19)
+boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
+beeswarm(f1,add=TRUE,col="black",pch=19)
+
+png("fig3_box.png", width=7,height=5,units="in",res=150,pointsize=12)
+par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
+par(mfrow=c(1,3))
+boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
+beeswarm(prec,add=TRUE,col="black",pch=19)
+boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
+beeswarm(rec,add=TRUE,col="black",pch=19)
+boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
+beeswarm(f1,add=TRUE,col="black",pch=19)
+dev.off()
+
+pdf("fig3_bars.pdf", width=7,height=5,pointsize=12)
+par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
+par(mfrow=c(1,3))
+boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
+beeswarm(prec,add=TRUE,col="black",pch=19)
+boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
+beeswarm(rec,add=TRUE,col="black",pch=19)
+boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
+beeswarm(f1,add=TRUE,col="black",pch=19)
+dev.off()
+
 ```

### 2024-09-11 analysis/sim/run_simpw.Rmd
bugfix

@@
 vioplot(fo2[["r"]],xlab="introduced noise",ylab="recall",main="fora") ; grid()
-vioplot(fo2[["x"]],xlab="introduced noise",ylab="F1",main="fora") ; grid()
+vioplot(fo2[["f"]],xlab="introduced noise",ylab="F1",main="fora") ; grid()
 

### 2024-09-11 analysis/sim/run_simpw.Rmd
fixnames

@@
 
-pdf("fig3_bars.pdf", width=7,height=5,pointsize=12)
+pdf("fig3_box.pdf", width=7,height=5,pointsize=12)
 par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))

### 2024-09-12 analysis/analyse_rnaseq1.Rmd
adding mean to boxplot

@@
 beeswarm(unsl,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+beeswarm(as.list(colMeans(unsl)),horizontal=TRUE,add=TRUE,col="black",pch="x",cex=1.0)
 
@@
 beeswarm(unsl1,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+beeswarm(as.list(colMeans(unsl1)),horizontal=TRUE,add=TRUE,col="black",pch="x",cex=1.0)
 
@@
 beeswarm(unsl1,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+beeswarm(as.list(colMeans(unsl1)),horizontal=TRUE,add=TRUE,col="black",pch="x",cex=1.0)
 legend("bottomright", inset=.02, legend=c("corrected","original"),

### 2024-09-12 analysis/analyse_rnaseq2.Rmd
add means to boxplot

@@
 beeswarm(res3d,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+beeswarm(as.list(colMeans(res3d)),horizontal=TRUE,add=TRUE,col="black",pch="x",cex=1.0)
 
@@
 beeswarm(res3d1,horizontal=TRUE,add=TRUE,col=c("red","blue"),pch=19,cex=0.6)
+beeswarm(as.list(colMeans(res3d1)),horizontal=TRUE,add=TRUE,col="black",pch="x",cex=1.0)
 

### 2024-09-12 docker_results/analysis/sim/run_simpw.Rmd
better boxplots with means

--- 2024-09-11 analysis/sim/run_simpw.Rmd
@@
 
-SIMS=32
+SIMS=1000
 #run 1000 sims
@@
 
+if (file.exists("res.Rds"") {
+
+  res <- readRDS("res.Rds")
+
+} else {
+
 res <- lapply(1:nrow(mygrid), function(i) {
@@
 saveRDS(res,file="res.Rds")
+
+}
 
@@
 boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
-beeswarm(prec,add=TRUE,col="black",pch=19)
+beeswarm(prec,add=TRUE,col="darkgray",pch=19)
+beeswarm(lapply(prec,mean),add=TRUE,col="black",pch="x")
 boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
-beeswarm(rec,add=TRUE,col="black",pch=19)
+beeswarm(rec,add=TRUE,col="darkgray",pch=19)
+beeswarm(lapply(rec,mean),add=TRUE,col="black",pch="x")
 boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
-beeswarm(f1,add=TRUE,col="black",pch=19)
+beeswarm(f1,add=TRUE,col="darkgray",pch=19)
+beeswarm(lapply(prec,f1),add=TRUE,col="black",pch="x")
 
@@
 boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
-beeswarm(prec,add=TRUE,col="black",pch=19)
+beeswarm(prec,add=TRUE,col="darkgray",pch=19)
+beeswarm(lapply(prec,mean),add=TRUE,col="black",pch="x")
 boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
-beeswarm(rec,add=TRUE,col="black",pch=19)
+beeswarm(rec,add=TRUE,col="darkgray",pch=19)
+beeswarm(lapply(rec,mean),add=TRUE,col="black",pch="x")
 boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
-beeswarm(f1,add=TRUE,col="black",pch=19)
+beeswarm(f1,add=TRUE,col="darkgray",pch=19)
+beeswarm(lapply(prec,f1),add=TRUE,col="black",pch="x")
 dev.off()
@@
 
+save.image("run_simpw.Rdata")
+
 sessionInfo()

### 2024-09-12 analysis/sim/run_simpw.Rmd
more sims
sims
savedata at end
fig3 boxplot with mean

@@
 boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
-beeswarm(prec,add=TRUE,col="black",pch=19)
+beeswarm(prec,add=TRUE,col="darkgray",pch=19)
+beeswarm(lapply(prec,mean),add=TRUE,col="black",pch="x",cex=1.0)
 boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
-beeswarm(rec,add=TRUE,col="black",pch=19)
+beeswarm(rec,add=TRUE,col="darkgray",pch=19)
+beeswarm(lapply(rec,mean),add=TRUE,col="black",pch="x",cex=1.0)
 boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
-beeswarm(f1,add=TRUE,col="black",pch=19)
+beeswarm(f1,add=TRUE,col="darkgray",pch=19)
+beeswarm(lapply(f1,mean),add=TRUE,col="black",pch="x",cex=1.0)
 
@@
 boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
-beeswarm(prec,add=TRUE,col="black",pch=19)
+beeswarm(prec,add=TRUE,col="darkgray",pch=19)
+beeswarm(lapply(prec,mean),add=TRUE,col="black",pch="x",cex=1.0)
 boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
-beeswarm(rec,add=TRUE,col="black",pch=19)
+beeswarm(rec,add=TRUE,col="darkgray",pch=19)
+beeswarm(lapply(rec,mean),add=TRUE,col="black",pch="x",cex=1.0)
 boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
-beeswarm(f1,add=TRUE,col="black",pch=19)
+beeswarm(f1,add=TRUE,col="darkgray",pch=19)
+beeswarm(lapply(f1,mean),add=TRUE,col="black",pch="x",cex=1.0)
 dev.off()

### 2024-09-13 analysis/GSE158420_rna.Rmd
png
ms
nice violin plot for downsampling

@@
   library("edgeR")
+  library("vioplot")
 })
@@
 
-This will run the FGSEA analysis at group sizes of 2, 3, 6, 10, 16, 22 and 30,
+This will run the FGSEA analysis at group sizes of 2, 3, 5, 10, 16, 22 and 30,
 50 times each, with pseudorandom seeds from 100 to 5000 jumping up by 100.
@@
 colnames(df) <- c("Sample size", "fora", "fgsea")
+
+df
 
@@
 
-df
-
 MAX=length(fsets)
@@
 
-df
-
 MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
@@
 
-df
+MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="proportion",main="Proportion of inconsistent pathways",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+grid()
+dev.off()
+
+### PNG
+png("fig4.png", width=7,height=5,units="in",res=150,pointsize=12)
+par(mfrow=c(1,3))
+
+dots <- lapply(fgres,function(x) { x[,"TP"] })
+fgres2 <- do.call(rbind,lapply(dots,summary))
+
+dots <- lapply(fores,function(x) { x[,"TP"] })
+fores2 <- do.call(rbind,lapply(dots,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+MAX=length(fsets)
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways", main="no. consistent pathways FDR<0.05",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+abline(h=length(osets),col="red")
+abline(h=length(fsets),col="blue")
+grid()
+
+legend("bottomright", legend=c("fgsea","fora"),
+       col=c("blue","red"), lty=1, pch=19,  cex=1)
+
+dots <- lapply(fgres,function(x) { x[,"FP"] })
+fgres2 <- do.call(rbind,lapply(dots,summary))
+
+dots <- lapply(fores,function(x) { x[,"FP"] })
+fores2 <- do.call(rbind,lapply(dots,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
+
+MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
+plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways",main="no. inconsistent pathways FDR<0.05",ylim=c(0,MAX))
+points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
+grid()
+
+dots <- lapply(fores,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(fores,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+fores2 <- do.call(rbind,lapply(ratio,summary))
+
+dots <- lapply(fgres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(fgres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+fgres2 <- do.call(rbind,lapply(ratio,summary))
+
+df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
+colnames(df) <- c("Sample size", "fora", "fgsea")
 
@@
 
+## Make violin plot
+
+```{r,vioplot}
+
+length(osets)
+length(fsets)
+
+names(fores) <- names(fgres) <- as.character(c(2,3,5,10,15,20,25,30))
+
+par(mfrow=c(1,3))
+
+vioplot(lapply(fores,function(mx) { mx[,2] } ) , side="left" , col="white" , border = "red",
+  plotCentre = "line", ylim=c(0,457), xlab="Sample size", ylab="no. pathways")
+  mtext("no. consistent pathways FDR<0.05" )
+vioplot(lapply(fgres,function(mx) { mx[,2] } ) , side="right" , col="white", border="blue" ,
+   add = TRUE, plotCentre = "line")
+
+abline(h=length(osets),col="red")
+abline(h=length(fsets),col="blue")
+
+vioplot(lapply(fores,function(mx) { mx[,3] } ) , side="left" , col="white" , border="red",
+  plotCentre = "line", ylim=c(0,210), xlab="Sample size", ylab="no. pathways")
+  mtext("no. inconsistent pathways FDR<0.05" )
+vioplot(lapply(fgres,function(mx) { mx[,3] } ) , side="right" , col="white", border="blue",
+  add = TRUE , plotCentre = "line")
+
+fg1 <- lapply(fgres,function(mx) { TP <- mx[,2] ; FP <- mx[,3]  ; FP / (TP + FP) } )
+fo1 <- lapply(fores,function(mx) { TP <- mx[,2] ; FP <- mx[,3]  ; FP / (TP + FP) } )
+
+vioplot(fo1 , side="left" , col="white", border="red", plotCentre = "line",
+  ylim=c(0,0.5), xlab="Sample size", ylab="no. pathways")
+  mtext("proportion inconsistent pathways" )
+vioplot(fg1 , side="right" , col="white", border="blue", add = TRUE , plotCentre = "line")
+
+### PDF CHART
+pdf("sens_vio.pdf",height=4,width=8)
+par(mfrow=c(1,3))
+vioplot(lapply(fores,function(mx) { mx[,2] } ) , side="left" , col="white" , border = "red",
+  plotCentre = "line", ylim=c(0,457), xlab="Sample size", ylab="no. pathways")
+  mtext("no. consistent pathways FDR<0.05" ,cex=0.8)
+vioplot(lapply(fgres,function(mx) { mx[,2] } ) , side="right" , col="white", border="blue" ,
+   add = TRUE, plotCentre = "line")
+
+abline(h=length(osets),col="red")
+abline(h=length(fsets),col="blue")
+
+vioplot(lapply(fores,function(mx) { mx[,3] } ) , side="left" , col="white" , border="red",
+  plotCentre = "line", ylim=c(0,210), xlab="Sample size", ylab="no. pathways")
+  mtext("no. inconsistent pathways FDR<0.05" , cex=0.8)
+vioplot(lapply(fgres,function(mx) { mx[,3] } ) , side="right" , col="white", border="blue",
+  add = TRUE , plotCentre = "line")
+
+fg1 <- lapply(fgres,function(mx) { TP <- mx[,2] ; FP <- mx[,3]  ; FP / (TP + FP) } )
+fo1 <- lapply(fores,function(mx) { TP <- mx[,2] ; FP <- mx[,3]  ; FP / (TP + FP) } )
+
+vioplot(fo1 , side="left" , col="white", border="red", plotCentre = "line",
+  ylim=c(0,0.5), xlab="Sample size", ylab="no. pathways")
+  mtext("proportion inconsistent pathways",, cex=0.8)
+vioplot(fg1 , side="right" , col="white", border="blue", add = TRUE , plotCentre = "line")
+dev.off()
+
+```
+
 ## Session information

### 2024-09-13 analysis/GSE158420_rna.Rmd
getting some figures for inclusion in the paper

@@
 
+# output some data for the paper
+head(fores[[1]])
+message(
+message("FORA TP") ; lapply(fores,function(mx) { summary( mx[,2] ) } )
+message("FGSEA TP") ; lapply(fgres,function(mx) { summary( mx[,2] ) } )
+message("FORA FP") ; lapply(fores,function(mx) { summary( mx[,3] ) } )
+message("FGSEA FP") ; lapply(fores,function(mx) { summary( mx[,3] ) } )
+message("FORA FP proportion") ; lapply(fo1,summary)
+message("FGSEA FP proportion") ; lapply(fg1,summary)
+
 ### PDF CHART

### 2024-09-13 analysis/GSE158420_rna.Rmd
bugfix

--- 2024-09-13 analysis/GSE158420_rna.Rmd
@@
 
+# output some data for the paper
+head(fores[[1]])
+message("FORA TP")lapply(fores,function(mx) { summary( mx[,2] ) } )
+message("FGSEA TP") ; lapply(fgres,function(mx) { summary( mx[,2] ) } )
+message("FORA FP") ; lapply(fores,function(mx) { summary( mx[,3] ) } )
+message("FGSEA FP") ; lapply(fores,function(mx) { summary( mx[,3] ) } )
+message("FORA FP proportion") ; lapply(fo1,summary)
+message("FGSEA FP proportion") ; lapply(fg1,summary)
+
 ### PDF CHART

### 2024-09-13 analysis/GSE158420_rna.Rmd
bugfix

--- 2024-09-13 analysis/GSE158420_rna.Rmd
@@
 
+# output some data for the paper
+head(fores[[1]])
+message("FORA TP") ; lapply(fores,function(mx) { summary( mx[,2] ) } )
+message("FGSEA TP") ; lapply(fgres,function(mx) { summary( mx[,2] ) } )
+message("FORA FP") ; lapply(fores,function(mx) { summary( mx[,3] ) } )
+message("FGSEA FP") ; lapply(fores,function(mx) { summary( mx[,3] ) } )
+message("FORA FP proportion") ; lapply(fo1,summary)
+message("FGSEA FP proportion") ; lapply(fg1,summary)
+
 ### PDF CHART

### 2024-09-14 analysis/docker_results/GSE158420_rna.Rmd
ylim

--- 2024-09-13 analysis/GSE158420_rna.Rmd
@@
 vioplot(lapply(fores,function(mx) { mx[,2] } ) , side="left" , col="white" , border = "red",
-  plotCentre = "line", ylim=c(0,457), xlab="Sample size", ylab="no. pathways")
+  plotCentre = "line", ylim=c(0,length(fsets)), xlab="Sample size", ylab="no. pathways")
   mtext("no. consistent pathways FDR<0.05" ,cex=0.8)

### 2024-09-14 analysis/GSE158420_rna.Rmd
legend

@@
 
+legend("bottomright", legend=c("fgsea","fora"), pch=0,
+       col=c("blue","red"), cex=1)
+
 vioplot(lapply(fores,function(mx) { mx[,3] } ) , side="left" , col="white" , border="red",
@@
 abline(h=length(fsets),col="blue")
+
+legend("bottomright", legend=c("fgsea","fora"), pch=0,
+       col=c("blue","red"), cex=1)
 

### 2024-09-14 analysis/sim/run_simpw.Rmd
use data if saved
aspect ratio

@@
 
+if (file.exists("res.Rds") ) {
+
+  res <- readRDS("res.Rds")
+
+} else {
+
 res <- lapply(1:nrow(mygrid), function(i) {
@@
 
-png("fig3_sim.png", width=7,height=5,units="in",res=150,pointsize=12)
+png("fig3_sim.png", width=8,height=4,units="in",res=150,pointsize=12)
 par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
@@
 
-pdf("fig3_sim.pdf", width=7,height=5,pointsize=12)
+pdf("fig3_sim.pdf", width=8,height=4,pointsize=12)
 par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
@@
 
-png("fig3_bars.png", width=7,height=5,units="in",res=150,pointsize=12)
+png("fig3_bars.png", width=8,height=4,units="in",res=150,pointsize=12)
 par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
@@
 #################################
-pdf("fig3_bars.pdf", width=7,height=5,pointsize=12)
+pdf("fig3_bars.pdf", width=8,height=4,pointsize=12)
 par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
@@
 
-png("fig3_box.png", width=7,height=5,units="in",res=150,pointsize=12)
+png("fig3_box.png", width=8,height=4,units="in",res=150,pointsize=12)
 par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
@@
 
-pdf("fig3_box.pdf", width=7,height=5,pointsize=12)
+pdf("fig3_box.pdf", width=8,height=4,pointsize=12)
 par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))

### 2024-09-16 manuscript/manuscript.Rmd
updates for reviewer comments

@@
 superior to ORA over a range of experimental designs [@Kaspi2020-rn].
-Despite this performance difference, a survey of articles published in 2019 showed that
-seven of the top eight most popular tools were based on ORA [@Wijesooriya2022-fw].
+A study of 1366 PubMed Central articles featuring enrichment analysis published in 2019
+indicates that seven of the top eight most popular tools were based on ORA (according to the
+supplementary data)[@Wijesooriya2022-fw].
 The discrepancy between performance and popularity is likely due to the relative ease of
@@
 [@Ziemann2019-ux].
+The criteria for selection included human as the species, presence in the DEE2 database,
+passing most DEE2 quality control measures, a simple control-case experiment design with
+two or three replicates per group, and with 50 or more differentially expressed genes
+(FDR<0.05).
 Raw data for these seven studies are available from NCBI Sequence Read Archive under the
@@
 Gene symbols were then fetched using biomaRt v2.60.0, based on Ensembl version 112 [@Durinck2009-dr].
+
+| SRA accession | Control accessions | Case accessions | Comparison | No. detected genes | No. genes with FDR<0.05 | Reference |
+| --- | --- | --- | --- | --- | --- | --- |
+| SRP128998 | SRR6467485, SRR6467486, SRR6467487 | SRR6467479, SRR6467480, SRR6467481 | Normal glucose versus high glucose | 15,635 | 3,472  | [@Felisbino2021-hc] |
+| SRP038101 | SRR1171523, SRR1171524, SRR1171525 | SRR1171526, SRR1171527, SRR1171528 | Control versus azacytidine treated cells | 13,926 | 3,590 | [@Lund2014-ds] |
+| SRP037718 | SRR1168228, SRR1168229, SRR1168230 | SRR1168225, SRR1168226, SRR1168227 | Control versus SAHA treated cells | 15,477 | 4,988 | [@Rafehi2014-rv] |
+| SRP096177 | SRR5150595, SRR5150596, SRR5150597 | SRR5150592, SRR5150593, SRR5150594 | Control versus Set7 knock-down cells | 15,607 | 5,152 | [@Keating2014-ab] |
+| SRP097759 | SRR5201525, SRR5201526 | SRR5201527, SRR5201528 | GFP control vs SAHH overexpression | 19,139 | 62 | Unpublished |
+| SRP253951 | SRR11517674, SRR11517675, SRR11517676 | SRR11517677, SRR11517678, SRR11517679 | Mock vs SARS-CoV-2 infection | 15,182 | 8,588 | [@Blanco-Melo2020-hh] |
+| SRP068733 | SRR3112216, SRR3112217, SRR3112218 | SRR3112219, SRR3112220, SRR3112221 | Control vs EP300 knock-down | 14,255 | 7,365 | [@Rafehi2017-qn] |
+
+: Table 1. Seven selected RNA-seq datasets for benchmarking.
 
@@
 This dataset was selected due to its high number of assigned reads (367M).
-A library of 200 random gene sets were generated, each containing 30 genes.
+A library of 200 random gene sets, each containing 30 genes, was generated.
 To generate differential expression profiles, 5% of gene sets were selected to be
@@
 ClusterProfiler is not the only tool to suffer from these problems, they are widespread
-(**Table 1**). Still, there are some tools free of such problems.
+(**Table 2**). Still, there are some tools free of such problems.
 For example, fora is a good option for R-based workflows, however the function does not calculate
@@
 container option for local execution.
-One problem that all of these tools listed in Table 1 have is that a background list is optional.
+One problem that all of these tools listed in Table 2 have is that a background list is optional.
 Sometimes this feature is hidden under "advanced options".
@@
 
-: Table 1. Classification of several popular freely available ORA tools regarding the provision of information such as
+: Table 2. Classification of several popular freely available ORA tools regarding the provision of information such as
 enrichment scores and FDR values, as well as behaviours regarding the two algorithmic issues described above.
@@
 pathways relevant to the biological processes being studied.
-Secondly, many popular ORA tools listed in Table 1 do not provide enrichment scores in their
+Secondly, many popular ORA tools listed in Table 2 do not provide enrichment scores in their
 default output tables.

### 2024-09-16 analysis/analyse_rnaseq1.Rmd
savedata at end
added deepdive

@@
   library("RhpcBLASctl")
+  library("Hmisc")
 })
@@
 
+## Focus on dataset 1 down with HPO genesets
+
+This extreme example will help us understand the bias better.
+
+```{r,hpodn}
+
+orig <- orabad(d[[1]],gs[[7]])
+corr <- oragood(d[[1]],gs[[7]])
+orig_dn <- orig[[2]]
+corr_dn <- corr[[2]]
+mg <- merge(orig_dn,corr_dn,by=0)
+
+mg$size <- as.numeric(sapply(strsplit(mg$BgRatio.x,"/"),"[[",1))
+quantiles <- quantile(mg$size, prob = seq(0, 1, length = 6), type = 5)
+mg$levels <- cut2(mg$size, cuts = as.numeric(quantiles))
+cols <- gsub("5","red",gsub("4","orange",gsub("3","yellow",gsub("2","green",gsub("1","blue",as.character(as.numeric(mg$levels)))))))
+
+plot(-log10(mg$pvalue.x),-log10(mg$pvalue.y),
+  ylim=c(0,7), xlim=c(0,7), pch=19,
+  main="SRP128998 HPO down",
+  xlab="-log10 p-value (original)", ylab="-log10 p-value (corrected)")
+
+grid()
+abline(coef = c(0,1))
+
+points(-log10(mg$pvalue.x),-log10(mg$pvalue.y),
+  col=cols, pch=19,cex=0.7)
+
+pdf("fig1_HPOdn.pdf",width=7, height=7)
+plot(-log10(mg$pvalue.x),-log10(mg$pvalue.y),
+  ylim=c(0,7), xlim=c(0,7), pch=19,
+  main="SRP128998 HPO down",
+  xlab="-log10 p-value (original)", ylab="-log10 p-value (corrected)")
+
+grid()
+abline(coef = c(0,1))
+
+points(-log10(mg$pvalue.x),-log10(mg$pvalue.y),
+  col=cols, pch=19,cex=0.7)
+dev.off()
+
+```
+
 ## Table for reviewer
@@
 
+save.image("analyse_rnaseq1.Rdata")
+
 sessionInfo()

### 2024-09-16 analysis/analyse_rnaseq1.Rmd
add quantiles

@@
 quantiles <- quantile(mg$size, prob = seq(0, 1, length = 6), type = 5)
+quantiles
 mg$levels <- cut2(mg$size, cuts = as.numeric(quantiles))

### 2024-09-16 manuscript/manuscript.Rmd
ms work

@@
 Gene symbols were then fetched using biomaRt v2.60.0, based on Ensembl version 112 [@Durinck2009-dr].
+Descriptive information of the seven datasets is provided in Table 1.
 
@@
 No set seed was used, but results were similar when run on different computer systems.
-All analyses were undertaken with R v4.4.0 in a Docker container.
+All analyses were undertaken with R v4.4.1 in a Docker container.
 
@@
 gene set libraries.
-The mean number of statistically significant results was calculated for each of the gene set
+The number of statistically significant results was calculated for each of the gene set
 libraries before and after correcting the problem (**Figure 1A**).
-Each of the gene set libraries showed an uplift in the number of significant gene sets when correcting the background problem.
-The smallest increase was observed in Cellmarkers gene library (11.8%), while the largest
-increase was observed for HPO (224%).
+Each study was affected to a similar degree (mean uplift 125% ±199%), except for #5 which saw the number of
+significant GO terms increase from 8 to 54 after correction (**Table S1**).
 The similarity between results, as quantified with the Jaccard similarity, was highest for
@@
 more severely.
+We investigated discrepant results for data set #1 (SRP128998) downregulated genes in HPO gene sets.
+A scatterplot of significance scores (**Figure 1C**) shows severe deviation in p-values in gene sets in the largest
+quintile (109-1382 members detected), suggesting larger gene sets are more severely affected by the background
+problem.
 
-![**Figure 1A. Effect of correcting the background problem.** Figures represent the mean number
-of statistically significant sets (FDR<0.05) over seven independent datasets detected with
-clusterProfiler with and without correcting the background problem.
-Separate tests for up- and down-regulated gene lists were conducted.
-The percent increase in significant sets are shown.](../analysis/fig1_ORAsig.png "Figure 1A
+![**Figure 1A. Number of statistically significant pathways (FDR<0.05) in seven independent RNA-seq studies with and
+without correction of the background problem.
+Separate tests were conducted for up- and down-regulated gene lists.
+Mean values are depicted as an “x”. The percent increase in the mean number of significant sets are shown.
+“*” indicates that 1 has been added to all values before log transformation.](
+../analysis/fig1_ORAsig.png "Figure 1A
 shows that correcting the background bug improves the number of significant sets between 12 and
@@
 
-![**Figure 1B. Impact of the background problem is worse for smaller gene set libraries.** Mean
-Jaccard similarity index was calculated for original and corrected ORA for seven independent
-datasets. The number of genes represented in one or more functional sets is shown on the x-axis,
-and the mean Jaccard index is shown on the y-axis.](../analysis/fig1_ORAjac.png "Figure 1B
-shows that the background bug impacts smaller gene sets libraries like KEGG and HPO more
-severely.")
+![**Figure 1B. Impact of the background problem is worse for smaller gene set libraries.**
+Mean Jaccard similarity index was calculated for original and corrected ORA for seven independent datasets.
+The number of genes represented in one or more functional sets is shown on the x-axis, and the mean Jaccard index is
+shown on the y-axis; whiskers represent the standard deviation.](
+../analysis/fig1_ORAjac.png "Figure 1B shows that the background bug impacts smaller gene sets libraries like KEGG
+and HPO more severely.")
+
+![**Figure 1C. Significance values (-log10 p-values) for dataset #1 (SRP128998) enrichment analysis
+using HPO gene sets with and without correction of the background problem.**
+Gene sets were divided into quintiles based on the number of genes detected in the dataset.](
+../analysis/fig1_HPOdn.png "Figure 1C shows that larger gene sets are affect the p-values more severely by the
+background bug.")
 
@@
 original and corrected analyses (**Figure 2A**).
+The impact of the FDR problem was similar across these seven datasets (**Table S2**).
 
@@
 
-![**Figure 2A. Effect of correcting the FDR problem.** Figures represent the mean number of
-significant sets with and without correction of the FDR problem over seven independent RNA-seq
-datasets.
-Foreground list consisted of 2000 genes with the smallest p-values](../analysis/fig2_ORAfdr.png "Figure 2A shows that correcting the FDR bug slightly reduces the number of significant sets.")
+![**Figure 2A. Effect of correcting the FDR problem.**
+Figures represent the mean number of significant sets with and without correction of the FDR problem over seven
+independent RNA-seq datasets.
+Foreground list consisted of 2000 genes with the smallest p-values.
+Mean values are depicted as an “x”.
+“*” indicates that 1 has been added to all values before log transformation.](
+../analysis/fig2_ORAfdr.png "Figure 2A shows that correcting the FDR bug slightly reduces the number of significant
+sets.")
 
-![**Figure 2B. Impact of the FDR problem is worse for shorter foreground gene lists.** Mean Jaccard
-index between original and corrected analysis were calculated for seven RNA-seq datasets and nine
-gene set libraries. Only GO, Reactome, Cellmarkers and TFT GTRD are shown.](
+![**Figure 2B. Impact of the FDR problem is worse for shorter foreground gene lists.**
+Mean Jaccard index between original and corrected analysis were calculated for seven RNA-seq datasets with gene sets
+from GO, Reactome, Cellmarkers and TFT GTRD; whiskers represent the standard deviation.](
 ../analysis/fig2_ORAsz.png "Figure 2B shows that the impact of the FDR bug is worse when
@@
 prototypical FCS method.
-These methods were tested with different levels of introduced variation and the mean
+These methods were tested with different levels of introduced variation and the
 precision, recall and F1 scores were recorded (**Figure 3A**).
-Mitigating the background problem improved recall (by 1.8%), however it caused
-a similar-sized decrease in precision (1.6% lower).
-Addressing the FDR problem improved precision by 9.9%, but the recall dropped by 7.7%.
-Addressing both problems improved precision by 8.2% but decreased recall by 5.4%.
+Mitigating the background problem improved mean recall (by 2.9%), however it also caused
+a decrease in mean precision (1.3% lower).
+Addressing the FDR problem improved mean precision by 11.7%, but mean recall dropped by 16.6%.
+Addressing both problems improved mean precision by 9.7% but decreased mean recall by 11.7%.
 Under these conditions, the FDR problem affected results more severely than the background
@@
 fora to three significant figures.
-fgsea recorded precision of 0.940, similar to the ORA methods, but recall was far superior,
-with fgsea scoring 22.9% higher than default clusterProfiler and 28.3% higher than fora.
-As a result, fgsea's overall accuracy as defined by F1 index was 21.0% higher than fora and
-18.6% higher than default clusterProfiler.
+fgsea recorded mean precision of 0.939, similar to the ORA methods, but recall was far superior,
+with fgsea scoring 44.9% higher than default clusterProfiler and 64.2% higher than fora.
+As a result, fgsea's overall accuracy as defined by F1 index was 35.4% higher than fora and
+28.9% higher than default clusterProfiler.
 
@@
 The precision of fgsea was relatively stable despite high levels of added noise, ranging from
-0.95 to 0.92.
+0.95 to 0.91.
 The recall of fgsea was superior to all ORA approaches across all levels of added noise.
@@
 data.**
-Precision, recall and F1 scores are the mean of trials at a range of added noise.
+(A) Precision, recall and F1 scores across simulations.
 CP; clusterProfiler, CP BG fix; clusterProfiler with a workaround to the background problem,
@@
 fgsea; an FCS method similar to GSEA.
+Mean values are depicted as an “x”.
 ](../analysis/sim/fig3_bars.png "Figure 3A shows that fora is more
@@
 levels of added noise.**
-Precision, recall and F1 score are shown with the addition of different amounts of random
-noise to expression levels.
-The purple line for clusterProfiler with mitigations for both problems is not visible as the
-performance was identical to fora.
+Precision, recall and F1 score are shown with the addition of different amounts of random noise to expression levels.
+The purple line for clusterProfiler with mitigations for both problems is not visible as the performance was
+identical to fora.
 ](../analysis/sim/fig3_sim.png "Figure 3B shows that ORA methods have deteriorating accuracy
@@
 In the full set of patient samples, fora (ORA) yielded 375 significantly differentially
-regulated Reactomes, while fgsea (FCS) gave 470 (FDR<0.05).
+regulated Reactomes, while fgsea (FCS) gave 480 (FDR<0.05).
 Fgsea yielded consistently more significant pathways than fora in downsampled data
@@
 Fgsea also identified a larger proportion of pathways in downsampled data compared to fora.
-For example at a sample size of 5, fgsea identified 74.5% of pathways from the full set as
+For example at a sample size of 5, fgsea identified 73.3% (median) of pathways from the full set as
 compared to 41.5% for fora, confirming better recall for FCS over ORA.
@@
 ![**Figure 4. Downsampling confirms higher sensitivity for FCS over ORA.**
-(A) The number of significant pathways identified in downsampled data that are consistent
-with findings in the full dataset, using FCS (fgsea) and ORA (fora) methods.
+(A) Violin plot showing the number of significant pathways identified in downsampled data that are consistent with
+findings in the full dataset, using FCS (fgsea) and ORA (fora) methods.
+Downsampling was repeated 50 times.
 The blue and red horizontal lines indicate the number of significant pathways identified
@@
 (C) The proportion of pathways in downsampled data that are inconsistent with the full
-dataset.
-All values are the median of 50 simulations.](../analysis/fig4.png "Figure 4A shows that fgsea is more sensitive than fora.
-Figure 4B shows fgsea has higher absule number of false positives, but Figure 4C shows that as a proportion of all findings fgsea has lower false positive rate than fora.")
+dataset.](
+../analysis/fig4.png "Figure 4A shows that fgsea is more sensitive than fora. Figure 4B shows fgsea has higher
+absule number of false positives, but Figure 4C shows that as a proportion of all findings fgsea has lower false
+positive rate than fora.")
 
@@
 ~2020 where confused users were wondering why the number of background genes changed so much.
+The effect of excluding genes without functional annotations causes the background list to appear
+smaller than it would otherwise be, causing enrichment scores to appear smaller.
 The results shown here indicate the background problem affects smaller gene set libraries
@@
 Users can upload their foreground and background lists, select their preferred gene set
-library and compare results from default and patched analyses.
+library and compare results from default and corrected analyses.
 
@@
 Sometimes this feature is hidden under "advanced options".
-Given the strong biological and technical biases that are present in contemporary omics like single cell and spatial transcriptomics,
-it should be a mandatory step for users to provide a custom background list.
+Given the strong biological and technical biases that are present in contemporary omics such as single cell
+and spatial transcriptomics, it should be a mandatory step for users to provide a custom background list.
 
@@
 [@Langfelder2008-hu], or with variant-harbouring genes associated with a disease [@White2019-ps].
+In these cases, we would recommend using a tool from Table 2 that reports FDR values and enrichment scores,
+and avoids background and FDR problems.
+Doubts also arise when choosing options and parameters for ORA.
+Most important is to use a background gene list that reflects the true universe of genes detected robustly in the
+assay.
+Secondly, significance of enrichment results must be based on the FDR-corrected p-values.
+Selection criteria for foreground genes is another crucial choice; for this we endorse the approach of
+Tarca et al, [@Tarca2013-xh] which accommodates cases where relatively few individual genes meet the FDR threshold.
+Some ORA tools have a filter for gene set size.
+We’d recommend a minimum size of 5-10 in order to avoid missing small gene sets that exhibit strong enrichments.
+Interpret with caution enrichment results where the intersection of input gene list and pathway is ≤3 genes.
+A maximum gene set filter is not required if end results are interpreted using enrichment scores together with FDR
+values.
+The choice of pathway database also influences the conclusions that can be drawn [@Mubeen2019-gw;@Karp2021-ue].
+Our suggestion is to use comprehensive and open-source pathway databases such as Reactome and Gene Ontology
+Biological Process.
+Proprietary databases like KEGG and Ingenuity are less comprehensive, have restrictive conditions for non-academic
+users and do not provide public access to historical database versions which inhibits future reproducibility.
 

### 2024-09-17 manuscript/manuscript.Rmd
more corrections

@@
 conducting ORA as compared to FCS.
-A minimal ORA might involve pasting a list of gene symbols into a text box on a website,
+A minimal ORA involves pasting a list of gene symbols into a text box on a website,
 with results appearing nearly instantly.
@@
 addition of some extra variation drawn from a normal distribution with mean of 1 and
-standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
-Three "control" and five "case" pseudosamples were generated.
+standard deviation varying between 0 and 0.6, this value selected randomly for each gene.
+Three "control" and "case" pseudosamples were generated.
 For the genes selected to be differentially expressed, expression values were multiplied by a
@@
 discovery rate.
-This was repeated 50 times at each sample size, and the median values were reported.
+This was repeated 50 times at each sample size, and the results were presented as a violin plot.
 
@@
 libraries before and after correcting the problem (**Figure 1A**).
-Each study was affected to a similar degree (mean uplift 125% ±199%), except for #5 which saw the number of
+Each study was affected to a similar degree (31-94% increase), except for #5 which saw the number of
 significant GO terms increase from 8 to 54 after correction (**Table S1**).
@@
 Firstly, all variations of ORA were less accurate (based on F1 score) as compared to FCS in this
-simulation work, confirming a previous report [@Kaspi2020-rn].
+simulation work (**Figure 3A**), confirming a previous report [@Kaspi2020-rn].
 Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS

### 2024-09-17 analysis/sim/run_simpw.Rmd
cores

@@
   VARIANCE=mygrid[i,"VARIANCE"]
-  agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets,cores=32)
+  agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets,cores=16)
 })

### 2024-09-17 analysis/sim/simpw_func.R
cores=16

@@ -159,7 +159,7 @@ RepParallel <- function(n, expr, simplify = "array",...) {
       else return(answer)
     }
 # RepParallel usage
-#xxx<-RepParallel(10,simrna(a,5,10000000,0.2,20), simplify=F, mc.cores = detectCores() )
+#xxx<-RepParallel(10,simrna(a,5,10000000,0.2,20), simplify=F, mc.cores = 16 )
 
 
 #################################################
@@ -190,7 +190,7 @@ deseq2<-function(x) {
 # clusterprofiler default function
 ##################################
 # Note that clusterprofiler requires different gene set format
-run_clusterprofiler_default <- function(x,DGE_FUNC,gsets,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,cores=32){
+run_clusterprofiler_default <- function(x,DGE_FUNC,gsets,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,cores=16){
 
 gsets2 <- stack(gsets)[,c(2,1)]
 colnames(gsets2) <- c("term","gene")
@@ -268,7 +268,7 @@ x
 # clusterprofiler BG fix function
 ##################################
 # Note that clusterprofiler requires different gene set format
-run_clusterprofiler_bgfix <- function(x,DGE_FUNC,gsets,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,cores=32){
+run_clusterprofiler_bgfix <- function(x,DGE_FUNC,gsets,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,cores=16){
 
 gsets2 <- stack(gsets)[,c(2,1)]
 colnames(gsets2) <- c("term","gene")
@@ -356,7 +356,7 @@ x
 # clusterprofiler FDR fix
 ##################################
 # Note that clusterprofiler requires different gene set format
[…]

### 2024-09-18 analysis/analyse_rnaseq1.Rmd
fig 1c panel png

@@
 
+# PNG
+png("fig1_HPOdn.png")
+
+plot(-log10(mg$pvalue.x),-log10(mg$pvalue.y),
+  ylim=c(0,7), xlim=c(0,7), pch=19,
+  main="SRP128998 HPO down",
+  xlab="-log10 p-value (original)", ylab="-log10 p-value (corrected)")
+
+grid()
+abline(coef = c(0,1))
+
+points(-log10(mg$pvalue.x),-log10(mg$pvalue.y),
+  col=cols, pch=19,cex=0.7)
+
+dev.off()
+
+# PDF
 pdf("fig1_HPOdn.pdf",width=7, height=7)
